---
title: "PSC study: Q4"
output:
  html_document:
    df_print: paged
    toc: true
    toc_float: true
    toc_depth: 6
editor_options: 
  markdown: 
    wrap: 72
---

# 3. Hypothesis: PSC disease is associated with Nancy index

## Introduction

An IBD vs. no-IBD comparison was performed within PSC patients (pre-LTx and rPSC individuals combined) to discover IBD-specific pattern. 

**IBD vs no_IBD analysis on merged data**

-   Alpha diversity –\> group effect, country effect, interaction effect

-   Beta diversity – PERMANOVA, PCA -\> group effect, country effect,
    interaction effect

-   DAA -\>

    -   o Group effect – linDA + MaAsLin2 intersection

    -   o Country effect – – taxa with significant interaction effect were excluded based on individual post-hoc analysis. Taxa had to have the same direction in both countries 

**IBD vs no_IBD** – finding IBD-specific pattern in PSC patients

*Importing libraries and custom functions built for this analysis*

```{r, warning=FALSE}
source("custom_functions.R")
```

## Data Import

Importing ASV, taxa and metadata tables for both Czech and Norway
samples.

**Czech**

*MICROBIOME*

```{r, warning=FALSE}
path = "../../data/analysis_ready_data/ikem/"
asv_tab_ikem <- as.data.frame(fread(file.path(path,"asv_table_ikem.csv"),
                                    check.names = FALSE))
taxa_tab_ikem <- as.data.frame(fread(file.path(path,"taxa_table_ikem.csv"),
                                     check.names = FALSE))
metadata_ikem <- as.data.frame(fread(file.path(path,"metadata_ikem.csv"),
                                     check.names = FALSE))
```

*Nancy_max info*

```{r, warning=FALSE}
# clinical metadata
metadata_clinical <- read.csv("../../data/clinical_data/clinical_metadata_cz.csv")
metadata_clinical$PatientID <- as.character(metadata_clinical$PatientID)
```

### Merging

**Clinical data**

```{r}
metadata_ikem <- merge(metadata_ikem,metadata_clinical[,c("SampleID","Nancy_max")],by="SampleID",all.x=TRUE)
```

**Terminal ileum**

```{r, warning=FALSE}
ileum_data <- merging_data(asv_tab_1=asv_tab_ikem,
                           asv_tab_2= NULL,
                           taxa_tab_1=taxa_tab_ikem,
                           taxa_tab_2=NULL,
                           metadata_1=metadata_ikem,
                           metadata_2=NULL,
                           segment="TI",Q="Q4")

ileum_asv_tab <- ileum_data[[1]]
ileum_taxa_tab <- ileum_data[[2]]
ileum_metadata <- ileum_data[[3]]

ileum_metadata <- merge(ileum_metadata,metadata_ikem[,c("SampleID","Nancy_max")],by="SampleID",all.x=TRUE) %>% dplyr::mutate(Nancy = case_when(
    Nancy_max %in% c(0,1) ~ "low",
    Nancy_max %in% c(2,3,4,5) ~ "high"))

ileum_metadata %<>% drop_na(Nancy)
ileum_asv_tab %<>% dplyr::select("SeqID",ileum_metadata$SampleID)
ileum_taxa_tab %<>% dplyr::filter(SeqID %in% ileum_asv_tab$SeqID)
```

**Colon**

```{r, warning=FALSE}
colon_data <- merging_data(asv_tab_1=asv_tab_ikem,
                           asv_tab_2=NULL,
                           taxa_tab_1=taxa_tab_ikem,
                           taxa_tab_2=NULL,
                           metadata_1=metadata_ikem,
                           metadata_2=NULL,
                           segment="colon",Q="Q4")

colon_asv_tab <- colon_data[[1]]
colon_taxa_tab <- colon_data[[2]]
colon_metadata <- colon_data[[3]]

colon_metadata <- merge(colon_metadata,metadata_ikem[,c("SampleID","Nancy_max")],by="SampleID",all.x=TRUE) %>% dplyr::mutate(Nancy = case_when(
    Nancy_max %in% c(0,1) ~ "low",
    Nancy_max %in% c(2,3,4,5) ~ "high"))

colon_metadata %<>% drop_na(Nancy)
colon_asv_tab %<>% dplyr::select("SeqID",colon_metadata$SampleID)
colon_taxa_tab %<>% dplyr::filter(SeqID %in% colon_asv_tab$SeqID)

```

## CHI SQUARE TEST

### pre_LTx, rPSC, non-rPSC

```{r, warning=FALSE}
cz <- metadata_ikem %>% dplyr::group_by(Nancy_max,Group) %>%
  distinct(Patient, .keep_all = TRUE) %>%
  count() %>% drop_na() %>% 
  dplyr::mutate(Nancy = case_when(
    Nancy_max %in% c(0,1) ~ "low",
    Nancy_max %in% c(2,3,4,5) ~ "high",
  )) %>% `colnames<-`(c("Nancy_max","Group","n","Nancy")) 

# Summarize by Nancy and Group
data <- cz %>%
  group_by(Nancy, Group) %>%
  summarise(n = sum(n), .groups = "drop") %>%
  pivot_wider(names_from = Group, values_from = n) %>%
  column_to_rownames("Nancy") %>%
  as.matrix()


chi_res <- chisq.test(data)
chi_res$expected
chi_res
```

### pre_LTx + rPSC (PSC patients), non-rPSC

```{r, warning=FALSE}
data2 <- data %>% as.data.frame() %>%
  dplyr::mutate(PSC=pre_ltx+rPSC) %>% 
  dplyr::select(PSC,`non-rPSC`) %>% as.matrix()

data2
chi_res <- chisq.test(data2)
chi_res$expected
chi_res
```

### rPSC, non-rPSC

```{r, warning=FALSE}
data3 <- data %>% as.data.frame() %>%
  dplyr::select(rPSC,`non-rPSC`) %>% as.matrix()

data3
chi_res <- chisq.test(data3)
chi_res$expected
chi_res
```

# Data Analysis - Terminal ileum

```{r, warning=FALSE}
segment="terminal_ileum"
```

## Filtering

Rules: 

  - sequencing depth \> 10000 
  
  - nearZeroVar() with default settings 

**Rarefaction Curve**

```{r, warning=FALSE}
path="../intermediate_files/rarecurves"
seq_depth_threshold <- 10000
```

```{r, eval = FALSE}
ps <- construct_phyloseq(ileum_asv_tab,ileum_taxa_tab,ileum_metadata)
rareres <- get_rarecurve(obj=ps, chunks=500)
save(rareres,file = file.path(path,"rarefaction_ileum.Rdata"))
```

```{r, warning=FALSE}
load(file.path(path,"rarefaction_ileum.Rdata"))

prare <- ggrarecurve(obj=rareres,
                      factorNames="Country",
                      indexNames=c("Observe")) + 
  theme_bw() +
  theme(axis.text=element_text(size=8), panel.grid=element_blank(),
        strip.background = element_rect(colour=NA,fill="grey"),
        strip.text.x = element_text(face="bold")) + 
  geom_vline(xintercept = seq_depth_threshold, 
             linetype="dashed", 
             color = "red") + 
  xlim(0, 20000)

prare
```

**Library size**

```{r, fig.width=5, fig.height=4, fig.fullwidth=TRUE, warning=FALSE}
read_counts(ileum_asv_tab, line = c(5000,10000))
```

### Sequencing depth

```{r, warning=FALSE}
data_filt <- seq_depth_filtering(ileum_asv_tab,
                                 ileum_taxa_tab,
                                 ileum_metadata,
                                 seq_depth_threshold = 10000)

filt_ileum_asv_tab <- data_filt[[1]]; alpha_ileum_asv_tab <- filt_ileum_asv_tab
filt_ileum_taxa_tab <- data_filt[[2]]; alpha_ileum_taxa_tab <- filt_ileum_taxa_tab
filt_ileum_metadata <- data_filt[[3]]; alpha_ileum_metadata <- filt_ileum_metadata

seq_step <- dim(filt_ileum_asv_tab)[1]
```

**Library size**

```{r, fig.width=5, fig.height=4, fig.fullwidth=TRUE, warning=FALSE}
read_counts(filt_ileum_asv_tab,line = c(5000,10000))
```

### NearZeroVar

```{r, warning=FALSE}
data_filt <- nearzerovar_filtering(filt_ileum_asv_tab, 
                                   filt_ileum_taxa_tab,
                                   filt_ileum_metadata)

filt_ileum_asv_tab <- data_filt[[1]]
filt_ileum_taxa_tab <- data_filt[[2]]
nearzero_step <- dim(filt_ileum_asv_tab)[1]
```

Library size

```{r, fig.width=5, fig.height=4, fig.fullwidth=TRUE, warning=FALSE}
read_counts(filt_ileum_asv_tab,line = c(5000,10000))
```

### Final Counts

```{r, warning=FALSE}
final_counts_filtering(ileum_asv_tab,
                       filt_ileum_asv_tab,
                       filt_ileum_metadata,
                       seq_step, 0, nearzero_step) %>% `colnames<-`("Count")
```

## Alpha diversity

```{r, warning=FALSE}
path = "../results/Q4/alpha_diversity"
```

**Calculation**

```{r, warning=FALSE}
# Construct MPSE object
alpha_ileum_metadata$Sample <- alpha_ileum_metadata$SampleID
ileum_mpse <- as.MPSE(construct_phyloseq(alpha_ileum_asv_tab,
                                         alpha_ileum_taxa_tab,
                                         alpha_ileum_metadata))

ileum_mpse %<>% mp_rrarefy(raresize = 10000,seed = 123)

# Calculate alpha diversity - rarefied counts
ileum_mpse %<>% mp_cal_alpha(.abundance=RareAbundance, force=TRUE)
```

```{r, warning=FALSE}

alpha_div_plots <- list()

# preparing data frame
alpha_data <- data.frame(SampleID=ileum_mpse$Sample.x,
                         Richness=ileum_mpse$Observe,
                         Shannon=ileum_mpse$Shannon,
                         Simpson=ileum_mpse$Simpson,
                         Pielou=ileum_mpse$Pielou,
                         Group=ileum_mpse$Nancy,
                         Country=ileum_mpse$Country,
                         Patient=ileum_mpse$Patient)

write.csv(alpha_data,file.path(path,paste0("alpha_indices_",segment,".csv")),
          row.names = FALSE)

alpha_data %<>% drop_na()
```

### Plots

```{r, fig.width=6, fig.height=4, fig.fullwidth=TRUE}
p_boxplot_alpha <- alpha_diversity_custom_2(alpha_data)

# save the results
alpha_div_plots[[paste(segment,"Custom")]] <- p_boxplot_alpha

# see the results
p_boxplot_alpha
```

```{r,results='hide', echo=FALSE}
pdf("../figures/Q4/alpha_diversity_terminal_ileum.pdf",
    height =4,width = 6)
p_boxplot_alpha
dev.off()
```

### Linear Model

```{r, warning=FALSE}
path = "../results/Q4/alpha_diversity"
alpha_data <- read.csv(file.path(path,paste0("alpha_indices_",segment,".csv")))
```

**Richness**

```{r, warning=FALSE}
results_model <- pairwise.lm(formula = "Richness ~ Group",
                             factors=alpha_data$Group,
                             data=alpha_data)

# check interaction
if (!is.data.frame(results_model)){
  results_model_observe <- results_model[[1]]
  results_model_observe_emeans <- results_model[[2]]
} else {
  results_model_observe <- results_model
  results_model_observe_emeans <- NA
}

# save the results
pc_observed <- list(); 
pc_observed[[segment]] <- results_model_observe
```

```{r, warning=FALSE}
# see the results
knitr::kable(results_model_observe,digits = 3,
caption = "Raw results of linear model of richness estimation.")

knitr::kable(results_model_observe_emeans,digits = 3,
caption = "Raw results of independent country analysis")
```

**Shannon**

```{r, warning=FALSE}
results_model <- pairwise.lm(formula = "Shannon ~ Group",
                             factors=alpha_data$Group,
                             data=alpha_data)

# check interaction
if (!is.data.frame(results_model)){
  results_model_shannon <- results_model[[1]]
  results_model_shannon_emeans <- results_model[[2]]
} else {
  results_model_shannon <- results_model
  results_model_shannon_emeans <- NA
}

# save the results
pc_shannon <- list(); 
pc_shannon[[segment]] <- as.data.frame(results_model_shannon)

```

```{r, warning=FALSE}
# see the results
knitr::kable(results_model_shannon,digits = 3,
caption = "Raw results of linear model of Shannon estimation.")

knitr::kable(results_model_shannon_emeans,digits = 3,
caption = "Raw results of independent country analysis")
```

**Simpson**

```{r, warning=FALSE}
results_model <- pairwise.lm(formula = "Simpson ~ Group",
                                     factors=alpha_data$Group,
                                     data=alpha_data)

# check interaction
if (!is.data.frame(results_model)){
  results_model_simpson <- results_model[[1]]
  results_model_simpson_emeans <- results_model[[2]]
} else {
  results_model_simpson <- results_model
  results_model_simpson_emeans <- NA
}


# save the results
pc_simpson <- list(); 
pc_simpson[[segment]] <- as.data.frame(results_model_simpson)
```

```{r, warning=FALSE}
# see the results
knitr::kable(results_model_simpson,digits = 3,
caption = "Raw results of linear model of Simpson estimation.")

knitr::kable(results_model_simpson_emeans,digits = 3,
caption = "Raw results of independent country analysis")
```

### Saving results

```{r, warning=FALSE}
alpha_list <- list(
  Richness=pc_observed[[segment]],
  Shannon=pc_shannon[[segment]])
                   
write.xlsx(alpha_list, 
           file = file.path(path,paste0("alpha_diversity_results_",segment,".xlsx")))
```


## Beta diversity

Calculating Aitchison distance (euclidean distance on clr-transformed
data), both at ASV and genus level.

### Main analysis

**Genus level, Aitchison distance**

```{r, warning=FALSE}
level="genus"
```

```{r, warning=FALSE}
path = "../results/Q4/beta_diversity"
```

```{r, warning=FALSE}
pairwise_aitchison_raw <- list()
pca_plots_list <- list()
```

Aggregation, filtering

```{r, warning=FALSE}
# Aggregation
genus_data <- aggregate_taxa(ileum_asv_tab,
                             ileum_taxa_tab,
                             taxonomic_level=level,
                             names=TRUE)
# Filtration
filt_data <- filtering_steps(genus_data[[1]],
                             genus_data[[2]],
                             ileum_metadata,
                             seq_depth_threshold=10000)

filt_ileum_genus_tab <- filt_data[[1]]
filt_ileum_genus_taxa <- filt_data[[2]]
filt_ileum_metadata <- filt_data[[3]]

filt_ileum_metadata %<>% dplyr::mutate(Group=Nancy)
```

##### PERMANOVA

```{r, warning=FALSE}
pairwise_df <- filt_ileum_genus_tab %>% column_to_rownames("SeqID") %>% t()

# main effect
pp_main <- pairwise.adonis(pairwise_df,
                           filt_ileum_metadata$Group,
                           covariate = NULL, sim.method = "robust.aitchison", p.adjust.m="BH")


# tidy the results
pp_factor <- pp_main

cols <- c("pairs","Df","SumsOfSqs", "F.Model","R2","p.value", "p.adjusted", "sig")
colnames(pp_factor) <- cols

# save raw results
pairwise_aitchison_raw[[paste(level, segment)]] <- pp_factor
```

```{r, warning=FALSE}
# see the results
knitr::kable(pp_factor,digits = 3,caption = "PERMANOVA, GROUP separation")
```

##### Plots

```{r, warning=FALSE}
p <- pca_plot_custom(filt_ileum_genus_tab,
                                 filt_ileum_genus_taxa,
                                 filt_ileum_metadata,
                                 show_boxplots = TRUE,
                                 variable = "Group", size=3, show_legend=FALSE)

# save the results
pca_plots_list[[paste(segment,level,"custom")]] <- p

# see the results
p
```

```{r,results='hide', echo=FALSE}
pdf("../figures/Q5/beta_diversity_terminal_ileum.pdf",
    height =5,width = 5)
p
dev.off()
```

#### Saving results

```{r, warning=FALSE}
write.xlsx(pairwise_aitchison_raw[[paste(level, segment)]], 
           file = file.path(path,
           paste0("beta_diversity_results_", segment,".xlsx")))
```

### Supplementary analysis

```{r, warning=FALSE}
supplements_beta <- list()
```

#### Genus level

```{r, warning=FALSE}
level="genus"
```

##### Bray-Curtis

**PERMANOVA**

```{r, warning=FALSE}
pairwise_df <- filt_ileum_genus_tab %>% column_to_rownames("SeqID") %>% t()

# main effect
pp_main <- pairwise.adonis(pairwise_df,filt_ileum_metadata$Group,covariate = NULL, sim.method = "bray", p.adjust.m="BH")


# tidy the results
pp_factor <- pp_main

cols <- c("pairs","Df","SumsOfSqs", "F.Model","R2","p.value", "p.adjusted", "sig")
colnames(pp_factor) <- cols

# save raw results
supplements_beta[[paste("bray",level,segment)]] <- pp_factor
```

```{r, warning=FALSE}
# see the results
knitr::kable(pp_factor,digits = 3,caption = "PERMANOVA, GROUP separation")

```


**Plots**

```{r, warning=FALSE}
p <- pca_plot_custom(filt_ileum_genus_tab,
                                 filt_ileum_genus_taxa,
                                 filt_ileum_metadata,
                                 measure = "bray",
                                 show_boxplots = TRUE,
                                 variable = "Group", size=3, show_legend=TRUE)

# save the results
supplements_beta[[paste("PCoA bray",level,segment)]] <- p

# see the results
p
```

##### Jaccard

**PERMANOVA**

```{r, warning=FALSE}
pairwise_df <- filt_ileum_genus_tab %>% column_to_rownames("SeqID") %>% t()

# main effect
pp_main <- pairwise.adonis(pairwise_df,filt_ileum_metadata$Group,covariate = NULL, sim.method = "jaccard", p.adjust.m="BH")


# tidy the results
pp_factor <- pp_main

cols <- c("pairs","Df","SumsOfSqs", "F.Model","R2","p.value", "p.adjusted", "sig")
colnames(pp_factor) <- cols

# save raw results
supplements_beta[[paste("jaccard",level,segment)]] <- pp_factor
```

```{r, warning=FALSE}
# see the results
knitr::kable(pp_factor,digits = 3,caption = "PERMANOVA, GROUP separation")
```



**Plots**

```{r, warning=FALSE}
p <- pca_plot_custom(filt_ileum_genus_tab,
                                 filt_ileum_genus_taxa,
                                 filt_ileum_metadata,
                                 measure = "jaccard",
                                 show_boxplots = TRUE,
                                 variable = "Group", size=3, show_legend=TRUE)

# save the results
supplements_beta[[paste("PCoA jaccard",level,segment)]] <- p

# see the results
p
```

#### Saving results

```{r, warning=FALSE}
write.xlsx(supplements_beta[!grepl("PCoA",names(supplements_beta))],
           file = file.path(path,
           paste0("supplements_beta_diversity_", segment,".xlsx")))
```

## Univariate Analysis

### Main analysis

```{r, warning=FALSE}
level="genus"
```

```{r, warning=FALSE}
# needed paths
path = "../results/Q4/univariate_analysis"
path_maaslin=file.path("../intermediate_files/maaslin/Q4",level)
```

```{r, warning=FALSE}
# variables
raw_linda_results_genus <- list();
raw_linda_results_genus[[segment]] <- list()
linda_results_genus <- list(); 
linda_results_genus[[segment]] <- list()

# country and interaction problems
list_intersections <- list()
list_venns <- list()
uni_statistics <- list()

# workbook for final df
wb <- createWorkbook()

# PSC effect
psc_effect <- list()
```

Aggregate taxa

```{r, warning=FALSE}
genus_data <- aggregate_taxa(ileum_asv_tab,
                             ileum_taxa_tab,
                             taxonomic_level = level)

ileum_genus_tab <- genus_data[[1]]
ileum_genus_taxa_tab <- genus_data[[2]]

ileum_genus_asv_taxa_tab <- create_asv_taxa_table(ileum_genus_tab,
                                                  ileum_genus_taxa_tab)
```

#### low vs high

```{r, warning=FALSE}
group <- c("low","high")
comparison_name <- paste0(group[1], " vs ",group[2])
```

```{r, echo=FALSE}
comparison_title <- sub("(\\w+) vs (\\w+)", "\\2 and \\1", comparison_name)
```

##### linDA

```{r, fig.width=15, fig.height=5, fig.fullwidth=TRUE}

ileum_metadata %<>% dplyr::mutate(Group=Nancy)
# prepare the data
linda_data <- binomial_prep(ileum_genus_tab,
                            ileum_genus_taxa_tab,
                            ileum_metadata,
                            group, usage="linDA")

filt_ileum_uni_data <- linda_data[[1]]
filt_ileum_uni_taxa <- linda_data[[2]]
filt_ileum_uni_metadata <- linda_data[[3]]

# fit the model
linda.obj <- linda(filt_ileum_uni_data, 
                   filt_ileum_uni_metadata, 
                   formula = '~ Group')

linda.output <- linda.obj$output
linda.output <- linda_renaming(linda.output, group)

# save the results
group1 <- paste0(group[1], " vs ","Group",group[2])

for (grp in c(group1)){
  raw_linda_results_genus[[segment]][[grp]] <- 
    rawlinda.df(linda.output,
                grp,
                filt_ileum_uni_data,
                filt_ileum_uni_taxa)
  
  linda_results_genus[[segment]][[grp]] <- 
    linda.df(linda.output,
             grp,
             filt_ileum_uni_data,
             filt_ileum_uni_taxa)
}

```

```{r, fig.width=5, fig.height=5, fig.fullwidth=TRUE}
# volcano plot
volcano_1 <- volcano_plot_linda(linda.output, 
                                group1, 
                                taxa_table = filt_ileum_uni_taxa) + 
            ggtitle(comparison_title)

volcano_1
```

##### MaAsLin2

```{r, echo=FALSE,results='hide',message=FALSE,warning=FALSE}
log_file <- file(tempfile(), open = "wt")
sink(log_file)  # Redirect standard output
sink(log_file, type = "message")  # Redirect messages and warnings

fit_data = Maaslin2(
    input_data = filt_ileum_uni_data, 
    input_metadata = filt_ileum_uni_metadata, min_abundance = 0,
    min_prevalence = 0,min_variance = 0,
    output = file.path(path_maaslin,group1), max_significance = 0.05,
    fixed_effects = c('Group'),correction = "BH")

sink()  # Restore standard output
sink(type = "message")  # Restore messages
```

```{r, warning=FALSE}
volcano1 <- volcano_plot_maaslin(fit_data,filt_ileum_uni_taxa) + 
            ggtitle(comparison_title)

volcano_1
```

##### Group - Intersection

```{r, warning=FALSE}
intersection_results <- group_intersection(group, list_intersections, list_venns,
                                           linda.output, fit_data,
                                           raw_linda_results_genus, 
                                           segment = segment,
                                           level=level)

list_intersections <- intersection_results[[1]]
list_venns <- intersection_results[[2]]
venn <- intersection_results[[3]]

# show the results
venn
```

##### Basic statistics

```{r, warning=FALSE}
uni_df <-  merge(basic_univariate_statistics(linda_data,group),
                 raw_linda_results_genus[[segment]][[group1]],
                 by="SeqID",all=TRUE)

uni_df[["final_sig"]] <- uni_df$SeqID %in% list_intersections[[paste(segment,level,comparison_name)]][["SeqID"]]
uni_statistics[[segment]][[paste(level,comparison_name)]] <- uni_df

# for comparison
new_name <- comparison_name
addWorksheet(wb, sheetName = new_name)
writeData(wb, sheet = new_name, uni_df, rowNames=FALSE)
```


#### Saving results

```{r, warning=FALSE}
# ALL DATA
saveWorkbook(wb,file.path(path,paste0("uni_analysis_wb_",segment,".xlsx")),
             overwrite = TRUE)

# SIGNIFICANT taxa

write.xlsx(list_intersections[grepl(segment,names(list_intersections))] %>%
            `names<-`(gsub(segment, "", names(
              list_intersections[grepl(segment,names(list_intersections))]))),
           file.path(path,paste0("significant_taxa_",segment,".xlsx")))
```

### Supplementary Analysis

```{r, warning=FALSE}
supplements_uni <- list()
supplements_wb <- createWorkbook()
```


#### Phylum level

```{r, warning=FALSE}
level="phylum"
```

```{r, warning=FALSE}
path_maaslin="../intermediate_files/maaslin/Q4/Phylum/"
```

```{r, warning=FALSE}
raw_linda_results_phylum <- list();
raw_linda_results_phylum[[segment]] <- list()
linda_results_phylum <- list(); 
linda_results_phylum[[segment]] <- list()
```

Aggregate taxa

```{r, warning=FALSE}
phylum_data <- aggregate_taxa(ileum_asv_tab,
                             ileum_taxa_tab,
                             taxonomic_level = "Phylum")

ileum_phylum_tab <- phylum_data[[1]]
ileum_phylum_taxa_tab <- phylum_data[[2]]

```

##### low vs high

```{r, warning=FALSE}
group <- c("low","high")
comparison_name <- paste0(group[1], " vs ",group[2])
```

```{r, echo=FALSE}
comparison_title <- sub("(\\w+) vs (\\w+)", "\\2 and \\1", comparison_name)
```

**linDA**

```{r, fig.width=15, fig.height=5, fig.fullwidth=TRUE}

# prepare the data
linda_data <- binomial_prep(ileum_phylum_tab,
                            ileum_phylum_taxa_tab,
                            ileum_metadata,
                            group, usage="linDA")

filt_ileum_uni_data <- linda_data[[1]]
filt_ileum_uni_taxa <- linda_data[[2]]
filt_ileum_uni_metadata <- linda_data[[3]]

# fit the model
linda.obj <- linda(filt_ileum_uni_data, 
                   filt_ileum_uni_metadata, 
                   formula = '~ Group')

linda.output <- linda.obj$output
linda.output <- linda_renaming(linda.output, group)

# save the results
group1 <- paste0(group[1], " vs ","Group",group[2])

for (grp in c(group1)){
  raw_linda_results_phylum[[segment]][[grp]] <- 
    rawlinda.df(linda.output,
                grp,
                filt_ileum_uni_data,
                filt_ileum_uni_taxa)
  
  linda_results_phylum[[segment]][[grp]] <- 
    linda.df(linda.output,
             grp,
             filt_ileum_uni_data,
             filt_ileum_uni_taxa)
}

```

```{r, fig.width=5, fig.height=5, fig.fullwidth=TRUE}
# volcano plot
volcano_1 <- volcano_plot_linda(linda.output, 
                                group1, 
                                taxa_table = filt_ileum_uni_taxa) + 
            ggtitle(comparison_title)

volcano_1
```

**MaAsLin2**

```{r, echo=FALSE,results='hide',message=FALSE,warning=FALSE}
log_file <- file(tempfile(), open = "wt")
sink(log_file)  # Redirect standard output
sink(log_file, type = "message")  # Redirect messages and warnings

fit_data = Maaslin2(
    input_data = filt_ileum_uni_data, 
    input_metadata = filt_ileum_uni_metadata, min_abundance = 0,
    min_prevalence = 0,min_variance = 0,
    output = file.path(path_maaslin,group1), max_significance = 0.05,
    fixed_effects = c('Group'),correction = "BH")

sink()  # Restore standard output
sink(type = "message")  # Restore messages
```

```{r, warning=FALSE}
volcano1 <- volcano_plot_maaslin(fit_data,filt_ileum_uni_taxa) + 
            ggtitle(comparison_title)

volcano1
```

Group - Intersection

```{r, warning=FALSE}
intersection_results <- group_intersection(group, list_intersections, list_venns,
                                           linda.output, fit_data,
                                           raw_linda_results_phylum, 
                                           segment = segment,
                                           level=level)

list_intersections <- intersection_results[[1]]
list_venns <- intersection_results[[2]]
venn <- intersection_results[[3]]

# show the results
venn
```


Basic statistics

```{r, warning=FALSE}
uni_df <-  merge(basic_univariate_statistics(linda_data,group),
                 raw_linda_results_phylum[[segment]][[group1]],
                 by="SeqID",all=TRUE)

uni_df[["final_sig"]] <- uni_df$SeqID %in% list_intersections[[paste(segment,level,comparison_name)]][["SeqID"]]
uni_statistics[[segment]][[paste(level,comparison_name)]] <- uni_df

# for comparison
new_name <- paste(level,comparison_name)
addWorksheet(supplements_wb, sheetName = new_name)
writeData(supplements_wb, sheet = new_name, uni_df, rowNames=FALSE)
```


## Machine learning

```{r, warning=FALSE}
path = "../results/Q4/models"
```

### ElasticNet

```{r, warning=FALSE}
model="enet"
```

#### ASV level

```{r, warning=FALSE}
level="ASV"
```

#####low vs high

```{r, warning=FALSE}
group <- c("low","high")
comparison_name <- paste0(group[1], " vs ",group[2])
```

```{r, warning=FALSE}
model_name <- paste(comparison_name,level,segment)

# prepare the data
filt_ileum_uni_data <- binomial_prep(ileum_asv_tab,
                                     ileum_taxa_tab,
                                     ileum_metadata,
                                     group, usage="ml_clr")

# fit the model
enet_model <- glmnet_binomial(filt_ileum_uni_data,
                              sample_method = "atypboot",
                              outcome="Group", 
                              N=10,
                              reuse=FALSE,
                              file=model_name,
                              Q="Q4")

# ROC curve
roc_c <- roc_curve(enet_model, group)

# save the results
models_summ <- list()
models_cm <- list()
betas <- list()
roc_cs <- list()

models_summ[[model_name]] <- enet_model$model_summary
models_cm[[model_name]] <- enet_model$conf_matrices$original
roc_cs[[model_name]] <- enet_model$kfold_rocobjs
betas[[model_name]] <- as.matrix(enet_model$betas)

# see the results
enet_model$model_summary %>% t()
enet_model$conf_matrices
enet_model$plot
roc_c
```

#### Genus level

```{r, warning=FALSE}
level="genus"
```

Aggregate taxa

```{r, warning=FALSE}
genus_data <- aggregate_taxa(ileum_asv_tab,
                             ileum_taxa_tab,
                             taxonomic_level = level)

ileum_genus_tab <- genus_data[[1]]
ileum_genus_taxa_tab <- genus_data[[2]]
```

##### low vs high

```{r, warning=FALSE}
group <- c("low","high")
comparison_name <- paste0(group[1], " vs ",group[2])
```

```{r, warning=FALSE}
model_name <- paste(comparison_name,level,segment)

# prepare the data
filt_ileum_uni_data <- binomial_prep(ileum_genus_tab,
                                     ileum_genus_taxa_tab,
                                     ileum_metadata,
                                     group, 
                                     usage="ml_clr")

# fit the model
enet_model <- glmnet_binomial(filt_ileum_uni_data,
                              sample_method = "atypboot",
                              outcome="Group", 
                              N=10,
                              reuse=FALSE,
                              file=model_name,
                              Q="Q4")

# ROC
roc_c <- roc_curve(enet_model, group)

# save the results
models_summ[[model_name]] <- enet_model$model_summary
models_cm[[model_name]] <- enet_model$conf_matrices$original
roc_cs[[model_name]] <- enet_model$kfold_rocobjs
betas[[model_name]] <- as.matrix(enet_model$betas)

# see the results
enet_model$model_summary %>% t()
enet_model$conf_matrices
enet_model$plot

roc_c
```

#### Saving results

```{r, warning=FALSE}
models_summ_df_ileum <- do.call(rbind, 
  models_summ[grep(segment,names(models_summ),value = TRUE)])

write.csv(models_summ_df_ileum,file.path(path,paste0("elastic_net_",segment,".csv")))
```

```{r}
k
```

### Supplementary models

```{r, warning=FALSE}
supplements_models <- list()
```

#### CLR-transformed data

##### kNN

```{r, warning=FALSE}
model="knn"
```

###### ASV level

```{r, warning=FALSE}
level="ASV"
```

***low vs high***

```{r, warning=FALSE}
group <- c("low","high")
comparison_name <- paste0(group[1], " vs ",group[2])
```

```{r, warning=FALSE}
model_name <- paste(comparison_name,level,segment)

# prepare the data
filt_ileum_uni_data <- binomial_prep(ileum_asv_tab,
                                     ileum_taxa_tab,
                                     ileum_metadata,
                                     group, usage="ml_clr")

# fit the model
knn_model <- knn_binomial(filt_ileum_uni_data,
                              sample_method = "atypboot",
                              outcome="Group", 
                              N=10,
                              reuse=FALSE,
                              file=model_name,
                              Q="Q4")

# ROC curve
roc_c <- roc_curve(knn_model, group)

# save the results
supplements_models[["models_summ"]][["knn_model"]][[model_name]] <- knn_model$model_summary
supplements_models[["roc_cs"]][["knn_model"]][[model_name]] <- knn_model$kfold_rocobjs

# see the results
knn_model$model_summary %>% t()
roc_c
```


###### Genus level

```{r, warning=FALSE}
level="genus"
```

Aggregate taxa

```{r, warning=FALSE}
genus_data <- aggregate_taxa(ileum_asv_tab,
                             ileum_taxa_tab,
                             taxonomic_level = level)

ileum_genus_tab <- genus_data[[1]]
ileum_genus_taxa_tab <- genus_data[[2]]
```

***low vs high***

```{r, warning=FALSE}
group <- c("low","high")
comparison_name <- paste0(group[1], " vs ",group[2])
```

```{r, warning=FALSE}
model_name <- paste(comparison_name,level,segment)

# prepare the data
filt_ileum_uni_data <- binomial_prep(ileum_genus_tab,
                                     ileum_genus_taxa_tab,
                                     ileum_metadata,
                                     group, 
                                     usage="ml_clr")

# fit the model
knn_model <- knn_binomial(filt_ileum_uni_data,
                              sample_method = "atypboot",
                              outcome="Group", 
                              N=10,
                              reuse=FALSE,
                              file=model_name,
                              Q="Q4")

# ROC curve
roc_c <- roc_curve(knn_model, group)

# save the results
supplements_models[["models_summ"]][["knn_model"]][[model_name]] <- knn_model$model_summary
supplements_models[["roc_cs"]][["knn_model"]][[model_name]] <- knn_model$kfold_rocobjs


# see the results
knn_model$model_summary %>% t()
roc_c
```


##### RF

```{r, warning=FALSE}
model="rf"
```

###### ASV level

```{r, warning=FALSE}
level="ASV"
```

***low vs high***

```{r, warning=FALSE}
group <- c("low","high")
comparison_name <- paste0(group[1], " vs ",group[2])
```

```{r, warning=FALSE}
model_name <- paste(comparison_name,level,segment)

# prepare the data
filt_ileum_uni_data <- binomial_prep(ileum_asv_tab,
                                     ileum_taxa_tab,
                                     ileum_metadata,
                                     group, usage="ml_clr")

# fit the model
rf_model <- rf_binomial(filt_ileum_uni_data,
                              sample_method = "atypboot",
                              outcome="Group", 
                              N=10,
                              reuse=FALSE,
                              file=model_name,
                              Q="Q4")

# ROC curve
roc_c <- roc_curve(rf_model, group)

# save the results
supplements_models[["models_summ"]][["rf_model"]][[model_name]] <- rf_model$model_summary
supplements_models[["roc_cs"]][["rf_model"]][[model_name]] <- rf_model$kfold_rocobjs

# see the results
rf_model$model_summary %>% t()
roc_c
```


###### Genus level

```{r, warning=FALSE}
level="genus"
```

Aggregate taxa

```{r, warning=FALSE}
genus_data <- aggregate_taxa(ileum_asv_tab,
                             ileum_taxa_tab,
                             taxonomic_level = level)

ileum_genus_tab <- genus_data[[1]]
ileum_genus_taxa_tab <- genus_data[[2]]
```

***low vs high***

```{r, warning=FALSE}
group <- c("low","high")
comparison_name <- paste0(group[1], " vs ",group[2])
```

```{r, warning=FALSE}
model_name <- paste(comparison_name,level,segment)

# prepare the data
filt_ileum_uni_data <- binomial_prep(ileum_genus_tab,
                                     ileum_genus_taxa_tab,
                                     ileum_metadata,
                                     group, 
                                     usage="ml_clr")

# fit the model
rf_model <- rf_binomial(filt_ileum_uni_data,
                              sample_method = "atypboot",
                              outcome="Group", 
                              N=10,
                              reuse=FALSE,
                              file=model_name,
                              Q="Q4")

# ROC curve
roc_c <- roc_curve(rf_model, group)

# save the results
supplements_models[["models_summ"]][["rf_model"]][[model_name]] <- rf_model$model_summary
supplements_models[["roc_cs"]][["rf_model"]][[model_name]] <- rf_model$kfold_rocobjs

# see the results
rf_model$model_summary %>% t()
roc_c
```


##### Gboost

```{r, warning=FALSE}
model="gb"
```

###### ASV level

```{r, warning=FALSE}
level="ASV"
```

***low vs high***

```{r, warning=FALSE}
group <- c("low","high")
comparison_name <- paste0(group[1], " vs ",group[2])
```

```{r, warning=FALSE}
model_name <- paste(comparison_name,level,segment)

# prepare the data
filt_ileum_uni_data <- binomial_prep(ileum_asv_tab,
                                     ileum_taxa_tab,
                                     ileum_metadata,
                                     group, usage="ml_clr")

# fit the model
gbm_model <- gbm_binomial(filt_ileum_uni_data,
                              sample_method = "atypboot",
                              outcome="Group", 
                              N=10,
                              reuse=FALSE,
                              file=model_name,
                              Q="Q4")

# ROC curve
roc_c <- roc_curve(gbm_model, group)

# save the results
supplements_models[["models_summ"]][["gbm_model"]][[model_name]] <- gbm_model$model_summary
supplements_models[["roc_cs"]][["gbm_model"]][[model_name]] <- gbm_model$kfold_rocobjs

# see the results
gbm_model$model_summary %>% t()
roc_c
```

###### Genus level

```{r, warning=FALSE}
level="genus"
```

Aggregate taxa

```{r, warning=FALSE}
genus_data <- aggregate_taxa(ileum_asv_tab,
                             ileum_taxa_tab,
                             taxonomic_level = level)

ileum_genus_tab <- genus_data[[1]]
ileum_genus_taxa_tab <- genus_data[[2]]
```

***low vs high***

```{r, warning=FALSE}
group <- c("low","high")
comparison_name <- paste0(group[1], " vs ",group[2])
```

```{r, warning=FALSE}
model_name <- paste(comparison_name,level,segment)

# prepare the data
filt_ileum_uni_data <- binomial_prep(ileum_genus_tab,
                                     ileum_genus_taxa_tab,
                                     ileum_metadata,
                                     group, 
                                     usage="ml_clr")

# fit the model
gbm_model <- gbm_binomial(filt_ileum_uni_data,
                              sample_method = "atypboot",
                              outcome="Group", 
                              N=10,
                             reuse=FALSE,
                              file=model_name,
                              Q="Q4")

# ROC curve
roc_c <- roc_curve(gbm_model, group)

# save the results
supplements_models[["models_summ"]][["gbm_model"]][[model_name]] <- gbm_model$model_summary
supplements_models[["roc_cs"]][["gbm_model"]][[model_name]] <- gbm_model$kfold_rocobjs

# see the results
gbm_model$model_summary %>% t()
roc_c
```


#### Saving results

```{r, warning=FALSE}
models_list <- list()

for (model_name in names(supplements_models$models_summ)){
  df <- do.call(rbind, supplements_models$models_summ[[model_name]])
  models_list[[model_name]] <- df
}

write.xlsx(models_list,
           file=file.path(path,paste0("supplements_models_",segment,".xlsx")),
           rowNames=TRUE)

```


## Results overview

#### Alpha diversity

```{r, warning=FALSE}
pc_observed[[segment]]
pc_shannon[[segment]]
pc_simpson[[segment]]
pc_pielou[[segment]]
```

Plots

```{r, fig.width=4, fig.height=5, fig.fullwidth=TRUE}
alpha_div_plots[[paste(segment,"Country")]]
```


#### Beta diversity

**Main results**

```{r, warning=FALSE}
pairwise_aitchison_raw[[paste("genus", segment)]]
```

*PCA*

```{r, fig.width=5, fig.height=3, fig.fullwidth=TRUE, eval=TRUE}
pca_plots_list[[paste(segment,"genus custom")]]
```

**Supplements**

```{r, warning=FALSE}
knitr::kable(supplements_beta[!grepl("PCoA",names(supplements_beta)) & (grepl("genus",names(supplements_beta)))],
             digits = 3,
             caption = "Supplementary PERMANOVA results: Bray-curtis, Jaccard distances")
```

*PCA*

```{r, fig.width=8, fig.height=7, fig.fullwidth=TRUE, eval=FALSE}
ggarrange(plotlist = supplements_beta[grepl("PCoA",names(supplements_beta))],
          labels=names(supplements_beta[grepl("PCoA",names(supplements_beta))]),
          font.label = list(size=5,face="plain"),
          ncol=2,nrow=3)
```

#### Univariate analysis

**Number of significant taxa**

```{r, warning=FALSE}
knitr::kable(as.data.frame(lapply(list_intersections,nrow)) %>% t() %>% 
  `colnames<-`("Count") %>% 
  `rownames<-`(c(names(list_intersections))),caption="Number of significant taxa")
```

#### Machine learning

**Main models**

```{r, warning=FALSE}
knitr::kable(models_summ_df_ileum %>% dplyr::select(
"alpha","lambda",
"auc_optimism_corrected",
"auc_optimism_corrected_CIL",
"auc_optimism_corrected_CIU"),
             digits=2,caption="Elastic net results")
```

**ROC - Genus level**

```{r, warning=FALSE}
roc_curve_all_custom(roc_cs[2],Q="Q4",
                     model_name="enet_model")
```

**Supplementary models**


```{r, warning=FALSE}
# Build final dataframe
models_list[["enet_model"]] <- models_summ_df_ileum
final_df <- tibble(row_names = rownames(models_list[[1]]))

# Loop through models and extract required values
for (model_name in names(models_list)) {
  model_df <- models_list[[model_name]]
  
  # Combine AUC_optimism_corrected with its CI values
  final_df[[model_name]] <- paste0(
    round(model_df$auc_optimism_corrected, 3), 
    " (", round(model_df$auc_optimism_corrected_CIL, 3), "; ", 
    round(model_df$auc_optimism_corrected_CIU, 3), ")"
  )
}

knitr::kable(final_df, caption="All models")
```

**ROC - genus**

```{r, fig.width=10, fig.height=8, fig.fullwidth=TRUE, warning=FALSE}
rocs_list <- supplements_models$roc_cs
rocs_list[["enet_model"]] <- roc_cs
plot_list <- list()

for (model_name in names(rocs_list)) {
  plot_list[[model_name]] <- roc_curve_all_custom(rocs_list[[model_name]][2],
                       Q="Q4",
                       model_name=model_name)
}

p <- ggarrange(plotlist = plot_list,labels = names(rocs_list),font.label = list(face="plain",size=5))
```

```{r,results='hide', echo=FALSE, warning=FALSE}
pdf("../figures/Q4/models_terminal_ileum.pdf",
    height =10,width = 10)
p
dev.off()
```

# Data Analysis - Colon

```{r, warning=FALSE}
segment="colon"
```

## Filtering

Rules: 

  - sequencing depth \> 10000 
  
  - nearZeroVar() with default settings 

**Rarefaction Curve**

```{r, warning=FALSE}
path="../intermediate_files/rarecurves"
seq_depth_threshold <- 10000
```

```{r, eval = FALSE}
ps <- construct_phyloseq(colon_asv_tab,colon_taxa_tab,colon_metadata)
rareres <- get_rarecurve(obj=ps, chunks=500)
save(rareres,file = file.path(path,"rarefaction_colon.Rdata"))
```

```{r, warning=FALSE}
load(file.path(path,"rarefaction_colon.Rdata"))

prare <- ggrarecurve(obj=rareres,
                      factorNames="Country",
                      indexNames=c("Observe")) + 
  theme_bw() +
  theme(axis.text=element_text(size=8), panel.grid=element_blank(),
        strip.background = element_rect(colour=NA,fill="grey"),
        strip.text.x = element_text(face="bold")) + 
  geom_vline(xintercept = seq_depth_threshold, 
             linetype="dashed", 
             color = "red") + 
  xlim(0, 20000)

prare
```

**Library size**

```{r, fig.width=5, fig.height=4, fig.fullwidth=TRUE, warning=FALSE}
read_counts(colon_asv_tab, line = c(5000,10000))
```

### Sequencing depth

```{r, warning=FALSE}
data_filt <- seq_depth_filtering(colon_asv_tab,
                                 colon_taxa_tab,
                                 colon_metadata,
                                 seq_depth_threshold = 10000)

filt_colon_asv_tab <- data_filt[[1]]; alpha_colon_asv_tab <- filt_colon_asv_tab
filt_colon_taxa_tab <- data_filt[[2]]; alpha_colon_taxa_tab <- filt_colon_taxa_tab
filt_colon_metadata <- data_filt[[3]]; alpha_colon_metadata <- filt_colon_metadata

seq_step <- dim(filt_colon_asv_tab)[1]
```

**Library size**

```{r, fig.width=5, fig.height=4, fig.fullwidth=TRUE, warning=FALSE}
read_counts(filt_colon_asv_tab,line = c(5000,10000))
```

### NearZeroVar

```{r, warning=FALSE}
data_filt <- nearzerovar_filtering(filt_colon_asv_tab, 
                                   filt_colon_taxa_tab,
                                   filt_colon_metadata)

filt_colon_asv_tab <- data_filt[[1]]
filt_colon_taxa_tab <- data_filt[[2]]
nearzero_step <- dim(filt_colon_asv_tab)[1]
```

Library size

```{r, fig.width=5, fig.height=4, fig.fullwidth=TRUE, warning=FALSE}
read_counts(filt_colon_asv_tab,line = c(5000,10000))
```

### Final Counts

```{r, warning=FALSE}
final_counts_filtering(colon_asv_tab,
                       filt_colon_asv_tab,
                       filt_colon_metadata,
                       seq_step, 0, nearzero_step) %>% `colnames<-`("Count")
```

## Alpha diversity

```{r, warning=FALSE}
path = "../results/Q4/alpha_diversity"
```

**Calculation**

```{r, warning=FALSE}
# Construct MPSE object
alpha_colon_metadata$Sample <- alpha_colon_metadata$SampleID
colon_mpse <- as.MPSE(construct_phyloseq(alpha_colon_asv_tab,
                                         alpha_colon_taxa_tab,
                                         alpha_colon_metadata))

colon_mpse %<>% mp_rrarefy(raresize = 10000,seed = 123)

# Calculate alpha diversity - rarefied counts
colon_mpse %<>% mp_cal_alpha(.abundance=RareAbundance, force=TRUE)
```

```{r, warning=FALSE}

# preparing data frame
alpha_data <- data.frame(SampleID=colon_mpse$Sample.x,
                         Richness=colon_mpse$Observe,
                         Shannon=colon_mpse$Shannon,
                         Simpson=colon_mpse$Simpson,
                         Pielou=colon_mpse$Pielou,
                         Group=colon_mpse$Nancy,
                         Country=colon_mpse$Country,
                         Patient=colon_mpse$Patient)

write.csv(alpha_data,file.path(path,paste0("alpha_indices_",segment,".csv")),
          row.names = FALSE)

alpha_data %<>% drop_na()
```

### Plots

```{r, fig.width=6, fig.height=4, fig.fullwidth=TRUE}
p_boxplot_alpha <- alpha_diversity_custom_2(alpha_data)

# save the results
alpha_div_plots[[paste(segment,"Custom")]] <- p_boxplot_alpha

# see the results
p_boxplot_alpha
```

```{r,results='hide', echo=FALSE}
pdf("../figures/Q4/alpha_diversity_colon.pdf",
    height =4,width = 6)
p_boxplot_alpha
dev.off()
```

### Linear Model

```{r, warning=FALSE}
path = "../results/Q4/alpha_diversity"
alpha_data <- read.csv(file.path(path,paste0("alpha_indices_",segment,".csv")))
```

**Richness**

```{r, warning=FALSE}
results_model <- pairwise.lm(formula = "Richness ~ Group + (1|Patient)",
                             factors=alpha_data$Group,
                             data=alpha_data)

# check interaction
if (!is.data.frame(results_model)){
  results_model_observe <- results_model[[1]]
  results_model_observe_emeans <- results_model[[2]]
} else {
  results_model_observe <- results_model
  results_model_observe_emeans <- NA
}

# save the results
pc_observed[[segment]] <- results_model_observe
```

```{r, warning=FALSE}
# see the results
knitr::kable(results_model_observe,digits = 3,
caption = "Raw results of linear model of richness estimation.")

knitr::kable(results_model_observe_emeans,digits = 3,
caption = "Raw results of independent country analysis")
```

**Shannon**

```{r, warning=FALSE}
results_model <- pairwise.lm(formula = "Shannon ~ Group + (1|Patient)",
                             factors=alpha_data$Group,
                             data=alpha_data)

# check interaction
if (!is.data.frame(results_model)){
  results_model_shannon <- results_model[[1]]
  results_model_shannon_emeans <- results_model[[2]]
} else {
  results_model_shannon <- results_model
  results_model_shannon_emeans <- NA
}

# save the results
pc_shannon[[segment]] <- as.data.frame(results_model_shannon)

```

```{r, warning=FALSE}
# see the results
knitr::kable(results_model_shannon,digits = 3,
caption = "Raw results of linear model of Shannon estimation.")

knitr::kable(results_model_shannon_emeans,digits = 3,
caption = "Raw results of independent country analysis")
```

**Simpson**

```{r, warning=FALSE}
results_model <- pairwise.lm(formula = "Simpson ~ Group + (1|Patient)",
                                     factors=alpha_data$Group,
                                     data=alpha_data)

# check interaction
if (!is.data.frame(results_model)){
  results_model_simpson <- results_model[[1]]
  results_model_simpson_emeans <- results_model[[2]]
} else {
  results_model_simpson <- results_model
  results_model_simpson_emeans <- NA
}


# save the results
pc_simpson[[segment]] <- as.data.frame(results_model_simpson)
```

```{r, warning=FALSE}
# see the results
knitr::kable(results_model_simpson,digits = 3,
caption = "Raw results of linear model of Simpson estimation.")

knitr::kable(results_model_simpson_emeans,digits = 3,
caption = "Raw results of independent country analysis")
```

### Saving results

```{r, warning=FALSE}
alpha_list <- list(
  Richness=pc_observed[[segment]],
  Shannon=pc_shannon[[segment]])
                   
write.xlsx(alpha_list, 
           file = file.path(path,paste0("alpha_diversity_results_",segment,".xlsx")))
```


## Beta diversity

Calculating Aitchison distance (euclidean distance on clr-transformed
data), both at ASV and genus level.

### Main analysis

**Genus level, Aitchison distance**

```{r, warning=FALSE}
level="genus"
```

```{r, warning=FALSE}
path = "../results/Q4/beta_diversity"
```

Aggregation, filtering

```{r, warning=FALSE}
# Aggregation
genus_data <- aggregate_taxa(colon_asv_tab,
                             colon_taxa_tab,
                             taxonomic_level=level,
                             names=TRUE)
# Filtration
filt_data <- filtering_steps(genus_data[[1]],
                             genus_data[[2]],
                             colon_metadata,
                             seq_depth_threshold=10000)

filt_colon_genus_tab <- filt_data[[1]]
filt_colon_genus_taxa <- filt_data[[2]]
filt_colon_metadata <- filt_data[[3]]

filt_colon_metadata %<>% dplyr::mutate(Group=Nancy)
```

##### PERMANOVA

```{r, warning=FALSE}
pairwise_df <- filt_colon_genus_tab %>% column_to_rownames("SeqID") %>% t()

# main effect
pp_main <- pairwise.adonis(pairwise_df,
                           filt_colon_metadata$Nancy,
                           covariate = NULL, sim.method = "robust.aitchison", p.adjust.m="BH",patients = filt_colon_metadata$Patient)


# tidy the results
pp_factor <- pp_main

cols <- c("pairs","Df","SumsOfSqs", "F.Model","R2","p.value", "p.adjusted", "sig")
colnames(pp_factor) <- cols

# save raw results
pairwise_aitchison_raw[[paste(level, segment)]] <- pp_factor
```

```{r, warning=FALSE}
# see the results
knitr::kable(pp_factor,digits = 3,caption = "PERMANOVA, GROUP separation")
```

##### Plots

```{r, warning=FALSE}
p <- pca_plot_custom(filt_colon_genus_tab,
                                 filt_colon_genus_taxa,
                                 filt_colon_metadata,
                                 show_boxplots = TRUE,
                                 variable = "Group", size=3, show_legend=FALSE)

# save the results
pca_plots_list[[paste(segment,level,"custom")]] <- p

# see the results
p
```

```{r,results='hide', echo=FALSE}
pdf("../figures/Q4/beta_diversity_colon.pdf",
    height =5,width = 5)
p
dev.off()
```

#### Saving results

```{r, warning=FALSE}
write.xlsx(pairwise_aitchison_raw[[paste(level, segment)]], 
           file = file.path(path,
           paste0("beta_diversity_results_", segment,".xlsx")))
```

### Supplementary analysis


#### Genus level

```{r, warning=FALSE}
level="genus"
```

##### Bray-Curtis

**PERMANOVA**

```{r, warning=FALSE}
pairwise_df <- filt_colon_genus_tab %>% column_to_rownames("SeqID") %>% t()

# main effect
pp_main <- pairwise.adonis(pairwise_df,filt_colon_metadata$Group,covariate = NULL, sim.method = "bray", p.adjust.m="BH", filt_colon_metadata$Patient)


# tidy the results
pp_factor <- pp_main

cols <- c("pairs","Df","SumsOfSqs", "F.Model","R2","p.value", "p.adjusted", "sig")
colnames(pp_factor) <- cols

# save raw results
supplements_beta[[paste("bray",level,segment)]] <- pp_factor
```

```{r, warning=FALSE}
# see the results
knitr::kable(pp_factor,digits = 3,caption = "PERMANOVA, GROUP separation")

```


**Plots**

```{r, warning=FALSE}
p <- pca_plot_custom(filt_colon_genus_tab,
                                 filt_colon_genus_taxa,
                                 filt_colon_metadata,
                                 measure = "bray",
                                 show_boxplots = TRUE,
                                 variable = "Group", size=3, show_legend=TRUE)

# save the results
supplements_beta[[paste("PCoA bray",level,segment)]] <- p

# see the results
p
```

##### Jaccard

**PERMANOVA**

```{r, warning=FALSE}
pairwise_df <- filt_colon_genus_tab %>% column_to_rownames("SeqID") %>% t()

# main effect
pp_main <- pairwise.adonis(pairwise_df,filt_colon_metadata$Group,covariate = NULL, sim.method = "jaccard", p.adjust.m="BH", filt_colon_metadata$Patient)


# tidy the results
pp_factor <- pp_main

cols <- c("pairs","Df","SumsOfSqs", "F.Model","R2","p.value", "p.adjusted", "sig")
colnames(pp_factor) <- cols

# save raw results
supplements_beta[[paste("jaccard",level,segment)]] <- pp_factor
```

```{r, warning=FALSE}
# see the results
knitr::kable(pp_factor,digits = 3,caption = "PERMANOVA, GROUP separation")
```



**Plots**

```{r, warning=FALSE}
p <- pca_plot_custom(filt_colon_genus_tab,
                                 filt_colon_genus_taxa,
                                 filt_colon_metadata,
                                 measure = "jaccard",
                                 show_boxplots = TRUE,
                                 variable = "Group", size=3, show_legend=TRUE)

# save the results
supplements_beta[[paste("PCoA jaccard",level,segment)]] <- p

# see the results
p
```

#### Saving results

```{r, warning=FALSE}
write.xlsx(supplements_beta[!grepl("PCoA",names(supplements_beta))],
           file = file.path(path,
           paste0("supplements_beta_diversity_", segment,".xlsx")))
```

## Univariate Analysis

### Main analysis

```{r, warning=FALSE}
level="genus"
```

```{r, warning=FALSE}
# needed paths
path = "../results/Q4/univariate_analysis"
path_maaslin=file.path("../intermediate_files/maaslin/Q4",level)
```

```{r, warning=FALSE}
# variables
raw_linda_results_genus[[segment]] <- list()
linda_results_genus[[segment]] <- list()

# country and interaction problems
list_intersections <- list()
list_venns <- list()
uni_statistics <- list()

# workbook for final df
wb <- createWorkbook()
```

Aggregate taxa

```{r, warning=FALSE}
genus_data <- aggregate_taxa(colon_asv_tab,
                             colon_taxa_tab,
                             taxonomic_level = level)

colon_genus_tab <- genus_data[[1]]
colon_genus_taxa_tab <- genus_data[[2]]

colon_genus_asv_taxa_tab <- create_asv_taxa_table(colon_genus_tab,
                                                  colon_genus_taxa_tab)
```

#### low vs high

```{r, warning=FALSE}
group <- c("low","high")
comparison_name <- paste0(group[1], " vs ",group[2])
```

```{r, echo=FALSE}
comparison_title <- sub("(\\w+) vs (\\w+)", "\\2 and \\1", comparison_name)
```

##### linDA

```{r, fig.width=15, fig.height=5, fig.fullwidth=TRUE}

colon_metadata %<>% dplyr::mutate(Group=Nancy)
# prepare the data
linda_data <- binomial_prep(colon_genus_tab,
                            colon_genus_taxa_tab,
                            colon_metadata,
                            group, usage="linDA")

filt_colon_uni_data <- linda_data[[1]]
filt_colon_uni_taxa <- linda_data[[2]]
filt_colon_uni_metadata <- linda_data[[3]]

# fit the model
linda.obj <- linda(filt_colon_uni_data, 
                   filt_colon_uni_metadata, 
                   formula = '~ Group + (1|Patient)')

linda.output <- linda.obj$output
linda.output <- linda_renaming(linda.output, group)

# save the results
group1 <- paste0(group[1], " vs ","Group",group[2])

for (grp in c(group1)){
  raw_linda_results_genus[[segment]][[grp]] <- 
    rawlinda.df(linda.output,
                grp,
                filt_colon_uni_data,
                filt_colon_uni_taxa)
  
  linda_results_genus[[segment]][[grp]] <- 
    linda.df(linda.output,
             grp,
             filt_colon_uni_data,
             filt_colon_uni_taxa)
}

```

```{r, fig.width=5, fig.height=5, fig.fullwidth=TRUE}
# volcano plot
volcano_1 <- volcano_plot_linda(linda.output, 
                                group1, 
                                taxa_table = filt_colon_uni_taxa) + 
            ggtitle(comparison_title)

volcano_1
```

##### MaAsLin2

```{r, echo=FALSE,results='hide',message=FALSE,warning=FALSE}
log_file <- file(tempfile(), open = "wt")
sink(log_file)  # Redirect standard output
sink(log_file, type = "message")  # Redirect messages and warnings

fit_data = Maaslin2(
    input_data = filt_colon_uni_data, 
    input_metadata = filt_colon_uni_metadata, min_abundance = 0,
    min_prevalence = 0,min_variance = 0,
    output = file.path(path_maaslin,group1), max_significance = 0.05,
    fixed_effects = c('Group'),fixed_effects = c("Patient"),correction = "BH")

sink()  # Restore standard output
sink(type = "message")  # Restore messages
```

```{r, warning=FALSE}
volcano1 <- volcano_plot_maaslin(fit_data,filt_colon_uni_taxa) + 
            ggtitle(comparison_title)

volcano_1
```

##### Group - Intersection

```{r, warning=FALSE}
intersection_results <- group_intersection(group, list_intersections, list_venns,
                                           linda.output, fit_data,
                                           raw_linda_results_genus, 
                                           segment = segment,
                                           level=level)

list_intersections <- intersection_results[[1]]
list_venns <- intersection_results[[2]]
venn <- intersection_results[[3]]

# show the results
venn
```

##### Basic statistics

```{r, warning=FALSE}
uni_df <-  merge(basic_univariate_statistics(linda_data,group),
                 raw_linda_results_genus[[segment]][[group1]],
                 by="SeqID",all=TRUE)

uni_df[["final_sig"]] <- uni_df$SeqID %in% list_intersections[[paste(segment,level,comparison_name)]][["SeqID"]]
uni_statistics[[segment]][[paste(level,comparison_name)]] <- uni_df

# for comparison
new_name <- comparison_name
addWorksheet(wb, sheetName = new_name)
writeData(wb, sheet = new_name, uni_df, rowNames=FALSE)
```


#### Saving results

```{r, warning=FALSE}
# ALL DATA
saveWorkbook(wb,file.path(path,paste0("uni_analysis_wb_",segment,".xlsx")),
             overwrite = TRUE)

# SIGNIFICANT taxa

write.xlsx(list_intersections[grepl(segment,names(list_intersections))] %>%
            `names<-`(gsub(segment, "", names(
              list_intersections[grepl(segment,names(list_intersections))]))),
           file.path(path,paste0("significant_taxa_",segment,".xlsx")))
```

### Supplementary Analysis

```{r, warning=FALSE}
supplements_uni <- list()
supplements_wb <- createWorkbook()
```


#### Phylum level

```{r, warning=FALSE}
level="phylum"
```

```{r, warning=FALSE}
path_maaslin="../intermediate_files/maaslin/Q4/Phylum/"
```

```{r, warning=FALSE}
raw_linda_results_phylum[[segment]] <- list()
linda_results_phylum[[segment]] <- list()
```

Aggregate taxa

```{r, warning=FALSE}
phylum_data <- aggregate_taxa(colon_asv_tab,
                             colon_taxa_tab,
                             taxonomic_level = "Phylum")

colon_phylum_tab <- phylum_data[[1]]
colon_phylum_taxa_tab <- phylum_data[[2]]

```

##### low vs high

```{r, warning=FALSE}
group <- c("low","high")
comparison_name <- paste0(group[1], " vs ",group[2])
```

```{r, echo=FALSE}
comparison_title <- sub("(\\w+) vs (\\w+)", "\\2 and \\1", comparison_name)
```

**linDA**

```{r, fig.width=15, fig.height=5, fig.fullwidth=TRUE}

# prepare the data
linda_data <- binomial_prep(colon_phylum_tab,
                            colon_phylum_taxa_tab,
                            colon_metadata,
                            group, usage="linDA")

filt_colon_uni_data <- linda_data[[1]]
filt_colon_uni_taxa <- linda_data[[2]]
filt_colon_uni_metadata <- linda_data[[3]]

# fit the model
linda.obj <- linda(filt_colon_uni_data, 
                   filt_colon_uni_metadata, 
                   formula = '~ Group + (1|Patient)')

linda.output <- linda.obj$output
linda.output <- linda_renaming(linda.output, group)

# save the results
group1 <- paste0(group[1], " vs ","Group",group[2])

for (grp in c(group1)){
  raw_linda_results_phylum[[segment]][[grp]] <- 
    rawlinda.df(linda.output,
                grp,
                filt_colon_uni_data,
                filt_colon_uni_taxa)
  
  linda_results_phylum[[segment]][[grp]] <- 
    linda.df(linda.output,
             grp,
             filt_colon_uni_data,
             filt_colon_uni_taxa)
}

```

```{r, fig.width=5, fig.height=5, fig.fullwidth=TRUE}
# volcano plot
volcano_1 <- volcano_plot_linda(linda.output, 
                                group1, 
                                taxa_table = filt_colon_uni_taxa) + 
            ggtitle(comparison_title)

volcano_1
```

**MaAsLin2**

```{r, echo=FALSE,results='hide',message=FALSE,warning=FALSE}
log_file <- file(tempfile(), open = "wt")
sink(log_file)  # Redirect standard output
sink(log_file, type = "message")  # Redirect messages and warnings

fit_data = Maaslin2(
    input_data = filt_colon_uni_data, 
    input_metadata = filt_colon_uni_metadata, min_abundance = 0,
    min_prevalence = 0,min_variance = 0,
    output = file.path(path_maaslin,group1), max_significance = 0.05,
    fixed_effects = c('Group'),random_effects = c("Patient"), correction = "BH")

sink()  # Restore standard output
sink(type = "message")  # Restore messages
```

```{r, warning=FALSE}
volcano1 <- volcano_plot_maaslin(fit_data,filt_colon_uni_taxa) + 
            ggtitle(comparison_title)

volcano1
```

Group - Intersection

```{r, warning=FALSE}
intersection_results <- group_intersection(group, list_intersections, list_venns,
                                           linda.output, fit_data,
                                           raw_linda_results_phylum, 
                                           segment = segment,
                                           level=level)

list_intersections <- intersection_results[[1]]
list_venns <- intersection_results[[2]]
venn <- intersection_results[[3]]

# show the results
venn
```


Basic statistics

```{r, warning=FALSE}
uni_df <-  merge(basic_univariate_statistics(linda_data,group),
                 raw_linda_results_phylum[[segment]][[group1]],
                 by="SeqID",all=TRUE)

uni_df[["final_sig"]] <- uni_df$SeqID %in% list_intersections[[paste(segment,level,comparison_name)]][["SeqID"]]
uni_statistics[[segment]][[paste(level,comparison_name)]] <- uni_df

# for comparison
new_name <- paste(level,comparison_name)
addWorksheet(supplements_wb, sheetName = new_name)
writeData(supplements_wb, sheet = new_name, uni_df, rowNames=FALSE)
```


## Machine learning

```{r, warning=FALSE}
path = "../results/Q4/models"
```

### ElasticNet

```{r, warning=FALSE}
model="enet"
```

#### ASV level

```{r, warning=FALSE}
level="ASV"
```

#####low vs high

```{r, warning=FALSE}
group <- c("low","high")
comparison_name <- paste0(group[1], " vs ",group[2])
```

```{r, warning=FALSE}
model_name <- paste(comparison_name,level,segment)

# prepare the data
filt_colon_uni_data <- binomial_prep(colon_asv_tab,
                                     colon_taxa_tab,
                                     colon_metadata,
                                     group, usage="ml_clr",patient = TRUE)

# fit the model
enet_model <- glmnet_binomial(filt_colon_uni_data,
                              sample_method = "atypboot",
                              outcome="Group", 
                              N=10,
                              reuse=FALSE,
                              file=model_name,clust_var = "Patient",
                              Q="Q4")

# ROC curve
roc_c <- roc_curve(enet_model, group)

# save the results
models_summ <- list()
models_cm <- list()
betas <- list()
roc_cs <- list()

models_summ[[model_name]] <- enet_model$model_summary
models_cm[[model_name]] <- enet_model$conf_matrices$original
roc_cs[[model_name]] <- enet_model$kfold_rocobjs
betas[[model_name]] <- as.matrix(enet_model$betas)

# see the results
enet_model$model_summary %>% t()
enet_model$conf_matrices
enet_model$plot
roc_c
```

#### Genus level

```{r, warning=FALSE}
level="genus"
```

Aggregate taxa

```{r, warning=FALSE}
genus_data <- aggregate_taxa(colon_asv_tab,
                             colon_taxa_tab,
                             taxonomic_level = level)

colon_genus_tab <- genus_data[[1]]
colon_genus_taxa_tab <- genus_data[[2]]
```

##### low vs high

```{r, warning=FALSE}
group <- c("low","high")
comparison_name <- paste0(group[1], " vs ",group[2])
```

```{r, warning=FALSE}
model_name <- paste(comparison_name,level,segment)

# prepare the data
filt_colon_uni_data <- binomial_prep(colon_genus_tab,
                                     colon_genus_taxa_tab,
                                     colon_metadata,
                                     group, 
                                     usage="ml_clr",patient = TRUE)

# fit the model
enet_model <- glmnet_binomial(filt_colon_uni_data,
                              sample_method = "atypboot",
                              outcome="Group", 
                              N=10,
                              reuse=FALSE,clust_var = "Patient",
                              file=model_name,
                              Q="Q4")

# ROC
roc_c <- roc_curve(enet_model, group)

# save the results
models_summ[[model_name]] <- enet_model$model_summary
models_cm[[model_name]] <- enet_model$conf_matrices$original
roc_cs[[model_name]] <- enet_model$kfold_rocobjs
betas[[model_name]] <- as.matrix(enet_model$betas)

# see the results
enet_model$model_summary %>% t()
enet_model$conf_matrices
enet_model$plot

roc_c
```

#### Saving results

```{r, warning=FALSE}
models_summ_df_colon <- do.call(rbind, 
  models_summ[grep(segment,names(models_summ),value = TRUE)])

write.csv(models_summ_df_colon,file.path(path,paste0("elastic_net_",segment,".csv")))
```

```{r}
k
```

### Supplementary models

#### CLR-transformed data

##### kNN

```{r, warning=FALSE}
model="knn"
```

###### ASV level

```{r, warning=FALSE}
level="ASV"
```

***low vs high***

```{r, warning=FALSE}
group <- c("low","high")
comparison_name <- paste0(group[1], " vs ",group[2])
```

```{r, warning=FALSE}
model_name <- paste(comparison_name,level,segment)

# prepare the data
filt_colon_uni_data <- binomial_prep(colon_asv_tab,
                                     colon_taxa_tab,
                                     colon_metadata,
                                     group, usage="ml_clr",patient = TRUE)

# fit the model
knn_model <- knn_binomial(filt_colon_uni_data,
                              sample_method = "atypboot",
                              outcome="Group", 
                              N=10,
                              reuse=FALSE,
                              file=model_name,
                              Q="Q4",clust_var = "Patient")

# ROC curve
roc_c <- roc_curve(knn_model, group)

# save the results
supplements_models[["models_summ"]][["knn_model"]][[model_name]] <- knn_model$model_summary
supplements_models[["roc_cs"]][["knn_model"]][[model_name]] <- knn_model$kfold_rocobjs

# see the results
knn_model$model_summary %>% t()
roc_c
```


###### Genus level

```{r, warning=FALSE}
level="genus"
```

Aggregate taxa

```{r, warning=FALSE}
genus_data <- aggregate_taxa(colon_asv_tab,
                             colon_taxa_tab,
                             taxonomic_level = level)

colon_genus_tab <- genus_data[[1]]
colon_genus_taxa_tab <- genus_data[[2]]
```

***low vs high***

```{r, warning=FALSE}
group <- c("low","high")
comparison_name <- paste0(group[1], " vs ",group[2])
```

```{r, warning=FALSE}
model_name <- paste(comparison_name,level,segment)

# prepare the data
filt_colon_uni_data <- binomial_prep(colon_genus_tab,
                                     colon_genus_taxa_tab,
                                     colon_metadata,
                                     group, 
                                     usage="ml_clr",patient = TRUE)

# fit the model
knn_model <- knn_binomial(filt_colon_uni_data,
                              sample_method = "atypboot",
                              outcome="Group", 
                              N=10,
                              reuse=FALSE,
                              file=model_name,
                              Q="Q4",clust_var = "Patient")

# ROC curve
roc_c <- roc_curve(knn_model, group)

# save the results
supplements_models[["models_summ"]][["knn_model"]][[model_name]] <- knn_model$model_summary
supplements_models[["roc_cs"]][["knn_model"]][[model_name]] <- knn_model$kfold_rocobjs


# see the results
knn_model$model_summary %>% t()
roc_c
```


##### RF

```{r, warning=FALSE}
model="rf"
```

###### ASV level

```{r, warning=FALSE}
level="ASV"
```

***low vs high***

```{r, warning=FALSE}
group <- c("low","high")
comparison_name <- paste0(group[1], " vs ",group[2])
```

```{r, warning=FALSE}
model_name <- paste(comparison_name,level,segment)

# prepare the data
filt_colon_uni_data <- binomial_prep(colon_asv_tab,
                                     colon_taxa_tab,
                                     colon_metadata,
                                     group, usage="ml_clr",patient = TRUE)

# fit the model
rf_model <- rf_binomial(filt_colon_uni_data,
                              sample_method = "atypboot",
                              outcome="Group", 
                              N=10,
                              reuse=FALSE,
                              file=model_name,
                              Q="Q4", clust_var = "Patient")

# ROC curve
roc_c <- roc_curve(rf_model, group)

# save the results
supplements_models[["models_summ"]][["rf_model"]][[model_name]] <- rf_model$model_summary
supplements_models[["roc_cs"]][["rf_model"]][[model_name]] <- rf_model$kfold_rocobjs

# see the results
rf_model$model_summary %>% t()
roc_c
```


###### Genus level

```{r, warning=FALSE}
level="genus"
```

Aggregate taxa

```{r, warning=FALSE}
genus_data <- aggregate_taxa(colon_asv_tab,
                             colon_taxa_tab,
                             taxonomic_level = level)

colon_genus_tab <- genus_data[[1]]
colon_genus_taxa_tab <- genus_data[[2]]
```

***low vs high***

```{r, warning=FALSE}
group <- c("low","high")
comparison_name <- paste0(group[1], " vs ",group[2])
```

```{r, warning=FALSE}
model_name <- paste(comparison_name,level,segment)

# prepare the data
filt_colon_uni_data <- binomial_prep(colon_genus_tab,
                                     colon_genus_taxa_tab,
                                     colon_metadata,
                                     group, 
                                     usage="ml_clr",patient = TRUE)

# fit the model
rf_model <- rf_binomial(filt_colon_uni_data,
                              sample_method = "atypboot",
                              outcome="Group", 
                              N=10,
                              reuse=FALSE,
                              file=model_name,clust_var = "Patient",
                              Q="Q4")

# ROC curve
roc_c <- roc_curve(rf_model, group)

# save the results
supplements_models[["models_summ"]][["rf_model"]][[model_name]] <- rf_model$model_summary
supplements_models[["roc_cs"]][["rf_model"]][[model_name]] <- rf_model$kfold_rocobjs

# see the results
rf_model$model_summary %>% t()
roc_c
```


##### Gboost

```{r, warning=FALSE}
model="gb"
```

###### ASV level

```{r, warning=FALSE}
level="ASV"
```

***low vs high***

```{r, warning=FALSE}
group <- c("low","high")
comparison_name <- paste0(group[1], " vs ",group[2])
```

```{r, warning=FALSE}
model_name <- paste(comparison_name,level,segment)

# prepare the data
filt_colon_uni_data <- binomial_prep(colon_asv_tab,
                                     colon_taxa_tab,
                                     colon_metadata,
                                     group, usage="ml_clr",patient = TRUE)

# fit the model
gbm_model <- gbm_binomial(filt_colon_uni_data,
                              sample_method = "atypboot",
                              outcome="Group", 
                              N=10,
                              reuse=FALSE,
                              file=model_name,
                              Q="Q4",clust_var = "Patient")

# ROC curve
roc_c <- roc_curve(gbm_model, group)

# save the results
supplements_models[["models_summ"]][["gbm_model"]][[model_name]] <- gbm_model$model_summary
supplements_models[["roc_cs"]][["gbm_model"]][[model_name]] <- gbm_model$kfold_rocobjs

# see the results
gbm_model$model_summary %>% t()
roc_c
```

###### Genus level

```{r, warning=FALSE}
level="genus"
```

Aggregate taxa

```{r, warning=FALSE}
genus_data <- aggregate_taxa(colon_asv_tab,
                             colon_taxa_tab,
                             taxonomic_level = level)

colon_genus_tab <- genus_data[[1]]
colon_genus_taxa_tab <- genus_data[[2]]
```

***low vs high***

```{r, warning=FALSE}
group <- c("low","high")
comparison_name <- paste0(group[1], " vs ",group[2])
```

```{r, warning=FALSE}
model_name <- paste(comparison_name,level,segment)

# prepare the data
filt_colon_uni_data <- binomial_prep(colon_genus_tab,
                                     colon_genus_taxa_tab,
                                     colon_metadata,
                                     group, 
                                     usage="ml_clr",patient = TRUE)

# fit the model
gbm_model <- gbm_binomial(filt_colon_uni_data,
                              sample_method = "atypboot",
                              outcome="Group", 
                              N=10,
                             reuse=FALSE,
                              file=model_name,
                              Q="Q4",
                          clust_var = "Patient")

# ROC curve
roc_c <- roc_curve(gbm_model, group)

# save the results
supplements_models[["models_summ"]][["gbm_model"]][[model_name]] <- gbm_model$model_summary
supplements_models[["roc_cs"]][["gbm_model"]][[model_name]] <- gbm_model$kfold_rocobjs

# see the results
gbm_model$model_summary %>% t()
roc_c
```


#### Saving results

```{r, warning=FALSE}
models_list <- list()

for (model_name in names(supplements_models$models_summ)){
  df <- do.call(rbind, supplements_models$models_summ[[model_name]])
  models_list[[model_name]] <- df
}

write.xlsx(models_list,
           file=file.path(path,paste0("supplements_models_",segment,".xlsx")),
           rowNames=TRUE)

```

## Results overview

#### Alpha diversity

```{r, warning=FALSE}
pc_observed[[segment]]
pc_shannon[[segment]]
pc_simpson[[segment]]
pc_pielou[[segment]]
```

Plots

```{r, fig.width=4, fig.height=5, fig.fullwidth=TRUE}
alpha_div_plots[[paste(segment,"Country")]]
```

#### Beta diversity

**Main results**

```{r, warning=FALSE}
pairwise_aitchison_raw[[paste("genus", segment)]]
```

PCA

```{r, fig.width=5, fig.height=3, fig.fullwidth=TRUE, eval=TRUE}
pca_plots_list[[paste(segment,"genus custom")]]
```

**Supplements**

```{r, warning=FALSE}
knitr::kable(supplements_beta[!grepl("PCoA",names(supplements_beta)) & (grepl("genus",names(supplements_beta)))],
             digits = 3,
             caption = "Supplementary PERMANOVA results: Bray-curtis, Jaccard distances")
```

PCA

```{r, fig.width=8, fig.height=7, fig.fullwidth=TRUE, eval=TRUE}
plot_list <- supplements_beta[grepl("PCoA",names(supplements_beta)) &
                              grepl(segment,names(supplements_beta))]

ggarrange(plotlist = plot_list,
          labels=names(plot_list),
          font.label = list(size=5,face="plain"),
          ncol=2,nrow=3)
```

#### Univariate analysis

**Number of significant taxa**

```{r, warning=FALSE}
knitr::kable(as.data.frame(lapply(list_intersections,nrow)) %>% t() %>% 
  `colnames<-`("Count") %>% 
  `rownames<-`(c(names(list_intersections))),caption="Number of significant taxa")
```


#### Machine learning

**Main models**

Summary

```{r, warning=FALSE}
knitr::kable(models_summ_df_colon %>% dplyr::select(
"alpha","lambda",
"auc_optimism_corrected",
"auc_optimism_corrected_CIL",
"auc_optimism_corrected_CIU"),
             digits=3,caption="Elastic net results")
```

**ROC - ASV level**

```{r, warning=FALSE}
roc_curve_all_custom(roc_cs[3], 
                     Q="Q4",
                     model_name="enet_model")
```

**ROC - Genus level**

```{r, warning=FALSE}
roc_curve_all_custom(roc_cs[4],Q="Q4",
                     model_name="enet_model")
```

**Supplementary models**


```{r, warning=FALSE}
# Build final dataframe
models_list[["enet_model"]] <- rbind(models_summ_df_colon,models_summ_df_colon)
final_df <- tibble(row_names = rownames(models_list[[1]]))

# Loop through models and extract required values
for (model_name in names(models_list)) {
  model_df <- models_list[[model_name]]
  
  # Combine AUC_optimism_corrected with its CI values
  final_df[[model_name]] <- paste0(
    round(model_df$auc_optimism_corrected, 3), 
    " (", round(model_df$auc_optimism_corrected_CIL, 3), "; ", 
    round(model_df$auc_optimism_corrected_CIU, 3), ")"
  )
}

knitr::kable(final_df, caption="All models")
write.csv(final_df,file=file.path(path,"AUC_all_models.csv"),row.names = FALSE)
```

**ROC - genus**

```{r, warning=FALSE, fig.width=10, fig.height=8, fig.fullwidth=TRUE}
rocs_list <- supplements_models$roc_cs
rocs_list[["enet_model"]] <- roc_cs
plot_list <- list()

for (model_name in names(rocs_list)) {
  plot_list[[model_name]] <- roc_curve_all_custom(rocs_list[[model_name]][4],
                       Q="Q4",
                       model_name=model_name)
}

p <- ggarrange(plotlist = plot_list,labels = names(rocs_list),font.label = list(face="plain",size=7))
```

```{r,results='hide', echo=FALSE, warning=FALSE}
pdf("../figures/Q4/models_colon.pdf",
    height =10,width = 10)
p
dev.off()
```

# Paper-ready visualizations

## Alpha diversity

```{r, warning=FALSE, fig.width=10, fig.height=4, fig.fullwidth=TRUE}
p_A <- alpha_div_plots$`terminal_ileum Country` +
  ggtitle("Terminal ileum")+
  theme(plot.title = element_text(hjust=0.5,face = "bold",size = 15)) 

p_B <-  alpha_div_plots$`colon Country` +
  ggtitle("Colon") +
  theme(plot.title = element_text(hjust=0.5,face = "bold",size = 15)) 

Q4_alpha <- ggarrange(p_A,ggplot() + theme_minimal(),p_B,nrow=1, ncol=3,
                      widths = c(1,0.1,1))
Q4_alpha
```

## Beta diversity


```{r, warning=FALSE, fig.width=10, fig.height=4, fig.fullwidth=TRUE}
pca_ti <- pca_plots_list$`terminal_ileum genus custom` 
pca_colon <- pca_plots_list$`colon genus custom` 

genus_Q4_beta <- ggarrange(pca_ti,
                           ggplot() + theme_minimal(),
                           pca_colon,ncol=3,
                           widths = c(1,0.1,1))
genus_Q4_beta
```

**Alpha + Beta diversity**

```{r, fig.width=10, fig.height=8, fig.fullwidth=TRUE}
alpha_beta <- ggarrange(Q4_alpha,genus_Q4_beta,
                        ncol = 1,nrow=2,labels = c("A","B"))
alpha_beta
```

## Elastic net

Genus level

```{r, warning=FALSE, fig.width=10, fig.height=4, fig.fullwidth=TRUE}
models_to_plot <- c("knn_model","rf_model","gbm_model","enet_model")
names(models_to_plot) <- c("kNN","RF","GBoost","ENET")

# ILEUM
plot_list_ileum <- list()
for (model_name in models_to_plot) {
  plot_list_ileum[[model_name]] <- 
    roc_curve_all_custom(rocs_list[[model_name]][2],
                       Q="Q4",
                       model_name=model_name,legend = FALSE) + 
    ggtitle(names(models_to_plot)[which(model_name==models_to_plot)]) + 
    theme(plot.title = element_text(face = "bold",size = 10)) 
}

roc_curve_ti <- ggarrange(plotlist = plot_list_ileum)

# COLON

plot_list_colon <- list()
for (model_name in models_to_plot) {
  plot_list_colon[[model_name]] <- 
    roc_curve_all_custom(rocs_list[[model_name]][c(4)],
                       Q="Q4",
                       model_name=model_name,legend = FALSE) + 
    ggtitle(names(models_to_plot)[which(model_name==models_to_plot)]) + 
    theme(plot.title = element_text(face = "bold",size = 10))  
}

roc_curve_colon <- ggarrange(plotlist = plot_list_colon)


roc_curve_plot <- ggarrange(roc_curve_ti,
                            ggplot() + theme_minimal(),
                            roc_curve_colon,
                            ncol=3, widths = c(1,0.1,1))
roc_curve_plot

```

```{r, fig.width=10, fig.height=12, fig.fullwidth=TRUE}
alpha_beta_elastic <- ggarrange(Q4_alpha,genus_Q4_beta,roc_curve_plot,
                        ncol = 1,nrow=3,labels = LETTERS,heights = c(1,1,1.2))
alpha_beta_elastic


alpha_beta_elastic <- ggarrange(alpha_beta_elastic,ggplot() + theme_minimal(),ncol=2,
                                widths = c(1,0.15))

```

```{r,results='hide', echo=FALSE}
pdf("../figures/Q4/supplement_1.pdf",paper = "a4",height = 10,width = 12)
alpha_beta_elastic
dev.off()
```

## DAA

```{r,  warning=FALSE, fig.width=10, fig.height=13, fig.fullwidth=TRUE}
if (exists("dot_heatmap_ileum")){
 p_ileum <- dot_heatmap_ileum + 
  ggtitle("Terminal ileum") +
  theme(plot.title = element_text(hjust=0.5,face = "bold",size = 15),
        legend.position = "none") 
}

if (exists("dot_heatmap_colon")){
  p_colon <- dot_heatmap_colon  +
  ggtitle("Colon") +
  theme(plot.title = element_text(hjust=0.5,face = "bold",size = 15),
        legend.position = "none")
}

if (exists("p_ileum") | exists("p_colon")){
  heatmap_plot <- ggarrange(p_ileum,
                          p_colon,
                          ncol = 2,labels=c("A","B"))
  heatmap_plot
}


```

```{r,results='hide', eval=FALSE, echo=FALSE}
pdf("../figures/Q4/supplement_2.pdf",
    height =10,width = 8,paper="a4")
heatmap_plot
dev.off()
```

# Session info

```{r, warning=FALSE}
sessionInfo()
```
