---
title: "Mucosal microbiota alterations in primary sclerosis cholangitis persist after liver transplantation and are associated with clinical features independently of geography"
subtitle: "Additional disease control"
format: 
  html:  
    embed-resources: true
    keep-md: false
    toc: true
    toc-expand: 3
    toc-depth: 6
    toc-location: left
    number-sections: true
    number-depth: 6
    code-fold: false
    code-tools: true
    code-summary: "Open code"
    grid: 
      body-width: 1000px
      sidebar-width: 600px
      margin-width: 100px
editor: source
project:
  type: default
  output-dir: output
fontsize: 12 px
---

# Introduction

**PSC vs ALD**

In previous analyses, we showed specific mucosal composition in PSC patients which persist after liver transplantation. Here, we added another positive control, to explore if this condition is disease specific, or it can be credited to (liver) disease in general.  

Fo this purpose, we conduct the univariate analysis on Czech ALD patients after LTx. By comparing post_LTx PSC and ALD patients, we can check if the proposed PSC-associated bacteria are at least partially overlapping.

If so, this would confirm our previous conclusions that these bacteria are specific for PSC.


**Importing libraries and custom functions built for this analysis**

```{r,warning=FALSE}
source("custom_functions.R")
```

# Data Import

Importing ASV, taxa and metadata tables for both Czech ALD samples
samples.

```{r,warning=FALSE}
path = "../../data/analysis_ready_data/ikem/ALD/"
asv_tab_ALD <- as.data.frame(fread(file.path(path,"asv_table_ALD.tsv"),
                                    check.names = FALSE))
taxa_tab_ALD <- as.data.frame(fread(file.path(path,"taxa_table_ALD.tsv"),
                                     check.names = FALSE))
metadata_ALD <- as.data.frame(fread(file.path(path,"metadata_ALD.tsv"),
                                     check.names = FALSE))
```

Importing previous Czech dataset with PSC patients.

```{r,warning=FALSE}
path = "../../data/analysis_ready_data/ikem/"
asv_tab_ikem <- as.data.frame(fread(file.path(path,"asv_table_ikem.csv"),
                                    check.names = FALSE))
taxa_tab_ikem <- as.data.frame(fread(file.path(path,"taxa_table_ikem.csv"),
                                     check.names = FALSE))
metadata_ikem <- as.data.frame(fread(file.path(path,"metadata_ikem.csv"),
                                     check.names = FALSE))
```

Importing proposed PSC-associated taxa.

```{r}
psc_genus_ti <- read.xlsx("../results/Q1/univariate_analysis/supplements_psc_effect_terminal_ileum.xlsx",
                     sheet = "terminal_ileum genus")

psc_genus_colon <- read.xlsx("../results/Q1/univariate_analysis/supplements_psc_effect_colon.xlsx",
                     sheet = "colon genus")
```

# Terminal ileum

```{r}
segment="TI"

ileum_data <- merging_data(asv_tab_1=asv_tab_ALD,
                           asv_tab_2=asv_tab_ikem,
                           taxa_tab_1=taxa_tab_ALD,
                           taxa_tab_2=taxa_tab_ikem,
                           metadata_1=metadata_ALD,
                           metadata_2=metadata_ikem,
                           segment=segment,Q="QALD")

ileum_asv_tab <- ileum_data[[1]]
ileum_taxa_tab <- ileum_data[[2]]
ileum_metadata <- ileum_data[[3]]
```

## Univariate Analysis

Similar to previous approaches, we will use two DAA tools - linDA and Maaslin. Only the intersection of sets
selected by each tool are shown to minimize false positives. As this analysis can only be conducted on Czech cohort, no interaction effect can be shown, therefore only Group effect is investigated. 

This analysis is performed at genus level only.

```{r,warning=FALSE}
level="genus"

# needed paths
path = "../results/QALD/univariate_analysis"
path_maaslin=file.path("../intermediate_files/maaslin/QALD",level)
```

```{r,warning=FALSE}
# variables
raw_linda_results_genus <- list();
raw_linda_results_genus[[segment]] <- list()
linda_results_genus <- list(); 
linda_results_genus[[segment]] <- list()

# country and interaction problems
list_country_union <- list()
list_intersections <- list()
list_venns <- list()
uni_statistics <- list()

# workbook for final df
wb <- createWorkbook()
```

Aggregate taxa

```{r,warning=FALSE}
genus_data <- aggregate_taxa(ileum_asv_tab,
                             ileum_taxa_tab,
                             taxonomic_level = level)

ileum_genus_tab <- genus_data[[1]]
ileum_genus_taxa_tab <- genus_data[[2]]

ileum_genus_asv_taxa_tab <- create_asv_taxa_table(ileum_genus_tab,
                                                  ileum_genus_taxa_tab)
```

### PSC vs ALD

We can assume, that the differentially abundant taxa between these two
groups are caused by PSC disease and other factors, like liver cirrhosis
and overall bad clinical condition.

```{r,warning=FALSE}
group <- c("ALD post_ltx","PSC post_ltx")
comparison_name <- paste0(group[1], " vs ",group[2])
```

```{r, echo=FALSE}
comparison_title <- sub('"(.+?)" vs "(.+?)"', '"\\2" and "\\1"', comparison_name)
```

#### linDA

```{r, warning=FALSE, fig.width=15, fig.height=5, fig.fullwidth=TRUE}

# prepare the data
linda_data <- binomial_prep(ileum_genus_tab,
                            ileum_genus_taxa_tab,
                            ileum_metadata,
                            group, usage="linDA")

filt_ileum_uni_data <- linda_data[[1]]
filt_ileum_uni_taxa <- linda_data[[2]]
filt_ileum_uni_metadata <- linda_data[[3]]

# fit the model
linda.obj <- linda(filt_ileum_uni_data, 
                   filt_ileum_uni_metadata, 
                   formula = '~ Group')

linda.output <- linda.obj$output
linda.output <- linda_renaming(linda.output, group)

# save the results
group1 <- paste0(group[1], " vs ","Group",group[2])

for (grp in c(group1)){
  raw_linda_results_genus[[segment]][[grp]] <- 
    rawlinda.df(linda.output,
                grp,
                filt_ileum_uni_data,
                filt_ileum_uni_taxa)
  
  linda_results_genus[[segment]][[grp]] <- 
    linda.df(linda.output,
             grp,
             filt_ileum_uni_data,
             filt_ileum_uni_taxa)
}

```

```{r, warning=FALSE, fig.width=5, fig.height=4, fig.fullwidth=TRUE}
# volcano plots
volcano <- volcano_plot_linda(linda.output, 
                                group1, 
                                taxa_table = filt_ileum_uni_taxa) + 
            ggtitle(comparison_title)

volcano
```

#### MaAsLin2

```{r, warning=FALSE, echo=FALSE,results='hide',message=FALSE,warning=FALSE}
log_file <- file(tempfile(), open = "wt")
sink(log_file)  # Redirect standard output
sink(log_file, type = "message")  # Redirect messages and warnings

fit_data <-  Maaslin2(
    input_data = filt_ileum_uni_data, 
    input_metadata = filt_ileum_uni_metadata, min_abundance = 0,
    min_prevalence = 0,min_variance = 0,
    output = file.path(path_maaslin,group1), max_significance = 0.05,
    fixed_effects = c('Group'),correction = "BH")

sink()  # Restore standard output
sink(type = "message")  # Restore messages
```

```{r,warning=FALSE}
volcano <- volcano_plot_maaslin(fit_data,filt_ileum_uni_taxa) + 
            ggtitle(comparison_title)


volcano
```

#### Group - Intersection

```{r,warning=FALSE}
intersection_results <- group_intersection(group, list_intersections, list_venns,
                                           linda.output, fit_data,
                                           raw_linda_results_genus, 
                                           segment = segment,
                                           level=level)

list_intersections <- intersection_results[[1]]
list_venns <- intersection_results[[2]]
venn <- intersection_results[[3]]

# show the results
venn
```

#### Basic statistics

```{r,warning=FALSE}
uni_df <-  merge(basic_univariate_statistics(linda_data,group),
                 raw_linda_results_genus[[segment]][[group1]],
                 by="SeqID",all=TRUE)

uni_df[["final_sig"]] <- uni_df$SeqID %in% list_intersections[[paste(segment,level,comparison_name)]][["SeqID"]]
uni_statistics[[segment]][[paste(level,comparison_name)]] <- uni_df

# for comparison
new_name <- comparison_name
addWorksheet(wb, sheetName = new_name)
writeData(wb, sheet = new_name, uni_df, rowNames=FALSE)
```

#### Comparison with PSC-associated taxa

```{r}
psc_ald_sig_ti <- uni_df %>%
  dplyr::filter(final_sig==TRUE) %>%
  dplyr::select(SeqID,Taxonomy,log2FoldChange)

psc_genus_ti <- psc_genus_ti
```

```{r}
list_core <- list()
list_core[["PSC vs ALD"]] <- psc_ald_sig_ti$Taxonomy
list_core[["PSC-associated taxa"]] <- psc_genus_ti$Taxonomy
  
# venn diagram
venn <- ggvenn(list_core, fill_color = c("blue", "red")) + 
    ggtitle("Intersection between PSC-associated taxa and (PSC vs ALD)") + 
    guides(fill = "none")

venn
```
```{r}
intersection_df <- intersect(list_core$`PSC vs ALD`,list_core$`PSC-associated taxa`)

intersection_df
```


### ALD vs HC

We can assume, that the differentially abundant taxa between these two
groups are caused by PSC disease and other factors, like liver cirrhosis
and overall bad clinical condition.

```{r,warning=FALSE}
group <- c("HC","ALD post_ltx")
comparison_name <- paste0(group[1], " vs ",group[2])
```

```{r, echo=FALSE}
comparison_title <- sub('"(.+?)" vs "(.+?)"', '"\\2" and "\\1"', comparison_name)
```

#### linDA

```{r, warning=FALSE, fig.width=15, fig.height=5, fig.fullwidth=TRUE}

# prepare the data
linda_data <- binomial_prep(ileum_genus_tab,
                            ileum_genus_taxa_tab,
                            ileum_metadata,
                            group, usage="linDA")

filt_ileum_uni_data <- linda_data[[1]]
filt_ileum_uni_taxa <- linda_data[[2]]
filt_ileum_uni_metadata <- linda_data[[3]]

# fit the model
linda.obj <- linda(filt_ileum_uni_data, 
                   filt_ileum_uni_metadata, 
                   formula = '~ Group')

linda.output <- linda.obj$output
linda.output <- linda_renaming(linda.output, group)

# save the results
group1 <- paste0(group[1], " vs ","Group",group[2])

for (grp in c(group1)){
  raw_linda_results_genus[[segment]][[grp]] <- 
    rawlinda.df(linda.output,
                grp,
                filt_ileum_uni_data,
                filt_ileum_uni_taxa)
  
  linda_results_genus[[segment]][[grp]] <- 
    linda.df(linda.output,
             grp,
             filt_ileum_uni_data,
             filt_ileum_uni_taxa)
}

```

```{r, warning=FALSE, fig.width=5, fig.height=4, fig.fullwidth=TRUE}
# volcano plots
volcano <- volcano_plot_linda(linda.output, 
                                group1, 
                                taxa_table = filt_ileum_uni_taxa) + 
            ggtitle(comparison_title)

volcano
```

#### MaAsLin2

```{r, warning=FALSE, echo=FALSE,results='hide',message=FALSE,warning=FALSE}
log_file <- file(tempfile(), open = "wt")
sink(log_file)  # Redirect standard output
sink(log_file, type = "message")  # Redirect messages and warnings

fit_data <-  Maaslin2(
    input_data = filt_ileum_uni_data, 
    input_metadata = filt_ileum_uni_metadata, min_abundance = 0,
    min_prevalence = 0,min_variance = 0,
    output = file.path(path_maaslin,group1), max_significance = 0.05,
    fixed_effects = c('Group'),correction = "BH")

sink()  # Restore standard output
sink(type = "message")  # Restore messages
```

```{r,warning=FALSE}
volcano <- volcano_plot_maaslin(fit_data,filt_ileum_uni_taxa) + 
            ggtitle(comparison_title)


volcano
```

#### Group - Intersection

```{r,warning=FALSE}
intersection_results <- group_intersection(group, list_intersections, list_venns,
                                           linda.output, fit_data,
                                           raw_linda_results_genus, 
                                           segment = segment,
                                           level=level)

list_intersections <- intersection_results[[1]]
list_venns <- intersection_results[[2]]
venn <- intersection_results[[3]]

# show the results
venn
```

#### Basic statistics

```{r,warning=FALSE}
uni_df <-  merge(basic_univariate_statistics(linda_data,group),
                 raw_linda_results_genus[[segment]][[group1]],
                 by="SeqID",all=TRUE)

uni_df[["final_sig"]] <- uni_df$SeqID %in% list_intersections[[paste(segment,level,comparison_name)]][["SeqID"]]
uni_statistics[[segment]][[paste(level,comparison_name)]] <- uni_df

# for comparison
new_name <- comparison_name
addWorksheet(wb, sheetName = new_name)
writeData(wb, sheet = new_name, uni_df, rowNames=FALSE)
```

#### Comparison with PSC-associated taxa

```{r}
hc_ald_sig_ti <- uni_df %>%
  dplyr::filter(final_sig==TRUE) %>%
  dplyr::select(SeqID,Taxonomy,log2FoldChange)

psc_genus_ti <- psc_genus_ti
```

```{r}
list_core <- list()
list_core[["ALD vs HC"]] <- hc_ald_sig_ti$Taxonomy
list_core[["PSC-associated taxa"]] <- psc_genus_ti$Taxonomy
  
# venn diagram
venn <- ggvenn(list_core, fill_color = c("blue", "red")) + 
    ggtitle("Intersection between PSC-associated taxa and (ALD vs HC)") + 
    guides(fill = "none")

venn
```

From this Venn diagram, we are interested in EXLUSIVE RIGH JOIN

```{r}
A <- list_core$`ALD vs HC`
B <- list_core$`PSC-associated taxa`
remaining_psc <- B[!(B %in% A)]

psc_genus_ti %>%
  dplyr::filter(Taxonomy %in% remaining_psc)

uni_df %>%
  dplyr::filter(Taxonomy %in% remaining_psc)
```

# Colon

```{r}
segment="colon"

colon_data <- merging_data(asv_tab_1=asv_tab_ALD,
                           asv_tab_2=asv_tab_ikem,
                           taxa_tab_1=taxa_tab_ALD,
                           taxa_tab_2=taxa_tab_ikem,
                           metadata_1=metadata_ALD,
                           metadata_2=metadata_ikem,
                           segment=segment,Q="QALD")

colon_asv_tab <- colon_data[[1]]
colon_taxa_tab <- colon_data[[2]]
colon_metadata <- colon_data[[3]]
```

## Univariate Analysis


```{r,warning=FALSE}
level="genus"
```

```{r,warning=FALSE}
# needed paths
path = "../results/QALD/univariate_analysis"
path_maaslin=file.path("../intermediate_files/maaslin/QALD",level)

# variables
raw_linda_results_genus[[segment]] <- list()
linda_results_genus[[segment]] <- list()

# country and interaction problems
list_country_union <- list()
list_venns <- list()

# workbook for final df
wb <- createWorkbook()

```

```{r,warning=FALSE}
level="genus"
```

Aggregate taxa

```{r,warning=FALSE}
genus_data <- aggregate_taxa(colon_asv_tab,
                             colon_taxa_tab,
                             taxonomic_level = level)

colon_genus_tab <- genus_data[[1]]
colon_genus_taxa_tab <- genus_data[[2]]

colon_genus_asv_taxa_tab <- create_asv_taxa_table(colon_genus_tab,
                                                  colon_genus_taxa_tab)
```

##### PSC vs ALD

###### linDA

```{r,warning=FALSE}
group <- c("ALD post_ltx","PSC post_ltx")
comparison_name <- paste0(group[1], " vs ",group[2])
```

```{r, echo=FALSE}
comparison_title <- sub('"(.+?)" vs "(.+?)"', '"\\2" and "\\1"', comparison_name)
```

```{r, warning=FALSE, fig.width=15, fig.height=5, fig.fullwidth=TRUE}

# prepare the data
linda_data <- binomial_prep(colon_genus_tab,
                            colon_genus_taxa_tab,
                            colon_metadata,
                            group, usage="linDA")

filt_colon_uni_data <- linda_data[[1]]
filt_colon_uni_taxa <- linda_data[[2]]
filt_colon_uni_metadata <- linda_data[[3]]

# fit the model
linda.obj <- linda(filt_colon_uni_data,
                   filt_colon_uni_metadata,
                   formula = '~ Group + (1|Patient)')

linda.output <- linda.obj$output
linda.output <- linda_renaming(linda.output, group)

# save the results
group1 <- paste0(group[1], " vs ","Group",group[2])

for (grp in c(group1)){
  raw_linda_results_genus[[segment]][[grp]] <- 
    rawlinda.df(linda.output,
                grp,
                filt_colon_uni_data,
                filt_colon_uni_taxa)
  
  linda_results_genus[[segment]][[grp]] <- 
    linda.df(linda.output,
             grp,
             filt_colon_uni_data,
             filt_colon_uni_taxa)
}
```

```{r, warning=FALSE, fig.width=5, fig.height=4, fig.fullwidth=TRUE}
# volcano plot
volcano <- volcano_plot_linda(linda.output, group1, 
                                taxa_table = filt_colon_uni_taxa) + 
            ggtitle(paste(group,collapse=" vs "))


# see the plot
volcano
```

###### MaAsLin2

```{r, warning=FALSE, echo=FALSE,results='hide',message=FALSE,warning=FALSE}

log_file <- file(tempfile(), open = "wt")
sink(log_file)  # Redirect standard output
sink(log_file, type = "message")  # Redirect messages and warnings

fit_data = Maaslin2(
    input_data = filt_colon_uni_data, 
    input_metadata = filt_colon_uni_metadata, min_abundance = 0,
    min_prevalence = 0,min_variance = 0,
    output = file.path(path_maaslin,group1), max_significance = 0.05,
    fixed_effects = c('Group'),random_effects = "Patient",
    correction = "BH")

sink()  # Restore standard output
sink(type = "message")  # Restore messages
```

```{r,warning=FALSE}
volcano <- volcano_plot_maaslin(fit_data,filt_colon_uni_taxa) + 
            ggtitle(paste(group[1], "vs", group[2]))

volcano
```

###### Group - Intersection

```{r,warning=FALSE}
intersection_results <- group_intersection(group, list_intersections, list_venns,
                                           linda.output, fit_data,
                                           raw_linda_results_genus, 
                                           segment = segment,
                                           level=level)

list_intersections <- intersection_results[[1]]
list_venns <- intersection_results[[2]]
venn <- intersection_results[[3]]

# show the results
venn
```


###### Basic statistics

```{r,warning=FALSE}
uni_df <-  merge(basic_univariate_statistics(linda_data,group),
                 raw_linda_results_genus[[segment]][[group1]],
                 by="SeqID",all=TRUE)

uni_df[["final_sig"]] <- uni_df$SeqID %in% list_intersections[[paste(segment,level,comparison_name)]][["SeqID"]]
uni_statistics[[segment]][[paste(level,comparison_name)]] <- uni_df

# for comparison
new_name <- comparison_name
addWorksheet(wb, sheetName = new_name)
writeData(wb, sheet = new_name, uni_df, rowNames=FALSE)
```

###### Comparison with PSC-associated taxa

```{r}
psc_ald_sig_colon <- uni_df %>%
  dplyr::filter(final_sig==TRUE) %>%
  dplyr::select(SeqID,Taxonomy,log2FoldChange)

psc_genus_colon <- psc_genus_colon
```

```{r}
list_core <- list()
list_core[["PSC vs ALD"]] <- psc_ald_sig_colon$Taxonomy
list_core[["PSC-associated taxa"]] <- psc_genus_colon$Taxonomy

# venn diagram
venn <- ggvenn(list_core, fill_color = c("blue", "red")) + 
    ggtitle("Intersection between PSC-associated taxa and (PSC vs ALD)") + 
    guides(fill = "none")

venn
```

```{r}
intersection_df <- intersect(list_core$`PSC vs ALD`,list_core$`PSC-associated taxa`)

intersection_df
```

##### ALD vs HC

###### linDA

```{r,warning=FALSE}
group <- c("HC","ALD post_ltx")
comparison_name <- paste0(group[1], " vs ",group[2])
```

```{r, echo=FALSE}
comparison_title <- sub('"(.+?)" vs "(.+?)"', '"\\2" and "\\1"', comparison_name)
```

```{r, warning=FALSE, fig.width=15, fig.height=5, fig.fullwidth=TRUE}

# prepare the data
linda_data <- binomial_prep(colon_genus_tab,
                            colon_genus_taxa_tab,
                            colon_metadata,
                            group, usage="linDA")

filt_colon_uni_data <- linda_data[[1]]
filt_colon_uni_taxa <- linda_data[[2]]
filt_colon_uni_metadata <- linda_data[[3]]

# fit the model
linda.obj <- linda(filt_colon_uni_data,
                   filt_colon_uni_metadata,
                   formula = '~ Group + (1|Patient)')

linda.output <- linda.obj$output
linda.output <- linda_renaming(linda.output, group)

# save the results
group1 <- paste0(group[1], " vs ","Group",group[2])

for (grp in c(group1)){
  raw_linda_results_genus[[segment]][[grp]] <- 
    rawlinda.df(linda.output,
                grp,
                filt_colon_uni_data,
                filt_colon_uni_taxa)
  
  linda_results_genus[[segment]][[grp]] <- 
    linda.df(linda.output,
             grp,
             filt_colon_uni_data,
             filt_colon_uni_taxa)
}
```

```{r, warning=FALSE, fig.width=5, fig.height=4, fig.fullwidth=TRUE}
# volcano plot
volcano <- volcano_plot_linda(linda.output, group1, 
                                taxa_table = filt_colon_uni_taxa) + 
            ggtitle(paste(group,collapse=" vs "))


# see the plot
volcano
```

###### MaAsLin2

```{r, warning=FALSE, echo=FALSE,results='hide',message=FALSE,warning=FALSE}

log_file <- file(tempfile(), open = "wt")
sink(log_file)  # Redirect standard output
sink(log_file, type = "message")  # Redirect messages and warnings

fit_data = Maaslin2(
    input_data = filt_colon_uni_data, 
    input_metadata = filt_colon_uni_metadata, min_abundance = 0,
    min_prevalence = 0,min_variance = 0,
    output = file.path(path_maaslin,group1), max_significance = 0.05,
    fixed_effects = c('Group'),random_effects = "Patient",
    correction = "BH")

sink()  # Restore standard output
sink(type = "message")  # Restore messages
```

```{r,warning=FALSE}
volcano <- volcano_plot_maaslin(fit_data,filt_colon_uni_taxa) + 
            ggtitle(paste(group[1], "vs", group[2]))

volcano
```

###### Group - Intersection

```{r,warning=FALSE}
intersection_results <- group_intersection(group, list_intersections, list_venns,
                                           linda.output, fit_data,
                                           raw_linda_results_genus, 
                                           segment = segment,
                                           level=level)

list_intersections <- intersection_results[[1]]
list_venns <- intersection_results[[2]]
venn <- intersection_results[[3]]

# show the results
venn
```


###### Basic statistics

```{r,warning=FALSE}
uni_df <-  merge(basic_univariate_statistics(linda_data,group),
                 raw_linda_results_genus[[segment]][[group1]],
                 by="SeqID",all=TRUE)

uni_df[["final_sig"]] <- uni_df$SeqID %in% list_intersections[[paste(segment,level,comparison_name)]][["SeqID"]]
uni_statistics[[segment]][[paste(level,comparison_name)]] <- uni_df

# for comparison
new_name <- comparison_name
addWorksheet(wb, sheetName = new_name)
writeData(wb, sheet = new_name, uni_df, rowNames=FALSE)
```

###### Comparison with PSC-associated taxa

```{r}
hc_ald_sig_colon <- uni_df %>%
  dplyr::filter(final_sig==TRUE) %>%
  dplyr::select(SeqID,Taxonomy,log2FoldChange)

psc_genus_colon <- psc_genus_colon
```

```{r}
list_core <- list()
list_core[["ALD vs HC"]] <- hc_ald_sig_colon$Taxonomy
list_core[["PSC-associated taxa"]] <- psc_genus_colon$Taxonomy

# venn diagram
venn <- ggvenn(list_core, fill_color = c("blue", "red")) + 
    ggtitle("Intersection between PSC-associated taxa and (ALD vs HC)") + 
    guides(fill = "none")

venn
```

From this Venn diagram, we are interested in EXLUSIVE RIGH JOIN

```{r}
A <- list_core$`ALD vs HC`
B <- list_core$`PSC-associated taxa`
remaining_psc <- B[!(B %in% A)]

psc_genus_colon %>%
  dplyr::filter(Taxonomy %in% remaining_psc)

uni_df %>%
  dplyr::filter(Taxonomy %in% remaining_psc)
```

