---
title: "PSC, clinical impact - CZECH"
output:
  html_document:
    df_print: paged
    toc: true
    toc_float: true
    toc_depth: 6
editor_options: 
  markdown: 
    wrap: 72
---

```{r}
library(openxlsx)
library(tidyverse)
library(reshape2)
library(picante)
library(tidyr)
library(ggpubr)
library(mice)
source("custom_functions.R")
```

# Functions

```{r}
clinical_boxplot <- function(metadata,variable){
  metadata_var <- metadata %>% dplyr::select(all_of(c("PatientID",
                                                       variable, "Group"))) %>% as.data.frame()
  metadata_var[,variable] <- as.numeric(metadata_var[,variable])
  metadata_var_without_na <- metadata_var[!is.na(metadata_var[,variable]),]
  colors <- c("#309f87","#f9c675","#F08080","#A00000")
  if (length(unique(metadata_var_without_na$Group))==1) colors <- c("#f9c675")
  if (!"healthy" %in% unique(metadata_var_without_na$Group)) colors <- c("#f9c675","#F08080","#A00000")
  
p <- ggplot(metadata_var_without_na) + 
  geom_boxplot(aes(x=Group, y=!!sym(variable)),outliers = FALSE) + 
    geom_jitter(width = 0.2,height = 0,aes(x=Group, y=!!sym(variable), color=Group),size=2) +
  scale_fill_manual(values=colors) + 
scale_color_manual(values=colors) + 
   theme_minimal() + 
 theme(panel.border = element_rect(color = "black", fill = NA, size = 0))
return(p)
}
```

```{r}
clinical_correlation <- function(metadata,variable,level,
                                 segment="ileum"){
  set.seed(123)
  level <- tolower(level)
  
  metadata_var <- metadata %>% 
    dplyr::select(all_of(c("SampleID", "PatientID",
                paste0("dys_filtered_",level),
                variable, "Group")))
  
  metadata_var[,variable] <- as.numeric(metadata_var[,variable])
  metadata_var <- drop_na(metadata_var)
  
  if (segment=="colon"){
    n_iterations <- 100
    corrs <- list()
    for (i in 1:n_iterations){
      random_sampled_df <- metadata_var %>%
      group_by(PatientID) %>%  # Group by PatientID
      slice_sample(n = 1) %>% # Randomly pick one row per PatientID
      ungroup()
      
      corr <- cor.table(
        random_sampled_df[,c(variable,paste0("dys_filtered_",level))], 
        cor.method="spearman"
        )
      
      corr$P <- corr$P[2]
      corr$r <- corr$r[2]
      corrs[[i]] <- corr
    }
    
    r_values_list <- unlist(map(corrs, ~ .x$r))  # List of r matrices
    p_values_list <- unlist(map(corrs, ~ .x$P))  # List of P matrices
    mean_r <- round(median(r_values_list),2)
    
    p_value <- quantile(p_values_list,0.9)
    corr <- list()
    corr$r <- mean_r
    corr$P <- p_value
  } else {
    corr <- cor.table(
      metadata_var[,c(variable,paste0("dys_filtered_",level))],
      cor.method="spearman")
    
    corr$P <- corr$P[2]
    corr$r <- round(corr$r[2],2)
  }
  
  return(corr)
}

clinical_correlation_abundances <- function(metadata,variable,taxon,level,
                                            segment="ileum"){
  set.seed(123)
  level <- tolower(level)
  
  metadata_var <- metadata %>% dplyr::select(all_of(c("SampleID", "PatientID",
                                                       variable,taxon,
                                                      "Group")))
  metadata_var[,variable] <- as.numeric(metadata_var[,variable])
  metadata_var <- drop_na(metadata_var)
  
  if (segment=="colon"){
    n_iterations <- 100
    corrs <- list()
    for (i in 1:n_iterations){
      random_sampled_df <- metadata_var %>%
      group_by(PatientID) %>%  # Group by PatientID
      slice_sample(n = 1) %>% # Randomly pick one row per PatientID
      ungroup()
      
      corr <- cor.table(random_sampled_df[,c(variable,taxon)], cor.method="spearman")
      corr$P <- corr$P[2]
      corr$r <- corr$r[2]
      corrs[[i]] <- corr
    }
     r_values_list <- unlist(map(corrs, ~ .x$r))  # List of r matrices
    p_values_list <- unlist(map(corrs, ~ .x$P))  # List of P matrices
    mean_r <- round(median(r_values_list),2)
    
    p_value <- quantile(p_values_list,0.9)
    #min_p <- min(p_values_list)
    #min_p <- round(min_p,2)
    #min_p <- ifelse(min_p<=0.05,"< 0.05",ifelse(min_p<=0.01,"< 0.01",ifelse(min_p<=0.001,"< 0.001",paste0("= ",min_p))))
    
    #max_p <- max(p_values_list)
    #max_p <- round(max_p,2)
    #max_p <- ifelse(max_p<=0.05,"< 0.05",ifelse(max_p<=0.01,"< 0.01",ifelse(max_p<=0.001,"< 0.001",max_p)))
    
    #if (min_p==max_p) combined_df <- min_p
    #else {
    #  max_p <- round(quantile(corr$P,0.9),2)
    #  combined_df <- ifelse(max_p<=0.05,"< 0.05",ifelse(max_p<=0.01,"< 0.01",ifelse(max_p<=0.001,"< 0.001",paste("=",max_p) )))
    #}
    corr <- list()
    corr$r <- mean_r
    #corr$P <- combined_df
    corr$P <- p_value
  } else {
    corr <- cor.table(metadata_var[,c(variable,taxon)], cor.method="spearman")
    #max_p <- round(corr$P[2],2)
    #max_p <- 
    #  max_p <- ifelse(max_p<=0.05,"< 0.05",ifelse(max_p<=0.01,"< 0.01",ifelse(max_p<=0.001,"< 0.001",paste0("= ",max_p))))
    #corr$P <- max_p
    corr$r <- round(corr$r[2],2)
    corr$P <- corr$P[2]
  }
  
  return(corr)
}

clinical_scatter <- function(corr,metadata,variable,level){
  level <- tolower(level)
  df_for_scatter_plot <- metadata %>% 
    dplyr::select(all_of(c("SampleID", "PatientID",
                           paste0("dys_filtered_",level),
                           variable, "Group")))
  
  df_for_scatter_plot[,variable] <- as.numeric(df_for_scatter_plot[,variable])
  
  df_for_scatter_plot <- drop_na(df_for_scatter_plot)
  
  colors <- c("#309f87","#f9c675","#F08080","#A00000")
  if (!"healthy" %in% unique(df_for_scatter_plot$Group)) colors <- c("#f9c675","#F08080","#A00000")
  
  r_corr <- corr$r
  p_corr <- corr$P
  p_corr <- ifelse(p_corr < 0.001, "<0.001", ifelse(p_corr < 0.01, "<0.01", ifelse(p_corr < 0.05, "<0.05", paste0("=",round(p_corr,3)))))
  
  x_var <- paste0("dys_filtered_",level)
  x_position <- (max(df_for_scatter_plot[,x_var])) - 0.2*(max(df_for_scatter_plot[,x_var]) - min(df_for_scatter_plot[,x_var]))
  
  y_position <- max(df_for_scatter_plot[,variable]) - 0.05*(max(df_for_scatter_plot[,variable]) - min(df_for_scatter_plot[,variable]))
  
  labels <- paste0("r = ",r_corr, "\\np ",p_corr)
  p <- ggplot(df_for_scatter_plot) + 
  geom_point(aes(x=!!sym(x_var), y=!!sym(variable), color=Group)) + 
  geom_smooth(aes(x=!!sym(x_var), y=!!sym(variable)),method=lm, se=TRUE) +
    geom_text(aes(x=x_position, y=y_position),
                  label=gsub("\\\\n", "\n", labels),hjust = 0) + 
  scale_color_manual(values=colors) + 
  theme_minimal() + 
 theme(panel.border = element_rect(color = "black", fill = NA, size = 0)) + 
  xlab("MDI")
  return(p)
}

clinical_scatter_abundances <- function(corr,metadata,variable,taxon,level,
                                        size=4){
  level <- tolower(level)
  df_for_scatter_plot <- metadata %>% dplyr::select(all_of(c("SampleID", "PatientID",
                                                       taxon,
                                                       variable, "Group")))
  df_for_scatter_plot[,variable] <- as.numeric(df_for_scatter_plot[,variable])
  df_for_scatter_plot <- drop_na(df_for_scatter_plot)
  colors <- c("#309f87","#f9c675","#F08080","#A00000")
  if (!"healthy" %in% unique(df_for_scatter_plot$Group)) colors <- c("#f9c675","#F08080","#A00000")
  
  x_var <- taxon
  r_corr <- corr$r
  p_corr <- corr$P
  
  x_position <- (max(df_for_scatter_plot[,x_var])) - 0.2*(max(df_for_scatter_plot[,x_var]) - min(df_for_scatter_plot[,x_var]))
  y_position <- max(df_for_scatter_plot[,variable]) - 0.05*(max(df_for_scatter_plot[,variable]) - min(df_for_scatter_plot[,variable]))
  labels <- paste0("r = ",r_corr, "\\np ",p_corr)
  p <- ggplot(df_for_scatter_plot) + 
  geom_point(aes(x=!!sym(x_var), y=!!sym(variable), color=Group)) + 
  geom_smooth(aes(x=!!sym(x_var), y=!!sym(variable)),method=lm, se=TRUE) +
    geom_text(aes(x=x_position, y=y_position,label=gsub("\\\\n", "\n", labels)),size=size) + 
  scale_color_manual(values=colors) + 
  theme_minimal() + 
 theme(panel.border = element_rect(color = "black", fill = NA, size = 0))
  return(p)
}

```

```{r}
clinical_contingency <- function(metadata,variable){
  metadata_var <- metadata %>% dplyr::select(all_of(c("PatientID",
                                                       variable, "Group"))) %>% as.data.frame()
  metadata_var_without_na <- metadata_var[!is.na(metadata_var[,variable]),]
  cont_table <- as.data.frame(table(metadata_var_without_na$Group, metadata_var_without_na[,variable]))
  
  p <- ggplot(cont_table, aes(x = Var1, y = Var2, fill = Freq)) +
  geom_tile(color = "white") +  # Tile background color
  geom_text(aes(label = Freq), color = "black", size = 6) +  # Add numbers
  scale_fill_gradient(low = "white", high = "blue") +  # Color gradient
  labs(title = "Contingency Table with Counts", 
       x = "Group", y = variable) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
  return(p)
}

clinical_boxplot_categorical <- function(metadata, variable,level){
  level=tolower(level)
  metadata_var <- metadata %>% dplyr::select(all_of(c("SampleID", "PatientID",
                                                       paste0("dys_filtered_",level),
                                                       variable, "Group")))
  metadata_var <- drop_na(metadata_var)
  metadata_var[,variable] <- as.factor(metadata_var[,variable])

  y_var <- paste0("dys_filtered_",level)
  p <- ggplot(metadata_var) + 
    geom_boxplot(aes(x=!!sym(variable), y=!!sym(y_var)),outliers = FALSE) + 
      geom_jitter(width = 0.2,height = 0,aes(x=!!sym(variable), y=!!sym(y_var),
                                             color=Group),size=2) +
    scale_fill_manual(values=c("#f9c675","#F08080","#A00000")) + 
  scale_color_manual(values=c("#f9c675","#F08080","#A00000")) + 
    theme_minimal() + 
 theme(panel.border = element_rect(color = "black", fill = NA, size = 0))
  return(p)
p
}
```

```{r}
subprepare_for_heatmap <- function(corrs_segment,MDI=TRUE){
    corrs_segment <- corrs_segment[!grepl("_log_",names(corrs_segment))]
    names(corrs_segment) <- gsub("score_","",names(corrs_segment))
    # Example list structure for demonstration
    # Initialize empty lists for r and P values
    r_values <- list()
    p_values <- list()
    
    # Extract segment and clinical variables
    for (name in names(corrs_segment)) {
      if (MDI) {
        clinical_var <- sub("^(.*?_){2}", "",  name)
        if (all(grepl("ileum",names(corrs_segment)))) {
        segment <- "MDI terminal ileum"
      } else if (all(grepl("colon",names(corrs_segment)))) {
        segment <- "MDI colon"
      } else (message("PROBLEM with segment"))
        
      }
      else {
        segment <- sub("^(.*?_){3}", "",  name)
        clinical_var <- sub(segment,"", sub("^(.*?_){2}", "",  name))
        clinical_var <- gsub("_","",clinical_var)
      }
  

      # Fill the r and P lists
      if (!is.null(corrs_segment[[name]]$r)) {
        r_values[[segment]][clinical_var] <- corrs_segment[[name]]$r
      }
      
      if (!is.null(corrs_segment[[name]]$P)) {
        p_values[[segment]][clinical_var] <- corrs_segment[[name]]$P
      }
    }

    # Convert the lists to data frames
    r_df <- do.call(rbind, lapply(r_values, function(x) setNames(as.data.frame(t(x)), names(x))))
    p_df <- do.call(rbind, lapply(p_values, function(x) setNames(as.data.frame(t(x)), names(x))))
    
    # Add row names as segment names
    rownames(r_df) <- names(r_values)
    rownames(p_df) <- names(p_values)
    
    if (MDI) pd_df_corrected <- as.data.frame(p.adjust(p_df,method="BH"))
    else pd_df_corrected <- as.data.frame(apply(p_df,2,function(x) p.adjust(x, method="BH")))
    p_df_sig <- pd_df_corrected
    p_df_sig[,] <- ""
    
    p_df_sig[pd_df_corrected < 0.05] <- "*"
    p_df_sig[pd_df_corrected < 0.01] <- "**"
    p_df_sig[pd_df_corrected < 0.001] <- "***"
    
    if (MDI) p_df_sig <- t(p_df_sig) %>% 
      as.data.frame()  %>% 
      `rownames<-`(segment)
    
    return(list(p_df_sig,r_df))
}

prepare_for_heatmap <- function(corrs_ileum,corrs_colon){

  ileum_list <- subprepare_for_heatmap(corrs_ileum)
  colon_list <- subprepare_for_heatmap(corrs_colon)

  p_df_sig <- rbind(ileum_list[[1]],colon_list[[1]])
  r_df <- rbind(ileum_list[[2]],colon_list[[2]])

  return(list(p_df_sig,r_df))
  
}

```



# Import clinical data

```{r}
# clinical metadata
metadata_clinical <- read.csv("../../data/clinical_data/clinical_metadata_cz.csv")
metadata_clinical$PatientID <- as.character(metadata_clinical$PatientID)

# DYSBIOSIS
metadata_dysbiosis <- read.csv("../../data/clinical_data/dysbiosis_metadata.csv") %>%
  dplyr::filter(Country=="CZ")

# ALPHA DIVERSITY
metadata_alpha_ileum <- read.csv(
  "../results/Q1/alpha_diversity/alpha_indices_terminal_ileum.csv") %>%
  dplyr::filter(Country=="CZ")

metadata_alpha_colon <- read.csv(
  "../results/Q1/alpha_diversity/alpha_indices_colon.csv") %>%
  dplyr::filter(Country=="CZ")

metadata_alpha <- rbind(metadata_alpha_ileum,metadata_alpha_colon) %>%
  mutate(PatientID=Patient) %>%
  dplyr::select(-c(Patient, Group))

# MERGING
metadata_cz <- full_join(metadata_clinical, metadata_dysbiosis, by=c("SampleID","Matrix","PatientID","Group","Country"))

metadata_cz <- full_join(metadata_cz, metadata_alpha, by=c("SampleID","PatientID","Country"))

#metadata_cz %<>% dplyr::filter(Group != "healthy") 
metadata_cz$Group <- factor(metadata_cz$Group,levels = c("healthy","pre_ltx","non-rPSC","rPSC"))
```

```{r}
variables <- c("PatientID", "Group",
               "Bilirubin", "Albumin",
               "ALP", "Platelets",
               "Fecal.calprotectin",
               "MAYO_PSC","AOM_score", 
               "APRI_score",
               "FIB4_score",
               "Nancy_max","eMAYO","MAYO_dai",
                     "Creatinine",
                     "Na",
                     "ALT",
                     "AST",
                     "GGT", "INR")
```

```{r}
metadata_cz$Fecal.calprotectin[metadata_cz$Group=="healthy"] <- NA
metadata_cz$INR[metadata_cz$Group=="healthy"] <- NA
metadata_cz$Fecal.calprotectin[metadata_cz$Fecal.calprotectin==">6000"] <- 6000
metadata_cz$Fecal.calprotectin <- as.numeric(metadata_cz$Fecal.calprotectin)
  
metadata_ileum <- metadata_cz %>% subset(Matrix=="TI") %>% as.data.frame()
metadata_colon <- metadata_cz %>% subset(Matrix!="TI") %>% as.data.frame()

# TO DO!!!!!!!!!!!!!! nefunguje ako ma
metadata_for_boxplots <- metadata_cz[,variables] %>%
  group_by(PatientID) %>%
  distinct(PatientID, .keep_all = TRUE) %>%
  as.data.frame()

```

# Import microbiome data

**Data Import**

Importing ASV, taxa and metadata tables for both Czech and Norway
samples.

*Czech*

```{r}
path = "../../data/analysis_ready_data/ikem/"
asv_tab_ikem <- as.data.frame(fread(file.path(path,"asv_table_ikem.csv"),
                                    check.names = FALSE))
taxa_tab_ikem <- as.data.frame(fread(file.path(path,"taxa_table_ikem.csv"),
                                     check.names = FALSE))
metadata_ikem <- as.data.frame(fread(file.path(path,"metadata_ikem.csv"),
                                     check.names = FALSE))
```

**Spliting to segments**

*Terminal ileum*

```{r}
ileum_data <- merging_data(asv_tab_1=asv_tab_ikem,
                           asv_tab_2=NULL,
                           taxa_tab_1=taxa_tab_ikem,
                           taxa_tab_2=NULL,
                           metadata_1=metadata_ikem,
                           metadata_2=NULL,
                           segment="TI",Q="clinical")

ileum_asv_tab <- ileum_data[[1]]
ileum_taxa_tab <- ileum_data[[2]]
ileum_metadata <- ileum_data[[3]]
```

*Colon*

```{r}
colon_data <- merging_data(asv_tab_1=asv_tab_ikem,
                           asv_tab_2=NULL,
                           taxa_tab_1=taxa_tab_ikem,
                           taxa_tab_2=NULL,
                           metadata_1=metadata_ikem,
                           metadata_2=NULL,
                           segment="colon",Q="clinical")

colon_asv_tab <- colon_data[[1]]
colon_taxa_tab <- colon_data[[2]]
colon_metadata <- colon_data[[3]]
```

**Filtering**

Rules: - prevalence \> 5% (per group) - nearZeroVar with default
settings - sequencing depth \> 5000 - taxonomic assignment at least
order

*Sequencing depth*

```{r}
data_filt <- seq_depth_filtering(ileum_asv_tab,
                                 ileum_taxa_tab,
                                 ileum_metadata,
                                 seq_depth_threshold = 10000)

filt_ileum_asv_tab <- data_filt[[1]]; alpha_ileum_asv_tab <- filt_ileum_asv_tab
filt_ileum_taxa_tab <- data_filt[[2]]; alpha_ileum_taxa_tab <- filt_ileum_taxa_tab
filt_ileum_metadata <- data_filt[[3]]; alpha_ileum_metadata <- filt_ileum_metadata

seq_step <- dim(filt_ileum_asv_tab)[1]
```

*NearZeroVar*

```{r}
data_filt <- nearzerovar_filtering(filt_ileum_asv_tab, 
                                   filt_ileum_taxa_tab,
                                   filt_ileum_metadata)

filt_ileum_asv_tab <- data_filt[[1]]
filt_ileum_taxa_tab <- data_filt[[2]]
nearzero_step <- dim(filt_ileum_asv_tab)[1]
```


# DYSBIOSIS INDEX

## Ileum

**All indices**

```{r,fig.width=12, fig.height=4}
metadata_ileum_melted <- melt(metadata_ileum %>% 
                                dplyr::select("Group", 
                                              "dys_unfiltered_asv",
                                              "dys_unfiltered_genus",
                                              "dys_filtered_asv",
                                              "dys_filtered_genus"))

p_il <- ggplot(metadata_ileum_melted) + 
  geom_boxplot(aes(x=Group, y=value),outliers = FALSE) + 
  geom_jitter(width = 0.2,height = 0,aes(x=Group, y=value, color=Group), 
              size=2) +
  facet_wrap(~variable, ncol = 4,scales = "free") + 
  theme_minimal() + 
 theme(panel.border = element_rect(color = "black", fill = NA, size = 0)) + 
  scale_fill_manual(values=c("#309f87","#f9c675","#F08080","#A00000")) + 
  scale_color_manual(values=c("#309f87","#f9c675","#F08080","#A00000"))

p_il    
```

**FILTERED GENUS**

```{r}
dys_limit <- max(metadata_ileum$dys_filtered_genus,na.rm = TRUE) + 0.6*max(metadata_ileum$dys_filtered_genus,na.rm = TRUE)
dys_min_limit <-  min(metadata_ileum$dys_filtered_genus)

p_il_dys_genus <- ggplot(metadata_ileum %>%
                           mutate(`MDI`=dys_filtered_genus)) + 
  geom_boxplot(aes(x=Group, y=`MDI`),
               outliers = FALSE) + 
  geom_jitter(width = 0.3,height = 0,
              aes(x=Group, y=`MDI`, color=Group,shape=Country),
              size=1.5) +
  theme_minimal() + 
 theme(panel.border = element_rect(color = "black", fill = NA, size = 0))  + 
  ylim(dys_min_limit,dys_limit) + 
  scale_fill_manual(values=c("#309f87","#f9c675","#F08080","#A00000")) + 
  scale_color_manual(values=c("#309f87","#f9c675","#F08080","#A00000")) + 
  theme(axis.text.x = element_text(angle = 45,face = "bold",vjust = 0.5))

p_il_dys_genus
```

## Colon

**All indices**

```{r,fig.width=12, fig.height=4}
colon_metadata_melted <- melt(metadata_colon %>% 
                                dplyr::select("Group", 
                                              "dys_unfiltered_asv",
                                              "dys_unfiltered_genus",
                                              "dys_filtered_asv",
                                              "dys_filtered_genus"))
p_col <- ggplot(colon_metadata_melted) + 
  geom_boxplot(aes(x=Group, y=value),outliers = FALSE) + 
  geom_jitter(width = 0.2,height = 0,
              aes(x=Group, y=value, color=Group),
              size=2) +
  facet_wrap(~variable, ncol = 4,scales = "free") + 
  theme_minimal() + 
 theme(panel.border = element_rect(color = "black", fill = NA, size = 0)) +
  scale_fill_manual(values=c("#309f87","#f9c675","#F08080","#A00000")) + 
  scale_color_manual(values=c("#309f87","#f9c675","#F08080","#A00000"))

p_col  
```

**FILTERED GENUS**

```{r}
dys_limit <- max(metadata_colon$dys_filtered_genus,na.rm = TRUE) + 0.6*max(metadata_colon$dys_filtered_genus,na.rm = TRUE)
dys_min_limit <-  min(metadata_colon$dys_filtered_genus)

p_col_dys_genus <- ggplot(metadata_colon %>%
                            mutate(`MDI`=dys_filtered_genus)) + 
  geom_boxplot(aes(x=Group, y=`MDI`),outliers = FALSE) + 
  geom_jitter(width = 0.3,height = 0,
              aes(x=Group, y=`MDI`, color=Group,shape=Country),
              size=1) +
  theme_minimal() + 
 theme(panel.border = element_rect(color = "black", fill = NA, size = 0))+
  ylim(dys_min_limit,dys_limit) + 
  scale_fill_manual(values=c("#309f87","#f9c675","#F08080","#A00000")) + 
  scale_color_manual(values=c("#309f87","#f9c675","#F08080","#A00000")) + 
  theme(axis.text.x = element_text(angle = 45,face = "bold",vjust = 0.5))

p_col_dys_genus
```

## Linear models

**Ileum**

```{r}
results_model <- pairwise.lm(
  formula = "dys_filtered_genus ~ Group",
  factors=metadata_ileum$Group,
  data=metadata_ileum)

knitr::kable(results_model[[1]][,c("Estimate", "p.adj","sig")],
             digits=3,caption="Results of linear modeling")
```

```{r}
results_model <- pairwise.lmer(
  formula = "dys_filtered_genus ~ Group + (1|PatientID)",
  factors=metadata_colon$Group,
  data=metadata_colon)

knitr::kable(results_model[[1]][,c("Estimate", "p.adj","sig")],
             digits=3,caption="Results of linear mixed-effect modeling")

```


```{r, eval=FALSE}
dys <- ggarrange(p_il + ggtitle("Ileum"),
                 p_col + ggtitle("Colon"),nrow = 2)

pdf("results/clinical/dysbiosis_index_boxplots.pdf",
    height = 8,width =15)

dys
dev.off()
```

# ALPHA DIVERSITY

```{r}
alpha_div_plots <- list()
alpha_boxplots <- list()
corrs <- list()
```

## Ileum

```{r}
ileum_metadata_alpha_final <- metadata_ileum %>% 
  dplyr::select("SampleID", "PatientID","Group",
                "Observe","Shannon","Simpson","Pielou",
                "dys_filtered_asv", "dys_filtered_genus")
```

```{r}
variables <- c("Observe","Shannon","Simpson","Pielou")

for (clinical_variable in variables){
  
  # ILEUM
  # boxplot
  p <- clinical_boxplot(ileum_metadata_alpha_final,
                      variable=clinical_variable)

  alpha_boxplots[[paste0("ileum_",clinical_variable)]] <- p
  
  # correlation
  ## ASV
  level="ASV"
  corr <- clinical_correlation(
    ileum_metadata_alpha_final,
    clinical_variable,
    level)
  
  corrs[[paste0("ileum_asv_",clinical_variable)]] <- corr
  
  p <- clinical_scatter(corr,
                        ileum_metadata_alpha_final,
                        clinical_variable,
                        level)
  
  alpha_div_plots[[paste0("ileum_asv_",clinical_variable)]] <- p

  ## Genus
  level="genus"
  corr <- clinical_correlation(ileum_metadata_alpha_final,
                               clinical_variable,
                               level)
  corrs[[paste0("ileum_genus_",clinical_variable)]] <- corr
  
  p <- clinical_scatter(corr,
                        ileum_metadata_alpha_final,
                        clinical_variable,
                        level)
  
  alpha_div_plots[[paste0("ileum_genus_",clinical_variable)]] <- p
}
```

### Boxplot

```{r,fig.width=12, fig.height=4}
alpha_boxplots_ileum <- ggarrange(
  plotlist = alpha_boxplots[grepl("ileum",names(alpha_boxplots))],
  ncol = 4,
  common.legend = TRUE, 
  legend = "right")

alpha_boxplots_ileum
```


## Colon

```{r}
colon_metadata_alpha_final <- metadata_colon %>% 
  dplyr::select("SampleID", "PatientID", "Group",
                "Observe","Shannon","Simpson","Pielou",
                "dys_filtered_asv","dys_filtered_genus")
```

```{r}
variables <- c("Observe","Shannon","Simpson","Pielou")

for (clinical_variable in variables){

  # COLUMN
  p <- clinical_boxplot(colon_metadata_alpha_final,
                        variable=clinical_variable)

  alpha_boxplots[[paste0("colon_",clinical_variable)]] <- p
  
  # correlation
  ## ASV
  level="ASV"
  corr <- clinical_correlation(colon_metadata_alpha_final,
                               clinical_variable,
                               level,
                               segment="colon")
  
  corrs[[paste0("colon_asv_",clinical_variable)]] <- corr
  
  p <- clinical_scatter(corr,
                        colon_metadata_alpha_final,
                        clinical_variable,
                        level)
  
  alpha_div_plots[[paste0("colon_asv_",clinical_variable)]] <- p

  ## Genus
  level="genus"
  corr <- clinical_correlation(colon_metadata_alpha_final,
                               clinical_variable,
                               level,
                               segment="colon")
  
  corrs[[paste0("colon_genus_",clinical_variable)]] <- corr
  
  p <- clinical_scatter(corr,
                        colon_metadata_alpha_final,
                        clinical_variable,
                        level)
  
  alpha_div_plots[[paste0("colon_genus_",clinical_variable)]] <- p
}

```

### Boxplot

```{r,fig.width=12, fig.height=4}
alpha_boxplots_colon <- ggarrange(
  plotlist = alpha_boxplots[grepl("colon",names(alpha_boxplots))],
  ncol = 4,
  common.legend = TRUE, 
  legend = "right")

alpha_boxplots_colon
```

```{r, eval=FALSE}
alpha_boxplot <- ggarrange(alpha_boxplots_ileum,
                           alpha_boxplots_colon,
                           nrow = 2)
pdf("results/clinical/alpha_diversity_boxplots.pdf",height = 7,width =17)
alpha_boxplot
dev.off()
```

## Scatterplots

**ASV**

```{r,fig.width=20, fig.height=8}
alpha_dys_asv <- ggarrange(
  plotlist = alpha_div_plots[grepl("asv",names(alpha_div_plots))],
  ncol=4,
  nrow=2,
  common.legend = TRUE,
  legend="right")

alpha_dys_asv
```

**Genus**

```{r,fig.width=20, fig.height=8}
alpha_dys_genus <- ggarrange(
  plotlist = alpha_div_plots[grepl("genus",names(alpha_div_plots))],
  ncol=4,
  nrow=2,
  common.legend = TRUE,
  legend="right")

alpha_dys_genus
```

**Saving plots**

```{r, eval=FALSE}
pdf("results/clinical/alpha_diversity.pdf",height = 8,width =18)
alpha_dys_asv
dev.off()
```

```{r, eval=FALSE}
pdf("results/clinical/alpha_diversity_genus.pdf",height = 8,width =18)
alpha_dys_genus
dev.off()
```

## Linear models

### Richness

**Ileum**

```{r}
results_model <- pairwise.lm(formula = "Observe ~ Group",
                                     factors=metadata_ileum$Group,
                                     data=metadata_ileum)

knitr::kable(results_model[[1]][,c("Estimate", "p.adj","sig")],
             digits=3,caption="Results of linear modeling")
```

**Colon**

```{r}
results_model <- pairwise.lmer(
  formula = "Observe ~ Group + (1|Patient)",
  factors=metadata_colon$Group,
  data=metadata_colon %>% mutate(Patient=PatientID))

knitr::kable(results_model[[1]][,c("Estimate", "p.adj","sig")],
             digits=3,caption="Results of linear modeling")
```

### Shannon

**Ileum**

```{r}
results_model <- pairwise.lm(formula = "Shannon ~ Group",
                                     factors=metadata_ileum$Group,
                                     data=metadata_ileum)

knitr::kable(results_model[[1]][,c("Estimate", "p.adj","sig")],
             digits=3,caption="Results of linear modeling")
```

**Colon**

```{r}
results_model <- pairwise.lmer(
  formula = "Shannon ~ Group + (1|Patient)",
  factors=metadata_colon$Group,
  data=metadata_colon %>% mutate(Patient=PatientID))

knitr::kable(results_model[[1]][,c("Estimate", "p.adj","sig")],
             digits=3,caption="Results of linear modeling")
```


# MDI and clinical variables

```{r}

variables <- c("ALP","log_ALP",
               "GGT","log_GGT",
               "ALT","log_ALT",
               "AST","log_AST",
               "Bilirubin","log_Bilirubin",
               "INR",
               "Albumin",
               "Platelets",
               "Creatinine","log_Creatinine",
               "FIB4_score",
               "APRI_score",
               "MAYO_PSC",
               "AOM_score",
               "Fecal.calprotectin",
               "Nancy_max",
               "eMAYO",
               "MAYO_dai"
               )


```


```{r, fig.width=5, fig.height=3}
boxplots_plots <- list()
corrs <- list()
clinical_plots <- list()


#metadata_ileum %<>% dplyr::filter(Group!="healthy")
#metadata_colon %<>% dplyr::filter(Group!="healthy")

for (clinical_variable in variables){
  
  if (grepl("log_",clinical_variable)) {
   metadata_for_boxplots[,clinical_variable] <- log(as.numeric(metadata_for_boxplots[,gsub("log_","",clinical_variable)]))
   
   metadata_ileum[,clinical_variable] <- log(as.numeric(metadata_ileum[,gsub("log_","",clinical_variable)]))
   metadata_colon[,clinical_variable] <- log(as.numeric(metadata_colon[,gsub("log_","",clinical_variable)]))
  }
  
  # boxplot
  p <- clinical_boxplot(metadata_for_boxplots,
                      variable=clinical_variable)

  boxplots_plots[[clinical_variable]] <- p
  
  # ILEUM
  # correlation
  
  ## ASV
  level="ASV"
  corr <- clinical_correlation(metadata_ileum,
                               clinical_variable,
                               level)
  
  corrs[[paste0("ileum_asv_",clinical_variable)]] <- corr
  
  p <- clinical_scatter(corr,
                        metadata_ileum,
                        clinical_variable,
                        level)
  clinical_plots[[paste0("ileum_asv_",clinical_variable)]] <- p

  ## Genus
  level="genus"
  corr <- clinical_correlation(metadata_ileum,
                               clinical_variable,
                               level)
  corrs[[paste0("ileum_genus_",clinical_variable)]] <- corr
  
  p <- clinical_scatter(corr,
                        metadata_ileum,
                        clinical_variable,
                        level)
  clinical_plots[[paste0("ileum_genus_",clinical_variable)]] <- p
  
  # correlation
  ## ASV
  level="ASV"
  corr <- clinical_correlation(metadata_colon,
                               clinical_variable,
                               level,
                               segment="colon")
  
  corrs[[paste0("colon_asv_",clinical_variable)]] <- corr
  
  p <- clinical_scatter(corr,
                        metadata_colon,
                        clinical_variable,
                        level)
  
  clinical_plots[[paste0("colon_asv_",clinical_variable)]] <- p

  ## Genus
  level="genus"
  corr <- clinical_correlation(metadata_colon,
                               clinical_variable,
                               level,
                               segment="colon")
  
  corrs[[paste0("colon_genus_",clinical_variable)]] <- corr
  
  p <- clinical_scatter(corr,
                        metadata_colon,
                        clinical_variable,
                        level)
  
  clinical_plots[[paste0("colon_genus_",clinical_variable)]] <- p
}

```

```{r, eval=FALSE}
variables <- c("Nancy_max","eMAYO","MAYO_dai")
```

```{r,eval=FALSE}
for (clinical_variable in variables){
  # CONTINGENCY TABLE
  p <- clinical_contingency(metadata_for_boxplots,
                            clinical_variable)
  boxplots_plots[[clinical_variable]] <- p
  
  # ILEUM
  ## ASV
  p <- clinical_boxplot_categorical(metadata_ileum,
                                    clinical_variable,
                                    level = "ASV")
  clinical_plots[[paste0("ileum_asv_",clinical_variable)]] <- p
  
  ## GENUS
  p <- clinical_boxplot_categorical(metadata_ileum,
                                    clinical_variable,
                                    level = "Genus")
  clinical_plots[[paste0("ileum_genus_",clinical_variable)]] <- p
  
  # COLON
  ## ASV
  p <- clinical_boxplot_categorical(metadata_ileum,
                                    clinical_variable,
                                    level = "ASV")
  clinical_plots[[paste0("colon_asv_",clinical_variable)]] <- p
  
  ## GENUS
  p <- clinical_boxplot_categorical(metadata_ileum,
                                    clinical_variable,
                                    level = "Genus")
  clinical_plots[[paste0("colon_genus_",clinical_variable)]] <- p
}

```

## Boxplots

```{r}
for (i in 1:length(boxplots_plots)){
  name <- names(boxplots_plots)[i]
  pl <- boxplots_plots[[i]]
  pl <- pl + ggtitle(name)
  boxplots_plots[[i]] <- pl
}
bx <- ggarrange(plotlist = boxplots_plots, ncol = 4,nrow=4)
```

```{r,fig.width=15, fig.height=12}
bx
```

```{r, eval=FALSE}
pdf("results/clinical/clinical_data_boxplots.pdf",height = 15,width =13)
bx
dev.off()
```

## Clinical plots - MDI ~ clinical variable

```{r}
for (i in 1:length(clinical_plots)){
  name <- names(clinical_plots)[i]
  pl <- clinical_plots[[i]]
  pl <- pl + ggtitle(name)
  clinical_plots[[i]] <- pl
}

cl_asv <- ggarrange(
  plotlist = clinical_plots[grepl("asv",names(clinical_plots))],
  ncol=4,nrow=8)

cl_genus <- ggarrange(plotlist = clinical_plots[grepl("genus",names(clinical_plots))],
          ncol=4,nrow=8)

```

```{r,fig.width=20, fig.height=22}
cl_asv
```

```{r,fig.width=20, fig.height=22}
cl_genus
```

```{r, eval=FALSE}
pdf("results/clinical/correlation.pdf",height = 22,width =18)
cl_asv
dev.off()
```


```{r, eval=FALSE}
pdf("results/clinical/correlation_genus.pdf",height = 22,width =18)
cl_genus
dev.off()

```

## HEATMAP

### ASV level

```{r}
corrs_ileum <- corrs[grepl("ileum_asv",names(corrs))]
corrs_colon <- corrs[grepl("colon_asv",names(corrs))]
```

```{r,fig.width=3,fig.height=7}
prepared_list <- prepare_for_heatmap(corrs_ileum,corrs_colon)
p_df_sig_mdi <- prepared_list[[1]]
r_df_mdi <- prepared_list[[2]]
```

```{r,fig.width=4,fig.height=7}
r_df_melt_mdi <- melt(r_df_mdi %>% rownames_to_column("SeqID"))
p_df_melt_mdi <- melt(p_df_sig_mdi %>% rownames_to_column("SeqID"),
                      id.vars = c("SeqID"))

r_df_melt_mdi$SeqID <- factor(r_df_melt_mdi$SeqID,
                              levels=unique(r_df_melt_mdi$SeqID))

p_df_melt_mdi$SeqID <- factor(p_df_melt_mdi$SeqID,
                              levels=unique(p_df_melt_mdi$SeqID))

r_df_melt_mdi$variable <- factor(r_df_melt_mdi$variable,
                              levels=rev(unique(r_df_melt_mdi$variable)))

p_df_melt_mdi$variable <- factor(p_df_melt_mdi$variable,
                              levels=rev(unique(p_df_melt_mdi$variable)))

p <- ggplot() + 
  geom_tile(data=r_df_melt_mdi, aes(SeqID, variable, fill= value))  + 
  theme_minimal() + 
  scale_fill_gradient2(name = "Rho", low = "blue", mid = "white", high = "red", midpoint = 0,limits = c(-0.7, 0.7)) + 
  geom_text(data=p_df_melt_mdi,aes(x=SeqID,y=variable,label=value,vjust=0.6)) + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  xlab("") + ylab("")  + 
  ggtitle("MDI")+ 
  theme(plot.title = element_text(hjust=0.5,face = "bold",size = 15))

p
```

### Genus level

```{r}
corrs_ileum <- corrs[grepl("ileum_genus",names(corrs))]
corrs_colon <- corrs[grepl("colon_genus",names(corrs))]
```

```{r}
prepared_list <- prepare_for_heatmap(corrs_ileum,corrs_colon)
p_df_sig_mdi <- prepared_list[[1]]
r_df_mdi <- prepared_list[[2]]
```

```{r,fig.width=4,fig.height=7}
r_df_melt_mdi <- melt(r_df_mdi %>% rownames_to_column("SeqID"))
p_df_melt_mdi <- melt(p_df_sig_mdi %>% rownames_to_column("SeqID"),
                      id.vars = c("SeqID"))

r_df_melt_mdi$SeqID <- factor(r_df_melt_mdi$SeqID,
                              levels=unique(r_df_melt_mdi$SeqID))

p_df_melt_mdi$SeqID <- factor(p_df_melt_mdi$SeqID,
                              levels=unique(p_df_melt_mdi$SeqID))

r_df_melt_mdi$variable <- factor(r_df_melt_mdi$variable,
                              levels=rev(unique(r_df_melt_mdi$variable)))

p_df_melt_mdi$variable <- factor(p_df_melt_mdi$variable,
                              levels=rev(unique(p_df_melt_mdi$variable)))

p_mdi <- ggplot() + 
  geom_tile(data=r_df_melt_mdi, aes(SeqID, variable, fill= value))  + 
  theme_minimal() + 
  scale_fill_gradient2(name = "Rho", low = "blue", mid = "white", high = "red", midpoint = 0,limits = c(-0.7, 0.7)) + 
  geom_text(data=p_df_melt_mdi,aes(x=SeqID,y=variable,label=value,vjust=0.6)) + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  xlab("") + ylab("")  + 
  ggtitle("MDI")+ 
  theme(plot.title = element_text(hjust=0.5,face = "bold",size = 15))

p_mdi
```

# Bacteria and significant clinical variables

## Terminal ileum

### Genus level

```{r}
level="genus"
```

Aggregation, filtering

```{r}
# Aggregation
genus_data <- aggregate_taxa(ileum_asv_tab,
                             ileum_taxa_tab,
                             taxonomic_level=level,
                             names=TRUE)
# Filtration
filt_data <- filtering_steps(genus_data[[1]],
                             genus_data[[2]],
                             ileum_metadata,
                             seq_depth_threshold=10000)

filt_ileum_genus_tab <- filt_data[[1]]
filt_ileum_genus_taxa <- filt_data[[2]]
filt_ileum_metadata_genus <- filt_data[[3]]
```

```{r}
psc_effect_ileum_genus <- read.xlsx(
  "../results/Q1_czech/univariate_analysis/supplements_psc_effect_terminal_ileum.xlsx",
  sheet = "terminal_ileum genus")
psc_taxa <- psc_effect_ileum_genus$SeqID
increased <- psc_effect_ileum_genus$ASV[psc_effect_ileum_genus$log2FoldChange >0]
decreased <- psc_effect_ileum_genus$ASV[psc_effect_ileum_genus$log2FoldChange <0]
```

**Increased + decreased **

```{r}
my_table <- filt_ileum_genus_tab %>% column_to_rownames("SeqID")
data_clr <- vegan::decostand(my_table,method = "clr", MARGIN = 2,pseudocount=0.5) %>% as.matrix()
data_clr <- data_clr[psc_taxa,] %>% t() %>% as.data.frame() %>% rownames_to_column("SampleID")

metadata_with_abundances <- data_clr %>% full_join(metadata_ileum,by="SampleID")
```

```{r}
# significant variables
variables <- c("ALP",
               "GGT",
               "APRI_score", 
               "FIB4_score",
               "Albumin",
               "Platelets")

```

```{r}
corrs_ileum <- corrs[grepl("ileum_genus",names(corrs))]
corrs_colon <- corrs[grepl("colon_genus",names(corrs))]
```

```{r,fig.width=3,fig.height=7}
prepared_list <- prepare_for_heatmap(corrs_ileum,corrs_colon)
p_df_sig_mdi <- prepared_list[[1]]

variables <- colnames(p_df_sig_mdi)[!apply(p_df_sig_mdi,2,function(x) {
  all(x == "")})]
```


```{r, fig.width=5, fig.height=3}
corrs <- list()
clinical_plots <- list()

for (clinical_variable in variables){
  for (taxon in psc_taxa){
    if (grepl("log_",clinical_variable)) {
   metadata_with_abundances[,clinical_variable] <- log(as.numeric(metadata_with_abundances[,gsub("log_","",clinical_variable)]))
  }
  
  # ILEUM
  # correlation
  corr <- clinical_correlation_abundances(metadata_with_abundances,clinical_variable,taxon,level="genus")
  corrs[[paste0("ileum_genus_",clinical_variable,"_", taxon)]] <- corr
  if (corr$P <0.05){
    p <- clinical_scatter_abundances(corr,metadata_with_abundances,clinical_variable,taxon,level="genus",size = 3)
  clinical_plots[[paste0("ileum_genus_",clinical_variable,"_",taxon)]] <- p
  } else print(corr$P)
  }
  
}
```

```{r}
prepared_list <- subprepare_for_heatmap(corrs,MDI = FALSE)
p_df_sig <- prepared_list[[1]]
r_df <- prepared_list[[2]]
```

```{r}
r_df_melt <- melt(r_df %>% rownames_to_column("SeqID"))
p_df_melt <- melt(p_df_sig %>% rownames_to_column("SeqID"),id.vars = c("SeqID"))

r_df_melt$variable <- factor(r_df_melt$variable, 
                             levels=rev(unique(r_df_melt$variable)))
p_df_melt$variable <- factor(p_df_melt$variable,
                             levels =rev(unique(p_df_melt$variable)))

r_df_melt$SeqID <- factor(r_df_melt$SeqID, levels = unique(r_df_melt$SeqID))
p_df_melt$SeqID <- factor(p_df_melt$SeqID, levels = unique(p_df_melt$SeqID))

p_ileum <- ggplot() + 
  geom_tile(data=r_df_melt, aes(variable, SeqID, fill= value))  + 
  theme_minimal() + 
  scale_fill_gradient2(name = "Rho", low = "blue", mid = "white", high = "red", midpoint = 0,limits = c(-0.5, 0.5)) + 
  geom_text(data=p_df_melt,aes(x=variable,y=SeqID,label=value,vjust=0.6)) + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  xlab("") + ylab("")  + 
  ggtitle("Terminal ileum")+ 
  theme(plot.title = element_text(hjust=0.5,face = "bold",size = 15))

p_ileum
```

## Colon

### Genus level

Aggregation, filtering

```{r}
genus_data <- aggregate_taxa(colon_asv_tab,
                             colon_taxa_tab,
                             taxonomic_level="Genus",
                             names=TRUE)

filt_colon_genus_tab <- genus_data[[1]]
filt_colon_genus_taxa <- genus_data[[2]]
filt_colon_metadata <- colon_metadata
```

```{r}
psc_effect_colon_genus <- read.xlsx("../results/Q1_czech/univariate_analysis/supplements_psc_effect_colon.xlsx",
                                    sheet = "colon genus")

psc_taxa <- psc_effect_colon_genus$SeqID
increased <- psc_effect_colon_genus$SeqID[psc_effect_colon_genus$log2FoldChange >0]
decreased <- psc_effect_colon_genus$SeqID[psc_effect_colon_genus$log2FoldChange <0]
```


```{r}
my_table <- filt_colon_genus_tab %>% column_to_rownames("SeqID")
data_clr <- vegan::decostand(my_table,method = "clr", MARGIN = 2,pseudocount=0.5) %>% as.matrix()
data_clr <- data_clr[psc_taxa,] %>% t() %>% as.data.frame() %>% rownames_to_column("SampleID")

metadata_with_abundances <- data_clr %>% full_join(metadata_colon,by="SampleID")
```

```{r}
variables <- c("ALP",
               "GGT",
               "APRI_score", 
               "FIB4_score",
               "Albumin",
               "Platelets")
```


```{r, fig.width=5, fig.height=3}
corrs <- list()
clinical_plots <- list()

for (clinical_variable in variables){
  for (taxon in psc_taxa){
    if (grepl("log_",clinical_variable)) {
   metadata_with_abundances[,clinical_variable] <- log(as.numeric(metadata_with_abundances[,gsub("log_","",clinical_variable)]))
  }
  
  # ILEUM
  # correlation
  corr <- clinical_correlation_abundances(metadata_with_abundances,clinical_variable,taxon,level="genus")
  corrs[[paste0("colon_genus_",clinical_variable,"_", taxon)]] <- corr
  if ( (corr$P == "< 0.05") | (corr$P == "< 0.01") | (corr$P == "< 0.01") ){
    p <- clinical_scatter_abundances(corr,metadata_with_abundances,clinical_variable,taxon,level="genus",size = 3)
  clinical_plots[[paste0("colon_genus_",clinical_variable,"_",taxon)]] <- p
  } else print(corr$P)
}
  

}
```

```{r}
clinical_plots_colon <- clinical_plots
corrs_colon <- corrs
```

```{r}
prepared_list <- subprepare_for_heatmap(corrs,MDI = FALSE)
p_df_sig <- prepared_list[[1]]
r_df <- prepared_list[[2]]
```

```{r}
r_df_melt <- melt(r_df %>% rownames_to_column("SeqID"))
p_df_melt <- melt(p_df_sig %>% rownames_to_column("SeqID"),id.vars = c("SeqID"))

r_df_melt$variable <- factor(r_df_melt$variable, 
                             levels=rev(unique(r_df_melt$variable)))
p_df_melt$variable <- factor(p_df_melt$variable,
                             levels =rev(unique(p_df_melt$variable)))

r_df_melt$SeqID <- factor(r_df_melt$SeqID, levels = unique(r_df_melt$SeqID))
p_df_melt$SeqID <- factor(p_df_melt$SeqID, levels = unique(p_df_melt$SeqID))

p_colon <- ggplot() + 
  geom_tile(data=r_df_melt, aes(variable, SeqID, fill= value))  + 
  theme_minimal() + 
  scale_fill_gradient2(name = "Rho", low = "blue", mid = "white", high = "red", midpoint = 0,limits = c(-0.5, 0.5)) + 
  geom_text(data=p_df_melt,aes(x=variable,y=SeqID,label=value,vjust=0.6)) + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  xlab("") + ylab("")  + 
  ggtitle("Colon")+ 
  theme(plot.title = element_text(hjust=0.5,face = "bold",size = 15))

p_colon
```

## heatmap plot

```{r,fig.width=15,fig.height=5}
ggarrange(p_mdi,p_ileum,p_colon,ncol = 3,widths = c(0.7,1,1))
```


# PCA and clinical variables fit

## Terminal ileum

**Analysis**

```{r}
segment="terminal_ileum"
```

### Main analysis - Genus, Aitchison

**Genus level, Aitchison distance**

```{r}
level="genus"
```

Aggregation, filtering

```{r,warning=FALSE}
# Aggregation
genus_data <- aggregate_taxa(ileum_asv_tab,
                             ileum_taxa_tab,
                             taxonomic_level=level,
                             names=TRUE)
# Filtration
filt_data <- filtering_steps(genus_data[[1]],
                             genus_data[[2]],
                             ileum_metadata,
                             seq_depth_threshold=10000)

filt_ileum_genus_tab <- filt_data[[1]]
filt_ileum_genus_taxa <- filt_data[[2]]
filt_ileum_metadata_genus <- filt_data[[3]]
```

**Plots**


```{r,warning=FALSE}
p <- pca_plot_custom(filt_ileum_genus_tab,
                                 filt_ileum_genus_taxa,
                                 filt_ileum_metadata_genus,
                                 show_boxplots = TRUE,
                                 variable = "Group", size=3, 
                     show_legend=TRUE,
                     clinical = TRUE,
                     clinical_metadata = metadata_cz)


# see the results
p
```


### Supplementary analysis

#### Genus level

```{r}
level="genus"
```

##### Bray-Curtis

**Plots**

```{r,warning=FALSE}
p <- pca_plot_custom(filt_ileum_genus_tab,
                                 filt_ileum_genus_taxa,
                                 filt_ileum_metadata_genus,
                                 measure = "bray",
                                 show_boxplots = TRUE,
                                 variable = "Group", size=3, show_legend=TRUE,
                     clinical=TRUE,
                     clinical_metadata=metadata_cz)

# see the results
p
```

##### Jaccard

**Plots**


```{r,warning=FALSE}
p <- pca_plot_custom(filt_ileum_genus_tab,
                                 filt_ileum_genus_taxa,
                                 filt_ileum_metadata_genus,
                                 measure = "jaccard",
                                 show_boxplots = TRUE,
                                 variable = "Group", size=3, show_legend=TRUE,
                     clinical=TRUE,
                     clinical_metadata = metadata_cz)

# see the results
p
```

#### ASV level

```{r}
level="ASV"
```

##### Aitchison 

**PCoA**

```{r,warning=FALSE}
p <- pca_plot_custom(filt_ileum_asv_tab,
                           filt_ileum_taxa_tab,
                           filt_ileum_metadata,
                           show_boxplots = TRUE,
                           variable = "Group", size=3, show_legend=TRUE,
                     clinical=TRUE,
                     clinical_metadata = metadata_cz)


# see the results
p
```

##### Bray-Curtis 

**PCoA**

```{r,warning=FALSE}
p <- pca_plot_custom(filt_ileum_asv_tab,
                     filt_ileum_taxa_tab,
                     filt_ileum_metadata,
                     measure = "bray",
                     show_boxplots = TRUE,
                     variable = "Group", size=3, show_legend=TRUE,
                     clinical=TRUE,
                     clinical_metadata=metadata_cz)

# see the results
p
```

##### Jaccard

**PCoA**

```{r,warning=FALSE}
p <- pca_plot_custom(filt_ileum_asv_tab,
                     filt_ileum_taxa_tab,
                     filt_ileum_metadata,
                     measure = "jaccard",
                     show_boxplots = TRUE,
                     variable = "Group", size=3, show_legend=TRUE,
                     clinical=TRUE,
                     clinical_metadata=metadata_cz)

# see the results
p
```

## Colon

```{r}
segment="colon"
```

**Filtering**

Rules: - prevalence \> 5% (per group) - nearZeroVar with default
settings - sequencing depth \> 5000 - taxonomic assignment at least
order

*Sequencing depth*

```{r}
data_filt <- seq_depth_filtering(colon_asv_tab,
                                 colon_taxa_tab,
                                 colon_metadata,
                                 seq_depth_threshold = 10000)

filt_colon_asv_tab <- data_filt[[1]]; alpha_colon_asv_tab <- filt_colon_asv_tab
filt_colon_taxa_tab <- data_filt[[2]]; alpha_colon_taxa_tab <- filt_colon_taxa_tab
filt_colon_metadata <- data_filt[[3]]; alpha_colon_metadata <- filt_colon_metadata

seq_step <- dim(filt_colon_asv_tab)[1]
```

*NearZeroVar*

```{r}
data_filt <- nearzerovar_filtering(filt_colon_asv_tab,
                                   filt_colon_taxa_tab,
                                   filt_colon_metadata)

filt_colon_asv_tab <- data_filt[[1]]
filt_colon_taxa_tab <- data_filt[[2]]
nearzero_step <- dim(filt_colon_asv_tab)[1]
```

*Check zero depth*

```{r}
data_filt <- check_zero_depth(filt_colon_asv_tab, 
                              filt_colon_taxa_tab, 
                              filt_colon_metadata)

filt_colon_asv_tab <- data_filt[[1]]; 
filt_colon_taxa_tab <- data_filt[[2]]; 
filt_colon_metadata <- data_filt[[3]]; 
```

Calculating Aitchison distance (euclidean distance on clr-transformed
data), both at ASV and genus level.

### Main analysis - Genus, Aitchison

**Genus level, Aitchison distance**

```{r}
level="genus"
```

Aggregation, filtering

```{r}
# Aggregation
genus_data <- aggregate_taxa(colon_asv_tab,
                             colon_taxa_tab,
                             taxonomic_level=level,
                             names=TRUE)
# Filtration
filt_data <- filtering_steps(genus_data[[1]],
                             genus_data[[2]],
                             colon_metadata,
                             seq_depth_threshold=10000)

filt_colon_genus_tab <- filt_data[[1]]
filt_colon_genus_taxa <- filt_data[[2]]
filt_colon_genus_metadata <- filt_data[[3]]
```

**PCoA**

```{r,warning=FALSE}
p <- pca_plot_custom(filt_colon_genus_tab,
                                 filt_colon_genus_taxa,
                                 filt_colon_genus_metadata,
                                 show_boxplots = TRUE,
                                 variable = "Group", size=3, show_legend=TRUE,
                     clinical=TRUE, clinical_metadata = metadata_cz)

# see the results
p
```

### Supplementary analysis

#### Genus level

```{r}
level="genus"
```

##### Bray-Curtis

**Plots**

```{r,warning=FALSE}
p <- pca_plot_custom(filt_colon_genus_tab,
                                 filt_colon_genus_taxa,
                                 filt_colon_genus_metadata,
                                 measure = "bray",
                                 show_boxplots = TRUE,
                                 variable = "Group", size=3, show_legend=TRUE,
                     clinical=TRUE, clinical_metadata = metadata_cz)


# see the results
p
```

##### Jaccard

**Plots**

```{r,warning=FALSE}
p <- pca_plot_custom(filt_colon_genus_tab,
                                 filt_colon_genus_taxa,
                                 filt_colon_genus_metadata,
                                 measure = "jaccard",
                                 show_boxplots = TRUE,
                                 variable = "Group", size=3, show_legend=TRUE,
                     clinical=TRUE, clinical_metadata = metadata_cz)


# see the results
p
```

#### ASV level

```{r}
level="ASV"
```

##### Aitchison 

**PCoA**

```{r,warning=FALSE}
p <- pca_plot_custom(filt_colon_asv_tab,
                           filt_colon_taxa_tab,
                           filt_colon_metadata,
                           show_boxplots = TRUE,
                           variable = "Group", size=3, show_legend=TRUE,
                     clinical=TRUE, clinical_metadata = metadata_cz)


# see the results
p
```

##### Bray-Curtis 

**PCoA**

```{r,warning=FALSE}
p <- pca_plot_custom(filt_colon_asv_tab,
                     filt_colon_taxa_tab,
                     filt_colon_metadata,
                     measure = "bray",
                     show_boxplots = TRUE,
                     variable = "Group", size=3, show_legend=TRUE,
                     clinical=TRUE, clinical_metadata = metadata_cz)

# see the results
p
```

##### Jaccard

**PCoA**

```{r,warning=FALSE}
p <- pca_plot_custom(filt_colon_asv_tab,
                     filt_colon_taxa_tab,
                     filt_colon_metadata,
                     measure = "jaccard",
                     show_boxplots = TRUE,
                     variable = "Group", size=3, show_legend=TRUE,
                      clinical=TRUE, clinical_metadata = metadata_cz)

# see the results
p
```

