---
title: "Mucosal microbiota alterations in primary sclerosis cholangitis persist after liver transplantation and are associated with clinical features independently of geography"
subtitle: "2. Hypothesis: PSC recurrence is associated with a specific composition of the gut microbiome"
format: 
  html:  
    embed-resources: true
    keep-md: false
    toc: true
    toc-expand: 3
    toc-depth: 6
    toc-location: left
    number-sections: true
    number-depth: 6
    code-fold: false
    code-tools: true
    code-summary: "Open code"
    grid: 
      body-width: 1000px
      sidebar-width: 600px
      margin-width: 100px
editor: source
project:
  type: default
  output-dir: output
fontsize: 12 px
---

-----------------------------------------------------------

# Introduction

## About

**rPSC vs non-rPSC vs Healthy analysis on merged data**

-   Alpha diversity –\> group effect, country effect, interaction effect

-   Beta diversity – PERMANOVA, PCA -\> group effect, country effect,
    interaction effect

-   DAA -\>

    -   Group effect – linDA + MaAsLin2 intersection

    -   Country effect – taxa with significant interaction effect were excluded based on individual post-hoc analysis. Taxa had to have the same direction in both countries 

-   ML models, trained and validated using out-of-sample bootstrap (500
    resamplings):

   - o ENET

   - o kNN

   - o GBoost

   - o RF

------------------------------------------------------

Pairwise comparisons:

**rPSC vs non-rPSC** –by this comparison, we will find the effect of
recurrence (whether the microbial composition is associated with it)

**non-rPSC vs Healthy** – by this comparison, we will find if the
non-rPSC samples are „closer“ to the healthy samples then rPSC samples
in terms of the microbial composition. In other words, does non-rPSC
samples have „healthy microbiome“?

**rPSC vs Healthy** – by this comparison, we will find how much are the
rPSC samples different from healthy samples.

However, the rPSC and non-rPSC samples are not different at all (see
results). Therefore, we can do EXCLUSIVE LEFT JOIN of differentially
abundant taxa from the analyses (rPSC vs Healthy, non-rPSC vs Healthy)

*Importing libraries and custom functions built for this analysis*

```{r, warning=FALSE}
source("custom_functions.R")
```

## Data Import

Importing ASV, taxa and metadata tables for both Czech and Norway
samples.

**Czech**

```{r, warning=FALSE}
path = "../../data/analysis_ready_data/ikem/"
asv_tab_ikem <- as.data.frame(fread(file.path(path,"asv_table_ikem.csv"),
                                    check.names = FALSE))
taxa_tab_ikem <- as.data.frame(fread(file.path(path,"taxa_table_ikem.csv"),
                                     check.names = FALSE))
metadata_ikem <- as.data.frame(fread(file.path(path,"metadata_ikem.csv"),
                                     check.names = FALSE))
```

**Norway**

```{r, warning=FALSE}
path = "../../data/analysis_ready_data/norway/"
asv_tab_norway <- as.data.frame(fread(file.path(path,"asv_table_norway.csv"),
                                    check.names = FALSE))
taxa_tab_norway <- as.data.frame(fread(file.path(path,"taxa_table_norway.csv"),
                                    check.names = FALSE))
metadata_norway <- as.data.frame(fread(file.path(path,"metadata_norway.csv"),
                                    check.names = FALSE))
```

## Merging data

Merging two countries to create whole dataset

```{r, warning=FALSE}
asv_tab <- merge(asv_tab_ikem,asv_tab_norway,by="SeqID",all=TRUE)
taxa_tab <- merging_taxa_tables(taxa_tab_ikem,taxa_tab_norway)
```

Merging two countries based on the different matrices - Ileum, Colon.

**Terminal ileum**

```{r, warning=FALSE}
ileum_data <- merging_data(asv_tab_1=asv_tab_ikem,
                           asv_tab_2=asv_tab_norway,
                           taxa_tab_1=taxa_tab_ikem,
                           taxa_tab_2=taxa_tab_norway,
                           metadata_1=metadata_ikem,
                           metadata_2=metadata_norway,
                           segment="TI",Q="Q2")

ileum_asv_tab <- ileum_data[[1]]
ileum_taxa_tab <- ileum_data[[2]]
ileum_metadata <- ileum_data[[3]]
```

**Right colon**

```{r, warning=FALSE}
rcolon_data <- merging_data(asv_tab_1=asv_tab_ikem,
                           asv_tab_2=asv_tab_norway,
                           taxa_tab_1=taxa_tab_ikem,
                           taxa_tab_2=taxa_tab_norway,
                           metadata_1=metadata_ikem,
                           metadata_2=metadata_norway,
                           segment="right_colon",Q="Q2")

rcolon_asv_tab <- rcolon_data[[1]]
rcolon_taxa_tab <- rcolon_data[[2]]
rcolon_metadata <- rcolon_data[[3]]
```

**Left colon**

```{r, warning=FALSE}
lcolon_data <- merging_data(asv_tab_1=asv_tab_ikem,
                           asv_tab_2=asv_tab_norway,
                           taxa_tab_1=taxa_tab_ikem,
                           taxa_tab_2=taxa_tab_norway,
                           metadata_1=metadata_ikem,
                           metadata_2=metadata_norway,
                           segment="left_colon",Q="Q2")

lcolon_asv_tab <- lcolon_data[[1]]
lcolon_taxa_tab <- lcolon_data[[2]]
lcolon_metadata <- lcolon_data[[3]]
```

# Data Analysis - Terminal ileum

```{r, warning=FALSE}
segment="terminal_ileum"
```

## Filtering

Rules: 

  - sequencing depth \> 10000 
  
  - nearZeroVar() with default settings 

**Rarefaction Curve**

```{r, warning=FALSE}
path="../intermediate_files/rarecurves"
seq_depth_threshold <- 10000
```

```{r, warning=FALSE,  eval = FALSE}
ps <- construct_phyloseq(ileum_asv_tab,ileum_taxa_tab,ileum_metadata)
rareres <- get_rarecurve(obj=ps, chunks=500)
save(rareres,file = file.path(path,"rarefaction_ileum.Rdata"))
```

```{r, warning=FALSE}
load(file.path(path,"rarefaction_ileum.Rdata"))

prare <- ggrarecurve(obj=rareres,
                      factorNames="Country",
                      indexNames=c("Observe")) + 
  theme_bw() +
  theme(axis.text=element_text(size=8), panel.grid=element_blank(),
        strip.background = element_rect(colour=NA,fill="grey"),
        strip.text.x = element_text(face="bold")) + 
  geom_vline(xintercept = seq_depth_threshold, 
             linetype="dashed", 
             color = "red") + 
  xlim(0, 20000)

prare
```

**Library size**

```{r, warning=FALSE,  fig.width=5, fig.height=4, fig.fullwidth=TRUE, warning=FALSE}
read_counts(ileum_asv_tab, line = c(5000,10000))
```

### Sequencing depth

```{r, warning=FALSE}
data_filt <- seq_depth_filtering(ileum_asv_tab,
                                 ileum_taxa_tab,
                                 ileum_metadata,
                                 seq_depth_threshold = 10000)

filt_ileum_asv_tab <- data_filt[[1]]; alpha_ileum_asv_tab <- filt_ileum_asv_tab
filt_ileum_taxa_tab <- data_filt[[2]]; alpha_ileum_taxa_tab <- filt_ileum_taxa_tab
filt_ileum_metadata <- data_filt[[3]]; alpha_ileum_metadata <- filt_ileum_metadata

seq_step <- dim(filt_ileum_asv_tab)[1]
```

**Library size**

```{r, warning=FALSE,  fig.width=5, fig.height=4, fig.fullwidth=TRUE, warning=FALSE}
read_counts(filt_ileum_asv_tab,line = c(5000,10000))
```

### NearZeroVar

```{r, warning=FALSE}
data_filt <- nearzerovar_filtering(filt_ileum_asv_tab, 
                                   filt_ileum_taxa_tab,
                                   filt_ileum_metadata)

filt_ileum_asv_tab <- data_filt[[1]]
filt_ileum_taxa_tab <- data_filt[[2]]
nearzero_step <- dim(filt_ileum_asv_tab)[1]
```

**Library size**

```{r, warning=FALSE,  fig.width=5, fig.height=4, fig.fullwidth=TRUE, warning=FALSE}
read_counts(filt_ileum_asv_tab,line = c(5000,10000))
```

### Final Counts

```{r, warning=FALSE}
final_counts_filtering(ileum_asv_tab,
                       filt_ileum_asv_tab,
                       filt_ileum_metadata,
                       seq_step, 0, nearzero_step) %>% `colnames<-`("Count")
```

## Alpha diversity

```{r, warning=FALSE}
path = "../results/Q2/alpha_diversity"
```

**Calculation**

```{r, warning=FALSE}
# Construct MPSE object
alpha_ileum_metadata$Sample <- alpha_ileum_metadata$SampleID
ileum_mpse <- as.MPSE(construct_phyloseq(alpha_ileum_asv_tab,
                                         alpha_ileum_taxa_tab,
                                         alpha_ileum_metadata))

ileum_mpse %<>% mp_rrarefy(raresize = 10000,seed = 123)

# Calculate alpha diversity - rarefied counts
ileum_mpse %<>% mp_cal_alpha(.abundance=RareAbundance, force=TRUE)
```

```{r, warning=FALSE}
alpha_div_plots <- list()

# preparing data frame
alpha_data <- data.frame(SampleID=ileum_mpse$Sample.x,
                         Observe=ileum_mpse$Observe,
                         Shannon=ileum_mpse$Shannon,
                         Simpson=ileum_mpse$Simpson,
                         Pielou=ileum_mpse$Pielou,
                         Group=ileum_mpse$Group,
                         Country=ileum_mpse$Country,
                         Patient=ileum_mpse$Patient)

write.csv(alpha_data,file.path(path,paste0("alpha_indices_",segment,".csv")),
          row.names = FALSE)
```

### Plots

```{r, warning=FALSE, fig.width=6, fig.height=4, fig.fullwidth=TRUE}

p_boxplot_alpha <- alpha_diversity_countries(alpha_data)

# save the results
alpha_div_plots[[paste(segment,"Country")]] <- p_boxplot_alpha

# see the results
p_boxplot_alpha
```

```{r, warning=FALSE, results='hide', echo=FALSE}
pdf("../figures/Q2/alpha_diversity_terminal_ileum.pdf",
    height =4,width = 7)
p_boxplot_alpha
dev.off()
```


### Linear models

```{r, warning=FALSE}
path = "../results/Q2/alpha_diversity"
alpha_data <- read.csv(file.path(path,paste0("alpha_indices_",segment,".csv")))
```

**Richness**

```{r, warning=FALSE}
results_model <- pairwise.lm(formula = "Observe ~ Group * Country",
                             factors=alpha_data$Group,
                             data=alpha_data)

# check interaction
if (!is.data.frame(results_model)){
  results_model_observe <- results_model[[1]]
  results_model_observe_emeans <- results_model[[2]]
} else {
  results_model_observe <- results_model
  results_model_observe_emeans <- NA
}

# save the results
pc_observed <- list(); 
pc_observed[[segment]] <- results_model_observe
```

```{r, warning=FALSE}
# see the results
knitr::kable(results_model_observe,digits = 3,
caption = "Raw results of linear model of richness estimation.")

knitr::kable(results_model_observe_emeans,digits = 3,
caption = "Raw results of independent country analysis")
```

**Shannon**

```{r, warning=FALSE}
results_model <- pairwise.lm(formula = "Shannon ~ Group * Country",
                             factors=alpha_data$Group,
                             data=alpha_data)

# check interaction
if (!is.data.frame(results_model)){
  results_model_shannon <- results_model[[1]]
  results_model_shannon_emeans <- results_model[[2]]
} else {
  results_model_shannon <- results_model
  results_model_shannon_emeans <- NA
}

# save the results
pc_shannon <- list(); 
pc_shannon[[segment]] <- as.data.frame(results_model_shannon)

```

```{r, warning=FALSE}
# see the results
knitr::kable(results_model_shannon,digits = 3,
caption = "Raw results of linear model of Shannon estimation.")

knitr::kable(results_model_shannon_emeans,digits = 3,
caption = "Raw results of independent country analysis")
```

**Simpson**

```{r, warning=FALSE}
results_model <- pairwise.lm(formula = "Simpson ~ Group * Country",
                                     factors=alpha_data$Group,
                                     data=alpha_data)

# check interaction
if (!is.data.frame(results_model)){
  results_model_simpson <- results_model[[1]]
  results_model_simpson_emeans <- results_model[[2]]
} else {
  results_model_simpson <- results_model
  results_model_simpson_emeans <- NA
}

# save the results
pc_simpson <- list(); 
pc_simpson[[segment]] <- as.data.frame(results_model_simpson)
```

```{r, warning=FALSE}
# see the results
knitr::kable(results_model_simpson,digits = 3,
caption = "Raw results of linear model of Simpson estimation.")

knitr::kable(results_model_simpson_emeans,digits = 3,
caption = "Raw results of independent country analysis")
```

**Pielou**

```{r, warning=FALSE}
results_model <- pairwise.lm(formula = "Pielou ~ Group * Country",
                                     factors=alpha_data$Group,
                                     data=alpha_data)

# check interaction

if (!is.data.frame(results_model)){
  results_model_pielou <- results_model[[1]]
  results_model_pielou_emeans <- results_model[[2]]
} else {
  results_model_pielou <- results_model
  results_model_pielou_emeans <- NA
}

# save the results
pc_pielou <- list(); 
pc_pielou[[segment]] <- as.data.frame(results_model_pielou)
```

```{r, warning=FALSE}
# see the results
knitr::kable(results_model_pielou,digits = 3,
caption = "Raw results of linear model of Pielou estimation.")

knitr::kable(results_model_pielou_emeans,digits = 3,
caption = "Raw results of independent country analysis")
```

### Saving results

```{r, warning=FALSE}
alpha_list <- list(
  Richness=pc_observed[[segment]] %>% rownames_to_column("Comparison"),
  Shannon=pc_shannon[[segment]] %>% rownames_to_column("Comparison"),
  Simpson=pc_simpson[[segment]] %>% rownames_to_column("Comparison"),
  Pielou=pc_pielou[[segment]] %>% rownames_to_column("Comparison"))
                   
write.xlsx(alpha_list, 
           file = file.path(path,paste0("alpha_diversity_results_",segment,".xlsx")))
```

## Beta diversity

Calculating Aitchison distance (euclidean distance on clr-transformed
data), both at ASV and genus level.

### Main analysis

**Genus level, Aitchison distance**

```{r, warning=FALSE}
level="genus"
path = "../results/Q2/beta_diversity"

pairwise_aitchison_raw <- list()
pca_plots_list <- list()
```

Aggregation, filtering

```{r, warning=FALSE}
# Aggregation
genus_data <- aggregate_taxa(ileum_asv_tab,
                             ileum_taxa_tab,
                             taxonomic_level=level,
                             names=TRUE)
# Filtration
filt_data <- filtering_steps(genus_data[[1]],
                             genus_data[[2]],
                             ileum_metadata,
                             seq_depth_threshold=10000)

filt_ileum_genus_tab <- filt_data[[1]]
filt_ileum_genus_taxa <- filt_data[[2]]
filt_ileum_metadata <- filt_data[[3]]
```

##### PERMANOVA

```{r, warning=FALSE}
pairwise_df <- filt_ileum_genus_tab %>% column_to_rownames("SeqID") %>% t()

# main effect
pp_main <- pairwise.adonis(pairwise_df,
                           filt_ileum_metadata$Group,
                           covariate = filt_ileum_metadata$Country, sim.method = "robust.aitchison", p.adjust.m="BH")

# interaction
pp_int <- pairwise.adonis(pairwise_df,filt_ileum_metadata$Group,covariate = filt_ileum_metadata$Country, interaction = TRUE, sim.method = "robust.aitchison", p.adjust.m="BH")

# tidy the results
pp_factor <- pp_main[[1]]
pp_cov <- pp_main[[2]]
pp_fac.cov <- pp_int[[3]]

cols <- c("pairs","Df","SumsOfSqs", "F.Model","R2","p.value", "p.adjusted", "sig")
colnames(pp_factor) <- cols; colnames(pp_cov) <- cols; colnames(pp_fac.cov) <- cols; 

# save raw results
pairwise_aitchison_raw[[paste(level, segment)]] <- rbind(pp_factor,pp_cov,pp_fac.cov)
```

```{r, warning=FALSE}
# see the results
knitr::kable(pp_factor,digits = 3,caption = "PERMANOVA, GROUP separation")
knitr::kable(pp_cov,digits = 3,caption = "PERMANOVA, COUNTRY separation")
knitr::kable(pp_fac.cov,digits = 3,caption = "PERMANOVA, INTERACTION GROUP:Country")
```

Interaction check

```{r, warning=FALSE}
interaction_sig <- pp_fac.cov$pairs[pp_fac.cov$p.adjusted < 0.05]

if (length(interaction_sig)>0){
  for (i in 1:length(interaction_sig)){
  group1 <- unlist(strsplit(interaction_sig[i],split = " vs "))[1]
  group2 <- unlist(strsplit(interaction_sig[i],split = " vs "))[2]
  group2 <- unlist(strsplit(group2,split = " : "))[1]
  
  result_list <- adonis_postanalysis(x=pairwise_df,
                      factors = filt_ileum_metadata$Group,
                      covariate = filt_ileum_metadata$Country, 
                      group1 = group1,
                      group2 = group2)
  print(result_list)
}
}

```

##### Plots

```{r, warning=FALSE}
p <- pca_plot_custom(filt_ileum_genus_tab,
                                 filt_ileum_genus_taxa,
                                 filt_ileum_metadata,
                                 show_boxplots = TRUE,
                                 variable = "Group", size=3, show_legend=FALSE)

# save the results
pca_plots_list[[paste(segment,level,"custom")]] <- p

# see the results
p
```

```{r, warning=FALSE, results='hide', echo=FALSE}
pdf("../figures/Q2/beta_diversity_terminal_ileum.pdf",
    height =5,width = 5)
p
dev.off()
```

#### Saving results

```{r, warning=FALSE}
write.xlsx(pairwise_aitchison_raw[[paste(level, segment)]], 
           file = file.path(path,
           paste0("beta_diversity_results_", segment,".xlsx")))
```

## Univariate Analysis

### Main analysis

```{r, warning=FALSE}
level="genus"

# needed paths
path = "../results/Q2/univariate_analysis"
path_maaslin=file.path("../intermediate_files/maaslin/Q2",level)

# variables
raw_linda_results_genus <- list();
raw_linda_results_genus[[segment]] <- list()
linda_results_genus <- list(); 
linda_results_genus[[segment]] <- list()

# country and interaction problems
list_country_union <- list()
list_intersections <- list()
list_venns <- list()
uni_statistics <- list()

# workbook for final df
wb <- createWorkbook()

# rPSC effect
rpsc_effect <- list()
```

Aggregate taxa

```{r, warning=FALSE}
genus_data <- aggregate_taxa(ileum_asv_tab,
                             ileum_taxa_tab,
                             taxonomic_level = level)

ileum_genus_tab <- genus_data[[1]]
ileum_genus_taxa_tab <- genus_data[[2]]

ileum_genus_asv_taxa_tab <- create_asv_taxa_table(ileum_genus_tab,
                                                  ileum_genus_taxa_tab)
```

#### rPSC vs non-rPSC

```{r, warning=FALSE}
group <- c("non-rPSC","rPSC")
comparison_name <- paste0(group[1], " vs ",group[2])
```

```{r, echo=FALSE}
comparison_title <- sub('"(.+?)" vs "(.+?)"', '"\\2" and "\\1"', comparison_name)
```

##### linDA

```{r, warning=FALSE,  fig.width=15, fig.height=5, fig.fullwidth=TRUE}

taxa_to_test <- read.xlsx("../results/Q1/univariate_analysis/significant_taxa_terminal_ileum.xlsx", sheet=" genus healthy vs pre_ltx")$SeqID

ileum_genus_tab_2 <- ileum_genus_tab %>% dplyr::filter(SeqID %in% taxa_to_test)
ileum_genus_taxa_tab_2 <- ileum_genus_taxa_tab %>% dplyr::filter(SeqID %in% taxa_to_test)

# prepare the data
linda_data <- binomial_prep(ileum_genus_tab_2,
                            ileum_genus_taxa_tab_2,
                            ileum_metadata,
                            group, usage="linDA", filtering = FALSE)

filt_ileum_uni_data <- linda_data[[1]]
filt_ileum_uni_taxa <- linda_data[[2]]
filt_ileum_uni_metadata <- linda_data[[3]]

# fit the model
linda.obj <- linda(filt_ileum_uni_data, 
                   filt_ileum_uni_metadata, 
                   formula = '~ Group * Country')

linda.output <- linda.obj$output
linda.output <- linda_renaming(linda.output, group)

# save the results
group1 <- paste0(group[1], " vs ","Group",group[2])
group2 <- paste0(group[1], " , ",group[2], " - CZ vs NO") 
group3 <- paste0(group[1], " vs ","Group",group[2], ":CountryNO")

for (grp in c(group1,group2,group3)){
  raw_linda_results_genus[[segment]][[grp]] <- 
    rawlinda.df(linda.output,
                grp,
                filt_ileum_uni_data,
                filt_ileum_uni_taxa)
  
  linda_results_genus[[segment]][[grp]] <- 
    linda.df(linda.output,
             grp,
             filt_ileum_uni_data,
             filt_ileum_uni_taxa)
}

```

```{r, warning=FALSE,  fig.width=15, fig.height=5, fig.fullwidth=TRUE}
# volcano plot
volcano_1 <- volcano_plot_linda(linda.output, 
                                group1, 
                                taxa_table = filt_ileum_uni_taxa) + 
            ggtitle(comparison_title)

volcano_2  <- volcano_plot_linda(linda.output, group2, 
                                taxa_table = filt_ileum_uni_taxa) + 
            ggtitle("NO vs CZ") 

volcano_3 <- volcano_plot_linda(linda.output, group3, 
                                taxa_table = filt_ileum_uni_taxa) +
            ggtitle("Interaction effect")

volcano <- ggarrange(volcano_1,volcano_2,volcano_3, ncol=3)
volcano
```

##### MaAsLin2

```{r, warning=FALSE,  echo=FALSE,results='hide',message=FALSE,warning=FALSE}
log_file <- file(tempfile(), open = "wt")
sink(log_file)  # Redirect standard output
sink(log_file, type = "message")  # Redirect messages and warnings

fit_data = Maaslin2(
    input_data = filt_ileum_uni_data, 
    input_metadata = filt_ileum_uni_metadata, min_abundance = 0,
    min_prevalence = 0,min_variance = 0,
    output = file.path(path_maaslin,group1), max_significance = 0.05,
    fixed_effects = c('Group', 'Country'),correction = "BH")

sink()  # Restore standard output
sink(type = "message")  # Restore messages
```

```{r, warning=FALSE}
volcano1 <- volcano_plot_maaslin(fit_data,filt_ileum_uni_taxa) + 
            ggtitle(comparison_title)

volcano2 <- volcano_plot_maaslin(fit_data,filt_ileum_uni_taxa,variable="Country") + 
            ggtitle("NO vs CZ")

volcano <- ggarrange(volcano1,volcano2, ncol=2)

volcano
```

##### Group - Intersection

```{r, warning=FALSE}
intersection_results <- group_intersection(group, list_intersections, list_venns,
                                           linda.output, fit_data,
                                           raw_linda_results_genus, 
                                           segment = segment,
                                           level=level)

list_intersections <- intersection_results[[1]]
list_venns <- intersection_results[[2]]
venn <- intersection_results[[3]]

# show the results
venn
```


##### Interaction effect

```{r, warning=FALSE}
list_interaction_significant <- country_interaction(group,
                                                    linda.output, 
                                                    list_intersections,
                                          filt_ileum_uni_data,
                                          filt_ileum_uni_metadata,
                                          segment=segment,
                                          level=level)

# see the result
## significant interaction effect
list_interaction_significant[[1]]

## results for czech cohort
list_interaction_significant[[2]]

## results for norwegian cohort
list_interaction_significant[[3]]
```

Removing problematic taxa

```{r, warning=FALSE}
list_intersections <- removing_interaction_problems(group,
                                                    list_interaction_significant,
                                                    list_intersections,
                                                    segment=segment,
                                                    level=level)
```

##### Basic statistics

```{r, warning=FALSE}
uni_df <-  merge(basic_univariate_statistics(linda_data,group),
                 raw_linda_results_genus[[segment]][[group1]],
                 by="SeqID",all=TRUE)

uni_df[["final_sig"]] <- uni_df$SeqID %in% list_intersections[[paste(segment,level,comparison_name)]][["SeqID"]]
uni_statistics[[segment]][[paste(level,comparison_name)]] <- uni_df

# for comparison
new_name <- comparison_name
addWorksheet(wb, sheetName = new_name)
writeData(wb, sheet = new_name, uni_df, rowNames=FALSE)
```

##### Prevalence testing

```{r}
uni_df$prevalence_diff <- abs(uni_df$`PREVALENCE_ rPSC` - uni_df$`PREVALENCE_ non-rPSC`)

#uni_df %<>% dplyr::filter(prevalence_diff >=0.2)
#filt_colon_uni_data <- filt_colon_uni_data[uni_df$SeqID,]
filt_ileum_uni_data <- ifelse(filt_ileum_uni_data>0,"YES","NO")
filt_ileum_uni_data %<>% t() %>% as.data.frame()
#filt_colon_uni_taxa %<>% dplyr::filter(SeqID %in% uni_df$SeqID)

filt_ileum_uni_data <- merge(filt_ileum_uni_data %>% rownames_to_column("SampleID"),filt_ileum_metadata,by="SampleID",all.x=TRUE) %>% column_to_rownames("SampleID")

p_values <- c()
for (taxon in uni_df$SeqID){
data <- filt_ileum_uni_data %>%
  group_by(!!sym(taxon), Group) %>%
  count() %>%
  group_by(!!sym(taxon), Group) %>%
  summarise(n = sum(n), .groups = "drop") %>%
  pivot_wider(names_from = Group, values_from = n) %>%
  column_to_rownames(taxon) %>%
  as.matrix()

if (nrow(data)>1){
  data[is.na(data)] <- 0
chi_res <- chisq.test(data)
chi_res$expected
#p_values <- c(p_values,chi_res$p.value)

test <- fisher.test(data)
test$p.value
p_values <- c(p_values,test$p.value)
} else p_values <- c(p_values,NA)

}

p_values_adjusted <- p.adjust(p_values,method = "BH")
names(p_values_adjusted) <- uni_df$SeqID
names(p_values) <- uni_df$SeqID
p_values_adjusted[which(p_values_adjusted<0.05)]
```

```{r}
asv_tab <- as.data.frame(rrarefy(
    linda_data[[1]] %>% t(), 
    sample=10000)) %>% t() %>% as.data.frame()

asv_tab <- ifelse(asv_tab>10,"YES","NO")
asv_tab %<>% t() %>% as.data.frame()

asv_tab <- merge(asv_tab %>% rownames_to_column("SampleID"),filt_ileum_metadata,by="SampleID",all.x=TRUE) %>% column_to_rownames("SampleID")

p_values <- c()
for (taxon in uni_df$SeqID){
data <- asv_tab %>%
  group_by(!!sym(taxon), Group) %>%
  count() %>%
  group_by(!!sym(taxon), Group) %>%
  summarise(n = sum(n), .groups = "drop") %>%
  pivot_wider(names_from = Group, values_from = n) %>%
  column_to_rownames(taxon) %>%
  as.matrix()

if (nrow(data)>1){
data[is.na(data)] <- 0
chi_res <- chisq.test(data)
chi_res$expected
test <- fisher.test(data)
test$p.value
chi_res$p.value
p_values <- c(p_values,test$p.value)
} else p_values <- c(p_values,NA)
}

p_values_adjusted <- p.adjust(p_values,method = "BH")
names(p_values_adjusted) <- uni_df$SeqID
names(p_values) <- uni_df$SeqID
p_values_adjusted[which(p_values_adjusted<0.05)]

df <- data.frame(SeqID=uni_df$SeqID,
                 p=p_values,
                 p_adj=p_values_adjusted)
```

```{r}
prevalence_df <- uni_df[,c("SeqID","PREVALENCE_ non-rPSC","PREVALENCE_ rPSC")]
prevalence_df <- merge(prevalence_df,df,by="SeqID",all=TRUE)
write.xlsx(prevalence_df,file.path(path,paste0("supplements_prevalence_",segment,".xlsx")))
```

#### rPSC vs healthy

```{r, warning=FALSE}
group <- c("healthy","rPSC")
comparison_name <- paste0(group[1], " vs ",group[2])
```

```{r, echo=FALSE}
comparison_title <- sub('"(.+?)" vs "(.+?)"', '"\\2" and "\\1"', comparison_name)
```

##### linDA

```{r, warning=FALSE,  fig.width=15, fig.height=5, fig.fullwidth=TRUE}

# prepare the data
linda_data <- binomial_prep(ileum_genus_tab,
                            ileum_genus_taxa_tab,
                            ileum_metadata,
                            group, usage="linDA")

filt_ileum_uni_data <- linda_data[[1]]
filt_ileum_uni_taxa <- linda_data[[2]]
filt_ileum_uni_metadata <- linda_data[[3]]

# fit the model
linda.obj <- linda(filt_ileum_uni_data, 
                   filt_ileum_uni_metadata, 
                   formula = '~ Group * Country')

linda.output <- linda.obj$output
linda.output <- linda_renaming(linda.output, group)

# save the results
group1 <- paste0(group[1], " vs ","Group",group[2])
group2 <- paste0(group[1], " , ",group[2], " - CZ vs NO") 
group3 <- paste0(group[1], " vs ","Group",group[2], ":CountryNO")

for (grp in c(group1,group2,group3)){
  raw_linda_results_genus[[segment]][[grp]] <- 
    rawlinda.df(linda.output,
                grp,
                filt_ileum_uni_data,
                filt_ileum_uni_taxa)
  
  linda_results_genus[[segment]][[grp]] <- 
    linda.df(linda.output,
             grp,
             filt_ileum_uni_data,
             filt_ileum_uni_taxa)
}

```

```{r, warning=FALSE,  fig.width=15, fig.height=5, fig.fullwidth=TRUE}
# volcano plot
volcano_1 <- volcano_plot_linda(linda.output, 
                                group1, 
                                taxa_table = filt_ileum_uni_taxa) + 
            ggtitle(comparison_title)

volcano_2  <- volcano_plot_linda(linda.output, group2, 
                                taxa_table = filt_ileum_uni_taxa) + 
            ggtitle("NO vs CZ") 

volcano_3 <- volcano_plot_linda(linda.output, group3, 
                                taxa_table = filt_ileum_uni_taxa) +
            ggtitle("Interaction effect")

volcano <- ggarrange(volcano_1,volcano_2,volcano_3, ncol=3)

```

##### MaAsLin2

```{r, warning=FALSE,  echo=FALSE,results='hide',message=FALSE,warning=FALSE}
log_file <- file(tempfile(), open = "wt")
sink(log_file)  # Redirect standard output
sink(log_file, type = "message")  # Redirect messages and warnings

fit_data = Maaslin2(
    input_data = filt_ileum_uni_data, 
    input_metadata = filt_ileum_uni_metadata, min_abundance = 0,
    min_prevalence = 0,min_variance = 0,
    output = file.path(path_maaslin,group1), max_significance = 0.05,
    fixed_effects = c('Group', 'Country'),correction = "BH")

sink()  # Restore standard output
sink(type = "message")  # Restore messages
```

```{r, warning=FALSE}
volcano1 <- volcano_plot_maaslin(fit_data,filt_ileum_uni_taxa) + 
            ggtitle(comparison_title)

volcano2 <- volcano_plot_maaslin(fit_data,filt_ileum_uni_taxa,variable="Country") + 
            ggtitle("NO vs CZ")

volcano <- ggarrange(volcano1,volcano2, ncol=2)
volcano
```

##### Group - Intersection

```{r, warning=FALSE}
intersection_results <- group_intersection(group, list_intersections, list_venns,
                                           linda.output, fit_data,
                                           raw_linda_results_genus, 
                                           segment = segment,
                                           level=level)

list_intersections <- intersection_results[[1]]
list_venns <- intersection_results[[2]]
venn <- intersection_results[[3]]

# show the results
venn
```

##### Interaction effect

```{r, warning=FALSE}
list_interaction_significant <- country_interaction(group,
                                                    linda.output, 
                                                    list_intersections,
                                          filt_ileum_uni_data,
                                          filt_ileum_uni_metadata,
                                          segment=segment,
                                          level=level)

# see the result
## significant interaction effect
list_interaction_significant[[1]]

## results for czech cohort
list_interaction_significant[[2]]

## results for norwegian cohort
list_interaction_significant[[3]]
```

Removing problematic taxa

```{r, warning=FALSE}
list_intersections <- removing_interaction_problems(group,
                                                    list_interaction_significant,
                                                    list_intersections,
                                                    segment=segment,
                                                    level=level)
```

##### Basic statistics

```{r, warning=FALSE}
uni_df <-  merge(basic_univariate_statistics(linda_data,group),
                 raw_linda_results_genus[[segment]][[group1]],
                 by="SeqID",all=TRUE)

uni_df[["final_sig"]] <- uni_df$SeqID %in% list_intersections[[paste(segment,level,comparison_name)]][["SeqID"]]
uni_statistics[[segment]][[paste(level,comparison_name)]] <- uni_df

# for comparison
new_name <- comparison_name
addWorksheet(wb, sheetName = new_name)
writeData(wb, sheet = new_name, uni_df, rowNames=FALSE)
```

#### non-rPSC vs healthy

```{r, warning=FALSE}
group <- c("healthy","non-rPSC")
comparison_name <- paste0(group[1], " vs ",group[2])
```

```{r, echo=FALSE}
comparison_title <- sub('"(.+?)" vs "(.+?)"', '"\\2" and "\\1"', comparison_name)
```

##### linDA

```{r, warning=FALSE,  fig.width=15, fig.height=5, fig.fullwidth=TRUE}

# prepare the data
linda_data <- binomial_prep(ileum_genus_tab,
                            ileum_genus_taxa_tab,
                            ileum_metadata,
                            group, usage="linDA")

filt_ileum_uni_data <- linda_data[[1]]
filt_ileum_uni_taxa <- linda_data[[2]]
filt_ileum_uni_metadata <- linda_data[[3]]

# fit the model
linda.obj <- linda(filt_ileum_uni_data, 
                   filt_ileum_uni_metadata, 
                   formula = '~ Group * Country')

linda.output <- linda.obj$output
linda.output <- linda_renaming(linda.output, group)

# save the results
group1 <- paste0(group[1], " vs ","Group",group[2])
group2 <- paste0(group[1], " , ",group[2], " - CZ vs NO") 
group3 <- paste0(group[1], " vs ","Group",group[2], ":CountryNO")

for (grp in c(group1,group2,group3)){
  raw_linda_results_genus[[segment]][[grp]] <- 
    rawlinda.df(linda.output,
                grp,
                filt_ileum_uni_data,
                filt_ileum_uni_taxa)
  
  linda_results_genus[[segment]][[grp]] <- 
    linda.df(linda.output,
             grp,
             filt_ileum_uni_data,
             filt_ileum_uni_taxa)
}

```

```{r, warning=FALSE,  fig.width=15, fig.height=5, fig.fullwidth=TRUE}
# volcano plot
volcano_1 <- volcano_plot_linda(linda.output, 
                                group1, 
                                taxa_table = filt_ileum_uni_taxa) + 
            ggtitle(comparison_title)

volcano_2  <- volcano_plot_linda(linda.output, group2, 
                                taxa_table = filt_ileum_uni_taxa) + 
            ggtitle("NO vs CZ") 

volcano_3 <- volcano_plot_linda(linda.output, group3, 
                                taxa_table = filt_ileum_uni_taxa) +
            ggtitle("Interaction effect")

volcano <- ggarrange(volcano_1,volcano_2,volcano_3, ncol=3)
volcano
```

##### MaAsLin2

```{r, warning=FALSE,  echo=FALSE,results='hide',message=FALSE,warning=FALSE}
log_file <- file(tempfile(), open = "wt")
sink(log_file)  # Redirect standard output
sink(log_file, type = "message")  # Redirect messages and warnings

fit_data = Maaslin2(
    input_data = filt_ileum_uni_data, 
    input_metadata = filt_ileum_uni_metadata, min_abundance = 0,
    min_prevalence = 0,min_variance = 0,
    output = file.path(path_maaslin,group1), max_significance = 0.05,
    fixed_effects = c('Group', 'Country'),correction = "BH")

sink()  # Restore standard output
sink(type = "message")  # Restore messages
```

```{r, warning=FALSE}
volcano1 <- volcano_plot_maaslin(fit_data,filt_ileum_uni_taxa) + 
            ggtitle(comparison_title)

volcano2 <- volcano_plot_maaslin(fit_data,filt_ileum_uni_taxa,variable="Country") + 
            ggtitle("NO vs CZ")

volcano <- ggarrange(volcano1,volcano2, ncol=2)
volcano
```

##### Group - Intersection

```{r, warning=FALSE}
intersection_results <- group_intersection(group, list_intersections, list_venns,
                                           linda.output, fit_data,
                                           raw_linda_results_genus, 
                                           segment = segment,
                                           level=level)

list_intersections <- intersection_results[[1]]
list_venns <- intersection_results[[2]]
venn <- intersection_results[[3]]

# show the results
venn
```


##### Interaction effect

```{r, warning=FALSE}
list_interaction_significant <- country_interaction(group,
                                                    linda.output, 
                                                    list_intersections,
                                          filt_ileum_uni_data,
                                          filt_ileum_uni_metadata,
                                          segment=segment,
                                          level=level)

# see the result
## significant interaction effect
list_interaction_significant[[1]]

## results for czech cohort
list_interaction_significant[[2]]

## results for norwegian cohort
list_interaction_significant[[3]]
```

Removing problematic taxa

```{r, warning=FALSE}
list_intersections <- removing_interaction_problems(group,
                                                    list_interaction_significant,
                                                    list_intersections,
                                                    segment=segment,
                                                    level=level)
```

##### Basic statistics

```{r, warning=FALSE}
uni_df <-  merge(basic_univariate_statistics(linda_data,group),
                 raw_linda_results_genus[[segment]][[group1]],
                 by="SeqID",all=TRUE)

uni_df[["final_sig"]] <- uni_df$SeqID %in% list_intersections[[paste(segment,level,comparison_name)]][["SeqID"]]
uni_statistics[[segment]][[paste(level,comparison_name)]] <- uni_df

# for comparison
new_name <- comparison_name
addWorksheet(wb, sheetName = new_name)
writeData(wb, sheet = new_name, uni_df, rowNames=FALSE)
```

#### Visualization

Heatmap visualizing the linDA's logFoldChange for taxa with p \< 0.1.

```{r, warning=FALSE,  fig.width=10, fig.height=17, fig.fullwidth=TRUE}
list_heatmap <- list_intersections[grep(paste(segment,level),
                                  names(list_intersections),value=TRUE)]

p_heatmap_linda <- heatmap_linda(list_heatmap,ileum_taxa_tab)
p_heatmap_linda
```

**Dot heatmap**

```{r, warning=FALSE,  fig.width=5, fig.height=10, fig.fullwidth=TRUE}
dotheatmap_linda <- dot_heatmap_linda(
  list_heatmap,                               uni_statistics$terminal_ileum[grepl(level,names(uni_statistics$terminal_ileum))],
                                      ileum_taxa_tab) + xlab("") + ylab("")
dotheatmap_linda
```

**Horizontal bar plot**

```{r, warning=FALSE, fig.height=10, fig.width=4}
p_prevalence <- horizontal_barplot(wb,taxa=levels(dotheatmap_linda$data$SeqID))

p_prevalence_final <- ggarrange(p_prevalence,
                                ggplot() + theme_minimal(),
                                nrow = 2,heights = c(1,0.09))

p <- ggarrange(dotheatmap_linda + theme(legend.position = "none"),p_prevalence_final,ncol=2,widths = c(1,0.3))

p

dot_heatmap_ileum <- p
```

```{r, warning=FALSE, results='hide', echo=FALSE}
pdf("../figures/Q2/dotplot_terminal_ileum.pdf",
    height =10,width = 4)
p
dev.off()
```


#### rPSC effect

**pre_LTx vs Healthy and Post_LTx vs Healthy intersection**


```{r, warning=FALSE}
A <- list_intersections[[paste(segment,level,"healthy vs rPSC")]]
B <- list_intersections[[paste(segment,level,"healthy vs non-rPSC")]]
df <- A[!(A$SeqID %in% B$SeqID),]


rpsc_effect[[paste(segment,level)]] <- df
  
# see the results
rpsc_effect[[paste(segment,level)]] 
```

#### Saving results

```{r, warning=FALSE}
# ALL DATA
saveWorkbook(wb,file.path(path,paste0("uni_analysis_wb_",segment,".xlsx")),
             overwrite = TRUE)

# PSC effect
write.xlsx(rpsc_effect[[paste(segment,level)]],file.path(path,paste0("rpsc_effect_",segment,".xlsx")))

# SIGNIFICANT taxa

write.xlsx(list_intersections[grepl(segment,names(list_intersections))] %>%
            `names<-`(gsub(segment, "", names(
              list_intersections[grepl(segment,names(list_intersections))]))),
           file.path(path,paste0("significant_taxa_",segment,".xlsx")))
```

## Machine learning

```{r, warning=FALSE}
path = "../results/Q2/models"
```

### ElasticNet

```{r, warning=FALSE}
model="enet"
```

#### Genus level

```{r, warning=FALSE}
level="genus"
```

Aggregate taxa

```{r, warning=FALSE}
genus_data <- aggregate_taxa(ileum_asv_tab,
                             ileum_taxa_tab,
                             taxonomic_level = level)

ileum_genus_tab <- genus_data[[1]]
ileum_genus_taxa_tab <- genus_data[[2]]
```

##### rPSC vs non-rPSC - all

```{r, warning=FALSE}
group <- c("rPSC","non-rPSC")
comparison_name <- paste0(group[1], " vs ",group[2])
```

```{r, warning=FALSE}
model_name <- paste(comparison_name,level,segment)

# prepare the data
filt_ileum_uni_data <- binomial_prep(ileum_genus_tab,
                                     ileum_genus_taxa_tab,
                                     ileum_metadata,
                                     group, 
                                     usage="ml_clr")

# fit the model
enet_model <- glmnet_binomial(filt_ileum_uni_data,
                              sample_method = "atypboot",
                              outcome="Group", 
                              N=500,
                              reuse=TRUE,
                              file=model_name,
                              Q="Q2")

# ROC
roc_c <- roc_curve(enet_model, group)

# save the results
models_summ <- list()
models_cm <- list()
roc_cs <- list()
betas <- list()

models_summ[[model_name]] <- enet_model$model_summary
models_cm[[model_name]] <- enet_model$conf_matrices$original
roc_cs[[model_name]] <- enet_model$kfold_rocobjs
betas[[model_name]] <- as.matrix(enet_model$betas)

# see the results
enet_model$model_summary %>% t()
enet_model$conf_matrices
enet_model$plot

roc_c
```

##### rPSC vs non-rPSC - subset

```{r, warning=FALSE}
group <- c("rPSC","non-rPSC")
comparison_name <- paste0(group[1], " vs ",group[2]," subset")
```

```{r, warning=FALSE}
taxa_to_test <- read.xlsx("../results/Q1/univariate_analysis/significant_taxa_terminal_ileum.xlsx", sheet=" genus healthy vs pre_ltx")

model_name <- paste(comparison_name,level,segment)

# prepare the data
filt_ileum_uni_data <- binomial_prep_psc_effect(ileum_genus_tab,
                                                ileum_genus_taxa_tab, 
                                                ileum_metadata,
                                                taxa_to_test,
                                                usage="ml_clr")

# fit the model
enet_model <- glmnet_binomial(filt_ileum_uni_data,
                              sample_method = "atypboot",
                              outcome="Group", 
                              N=500,
                              reuse=TRUE,
                              file=model_name,
                              Q="Q2")

# ROC
roc_c <- roc_curve(enet_model, group)

# save the results
models_summ[[model_name]] <- enet_model$model_summary
models_cm[[model_name]] <- enet_model$conf_matrices$original
roc_cs[[model_name]] <- enet_model$kfold_rocobjs
betas[[model_name]] <- as.matrix(enet_model$betas)

# see the results
enet_model$model_summary %>% t()
enet_model$conf_matrices
enet_model$plot

roc_c
```

#### Saving results

```{r, warning=FALSE}
models_summ_df_ileum <- do.call(rbind, 
  models_summ[grep(segment,names(models_summ),value = TRUE)])

write.csv(models_summ_df_ileum,file.path(path,paste0("elastic_net_",segment,".csv")))
```

## Results overview

#### Alpha diversity

**Models**

```{r,warning=FALSE}
knitr::kable(pc_observed[[segment]],
             digits = 3,
             caption = "Results of linear model testing ASV Richness")

knitr::kable(pc_shannon[[segment]],
             digits = 3,
             caption = "Results of linear model testing Shannon index")

knitr::kable(pc_simpson[[segment]],
             digits = 3,
             caption = "Results of linear model testing Simpson index")

knitr::kable(pc_pielou[[segment]],
             digits = 3,
             caption = "Results of linear model testing Pielou index")
```

**Plots**

```{r, warning=FALSE, fig.width=6, fig.height=4, fig.fullwidth=TRUE}
alpha_div_plots[[paste(segment,"Country")]]
```

#### Beta diversity

**Main results**

```{r,warning=FALSE}
knitr::kable(pairwise_aitchison_raw[[paste("genus", segment)]],
             digits = 3,
             caption = "Results of PERMANOVA - robust aitchison distance")
```

**PCA**

```{r, warning=FALSE,  fig.width=5, fig.height=3, fig.fullwidth=TRUE, eval=TRUE}
pca_plots_list[[paste(segment,"genus custom")]]
```


#### Univariate analysis

**Number of significant taxa**

```{r, warning=FALSE}
knitr::kable(cbind(as.data.frame(lapply(list_intersections,nrow)),
      as.data.frame(lapply(rpsc_effect,nrow))) %>% t() %>% 
  `colnames<-`("Count") %>% 
  `rownames<-`(c(names(list_intersections),"PSC effect Genus")),caption="Number of significant taxa")
```

#### Machine learning

**Main models**

```{r, warning=FALSE}
knitr::kable(models_summ_df_ileum %>% dplyr::select(
"alpha","lambda",
"auc_optimism_corrected",
"auc_optimism_corrected_CIL",
"auc_optimism_corrected_CIU"),
             digits=2,caption="Elastic net results")
```

**ROC - Genus level**

```{r, warning=FALSE,  warning=FALSE}
roc_curve_all_custom(roc_cs[c(1:2)],Q="Q2",
                     model_name="enet_model",legend = FALSE)
```


```{r, warning=FALSE, results='hide', echo=FALSE}
pdf("../figures/Q2/models_ileum.pdf",
    height =10,width = 10)
p
dev.off()
```

# Data Analysis - Right colon

```{r, warning=FALSE}
segment="right_colon"
```

## Filtering

Rules: 
  - sequencing depth \> 10000 
  
  - nearZeroVar() with default settings 
  
**Rarefaction Curve**

```{r, warning=FALSE}
path="../intermediate_files/rarecurves"
seq_depth_threshold <- 10000
```

```{r, warning=FALSE,  eval = FALSE}
ps <- construct_phyloseq(rcolon_asv_tab,rcolon_taxa_tab,rcolon_metadata)
rareres <- get_rarecurve(obj=ps, chunks=500)
save(rareres,file = file.path(path,"rarefaction_colon.Rdata"))
```

```{r, warning=FALSE}
load(file.path(path,"rarefaction_colon.Rdata"))
seq_depth_threshold <- 10000
prare <- ggrarecurve(obj=rareres,
                      factorNames="Country",
                      indexNames=c("Observe")) + 
        theme_bw()+
        theme(axis.text=element_text(size=8), 
              panel.grid=element_blank(),
              strip.background = element_rect(colour=NA,fill="grey"),
              strip.text.x = element_text(face="bold")) + 
        geom_vline(xintercept = seq_depth_threshold, 
                   linetype="dashed", color = "red") + 
        xlim(0, 20000)

prare
```

Library size

```{r, warning=FALSE,  fig.width=5, fig.height=4, fig.fullwidth=TRUE}
read_counts(rcolon_asv_tab, line = c(5000,10000))
```

### Sequencing depth

```{r, warning=FALSE}
data_filt <- seq_depth_filtering(rcolon_asv_tab,
                                 rcolon_taxa_tab,
                                 rcolon_metadata,
                                 seq_depth_threshold = 10000)

filt_rcolon_asv_tab <- data_filt[[1]]; alpha_rcolon_asv_tab <- filt_rcolon_asv_tab
filt_rcolon_taxa_tab <- data_filt[[2]]; alpha_rcolon_taxa_tab <- filt_rcolon_taxa_tab
filt_rcolon_metadata <- data_filt[[3]]; alpha_rcolon_metadata <- filt_rcolon_metadata

seq_step <- dim(filt_rcolon_asv_tab)[1]
```

Library size

```{r, warning=FALSE,  fig.width=5, fig.height=4, fig.fullwidth=TRUE}
read_counts(filt_rcolon_asv_tab,line = c(10000))
```

### NearZeroVar

```{r, warning=FALSE}
data_filt <- nearzerovar_filtering(filt_rcolon_asv_tab,
                                   filt_rcolon_taxa_tab,
                                   filt_rcolon_metadata)

filt_rcolon_asv_tab <- data_filt[[1]]
filt_rcolon_taxa_tab <- data_filt[[2]]
nearzero_step <- dim(filt_rcolon_asv_tab)[1]
```

Library size

```{r, warning=FALSE,  fig.width=5, fig.height=4, fig.fullwidth=TRUE}
read_counts(filt_rcolon_asv_tab,line = c(5000,10000))
```

Check zero depth

```{r, warning=FALSE}
data_filt <- check_zero_depth(filt_rcolon_asv_tab, 
                              filt_rcolon_taxa_tab, 
                              filt_rcolon_metadata)

filt_rcolon_asv_tab <- data_filt[[1]]; 
filt_rcolon_taxa_tab <- data_filt[[2]]; 
filt_rcolon_metadata <- data_filt[[3]]; 
```

Library size

```{r, warning=FALSE,  fig.width=5, fig.height=4, fig.fullwidth=TRUE}
read_counts(filt_rcolon_asv_tab,line = c(5000,10000))
```

### Final Counts

```{r, warning=FALSE}
final_counts_filtering(rcolon_asv_tab,
                       filt_rcolon_asv_tab,
                       filt_rcolon_metadata,
                       seq_step, 0, nearzero_step)
```


## Alpha diversity

```{r, warning=FALSE}
path = "../results/Q2/alpha_diversity"
```

**Calculation**

```{r, warning=FALSE}
# Construct MPSE object
alpha_rcolon_metadata$Sample <- alpha_rcolon_metadata$SampleID
rcolon_mpse <- as.MPSE(construct_phyloseq(alpha_rcolon_asv_tab,
                                         alpha_rcolon_taxa_tab,
                                         alpha_rcolon_metadata))

rcolon_mpse %<>% mp_rrarefy(raresize = 10000,seed = 123)

# Calculate alpha diversity - rarefied counts
rcolon_mpse %<>% mp_cal_alpha(.abundance=RareAbundance, force=TRUE)
```

```{r, warning=FALSE}
alpha_data <- data.frame(SampleID=rcolon_mpse$Sample.x,
                         Observe=rcolon_mpse$Observe,
                         Shannon=rcolon_mpse$Shannon,
                         Simpson=rcolon_mpse$Simpson,
                         Pielou=rcolon_mpse$Pielou,
                         Group=rcolon_mpse$Group,
                         Country=rcolon_mpse$Country,
                         Patient=rcolon_mpse$Patient)

write.csv(alpha_data,file.path(path,paste0("alpha_indices_",segment,".csv")),
          row.names = FALSE)
```

### Plots

```{r, warning=FALSE, fig.width=6, fig.height=4, fig.fullwidth=TRUE}

p_boxplot_alpha <- alpha_diversity_countries(alpha_data)

# save the results
alpha_div_plots[[paste(segment,"Country")]] <- p_boxplot_alpha

# see the results
p_boxplot_alpha
```

```{r, warning=FALSE, results='hide', echo=FALSE}
pdf("../figures/Q2/alpha_diversity_rcolon.pdf",
    height =4,width = 7)
p_boxplot_alpha
dev.off()
```

### Linear Models

```{r, warning=FALSE}
path = "../results/Q2/alpha_diversity"
alpha_data <- read.csv(file.path(path,paste0("alpha_indices_",segment,".csv")))
```

**Richness**

```{r, warning=FALSE}
results_model <- pairwise.lm(
  formula = "Observe ~ Group * Country",
  factors=alpha_data$Group,
  data=alpha_data)

# check interaction
if (!is.data.frame(results_model)){
  results_model_observe <- results_model[[1]]
  results_model_observe_detailed <- results_model[[2]]
} else {
  results_model_observe <- results_model
  results_model_observe_detailed <- NA
}

# save the results
pc_observed[[segment]] <- results_model_observe
```

```{r, warning=FALSE}
# see the results
knitr::kable(results_model_observe,digits = 3,
caption = "Raw results of linear model of richness estimation.")

knitr::kable(results_model_observe_detailed,digits = 3,
caption = "Raw results of independent country analysis")
```

**Shannon**

```{r, warning=FALSE}
results_model <- pairwise.lm(
  formula = "Shannon ~ Group * Country",
  factors=alpha_data$Group,
  data=alpha_data)

# check interaction
if (!is.data.frame(results_model)){
  results_model_shannon <- results_model[[1]]
  results_model_shannon_detailed <- results_model[[2]]
} else {
  results_model_shannon <- results_model
  results_model_shannon_detailed <- NA
}

# save the results
pc_shannon[[segment]] <- as.data.frame(results_model_shannon)

```

```{r, warning=FALSE}
# see the results
knitr::kable(results_model_shannon,digits = 3,
caption = "Raw results of linear model of Shannon estimation.")

knitr::kable(results_model_shannon_detailed,digits = 3,
caption = "Raw results of independent country analysis")
```

**Simpson**

```{r, warning=FALSE}
results_model <- pairwise.lm(
  formula = "Simpson ~ Group * Country",
  factors=alpha_data$Group,
  data=alpha_data)

# check interaction
if (!is.data.frame(results_model)){
  results_model_simpson <- results_model[[1]]
  results_model_simpson_detailed <- results_model[[2]]
} else {
  results_model_simpson <- results_model
  results_model_simpson_detailed <- NA
}

# save the results
pc_simpson[[segment]] <- as.data.frame(results_model_simpson)

```

```{r, warning=FALSE}
# see the results
knitr::kable(results_model_simpson,digits = 3,
caption = "Raw results of linear model of Simpson estimation.")

knitr::kable(results_model_simpson_detailed,digits = 3,
caption = "Raw results of independent country analysis")
```

**Pielou**

```{r, warning=FALSE}
results_model <- pairwise.lm(
  formula = "Pielou ~ Group * Country",
  factors=alpha_data$Group,
  data=alpha_data)

# check interaction
if (!is.data.frame(results_model)){
  results_model_pielou <- results_model[[1]]
  results_model_pielou_detailed <- results_model[[2]]
} else {
  results_model_pielou <- results_model
  results_model_pielou_detailed <- NA
}

# save the results
pc_pielou[[segment]] <- as.data.frame(results_model_pielou)

```

```{r, warning=FALSE}
# see the results
knitr::kable(results_model_pielou,digits = 3,
caption = "Raw results of linear model of Pielou estimation.")

knitr::kable(results_model_pielou_detailed,digits = 3,
caption = "Raw results of independent country analysis")
```

### Saving results

```{r, warning=FALSE}
alpha_list <- list(
  Richness=pc_observed[[segment]] %>% rownames_to_column("Comparison"),
  Shannon=pc_shannon[[segment]] %>% rownames_to_column("Comparison"),
  Simpson=pc_simpson[[segment]] %>% rownames_to_column("Comparison"),
  Pielou=pc_pielou[[segment]] %>% rownames_to_column("Comparison"))
                   
write.xlsx(alpha_list, 
           file = file.path(path,paste0("alpha_diversity_results_",segment,".xlsx")))
```

## Beta diversity

Calculating Aitchison distance (euclidean distance on clr-transformed
data), both at ASV and genus level.

### Main analysis

**Genus level, Aitchison distance**

```{r, warning=FALSE}
level="genus"

# needed path
path = "../results/Q2/beta_diversity"
```

Aggregation, filtering

```{r, warning=FALSE}
genus_data <- aggregate_taxa(rcolon_asv_tab,
                             rcolon_taxa_tab,
                             taxonomic_level=level,
                             names=TRUE)

filt_data <- filtering_steps(genus_data[[1]],
                             genus_data[[2]],
                             rcolon_metadata,
                            seq_depth_threshold=10000)

filt_rcolon_genus_tab <- filt_data[[1]]
filt_rcolon_genus_taxa <- filt_data[[2]]
filt_rcolon_genus_metadata <- filt_data[[3]]
```

##### PERMANOVA

```{r, warning=FALSE}
pairwise_df <- filt_rcolon_genus_tab %>% column_to_rownames("SeqID") %>% t()

# main effect
pp_main <- pairwise.adonis(pairwise_df,filt_rcolon_genus_metadata$Group,
                           covariate = filt_rcolon_genus_metadata$Country, 
                           sim.method = "robust.aitchison", p.adjust.m="BH")

# interaction
pp_int <- pairwise.adonis(pairwise_df,filt_rcolon_genus_metadata$Group,
                          covariate = filt_rcolon_genus_metadata$Country, 
                          interaction = TRUE, 
                          sim.method = "robust.aitchison", p.adjust.m="BH")

# tidy the results
pp_factor <- pp_main[[1]]
pp_cov <- pp_main[[2]]
pp_fac.cov <- pp_int[[3]]

cols <- c("pairs","Df","SumsOfSqs", "F.Model","R2","p.value", "p.adjusted", "sig")
colnames(pp_factor) <- cols; colnames(pp_cov) <- cols; colnames(pp_fac.cov) <- cols; 

# save raw results
pairwise_aitchison_raw[[paste(level, segment)]] <-rbind(pp_factor,pp_cov,pp_fac.cov)

```

```{r, warning=FALSE}
# see the results
knitr::kable(pp_factor,digits = 3,caption = "PERMANOVA, GROUP separation")
knitr::kable(pp_cov,digits = 3,caption = "PERMANOVA, COUNTRY separation")
knitr::kable(pp_fac.cov,digits = 3,caption = "PERMANOVA, INTERACTION GROUP:Country")
```

Interaction check

```{r, warning=FALSE}
interaction_sig <- pp_fac.cov$pairs[pp_fac.cov$p.adjusted < 0.05]

if (length(interaction_sig)>0){
 for (i in 1:length(interaction_sig)){
  group1 <- unlist(strsplit(interaction_sig[i],split = " vs "))[1]
  group2 <- unlist(strsplit(interaction_sig[i],split = " vs "))[2]
  group2 <- unlist(strsplit(group2,split = " : "))[1]
  
  result_list <- adonis_postanalysis(x=pairwise_df,
                      factors = filt_rcolon_genus_metadata$Group,
                      covariate = filt_rcolon_genus_metadata$Country, 
                      group1 = group1,
                      group2 = group2)
  print(result_list)
} 
}

```

##### Plots


```{r, warning=FALSE}
p <- pca_plot_custom(filt_rcolon_genus_tab,
                                 filt_rcolon_genus_taxa,
                                 filt_rcolon_genus_metadata,
                                 show_boxplots = TRUE,
                                 variable = "Group", size=2, 
                                 show_legend=FALSE)

# save the results
pca_plots_list[[paste(segment,level,"custom")]] <- p

# see the results
p

```

```{r, warning=FALSE, results='hide', echo=FALSE}
pdf("../figures/Q2/beta_diversity_rcolon.pdf",
    height =5,width = 5)
p
dev.off()
```

#### Saving results

```{r, warning=FALSE}
write.xlsx(pairwise_aitchison_raw[[paste(level, segment)]], 
           file = file.path(path,
           paste0("beta_diversity_results_", segment,".xlsx")))
```


## Univariate Analysis

### Main - Genus level

```{r, warning=FALSE}
level="genus"

# needed paths
path = "../results/Q2/univariate_analysis"
path_maaslin=file.path("../intermediate_files/maaslin/Q2",level)

# variables
raw_linda_results_genus[[segment]] <- list()
linda_results_genus[[segment]] <- list()

# country and interaction problems
list_country_union <- list()
list_venns <- list()

# workbook for final df
wb <- createWorkbook()

# PSC effect
rpsc_effect <- list()
```

#### Genus level

```{r, warning=FALSE}
level="genus"
```

Aggregate taxa

```{r, warning=FALSE}
genus_data <- aggregate_taxa(rcolon_asv_tab,
                             rcolon_taxa_tab,
                             taxonomic_level = level)

rcolon_genus_tab <- genus_data[[1]]
rcolon_genus_taxa_tab <- genus_data[[2]]

rcolon_genus_asv_taxa_tab <- create_asv_taxa_table(rcolon_genus_tab,
                                                  rcolon_genus_taxa_tab)
```

##### rPSC vs non-rPSC

###### linDA

```{r, warning=FALSE}
group <- c("non-rPSC","rPSC")
comparison_name <- paste0(group[1], " vs ",group[2])
```

```{r, echo=FALSE}
comparison_title <- sub('"(.+?)" vs "(.+?)"', '"\\2" and "\\1"', comparison_name)
```

```{r, warning=FALSE,  fig.width=15, fig.height=5, fig.fullwidth=TRUE}
taxa_to_test <- read.xlsx("../results/Q1/univariate_analysis/significant_taxa_right_colon.xlsx", sheet=" genus healthy vs pre_ltx")$SeqID

rcolon_genus_tab_2 <-  rcolon_genus_tab %>% dplyr::filter(SeqID %in% taxa_to_test)
rcolon_genus_taxa_tab_2 <-  rcolon_genus_taxa_tab %>% dplyr::filter(SeqID %in% taxa_to_test)

# prepare the data
linda_data <- binomial_prep(rcolon_genus_tab_2,
                            rcolon_genus_taxa_tab_2,
                            rcolon_metadata,
                            group, usage="linDA",filtering = FALSE)

filt_rcolon_uni_data <- linda_data[[1]]
filt_rcolon_uni_taxa <- linda_data[[2]]
filt_rcolon_uni_metadata <- linda_data[[3]]

# fit the model
linda.obj <- linda(filt_rcolon_uni_data,
                   filt_rcolon_uni_metadata,
                   formula = '~ Group * Country')

linda.output <- linda.obj$output
linda.output <- linda_renaming(linda.output, group)

# save the results
group1 <- paste0(group[1], " vs ","Group",group[2])
group2 <- paste0(group[1], " , ",group[2], " - CZ vs NO") 
group3 <- paste0(group[1], " vs ","Group",group[2], ":CountryNO")


for (grp in c(group1,group2,group3)){
  raw_linda_results_genus[[segment]][[grp]] <- 
    rawlinda.df(linda.output,
                grp,
                filt_rcolon_uni_data,
                filt_rcolon_uni_taxa)
  
  linda_results_genus[[segment]][[grp]] <- 
    linda.df(linda.output,
             grp,
             filt_rcolon_uni_data,
             filt_rcolon_uni_taxa)
}
```

```{r, warning=FALSE,  fig.width=15, fig.height=5, fig.fullwidth=TRUE}
# volcano plot
volcano_1 <- volcano_plot_linda(linda.output, group1, 
                                taxa_table = filt_rcolon_uni_taxa) + 
            ggtitle(paste(group,collapse=" vs "))

volcano_2  <- volcano_plot_linda(linda.output, group2, 
                                taxa_table = filt_rcolon_uni_taxa) + 
            ggtitle("NO vs CZ") 

volcano_3 <- volcano_plot_linda(linda.output, group3, 
                                taxa_table = filt_rcolon_uni_taxa) +
            ggtitle("Interaction effect")

volcano <- ggarrange(volcano_1,volcano_2,volcano_3, ncol=3)

# see the plot
volcano
```

###### MaAsLin2

```{r, warning=FALSE,  echo=FALSE,results='hide',message=FALSE,warning=FALSE}
log_file <- file(tempfile(), open = "wt")
sink(log_file)  # Redirect standard output
sink(log_file, type = "message")  # Redirect messages and warnings

fit_data = Maaslin2(
    input_data = filt_rcolon_uni_data, 
    input_metadata = filt_rcolon_uni_metadata, min_abundance = 0,
    min_prevalence = 0,min_variance = 0,
    output = file.path(path_maaslin,group1), max_significance = 0.05,
    fixed_effects = c('Group', 'Country'),
    correction = "BH")

sink()  # Restore standard output
sink(type = "message")  # Restore messages
```

```{r, warning=FALSE}
volcano1 <- volcano_plot_maaslin(fit_data,filt_rcolon_uni_taxa) + 
            ggtitle(paste(group[1], "vs", group[2]))

volcano2 <- volcano_plot_maaslin(fit_data,filt_rcolon_uni_taxa,variable="Country") + 
            ggtitle("NO vs CZ")

volcano <- ggarrange(volcano1,volcano2, ncol=2)

volcano
```

###### Group - Intersection

```{r, warning=FALSE}
intersection_results <- group_intersection(group, list_intersections, list_venns,
                                           linda.output, fit_data,
                                           raw_linda_results_genus, 
                                           segment = segment,
                                           level=level)

list_intersections <- intersection_results[[1]]
list_venns <- intersection_results[[2]]
venn <- intersection_results[[3]]

# show the results
venn
```

###### Interaction effect

```{r, warning=FALSE}
list_interaction_significant <- country_interaction(group,
                                                    linda.output, 
                                                    list_intersections,
                                          filt_rcolon_uni_data,
                                          filt_rcolon_uni_metadata,
                                          segment=segment,
                                          level=level)

# see the result
## significant interaction effect
list_interaction_significant[[1]]

## results for czech cohort
list_interaction_significant[[2]]

## results for norwegian cohort
list_interaction_significant[[3]]
```

Removing problematic taxa

```{r, warning=FALSE}
list_intersections <- removing_interaction_problems(group,
                                                    list_interaction_significant,
                                                    list_intersections,
                                                    segment=segment,
                                                    level=level)
```


##### Basic statistics

```{r, warning=FALSE}
uni_df <-  merge(basic_univariate_statistics(linda_data,group),
                 raw_linda_results_genus[[segment]][[group1]],
                 by="SeqID",all=TRUE)

uni_df[["final_sig"]] <- uni_df$SeqID %in% list_intersections[[paste(segment,level,comparison_name)]][["SeqID"]]
uni_statistics[[segment]][[paste(level,comparison_name)]] <- uni_df

# for comparison
new_name <- comparison_name
addWorksheet(wb, sheetName = new_name)
writeData(wb, sheet = new_name, uni_df, rowNames=FALSE)
```

##### Prevalence testing

```{r}
uni_df$prevalence_diff <- abs(uni_df$`PREVALENCE_ rPSC` - uni_df$`PREVALENCE_ non-rPSC`)

#uni_df %<>% dplyr::filter(prevalence_diff >=0.2)
#filt_rcolon_uni_data <- filt_rcolon_uni_data[uni_df$SeqID,]
filt_rcolon_uni_data <- ifelse(filt_rcolon_uni_data>0,"YES","NO")
filt_rcolon_uni_data %<>% t() %>% as.data.frame()
#filt_rcolon_uni_taxa %<>% dplyr::filter(SeqID %in% uni_df$SeqID)
filt_rcolon_uni_data <- merge(filt_rcolon_uni_data %>% rownames_to_column("SampleID"),filt_rcolon_metadata,by="SampleID",all.x=TRUE) %>% column_to_rownames("SampleID") %>% distinct(Patient, .keep_all = TRUE) 

p_values <- c()
for (taxon in uni_df$SeqID){
data <- filt_rcolon_uni_data %>%
  group_by(!!sym(taxon), Group) %>%
  count() %>%
  group_by(!!sym(taxon), Group) %>%
  summarise(n = sum(n), .groups = "drop") %>%
  pivot_wider(names_from = Group, values_from = n) %>%
  column_to_rownames(taxon) %>%
  as.matrix()

data
chi_res <- chisq.test(data)
chi_res$expected
#p_values <- c(p_values,chi_res$p.value)

test <- fisher.test(data)
test$p.value
p_values <- c(p_values,test$p.value)
}

p_values_adjusted <- p.adjust(p_values,method = "BH")
names(p_values_adjusted) <- uni_df$SeqID
names(p_values) <- uni_df$SeqID
p_values_adjusted[which(p_values_adjusted<0.05)]
```

```{r}
asv_tab <- as.data.frame(rrarefy(
    linda_data[[1]] %>% t(), 
    sample=10000)) %>% t() %>% as.data.frame()

asv_tab <- ifelse(asv_tab>10,"YES","NO")
asv_tab %<>% t() %>% as.data.frame()

set.seed(123)
asv_tab <- merge(asv_tab %>% rownames_to_column("SampleID"),filt_rcolon_metadata,by="SampleID",all.x=TRUE) %>% column_to_rownames("SampleID") %>% distinct(Patient, .keep_all = TRUE) 

p_values <- c()
for (taxon in uni_df$SeqID){
data <- asv_tab %>%
  group_by(!!sym(taxon), Group) %>%
  count() %>%
  group_by(!!sym(taxon), Group) %>%
  summarise(n = sum(n), .groups = "drop") %>%
  pivot_wider(names_from = Group, values_from = n) %>%
  column_to_rownames(taxon) %>%
  as.matrix()

data2 <- asv_tab %>%
  group_by(!!sym(taxon), Group,Country) %>%
  count() 


data[is.na(data)] <- 0
chi_res <- chisq.test(data)
chi_res$expected
test <- fisher.test(data)
test$p.value
chi_res$p.value
p_values <- c(p_values,test$p.value)
}

p_values_adjusted <- p.adjust(p_values,method = "BH")
names(p_values_adjusted) <- uni_df$SeqID
names(p_values) <- uni_df$SeqID
p_values_adjusted[which(p_values_adjusted<0.05)]

df <- data.frame(SeqID=uni_df$SeqID,
                 p=p_values,
                 p_adj=p_values_adjusted)
```

```{r}
prevalence_df <- uni_df[,c("SeqID","PREVALENCE_ non-rPSC","PREVALENCE_ rPSC")]
prevalence_df <- merge(prevalence_df,df,by="SeqID",all=TRUE)
write.xlsx(prevalence_df,file.path(path,paste0("supplements_prevalence_",segment,".xlsx")))
```

#### rPSC vs healthy

```{r, warning=FALSE}
group <- c("healthy","rPSC")
comparison_name <- paste0(group[1], " vs ",group[2])
```

```{r, echo=FALSE}
comparison_title <- sub('"(.+?)" vs "(.+?)"', '"\\2" and "\\1"', comparison_name)
```

##### linDA

```{r, warning=FALSE,  fig.width=15, fig.height=5, fig.fullwidth=TRUE}

# prepare the data
linda_data <- binomial_prep(rcolon_genus_tab,
                            rcolon_genus_taxa_tab,
                            rcolon_metadata,
                            group, usage="linDA")

filt_rcolon_uni_data <- linda_data[[1]]
filt_rcolon_uni_taxa <- linda_data[[2]]
filt_rcolon_uni_metadata <- linda_data[[3]]

# fit the model
linda.obj <- linda(filt_rcolon_uni_data,
                   filt_rcolon_uni_metadata,
                   formula = '~ Group * Country')

linda.output <- linda.obj$output
linda.output <- linda_renaming(linda.output, group)

# save the results
group1 <- paste0(group[1], " vs ","Group",group[2])
group2 <- paste0(group[1], " , ",group[2], " - CZ vs NO") 
group3 <- paste0(group[1], " vs ","Group",group[2], ":CountryNO")

for (grp in c(group1,group2,group3)){
  raw_linda_results_genus[[segment]][[grp]] <- 
    rawlinda.df(linda.output,
                grp,
                filt_rcolon_uni_data,
                filt_rcolon_uni_taxa)
  
  linda_results_genus[[segment]][[grp]] <- 
    linda.df(linda.output,
             grp,
             filt_rcolon_uni_data,
             filt_rcolon_uni_taxa)
}

```

```{r, warning=FALSE,  fig.width=15, fig.height=5, fig.fullwidth=TRUE}
# volcano plot
volcano_1 <- volcano_plot_linda(linda.output, group1, 
                                taxa_table = filt_rcolon_uni_taxa) + 
            ggtitle(paste(group,collapse=" vs "))

volcano_2  <- volcano_plot_linda(linda.output, group2, 
                                taxa_table = filt_rcolon_uni_taxa) + 
            ggtitle("NO vs CZ") 

volcano_3 <- volcano_plot_linda(linda.output, group3, 
                                taxa_table = filt_rcolon_uni_taxa) +
            ggtitle("Interaction effect")

volcano <- ggarrange(volcano_1,volcano_2,volcano_3, ncol=3)

# see the plot
volcano
```

###### MaAsLin2

```{r, warning=FALSE,  echo=FALSE,results='hide',message=FALSE,warning=FALSE}
log_file <- file(tempfile(), open = "wt")
sink(log_file)  # Redirect standard output
sink(log_file, type = "message")  # Redirect messages and warnings

fit_data = Maaslin2(
    input_data = filt_rcolon_uni_data, 
    input_metadata = filt_rcolon_uni_metadata, min_abundance = 0,
    min_prevalence = 0,min_variance = 0,
    output = file.path(path_maaslin,group1), max_significance = 0.05,
    fixed_effects = c('Group', 'Country'),
    correction = "BH")

sink()  # Restore standard output
sink(type = "message")  # Restore messages
```

```{r, warning=FALSE}
volcano1 <- volcano_plot_maaslin(fit_data,filt_rcolon_uni_taxa) + 
            ggtitle(paste(group[1], "vs", group[2]))

volcano2 <- volcano_plot_maaslin(fit_data,filt_rcolon_uni_taxa,variable="Country") + 
            ggtitle("NO vs CZ")

volcano <- ggarrange(volcano1,volcano2, ncol=2)
volcano
```

###### Group - Intersection

```{r, warning=FALSE}
intersection_results <- group_intersection(group, 
                                           list_intersections, 
                                           list_venns,
                                           linda.output, fit_data,
                                           raw_linda_results_genus, 
                                           segment = segment,
                                           level=level)

list_intersections <- intersection_results[[1]]
list_venns <- intersection_results[[2]]
venn <- intersection_results[[3]]

# show the results
venn
```

###### Interaction effect

```{r, warning=FALSE}
list_interaction_significant <- country_interaction(group,
                                                    linda.output, 
                                                    list_intersections,
                                          filt_rcolon_uni_data,
                                          filt_rcolon_uni_metadata,
                                          segment=segment,
                                          level=level)

# see the result
## significant interaction effect
list_interaction_significant[[1]]

## results for czech cohort
list_interaction_significant[[2]]

## results for norwegian cohort
list_interaction_significant[[3]]
```

Removing problematic taxa

```{r, warning=FALSE}
list_intersections <- removing_interaction_problems(group,
                                                    list_interaction_significant,
                                                    list_intersections,
                                                    segment=segment,
                                                    level=level)
```

##### Basic statistics

```{r, warning=FALSE}
uni_df <-  merge(basic_univariate_statistics(linda_data,group),
                 raw_linda_results_genus[[segment]][[group1]],
                 by="SeqID",all=TRUE)

uni_df[["final_sig"]] <- uni_df$SeqID %in% list_intersections[[paste(segment,level,comparison_name)]][["SeqID"]]
uni_statistics[[segment]][[paste(level,comparison_name)]] <- uni_df

# for comparison
new_name <- comparison_name
addWorksheet(wb, sheetName = new_name)
writeData(wb, sheet = new_name, uni_df, rowNames=FALSE)
```

##### non-rPSC vs healthy

```{r, warning=FALSE}
group <- c("healthy","non-rPSC")
comparison_name <- paste0(group[1], " vs ",group[2])
```

```{r, echo=FALSE}
comparison_title <- sub('"(.+?)" vs "(.+?)"', '"\\2" and "\\1"', comparison_name)
```

###### linDA

```{r, warning=FALSE,  fig.width=15, fig.height=5, fig.fullwidth=TRUE}

# prepare the data
linda_data <- binomial_prep(rcolon_genus_tab,
                            rcolon_genus_taxa_tab,
                            rcolon_metadata,group,
                            usage="linDA")

filt_rcolon_uni_data <- linda_data[[1]]
filt_rcolon_uni_taxa <- linda_data[[2]]
filt_rcolon_uni_metadata <- linda_data[[3]]

# fit the model
linda.obj <- linda(filt_rcolon_uni_data,
                   filt_rcolon_uni_metadata,
                   formula = '~ Group * Country')

linda.output <- linda.obj$output
linda.output <- linda_renaming(linda.output, group)

# save the results
group1 <- paste0(group[1], " vs ","Group",group[2])
group2 <- paste0(group[1], " , ",group[2], " - CZ vs NO") 
group3 <- paste0(group[1], " vs ","Group",group[2], ":CountryNO")

for (grp in c(group1,group2,group3)){
  raw_linda_results_genus[[segment]][[grp]] <- 
    rawlinda.df(linda.output,
                grp,
                filt_rcolon_uni_data,
                filt_rcolon_uni_taxa)
  
  linda_results_genus[[segment]][[grp]] <- 
    linda.df(linda.output,
             grp,
             filt_rcolon_uni_data,
             filt_rcolon_uni_taxa)
}

```

```{r, warning=FALSE,  fig.width=15, fig.height=5, fig.fullwidth=TRUE}
# volcano plot
volcano_1 <- volcano_plot_linda(linda.output, group1, 
                                taxa_table = filt_rcolon_uni_taxa) + 
            ggtitle(paste(group,collapse=" vs "))

volcano_2  <- volcano_plot_linda(linda.output, group2, 
                                taxa_table = filt_rcolon_uni_taxa) + 
            ggtitle("NO vs CZ") 

volcano_3 <- volcano_plot_linda(linda.output, group3, 
                                taxa_table = filt_rcolon_uni_taxa) +
            ggtitle("Interaction effect")

volcano <- ggarrange(volcano_1,volcano_2,volcano_3, ncol=3)

# see the plot
volcano
```

###### MaAsLin2

```{r, warning=FALSE,  echo=FALSE,results='hide',message=FALSE,warning=FALSE}
log_file <- file(tempfile(), open = "wt")
sink(log_file)  # Redirect standard output
sink(log_file, type = "message")  # Redirect messages and warnings

fit_data = Maaslin2(
    input_data = filt_rcolon_uni_data, 
    input_metadata = filt_rcolon_uni_metadata, min_abundance = 0,
    min_prevalence = 0,min_variance = 0,
    output = file.path(path_maaslin,group1), max_significance = 0.05,
    fixed_effects = c('Group', 'Country'),
    correction = "BH")

sink()  # Restore standard output
sink(type = "message")  # Restore messages
```

```{r, warning=FALSE}
volcano1 <- volcano_plot_maaslin(fit_data,filt_rcolon_uni_taxa) + 
            ggtitle(paste(group[1], "vs", group[2]))

volcano2 <- volcano_plot_maaslin(fit_data,filt_rcolon_uni_taxa,variable="Country") + 
            ggtitle("NO vs CZ")

volcano <- ggarrange(volcano1,volcano2, ncol=2)
volcano
```

###### Group - Intersection

```{r, warning=FALSE}
intersection_results <- group_intersection(group, list_intersections, list_venns,
                                           linda.output, fit_data,
                                           raw_linda_results_genus, 
                                           segment = segment,
                                           level=level)

list_intersections <- intersection_results[[1]]
list_venns <- intersection_results[[2]]
venn <- intersection_results[[3]]

# show the results
venn
```


###### Interaction effect

```{r, warning=FALSE}
list_interaction_significant <- country_interaction(group,
                                                    linda.output, 
                                                    list_intersections,
                                          filt_rcolon_uni_data,
                                          filt_rcolon_uni_metadata,
                                          segment=segment,
                                          level=level)

# see the result
## significant interaction effect
list_interaction_significant[[1]]

## results for czech cohort
list_interaction_significant[[2]]

## results for norwegian cohort
list_interaction_significant[[3]]
```

Removing problematic taxa

```{r, warning=FALSE}
list_intersections <- removing_interaction_problems(group,
                                                    list_interaction_significant,
                                                    list_intersections,
                                                    segment=segment,
                                                    level=level)
```

##### Basic statistics

```{r, warning=FALSE}
uni_df <-  merge(basic_univariate_statistics(linda_data,group),
                 raw_linda_results_genus[[segment]][[group1]],
                 by="SeqID",all=TRUE)

uni_df[["final_sig"]] <- uni_df$SeqID %in% list_intersections[[paste(segment,level,comparison_name)]][["SeqID"]]
uni_statistics[[segment]][[paste(level,comparison_name)]] <- uni_df

# for comparison
new_name <- comparison_name
addWorksheet(wb, sheetName = new_name)
writeData(wb, sheet = new_name, uni_df, rowNames=FALSE)
```

##### Visualization

Heatmap visualizing the linDA's logFoldChange for taxa with p \< 0.1.

```{r, warning=FALSE,  fig.width=10, fig.height=17, fig.fullwidth=TRUE}
list_heatmap <- list_intersections[grep(paste(segment,level),
                                  names(list_intersections),value=TRUE)]

p_heatmap_linda <- heatmap_linda(list_heatmap,rcolon_taxa_tab)
p_heatmap_linda
```

Dot heatmap

```{r, warning=FALSE,  fig.width=5, fig.height=17, fig.fullwidth=TRUE}
dotheatmap_linda <- dot_heatmap_linda(list_heatmap,
                                      uni_statistics$right_colon[grepl(level,names(uni_statistics$right_colon))],
                                      rcolon_taxa_tab) + xlab("") + ylab("")
dotheatmap_linda
```

**Horizontal bar plot**

```{r, warning=FALSE, fig.height=10, fig.width=4}
p_prevalence <- horizontal_barplot(wb,taxa=levels(dotheatmap_linda$data$SeqID))

p_prevalence_final <- ggarrange(p_prevalence,
                                ggplot() + theme_minimal(),
                                nrow = 2,heights = c(1,0.09))
p <- ggarrange(dotheatmap_linda + theme(legend.position = "none"),p_prevalence_final,ncol=2,widths = c(1,0.3))
p

dot_heatmap_rcolon <- p
```

```{r, warning=FALSE, results='hide', echo=FALSE}
pdf("../figures/Q2/dotplot_rcolon.pdf",
    height =10,width = 4)
p
dev.off()
```

<!-- ##### rPSC effect -->

<!-- **pre_LTx vs Healthy and Post_LTx vs Healthy intersection** -->


<!-- ```{r, warning=FALSE} -->
<!-- A <- list_intersections[[paste(segment,level,"healthy vs rPSC")]] -->
<!-- B <- list_intersections[[paste(segment,level,"healthy vs non-rPSC")]] -->
<!-- df <- A[!(A$SeqID %in% B$SeqID),] -->

<!-- rpsc_effect[[paste(segment,level)]] <- df -->

<!-- # see the results -->
<!-- rpsc_effect[[paste(segment,level)]]  -->
<!-- ``` -->

#### Saving results

```{r, warning=FALSE}
# ALL DATA
saveWorkbook(wb,file.path(path,paste0("uni_analysis_wb_",segment,".xlsx")),
             overwrite = TRUE)
```

## Machine learning

```{r, warning=FALSE}
path = "../results/Q2/models"
```

### ElasticNet

```{r, warning=FALSE}
model="enet"
```

#### Genus level

```{r, warning=FALSE}
level="genus"
```

Aggregate taxa

```{r, warning=FALSE}
genus_data <- aggregate_taxa(rcolon_asv_tab,
                             rcolon_taxa_tab,
                             taxonomic_level = level)

rcolon_genus_tab <- genus_data[[1]]
rcolon_genus_taxa_tab <- genus_data[[2]]
```

##### rPSC vs non-rPSC - all

```{r, warning=FALSE}
group <- c("rPSC","non-rPSC")
comparison_name <- paste0(group[1], " vs ",group[2])
```

```{r, warning=FALSE}
model_name <- paste(comparison_name,level,segment)

# prepare the data
filt_rcolon_uni_data <- binomial_prep(rcolon_genus_tab,
                                     rcolon_genus_taxa_tab,
                                     rcolon_metadata,
                                     group, 
                                     usage="ml_clr")

# fit the model
enet_model <- glmnet_binomial(filt_rcolon_uni_data,
                              sample_method = "atypboot",
                              outcome="Group", 
                              N=10,
                              reuse=TRUE,
                              file=model_name,
                              Q="Q2")

# ROC
roc_c <- roc_curve(enet_model, group)

# save the results
models_summ[[model_name]] <- enet_model$model_summary
models_cm[[model_name]] <- enet_model$conf_matrices$original
roc_cs[[model_name]] <- enet_model$kfold_rocobjs
betas[[model_name]] <- as.matrix(enet_model$betas)

# see the results
enet_model$model_summary %>% t()
enet_model$conf_matrices
enet_model$plot

roc_c
```

##### rPSC vs non-rPSC - subset

```{r, warning=FALSE}
group <- c("rPSC","non-rPSC")
comparison_name <- paste0(group[1], " vs ",group[2]," subset")
```

```{r, warning=FALSE}
taxa_to_test <- read.xlsx("../results/Q1/univariate_analysis/significant_taxa_right_colon.xlsx", sheet=" genus healthy vs pre_ltx")

model_name <- paste(comparison_name,level,segment)

# prepare the data
filt_rcolon_uni_data <- binomial_prep_psc_effect(rcolon_genus_tab,
                                                rcolon_genus_taxa_tab, 
                                                rcolon_metadata,
                                                taxa_to_test,
                                                usage="ml_clr")

# fit the model
enet_model <- glmnet_binomial(filt_rcolon_uni_data,
                              sample_method = "atypboot",
                              outcome="Group", 
                              N=10,
                              reuse=TRUE,
                              file=model_name,
                              Q="Q2")

# ROC
roc_c <- roc_curve(enet_model, group)

# save the results
models_summ[[model_name]] <- enet_model$model_summary
models_cm[[model_name]] <- enet_model$conf_matrices$original
roc_cs[[model_name]] <- enet_model$kfold_rocobjs
betas[[model_name]] <- as.matrix(enet_model$betas)

# see the results
enet_model$model_summary %>% t()
enet_model$conf_matrices
enet_model$plot

roc_c
```

#### Saving results

```{r, warning=FALSE}
models_summ_df_colon <- do.call(rbind, 
  models_summ[grep(segment,names(models_summ),value = TRUE)])

write.csv(models_summ_df_colon,file.path(path,paste0("elastic_net_",segment,".csv")))
```

## Results overview

#### Alpha diversity

**Linear models**

```{r,warning=FALSE}
knitr::kable(pc_observed[[segment]],
             digits = 3,
             caption = "Results of linear model testing ASV Richness")

knitr::kable(pc_shannon[[segment]],
             digits = 3,
             caption = "Results of linear model testing Shannon index")

knitr::kable(pc_simpson[[segment]],
             digits = 3,
             caption = "Results of linear model testing Simpson index")

knitr::kable(pc_pielou[[segment]],
             digits = 3,
             caption = "Results of linear model testing Pielou index")
```

**Plots**

```{r, warning=FALSE, fig.width=6, fig.height=4, fig.fullwidth=TRUE}
alpha_div_plots[[paste(segment,"Country")]]
```

#### Beta diversity

**Main results**

```{r,warning=FALSE}
knitr::kable(pairwise_aitchison_raw[[paste("genus", segment)]],
             digits = 3,
             caption = "Results of PERMANOVA - robust aitchison distance")
```

**PCA**

```{r, warning=FALSE,  fig.width=5, fig.height=3, fig.fullwidth=TRUE, eval=TRUE}
pca_plots_list[[paste(segment,"genus custom")]]
```


#### Univariate analysis

**Number of significant taxa**

```{r, warning=FALSE}
knitr::kable(as.data.frame(lapply(list_intersections,nrow)) %>% t() %>% 
  `colnames<-`("Count") )
```


#### Machine learning

**Main models**


```{r, warning=FALSE}
knitr::kable(models_summ_df_colon %>% dplyr::select(
"alpha","lambda",
"auc_optimism_corrected",
"auc_optimism_corrected_CIL",
"auc_optimism_corrected_CIU"),
             digits=3,caption="Elastic net results")
```

**ROC - Genus level**

```{r, warning=FALSE,  warning=FALSE}
roc_curve_all_custom(roc_cs[3:4],Q="Q2",
                     model_name="enet_model",legend=FALSE)
```


# Data Analysis - Left colon

```{r, warning=FALSE}
segment="left_colon"
```

## Filtering

Rules: 
  - sequencing depth \> 10000 
  
  - nearZeroVar() with default settings 
  
**Rarefaction Curve**

```{r, warning=FALSE}
path="../intermediate_files/rarecurves"
seq_depth_threshold <- 10000
```

```{r, warning=FALSE,  eval = FALSE}
ps <- construct_phyloseq(lcolon_asv_tab,lcolon_taxa_tab,lcolon_metadata)
rareres <- get_rarecurve(obj=ps, chunks=500)
save(rareres,file = file.path(path,"rarefaction_colon.Rdata"))
```

```{r, warning=FALSE}
load(file.path(path,"rarefaction_colon.Rdata"))
seq_depth_threshold <- 10000
prare <- ggrarecurve(obj=rareres,
                      factorNames="Country",
                      indexNames=c("Observe")) + 
        theme_bw()+
        theme(axis.text=element_text(size=8), 
              panel.grid=element_blank(),
              strip.background = element_rect(colour=NA,fill="grey"),
              strip.text.x = element_text(face="bold")) + 
        geom_vline(xintercept = seq_depth_threshold, 
                   linetype="dashed", color = "red") + 
        xlim(0, 20000)

prare
```

Library size

```{r, warning=FALSE,  fig.width=5, fig.height=4, fig.fullwidth=TRUE}
read_counts(lcolon_asv_tab, line = c(5000,10000))
```

### Sequencing depth

```{r, warning=FALSE}
data_filt <- seq_depth_filtering(lcolon_asv_tab,
                                 lcolon_taxa_tab,
                                 lcolon_metadata,
                                 seq_depth_threshold = 10000)

filt_lcolon_asv_tab <- data_filt[[1]]; alpha_lcolon_asv_tab <- filt_lcolon_asv_tab
filt_lcolon_taxa_tab <- data_filt[[2]]; alpha_lcolon_taxa_tab <- filt_lcolon_taxa_tab
filt_lcolon_metadata <- data_filt[[3]]; alpha_lcolon_metadata <- filt_lcolon_metadata

seq_step <- dim(filt_lcolon_asv_tab)[1]
```

Library size

```{r, warning=FALSE,  fig.width=5, fig.height=4, fig.fullwidth=TRUE}
read_counts(filt_lcolon_asv_tab,line = c(10000))
```

### NearZeroVar

```{r, warning=FALSE}
data_filt <- nearzerovar_filtering(filt_lcolon_asv_tab,
                                   filt_lcolon_taxa_tab,
                                   filt_lcolon_metadata)

filt_lcolon_asv_tab <- data_filt[[1]]
filt_lcolon_taxa_tab <- data_filt[[2]]
nearzero_step <- dim(filt_lcolon_asv_tab)[1]
```

Library size

```{r, warning=FALSE,  fig.width=5, fig.height=4, fig.fullwidth=TRUE}
read_counts(filt_lcolon_asv_tab,line = c(5000,10000))
```

Check zero depth

```{r, warning=FALSE}
data_filt <- check_zero_depth(filt_lcolon_asv_tab, 
                              filt_lcolon_taxa_tab, 
                              filt_lcolon_metadata)

filt_lcolon_asv_tab <- data_filt[[1]]; 
filt_lcolon_taxa_tab <- data_filt[[2]]; 
filt_lcolon_metadata <- data_filt[[3]]; 
```

Library size

```{r, warning=FALSE,  fig.width=5, fig.height=4, fig.fullwidth=TRUE}
read_counts(filt_lcolon_asv_tab,line = c(5000,10000))
```

### Final Counts

```{r, warning=FALSE}
final_counts_filtering(lcolon_asv_tab,
                       filt_lcolon_asv_tab,
                       filt_lcolon_metadata,
                       seq_step, 0, nearzero_step)
```


## Alpha diversity

```{r, warning=FALSE}
path = "../results/Q2/alpha_diversity"
```

**Calculation**

```{r, warning=FALSE}
# Construct MPSE object
alpha_lcolon_metadata$Sample <- alpha_lcolon_metadata$SampleID
lcolon_mpse <- as.MPSE(construct_phyloseq(alpha_lcolon_asv_tab,
                                         alpha_lcolon_taxa_tab,
                                         alpha_lcolon_metadata))

lcolon_mpse %<>% mp_rrarefy(raresize = 10000,seed = 123)

# Calculate alpha diversity - rarefied counts
lcolon_mpse %<>% mp_cal_alpha(.abundance=RareAbundance, force=TRUE)
```

```{r, warning=FALSE}
alpha_data <- data.frame(SampleID=lcolon_mpse$Sample.x,
                         Observe=lcolon_mpse$Observe,
                         Shannon=lcolon_mpse$Shannon,
                         Simpson=lcolon_mpse$Simpson,
                         Pielou=lcolon_mpse$Pielou,
                         Group=lcolon_mpse$Group,
                         Country=lcolon_mpse$Country,
                         Patient=lcolon_mpse$Patient)

write.csv(alpha_data,file.path(path,paste0("alpha_indices_",segment,".csv")),
          row.names = FALSE)
```

### Plots

```{r, warning=FALSE, fig.width=6, fig.height=4, fig.fullwidth=TRUE}

p_boxplot_alpha <- alpha_diversity_countries(alpha_data)

# save the results
alpha_div_plots[[paste(segment,"Country")]] <- p_boxplot_alpha

# see the results
p_boxplot_alpha
```

```{r, warning=FALSE, results='hide', echo=FALSE}
pdf("../figures/Q2/alpha_diversity_lcolon.pdf",
    height =4,width = 7)
p_boxplot_alpha
dev.off()
```

### Linear Models

```{r, warning=FALSE}
path = "../results/Q2/alpha_diversity"
alpha_data <- read.csv(file.path(path,paste0("alpha_indices_",segment,".csv")))
```

**Richness**

```{r, warning=FALSE}
results_model <- pairwise.lm(
  formula = "Observe ~ Group * Country",
  factors=alpha_data$Group,
  data=alpha_data,
  bootstraps=TRUE,N=100)

# check interaction
if (!is.data.frame(results_model)){
  results_model_observe <- results_model[[1]]
  results_model_observe_detailed <- results_model[[2]]
} else {
  results_model_observe <- results_model
  results_model_observe_detailed <- NA
}

# save the results
pc_observed[[segment]] <- results_model_observe
```

```{r, warning=FALSE}
# see the results
knitr::kable(results_model_observe,digits = 3,
caption = "Raw results of linear model of richness estimation.")

knitr::kable(results_model_observe_detailed,digits = 3,
caption = "Raw results of independent country analysis")
```

**Shannon**

```{r, warning=FALSE}
results_model <- pairwise.lm(
  formula = "Shannon ~ Group * Country",
  factors=alpha_data$Group,
  data=alpha_data,
  bootstraps=TRUE,N=100)

# check interaction
if (!is.data.frame(results_model)){
  results_model_shannon <- results_model[[1]]
  results_model_shannon_detailed <- results_model[[2]]
} else {
  results_model_shannon <- results_model
  results_model_shannon_detailed <- NA
}

# save the results
pc_shannon[[segment]] <- as.data.frame(results_model_shannon)

```

```{r, warning=FALSE}
# see the results
knitr::kable(results_model_shannon,digits = 3,
caption = "Raw results of linear model of Shannon estimation.")

knitr::kable(results_model_shannon_detailed,digits = 3,
caption = "Raw results of independent country analysis")
```

**Simpson**

```{r, warning=FALSE}
results_model <- pairwise.lm(
  formula = "Simpson ~ Group * Country",
  factors=alpha_data$Group,
  data=alpha_data,
  bootstraps=TRUE,N=100)

# check interaction
if (!is.data.frame(results_model)){
  results_model_simpson <- results_model[[1]]
  results_model_simpson_detailed <- results_model[[2]]
} else {
  results_model_simpson <- results_model
  results_model_simpson_detailed <- NA
}

# save the results
pc_simpson[[segment]] <- as.data.frame(results_model_simpson)

```

```{r, warning=FALSE}
# see the results
knitr::kable(results_model_simpson,digits = 3,
caption = "Raw results of linear model of Simpson estimation.")

knitr::kable(results_model_simpson_detailed,digits = 3,
caption = "Raw results of independent country analysis")
```

**Pielou**

```{r, warning=FALSE}
results_model <- pairwise.lm(
  formula = "Pielou ~ Group * Country",
  factors=alpha_data$Group,
  data=alpha_data,
  bootstraps=TRUE,N=100)

# check interaction
if (!is.data.frame(results_model)){
  results_model_pielou <- results_model[[1]]
  results_model_pielou_detailed <- results_model[[2]]
} else {
  results_model_pielou <- results_model
  results_model_pielou_detailed <- NA
}

# save the results
pc_pielou[[segment]] <- as.data.frame(results_model_pielou)

```

```{r, warning=FALSE}
# see the results
knitr::kable(results_model_pielou,digits = 3,
caption = "Raw results of linear model of Pielou estimation.")

knitr::kable(results_model_pielou_detailed,digits = 3,
caption = "Raw results of independent country analysis")
```

### Saving results

```{r, warning=FALSE}
alpha_list <- list(
  Richness=pc_observed[[segment]] %>% rownames_to_column("Comparison"),
  Shannon=pc_shannon[[segment]] %>% rownames_to_column("Comparison"),
  Simpson=pc_simpson[[segment]] %>% rownames_to_column("Comparison"),
  Pielou=pc_pielou[[segment]] %>% rownames_to_column("Comparison"))
                   
write.xlsx(alpha_list, 
           file = file.path(path,paste0("alpha_diversity_results_",segment,".xlsx")))
```

## Beta diversity

Calculating Aitchison distance (euclidean distance on clr-transformed
data), both at ASV and genus level.

### Main analysis

**Genus level, Aitchison distance**

```{r, warning=FALSE}
level="genus"

# needed path
path = "../results/Q2/beta_diversity"
```

Aggregation, filtering

```{r, warning=FALSE}
genus_data <- aggregate_taxa(lcolon_asv_tab,
                             lcolon_taxa_tab,
                             taxonomic_level=level,
                             names=TRUE)

filt_data <- filtering_steps(genus_data[[1]],
                             genus_data[[2]],
                             lcolon_metadata,
                            seq_depth_threshold=10000)

filt_lcolon_genus_tab <- filt_data[[1]]
filt_lcolon_genus_taxa <- filt_data[[2]]
filt_lcolon_genus_metadata <- filt_data[[3]]
```

##### PERMANOVA

```{r, warning=FALSE}
pairwise_df <- filt_lcolon_genus_tab %>% column_to_rownames("SeqID") %>% t()

# main effect
pp_main <- pairwise.adonis(pairwise_df,filt_lcolon_genus_metadata$Group,
                           covariate = filt_lcolon_genus_metadata$Country, 
                           sim.method = "robust.aitchison", p.adjust.m="BH",
                           patients = filt_lcolon_genus_metadata$Patient)

# interaction
pp_int <- pairwise.adonis(pairwise_df,filt_lcolon_genus_metadata$Group,
                          covariate = filt_lcolon_genus_metadata$Country, 
                          interaction = TRUE, 
                          sim.method = "robust.aitchison", p.adjust.m="BH",
                           patients = filt_lcolon_genus_metadata$Patient)

# tidy the results
pp_factor <- pp_main[[1]]
pp_cov <- pp_main[[2]]
pp_fac.cov <- pp_int[[3]]

cols <- c("pairs","Df","SumsOfSqs", "F.Model","R2","p.value", "p.adjusted", "sig")
colnames(pp_factor) <- cols; colnames(pp_cov) <- cols; colnames(pp_fac.cov) <- cols; 

# save raw results
pairwise_aitchison_raw[[paste(level, segment)]] <-rbind(pp_factor,pp_cov,pp_fac.cov)

```

```{r, warning=FALSE}
# see the results
knitr::kable(pp_factor,digits = 3,caption = "PERMANOVA, GROUP separation")
knitr::kable(pp_cov,digits = 3,caption = "PERMANOVA, COUNTRY separation")
knitr::kable(pp_fac.cov,digits = 3,caption = "PERMANOVA, INTERACTION GROUP:Country")
```

Interaction check

```{r, warning=FALSE}
interaction_sig <- pp_fac.cov$pairs[pp_fac.cov$p.adjusted < 0.05]

if (length(interaction_sig)>0){
 for (i in 1:length(interaction_sig)){
  group1 <- unlist(strsplit(interaction_sig[i],split = " vs "))[1]
  group2 <- unlist(strsplit(interaction_sig[i],split = " vs "))[2]
  group2 <- unlist(strsplit(group2,split = " : "))[1]
  
  result_list <- adonis_postanalysis(x=pairwise_df,
                      factors = filt_lcolon_genus_metadata$Group,
                      covariate = filt_lcolon_genus_metadata$Country, 
                      group1 = group1,
                      group2 = group2)
  print(result_list)
} 
}

```

##### Plots


```{r, warning=FALSE}
p <- pca_plot_custom(filt_lcolon_genus_tab,
                                 filt_lcolon_genus_taxa,
                                 filt_lcolon_genus_metadata,
                                 show_boxplots = TRUE,
                                 variable = "Group", size=2, 
                                 show_legend=FALSE)

# save the results
pca_plots_list[[paste(segment,level,"custom")]] <- p

# see the results
p

```

```{r, warning=FALSE, results='hide', echo=FALSE}
pdf("../figures/Q2/beta_diversity_lcolon.pdf",
    height =5,width = 5)
p
dev.off()
```

#### Saving results

```{r, warning=FALSE}
write.xlsx(pairwise_aitchison_raw[[paste(level, segment)]], 
           file = file.path(path,
           paste0("beta_diversity_results_", segment,".xlsx")))
```

## Univariate Analysis

### Main - Genus level

```{r, warning=FALSE}
level="genus"

# needed paths
path = "../results/Q2/univariate_analysis"
path_maaslin=file.path("../intermediate_files/maaslin/Q2",level)

# variables
raw_linda_results_genus[[segment]] <- list()
linda_results_genus[[segment]] <- list()

# country and interaction problems
list_country_union <- list()
list_venns <- list()

# workbook for final df
wb <- createWorkbook()

# PSC effect
rpsc_effect <- list()
```

#### Genus level

```{r, warning=FALSE}
level="genus"
```

Aggregate taxa

```{r, warning=FALSE}
genus_data <- aggregate_taxa(lcolon_asv_tab,
                             lcolon_taxa_tab,
                             taxonomic_level = level)

lcolon_genus_tab <- genus_data[[1]]
lcolon_genus_taxa_tab <- genus_data[[2]]

lcolon_genus_asv_taxa_tab <- create_asv_taxa_table(lcolon_genus_tab,
                                                  lcolon_genus_taxa_tab)
```

##### rPSC vs non-rPSC

###### linDA

```{r, warning=FALSE}
group <- c("non-rPSC","rPSC")
comparison_name <- paste0(group[1], " vs ",group[2])
```

```{r, echo=FALSE}
comparison_title <- sub('"(.+?)" vs "(.+?)"', '"\\2" and "\\1"', comparison_name)
```

```{r, warning=FALSE,  fig.width=15, fig.height=5, fig.fullwidth=TRUE}
taxa_to_test <- read.xlsx("../results/Q1/univariate_analysis/significant_taxa_right_colon.xlsx", sheet=" genus healthy vs pre_ltx")$SeqID

lcolon_genus_tab_2 <-  lcolon_genus_tab %>% dplyr::filter(SeqID %in% taxa_to_test)
lcolon_genus_taxa_tab_2 <-  lcolon_genus_taxa_tab %>% dplyr::filter(SeqID %in% taxa_to_test)

# prepare the data
linda_data <- binomial_prep(lcolon_genus_tab_2,
                            lcolon_genus_taxa_tab_2,
                            lcolon_metadata,
                            group, usage="linDA",filtering = FALSE)

filt_lcolon_uni_data <- linda_data[[1]]
filt_lcolon_uni_taxa <- linda_data[[2]]
filt_lcolon_uni_metadata <- linda_data[[3]]

# fit the model
linda.output <- bootstrap_linda(filt_lcolon_uni_data,
                                filt_lcolon_uni_metadata,
                                formula = '~ Group * Country',
                                group,
                                N = 3)

# save the results
group1 <- paste0(group[1], " vs ","Group",group[2])
group2 <- paste0(group[1], " , ",group[2], " - CZ vs NO") 
group3 <- paste0(group[1], " vs ","Group",group[2], ":CountryNO")


for (grp in c(group1,group2,group3)){
  raw_linda_results_genus[[segment]][[grp]] <- 
    rawlinda.df(linda.output,
                grp,
                filt_lcolon_uni_data,
                filt_lcolon_uni_taxa)
  
  linda_results_genus[[segment]][[grp]] <- 
    linda.df(linda.output,
             grp,
             filt_lcolon_uni_data,
             filt_lcolon_uni_taxa)
}
```

```{r, warning=FALSE,  fig.width=15, fig.height=5, fig.fullwidth=TRUE}
# volcano plot
volcano_1 <- volcano_plot_linda(linda.output, group1, 
                                taxa_table = filt_lcolon_uni_taxa) + 
            ggtitle(paste(group,collapse=" vs "))

volcano_2  <- volcano_plot_linda(linda.output, group2, 
                                taxa_table = filt_lcolon_uni_taxa) + 
            ggtitle("NO vs CZ") 

volcano_3 <- volcano_plot_linda(linda.output, group3, 
                                taxa_table = filt_lcolon_uni_taxa) +
            ggtitle("Interaction effect")

volcano <- ggarrange(volcano_1,volcano_2,volcano_3, ncol=3)

# see the plot
volcano
```

###### MaAsLin2

```{r, warning=FALSE,  echo=FALSE,results='hide',message=FALSE,warning=FALSE}
fit_fata <- bootstrap_maaslin(
    input_data = filt_lcolon_uni_data, 
    input_metadata = filt_lcolon_uni_metadata, 
    output = file.path(path_maaslin,group1), 
    fixed_effects = c('Group', 'Country'),
    N = 3)
```

```{r, warning=FALSE}
volcano1 <- volcano_plot_maaslin(fit_data,filt_lcolon_uni_taxa) + 
            ggtitle(paste(group[1], "vs", group[2]))

volcano2 <- volcano_plot_maaslin(fit_data,filt_lcolon_uni_taxa,variable="Country") + 
            ggtitle("NO vs CZ")

volcano <- ggarrange(volcano1,volcano2, ncol=2)

volcano
```

###### Group - Intersection

```{r, warning=FALSE}
intersection_results <- group_intersection(group, list_intersections, list_venns,
                                           linda.output, fit_data,
                                           raw_linda_results_genus, 
                                           segment = segment,
                                           level=level)

list_intersections <- intersection_results[[1]]
list_venns <- intersection_results[[2]]
venn <- intersection_results[[3]]

# show the results
venn
```

###### Interaction effect

```{r, warning=FALSE}
list_interaction_significant <- country_interaction(group,
                                                    linda.output, 
                                                    list_intersections,
                                          filt_lcolon_uni_data,
                                          filt_lcolon_uni_metadata,
                                          segment=segment,
                                          level=level)

# see the result
## significant interaction effect
list_interaction_significant[[1]]

## results for czech cohort
list_interaction_significant[[2]]

## results for norwegian cohort
list_interaction_significant[[3]]
```

Removing problematic taxa

```{r, warning=FALSE}
list_intersections <- removing_interaction_problems(group,
                                                    list_interaction_significant,
                                                    list_intersections,
                                                    segment=segment,
                                                    level=level)
```


###### Basic statistics

```{r, warning=FALSE}
uni_df <-  merge(basic_univariate_statistics(linda_data,group),
                 raw_linda_results_genus[[segment]][[group1]],
                 by="SeqID",all=TRUE)

uni_df[["final_sig"]] <- uni_df$SeqID %in% list_intersections[[paste(segment,level,comparison_name)]][["SeqID"]]
uni_statistics[[segment]][[paste(level,comparison_name)]] <- uni_df

# for comparison
new_name <- comparison_name
addWorksheet(wb, sheetName = new_name)
writeData(wb, sheet = new_name, uni_df, rowNames=FALSE)
```

###### Prevalence testing

```{r}
uni_df$prevalence_diff <- abs(uni_df$`PREVALENCE_ rPSC` - uni_df$`PREVALENCE_ non-rPSC`)

#uni_df %<>% dplyr::filter(prevalence_diff >=0.2)
#filt_lcolon_uni_data <- filt_lcolon_uni_data[uni_df$SeqID,]
filt_lcolon_uni_data <- ifelse(filt_lcolon_uni_data>0,"YES","NO")
filt_lcolon_uni_data %<>% t() %>% as.data.frame()
#filt_lcolon_uni_taxa %<>% dplyr::filter(SeqID %in% uni_df$SeqID)
filt_lcolon_uni_data <- merge(filt_lcolon_uni_data %>% rownames_to_column("SampleID"),filt_lcolon_metadata,by="SampleID",all.x=TRUE) %>% column_to_rownames("SampleID") %>% distinct(Patient, .keep_all = TRUE) 

p_values <- c()
for (taxon in uni_df$SeqID){
data <- filt_lcolon_uni_data %>%
  group_by(!!sym(taxon), Group) %>%
  count() %>%
  group_by(!!sym(taxon), Group) %>%
  summarise(n = sum(n), .groups = "drop") %>%
  pivot_wider(names_from = Group, values_from = n) %>%
  column_to_rownames(taxon) %>%
  as.matrix()

data
chi_res <- chisq.test(data)
chi_res$expected
#p_values <- c(p_values,chi_res$p.value)

test <- fisher.test(data)
test$p.value
p_values <- c(p_values,test$p.value)
}

p_values_adjusted <- p.adjust(p_values,method = "BH")
names(p_values_adjusted) <- uni_df$SeqID
names(p_values) <- uni_df$SeqID
p_values_adjusted[which(p_values_adjusted<0.05)]
```

```{r}
asv_tab <- as.data.frame(rrarefy(
    linda_data[[1]] %>% t(), 
    sample=10000)) %>% t() %>% as.data.frame()

asv_tab <- ifelse(asv_tab>10,"YES","NO")
asv_tab %<>% t() %>% as.data.frame()

set.seed(123)
asv_tab <- merge(asv_tab %>% rownames_to_column("SampleID"),filt_lcolon_metadata,by="SampleID",all.x=TRUE) %>% column_to_rownames("SampleID") %>% distinct(Patient, .keep_all = TRUE) 

p_values <- c()
for (taxon in uni_df$SeqID){
data <- asv_tab %>%
  group_by(!!sym(taxon), Group) %>%
  count() %>%
  group_by(!!sym(taxon), Group) %>%
  summarise(n = sum(n), .groups = "drop") %>%
  pivot_wider(names_from = Group, values_from = n) %>%
  column_to_rownames(taxon) %>%
  as.matrix()

data2 <- asv_tab %>%
  group_by(!!sym(taxon), Group,Country) %>%
  count() 


data[is.na(data)] <- 0
chi_res <- chisq.test(data)
chi_res$expected
test <- fisher.test(data)
test$p.value
chi_res$p.value
p_values <- c(p_values,test$p.value)
}

p_values_adjusted <- p.adjust(p_values,method = "BH")
names(p_values_adjusted) <- uni_df$SeqID
names(p_values) <- uni_df$SeqID
p_values_adjusted[which(p_values_adjusted<0.05)]

df <- data.frame(SeqID=uni_df$SeqID,
                 p=p_values,
                 p_adj=p_values_adjusted)
```

```{r}
prevalence_df <- uni_df[,c("SeqID","PREVALENCE_ non-rPSC","PREVALENCE_ rPSC")]
prevalence_df <- merge(prevalence_df,df,by="SeqID",all=TRUE)
write.xlsx(prevalence_df,file.path(path,paste0("supplements_prevalence_",segment,".xlsx")))
```

##### rPSC vs healthy

```{r, warning=FALSE}
group <- c("healthy","rPSC")
comparison_name <- paste0(group[1], " vs ",group[2])
```

```{r, echo=FALSE}
comparison_title <- sub('"(.+?)" vs "(.+?)"', '"\\2" and "\\1"', comparison_name)
```

###### linDA

```{r, warning=FALSE,  fig.width=15, fig.height=5, fig.fullwidth=TRUE}

# prepare the data
linda_data <- binomial_prep(lcolon_genus_tab,
                            lcolon_genus_taxa_tab,
                            lcolon_metadata,
                            group, usage="linDA")

filt_lcolon_uni_data <- linda_data[[1]]
filt_lcolon_uni_taxa <- linda_data[[2]]
filt_lcolon_uni_metadata <- linda_data[[3]]

# fit the model
linda.output <- bootstrap_linda(filt_lcolon_uni_data,
                                filt_lcolon_uni_metadata,
                                formula = '~ Group * Country',
                                group,
                                N = 3)

# save the results
group1 <- paste0(group[1], " vs ","Group",group[2])
group2 <- paste0(group[1], " , ",group[2], " - CZ vs NO") 
group3 <- paste0(group[1], " vs ","Group",group[2], ":CountryNO")

for (grp in c(group1,group2,group3)){
  raw_linda_results_genus[[segment]][[grp]] <- 
    rawlinda.df(linda.output,
                grp,
                filt_lcolon_uni_data,
                filt_lcolon_uni_taxa)
  
  linda_results_genus[[segment]][[grp]] <- 
    linda.df(linda.output,
             grp,
             filt_lcolon_uni_data,
             filt_lcolon_uni_taxa)
}

```

```{r, warning=FALSE,  fig.width=15, fig.height=5, fig.fullwidth=TRUE}
# volcano plot
volcano_1 <- volcano_plot_linda(linda.output, group1, 
                                taxa_table = filt_lcolon_uni_taxa) + 
            ggtitle(paste(group,collapse=" vs "))

volcano_2  <- volcano_plot_linda(linda.output, group2, 
                                taxa_table = filt_lcolon_uni_taxa) + 
            ggtitle("NO vs CZ") 

volcano_3 <- volcano_plot_linda(linda.output, group3, 
                                taxa_table = filt_lcolon_uni_taxa) +
            ggtitle("Interaction effect")

volcano <- ggarrange(volcano_1,volcano_2,volcano_3, ncol=3)

# see the plot
volcano
```

###### MaAsLin2

```{r, warning=FALSE,  echo=FALSE,results='hide',message=FALSE,warning=FALSE}

fit_fata <- bootstrap_maaslin(
    input_data = filt_lcolon_uni_data, 
    input_metadata = filt_lcolon_uni_metadata, 
    output = file.path(path_maaslin,group1), 
    fixed_effects = c('Group', 'Country'),
    N = 3)
```

```{r, warning=FALSE}
volcano1 <- volcano_plot_maaslin(fit_data,filt_lcolon_uni_taxa) + 
            ggtitle(paste(group[1], "vs", group[2]))

volcano2 <- volcano_plot_maaslin(fit_data,filt_lcolon_uni_taxa,variable="Country") + 
            ggtitle("NO vs CZ")

volcano <- ggarrange(volcano1,volcano2, ncol=2)
volcano
```

###### Group - Intersection

```{r, warning=FALSE}
intersection_results <- group_intersection(group, 
                                           list_intersections, 
                                           list_venns,
                                           linda.output, fit_data,
                                           raw_linda_results_genus, 
                                           segment = segment,
                                           level=level)

list_intersections <- intersection_results[[1]]
list_venns <- intersection_results[[2]]
venn <- intersection_results[[3]]

# show the results
venn
```

###### Interaction effect

```{r, warning=FALSE}
list_interaction_significant <- country_interaction(group,
                                                    linda.output, 
                                                    list_intersections,
                                          filt_lcolon_uni_data,
                                          filt_lcolon_uni_metadata,
                                          segment=segment,
                                          level=level)

# see the result
## significant interaction effect
list_interaction_significant[[1]]

## results for czech cohort
list_interaction_significant[[2]]

## results for norwegian cohort
list_interaction_significant[[3]]
```

Removing problematic taxa

```{r, warning=FALSE}
list_intersections <- removing_interaction_problems(group,
                                                    list_interaction_significant,
                                                    list_intersections,
                                                    segment=segment,
                                                    level=level)
```

###### Basic statistics

```{r, warning=FALSE}
uni_df <-  merge(basic_univariate_statistics(linda_data,group),
                 raw_linda_results_genus[[segment]][[group1]],
                 by="SeqID",all=TRUE)

uni_df[["final_sig"]] <- uni_df$SeqID %in% list_intersections[[paste(segment,level,comparison_name)]][["SeqID"]]
uni_statistics[[segment]][[paste(level,comparison_name)]] <- uni_df

# for comparison
new_name <- comparison_name
addWorksheet(wb, sheetName = new_name)
writeData(wb, sheet = new_name, uni_df, rowNames=FALSE)
```

##### non-rPSC vs healthy

```{r, warning=FALSE}
group <- c("healthy","non-rPSC")
comparison_name <- paste0(group[1], " vs ",group[2])
```

```{r, echo=FALSE}
comparison_title <- sub('"(.+?)" vs "(.+?)"', '"\\2" and "\\1"', comparison_name)
```

###### linDA

```{r, warning=FALSE,  fig.width=15, fig.height=5, fig.fullwidth=TRUE}

# prepare the data
linda_data <- binomial_prep(lcolon_genus_tab,
                            lcolon_genus_taxa_tab,
                            lcolon_metadata,group,
                            usage="linDA")

filt_lcolon_uni_data <- linda_data[[1]]
filt_lcolon_uni_taxa <- linda_data[[2]]
filt_lcolon_uni_metadata <- linda_data[[3]]

# fit the model
linda.output <- bootstrap_linda(filt_lcolon_uni_data,
                                filt_lcolon_uni_metadata,
                                formula = '~ Group * Country',
                                group,
                                N = 3)

# save the results
group1 <- paste0(group[1], " vs ","Group",group[2])
group2 <- paste0(group[1], " , ",group[2], " - CZ vs NO") 
group3 <- paste0(group[1], " vs ","Group",group[2], ":CountryNO")

for (grp in c(group1,group2,group3)){
  raw_linda_results_genus[[segment]][[grp]] <- 
    rawlinda.df(linda.output,
                grp,
                filt_lcolon_uni_data,
                filt_lcolon_uni_taxa)
  
  linda_results_genus[[segment]][[grp]] <- 
    linda.df(linda.output,
             grp,
             filt_lcolon_uni_data,
             filt_lcolon_uni_taxa)
}

```

```{r, warning=FALSE,  fig.width=15, fig.height=5, fig.fullwidth=TRUE}
# volcano plot
volcano_1 <- volcano_plot_linda(linda.output, group1, 
                                taxa_table = filt_lcolon_uni_taxa) + 
            ggtitle(paste(group,collapse=" vs "))

volcano_2  <- volcano_plot_linda(linda.output, group2, 
                                taxa_table = filt_lcolon_uni_taxa) + 
            ggtitle("NO vs CZ") 

volcano_3 <- volcano_plot_linda(linda.output, group3, 
                                taxa_table = filt_lcolon_uni_taxa) +
            ggtitle("Interaction effect")

volcano <- ggarrange(volcano_1,volcano_2,volcano_3, ncol=3)

# see the plot
volcano
```

###### MaAsLin2

```{r, warning=FALSE,  echo=FALSE,results='hide',message=FALSE,warning=FALSE}

fit_fata <- bootstrap_maaslin(
    input_data = filt_lcolon_uni_data, 
    input_metadata = filt_lcolon_uni_metadata, 
    output = file.path(path_maaslin,group1), 
    fixed_effects = c('Group', 'Country'),
    N = 3)
```

```{r, warning=FALSE}
volcano1 <- volcano_plot_maaslin(fit_data,filt_lcolon_uni_taxa) + 
            ggtitle(paste(group[1], "vs", group[2]))

volcano2 <- volcano_plot_maaslin(fit_data,filt_lcolon_uni_taxa,variable="Country") + 
            ggtitle("NO vs CZ")

volcano <- ggarrange(volcano1,volcano2, ncol=2)
volcano
```

###### Group - Intersection

```{r, warning=FALSE}
intersection_results <- group_intersection(group, list_intersections, list_venns,
                                           linda.output, fit_data,
                                           raw_linda_results_genus, 
                                           segment = segment,
                                           level=level)

list_intersections <- intersection_results[[1]]
list_venns <- intersection_results[[2]]
venn <- intersection_results[[3]]

# show the results
venn
```


###### Interaction effect

```{r, warning=FALSE}
list_interaction_significant <- country_interaction(group,
                                                    linda.output, 
                                                    list_intersections,
                                          filt_lcolon_uni_data,
                                          filt_lcolon_uni_metadata,
                                          segment=segment,
                                          level=level)

# see the result
## significant interaction effect
list_interaction_significant[[1]]

## results for czech cohort
list_interaction_significant[[2]]

## results for norwegian cohort
list_interaction_significant[[3]]
```

Removing problematic taxa

```{r, warning=FALSE}
list_intersections <- removing_interaction_problems(group,
                                                    list_interaction_significant,
                                                    list_intersections,
                                                    segment=segment,
                                                    level=level)
```

##### Basic statistics

```{r, warning=FALSE}
uni_df <-  merge(basic_univariate_statistics(linda_data,group),
                 raw_linda_results_genus[[segment]][[group1]],
                 by="SeqID",all=TRUE)

uni_df[["final_sig"]] <- uni_df$SeqID %in% list_intersections[[paste(segment,level,comparison_name)]][["SeqID"]]
uni_statistics[[segment]][[paste(level,comparison_name)]] <- uni_df

# for comparison
new_name <- comparison_name
addWorksheet(wb, sheetName = new_name)
writeData(wb, sheet = new_name, uni_df, rowNames=FALSE)
```

##### Visualization

Heatmap visualizing the linDA's logFoldChange for taxa with p \< 0.1.

```{r, warning=FALSE,  fig.width=10, fig.height=17, fig.fullwidth=TRUE}
list_heatmap <- list_intersections[grep(paste(segment,level),
                                  names(list_intersections),value=TRUE)]

p_heatmap_linda <- heatmap_linda(list_heatmap,lcolon_taxa_tab)
p_heatmap_linda
```

Dot heatmap

```{r, warning=FALSE,  fig.width=5, fig.height=17, fig.fullwidth=TRUE}
dotheatmap_linda <- dot_heatmap_linda(list_heatmap,
                                      uni_statistics$right_colon[grepl(level,names(uni_statistics$right_colon))],
                                      lcolon_taxa_tab) + xlab("") + ylab("")
dotheatmap_linda
```

**Horizontal bar plot**

```{r, warning=FALSE, fig.height=10, fig.width=4}
p_prevalence <- horizontal_barplot(wb,taxa=levels(dotheatmap_linda$data$SeqID))

p_prevalence_final <- ggarrange(p_prevalence,
                                ggplot() + theme_minimal(),
                                nrow = 2,heights = c(1,0.09))
p <- ggarrange(dotheatmap_linda + theme(legend.position = "none"), p_prevalence_final,ncol=2, widths = c(1,0.3))
p

dot_heatmap_lcolon <- p
```

```{r, warning=FALSE, results='hide', echo=FALSE}
pdf("../figures/Q2/dotplot_rcolon.pdf",
    height =10,width = 4)
p
dev.off()
```

<!-- ##### rPSC effect -->

<!-- **pre_LTx vs Healthy and Post_LTx vs Healthy intersection** -->


<!-- ```{r, warning=FALSE} -->
<!-- A <- list_intersections[[paste(segment,level,"healthy vs rPSC")]] -->
<!-- B <- list_intersections[[paste(segment,level,"healthy vs non-rPSC")]] -->
<!-- df <- A[!(A$SeqID %in% B$SeqID),] -->

<!-- rpsc_effect[[paste(segment,level)]] <- df -->

<!-- # see the results -->
<!-- rpsc_effect[[paste(segment,level)]]  -->
<!-- ``` -->

#### Saving results

```{r, warning=FALSE}
# ALL DATA
saveWorkbook(wb,file.path(path,paste0("uni_analysis_wb_",segment,".xlsx")),
             overwrite = TRUE)
```

## Machine learning

```{r, warning=FALSE}
path = "../results/Q2/models"
```

### ElasticNet

```{r, warning=FALSE}
model="enet"
```

#### Genus level

```{r, warning=FALSE}
level="genus"
```

Aggregate taxa

```{r, warning=FALSE}
genus_data <- aggregate_taxa(lcolon_asv_tab,
                             lcolon_taxa_tab,
                             taxonomic_level = level)

lcolon_genus_tab <- genus_data[[1]]
lcolon_genus_taxa_tab <- genus_data[[2]]
```

##### rPSC vs non-rPSC - all

```{r, warning=FALSE}
group <- c("rPSC","non-rPSC")
comparison_name <- paste0(group[1], " vs ",group[2])
```

```{r, warning=FALSE}
model_name <- paste(comparison_name,level,segment)

# prepare the data
filt_lcolon_uni_data <- binomial_prep(lcolon_genus_tab,
                                     lcolon_genus_taxa_tab,
                                     lcolon_metadata,
                                     group, 
                                     usage="ml_clr",
                                     patient = TRUE)

# fit the model
enet_model <- glmnet_binomial(filt_lcolon_uni_data,
                              sample_method = "atypboot",
                              outcome="Group", 
                              N=10,
                              reuse=TRUE,
                              file=model_name,
                              Q="Q2",
                              clust_var = "Patient")

# ROC
roc_c <- roc_curve(enet_model, group)

# save the results
models_summ[[model_name]] <- enet_model$model_summary
models_cm[[model_name]] <- enet_model$conf_matrices$original
roc_cs[[model_name]] <- enet_model$kfold_rocobjs
betas[[model_name]] <- as.matrix(enet_model$betas)

# see the results
enet_model$model_summary %>% t()
enet_model$conf_matrices
enet_model$plot

roc_c
```

##### rPSC vs non-rPSC - subset

```{r, warning=FALSE}
group <- c("rPSC","non-rPSC")
comparison_name <- paste0(group[1], " vs ",group[2]," subset")
```

```{r, warning=FALSE}
taxa_to_test <- read.xlsx("../results/Q1/univariate_analysis/significant_taxa_right_colon.xlsx", sheet=" genus healthy vs pre_ltx")

model_name <- paste(comparison_name,level,segment)

# prepare the data
filt_lcolon_uni_data <- binomial_prep_psc_effect(lcolon_genus_tab,
                                                lcolon_genus_taxa_tab, 
                                                lcolon_metadata,
                                                taxa_to_test,
                                                usage="ml_clr",
                                                patient = TRUE)

# fit the model
enet_model <- glmnet_binomial(filt_lcolon_uni_data,
                              sample_method = "atypboot",
                              outcome="Group", 
                              N=10,
                              reuse=TRUE,
                              file=model_name,
                              Q="Q2",clust_var = "Patient")

# ROC
roc_c <- roc_curve(enet_model, group)

# save the results
models_summ[[model_name]] <- enet_model$model_summary
models_cm[[model_name]] <- enet_model$conf_matrices$original
roc_cs[[model_name]] <- enet_model$kfold_rocobjs
betas[[model_name]] <- as.matrix(enet_model$betas)

# see the results
enet_model$model_summary %>% t()
enet_model$conf_matrices
enet_model$plot

roc_c

```

#### Saving results

```{r, warning=FALSE}
models_summ_df_colon <- do.call(rbind, 
  models_summ[grep(segment,names(models_summ),value = TRUE)])

write.csv(models_summ_df_colon,file.path(path,paste0("elastic_net_",segment,".csv")))
```

## Results overview

#### Alpha diversity

**Linear models**

```{r,warning=FALSE}
knitr::kable(pc_observed[[segment]],
             digits = 3,
             caption = "Results of linear model testing ASV Richness")

knitr::kable(pc_shannon[[segment]],
             digits = 3,
             caption = "Results of linear model testing Shannon index")

knitr::kable(pc_simpson[[segment]],
             digits = 3,
             caption = "Results of linear model testing Simpson index")

knitr::kable(pc_pielou[[segment]],
             digits = 3,
             caption = "Results of linear model testing Pielou index")
```

**Plots**

```{r, warning=FALSE, fig.width=6, fig.height=4, fig.fullwidth=TRUE}
alpha_div_plots[[paste(segment,"Country")]]
```

#### Beta diversity

**Main results**

```{r,warning=FALSE}
knitr::kable(pairwise_aitchison_raw[[paste("genus", segment)]],
             digits = 3,
             caption = "Results of PERMANOVA - robust aitchison distance")
```

**PCA**

```{r, warning=FALSE,  fig.width=5, fig.height=3, fig.fullwidth=TRUE, eval=TRUE}
pca_plots_list[[paste(segment,"genus custom")]]
```




#### Univariate analysis

**Number of significant taxa**

```{r, warning=FALSE}
knitr::kable(as.data.frame(lapply(list_intersections,nrow)) %>% t() %>% 
  `colnames<-`("Count") )
```


#### Machine learning

**Main models**


```{r, warning=FALSE}
knitr::kable(models_summ_df_colon %>% dplyr::select(
"alpha","lambda",
"auc_optimism_corrected",
"auc_optimism_corrected_CIL",
"auc_optimism_corrected_CIU"),
             digits=3,caption="Elastic net results")
```

**ROC - Genus level**

```{r, warning=FALSE,  warning=FALSE}
roc_curve_all_custom(roc_cs[3:4],Q="Q2",
                     model_name="enet_model",legend=FALSE)
```


# Paper-ready visualizations

## Alpha diversity

```{r, warning=FALSE, fig.width=10, fig.height=4, fig.fullwidth=TRUE}
p_A <- alpha_div_plots$`terminal_ileum Country` +
  ggtitle("Terminal ileum")+
  theme(plot.title = element_text(hjust=0.5,face = "bold",size = 15)) 

p_B <-  alpha_div_plots$`right_colon Country` +
  ggtitle("Right colon") +
  theme(plot.title = element_text(hjust=0.5,face = "bold",size = 15)) 

p_C <-  alpha_div_plots$`left_colon Country` +
  ggtitle("Left colon") +
  theme(plot.title = element_text(hjust=0.5,face = "bold",size = 15)) 


Q2_alpha <- ggarrange(p_A, ggplot() + theme_minimal(),
                      p_B, ggplot() + theme_minimal(), p_C,
                      nrow=1, ncol=5,widths = c(1,0.1,1,0.1,1))

Q2_alpha
```

## Beta diversity

```{r, warning=FALSE, fig.width=10, fig.height=4, fig.fullwidth=TRUE}
pca_ti <- pca_plots_list$`terminal_ileum genus custom` 
pca_right_colon <- pca_plots_list$`right_colon genus custom` 
pca_left_colon <- pca_plots_list$`left_colon genus custom` 

genus_Q2_beta <- ggarrange(pca_ti,
                           ggplot() + theme_minimal(),
                           pca_right_colon,
                           ggplot() + theme_minimal(),
                           pca_left_colon,
                           ncol=5,
                           widths = c(1,0.1,1,0.1,1))
genus_Q2_beta
```

**Alpha + Beta diversity**

```{r, warning=FALSE, fig.width=10, fig.height=8, fig.fullwidth=TRUE}
alpha_beta <- ggarrange(Q2_alpha,genus_Q2_beta,
                        ncol = 1,nrow=2,labels = c("A","B"))
alpha_beta
```

## Elastic net

Genus level

```{r, warning=FALSE, fig.width=10, fig.height=4, fig.fullwidth=TRUE, warning=FALSE}
#models_to_plot <- c("knn_model","rf_model","gbm_model","enet_model")
#names(models_to_plot) <- c("kNN","RF","GBoost","ENET")

rocs_list <- list()
rocs_list[["enet_model"]] <- roc_cs
models_to_plot <- c("enet_model")
names(models_to_plot) <- c("ENET")

# ILEUM
plot_list_ileum <- list()
for (model_name in models_to_plot) {
  plot_list_ileum[[model_name]] <- 
    roc_curve_all_custom(rocs_list[[model_name]][c(1:2)],
                       Q="Q2",
                       model_name=model_name,legend = FALSE) + 
    #ggtitle(names(models_to_plot)[which(model_name==models_to_plot)]) + 
    theme(plot.title = element_text(face = "bold",size = 15)) 
}

roc_curve_ti <- ggarrange(plotlist = plot_list_ileum)

# RIGHT COLON

plot_list_right_colon <- list()
for (model_name in models_to_plot) {
  plot_list_right_colon[[model_name]] <- 
    roc_curve_all_custom(rocs_list[[model_name]][c(3:4)],
                       Q="Q2",
                       model_name=model_name,legend = FALSE) + 
    #ggtitle(names(models_to_plot)[which(model_name==models_to_plot)]) + 
    theme(plot.title = element_text(face = "bold",size = 15))   
}

roc_curve_right_colon <- ggarrange(plotlist = plot_list_right_colon)

# LEFT COLON

plot_list_left_colon <- list()
for (model_name in models_to_plot) {
  plot_list_left_colon[[model_name]] <- 
    roc_curve_all_custom(rocs_list[[model_name]][c(5:6)],
                       Q="Q2",
                       model_name=model_name,legend = FALSE) + 
    #ggtitle(names(models_to_plot)[which(model_name==models_to_plot)]) + 
    theme(plot.title = element_text(face = "bold",size = 15))   
}

roc_curve_left_colon <- ggarrange(plotlist = plot_list_right_colon)

roc_curve_plot <- ggarrange(roc_curve_ti,
                            ggplot() + theme_minimal(),
                            roc_curve_right_colon,
                            ggplot() + theme_minimal(),
                            roc_curve_left_colon,
                            ncol=5, widths = c(1,0.1,1,0.1,1))
roc_curve_plot

```

## FIGURE 4

```{r, warning=FALSE, fig.width=10, fig.height=12, fig.fullwidth=TRUE}
alpha_beta_elastic <- ggarrange(Q2_alpha,genus_Q2_beta,roc_curve_plot,
                        ncol = 1,nrow=3,labels = LETTERS,heights = c(1,1,0.8))
alpha_beta_elastic

alpha_beta_elastic <- ggarrange(alpha_beta_elastic,ggplot() + theme_minimal(),ncol=2,
                                widths = c(1,0.15))

alpha_beta_elastic
```

```{r, warning=FALSE,results='hide',echo=FALSE}
pdf("../figures/Q2/FIGURE4.pdf",paper="a4r",height = 10,width = 16)
alpha_beta_elastic
dev.off()
```

## FIGURE5 - DAA

```{r, warning=FALSE, fig.width=8, fig.height=10, fig.fullwidth=TRUE}
p_ileum <- dot_heatmap_ileum + 
  ggtitle("Terminal ileum") +
  theme(plot.title = element_text(hjust=0.5,face = "bold",size = 15),
        legend.position = "none")


p_rcolon <- dot_heatmap_rcolon  +
  ggtitle("Right colon") +
  theme(plot.title = element_text(hjust=0.5,face = "bold",size = 15),
        legend.position = "none")

p_lcolon <- dot_heatmap_lcolon  +
  ggtitle("Left colon") +
  theme(plot.title = element_text(hjust=0.5,face = "bold",size = 15),
        legend.position = "none")

heatmap_plot <- ggarrange(p_ileum,
                          p_rcolon,
                          p_lcolon,
                          ncol = 3,labels=c("A","B","C"),
                          widths = c(1,1,1))


heatmap_plot 

```

```{r, warning=FALSE,results='hide',echo=FALSE}
pdf("../figures/Q2/FIGURE5.pdf",
    height =10,width = 14,paper="a4r")
heatmap_plot
dev.off()
```

# Session info

```{r, warning=FALSE}
sessionInfo()
```
