---
title: "Mucosal microbiota alterations in primary sclerosis cholangitis persist after liver transplantation and are associated with clinical features independently of geography"
subtitle: "2. Hypothesis: PSC recurrence is associated with a specific composition of the gut microbiome"
format: 
  html:  
    embed-resources: true
    keep-md: false
    toc: true
    toc-expand: 3
    toc-depth: 6
    toc-location: left
    number-sections: true
    number-depth: 6
    code-fold: false
    code-tools: true
    code-summary: "Open code"
    grid: 
      body-width: 1000px
      sidebar-width: 600px
      margin-width: 100px
editor: source
project:
  type: default
  output-dir: output
fontsize: 12 px
---

-----------------------------------------------------------

# Introduction

## About

**rPSC vs non-rPSC vs Healthy analysis on merged data**

-   Alpha diversity –\> group effect, country effect, interaction effect

-   Beta diversity – PERMANOVA, PCA -\> group effect, country effect,
    interaction effect

-   DAA -\>

    -   Group effect – linDA + MaAsLin2 intersection

    -   Country effect – taxa with significant interaction effect were excluded based on individual post-hoc analysis. Taxa had to have the same direction in both countries 

-   ML models, trained and validated using out-of-sample bootstrap (500
    resamplings):

   - o ENET

   - o kNN

   - o GBoost

   - o RF

------------------------------------------------------

Pairwise comparisons:

**rPSC vs non-rPSC** –by this comparison, we will find the effect of
recurrence (whether the microbial composition is associated with it)

**non-rPSC vs Healthy** – by this comparison, we will find if the
non-rPSC samples are „closer“ to the healthy samples then rPSC samples
in terms of the microbial composition. In other words, does non-rPSC
samples have „healthy microbiome“?

**rPSC vs Healthy** – by this comparison, we will find how much are the
rPSC samples different from healthy samples.

However, the rPSC and non-rPSC samples are not different at all (see
results). Therefore, we can do EXCLUSIVE LEFT JOIN of differentially
abundant taxa from the analyses (rPSC vs Healthy, non-rPSC vs Healthy)

*Importing libraries and custom functions built for this analysis*

```{r, warning=FALSE}
source("custom_functions.R")
```

## Data Import

Importing ASV, taxa and metadata tables for both Czech and Norway
samples.

**Czech**

```{r, warning=FALSE}
path = "../../data/analysis_ready_data/ikem/"
asv_tab_ikem <- as.data.frame(fread(file.path(path,"asv_table_ikem.csv"),
                                    check.names = FALSE))
taxa_tab_ikem <- as.data.frame(fread(file.path(path,"taxa_table_ikem.csv"),
                                     check.names = FALSE))
metadata_ikem <- as.data.frame(fread(file.path(path,"metadata_ikem.csv"),
                                     check.names = FALSE))
```

**Norway**

```{r, warning=FALSE}
path = "../../data/analysis_ready_data/norway/"
asv_tab_norway <- as.data.frame(fread(file.path(path,"asv_table_norway.csv"),
                                    check.names = FALSE))
taxa_tab_norway <- as.data.frame(fread(file.path(path,"taxa_table_norway.csv"),
                                    check.names = FALSE))
metadata_norway <- as.data.frame(fread(file.path(path,"metadata_norway.csv"),
                                    check.names = FALSE))
```

## Merging data

Merging two countries to create whole dataset

```{r, warning=FALSE}
asv_tab <- merge(asv_tab_ikem,asv_tab_norway,by="SeqID",all=TRUE)
taxa_tab <- merging_taxa_tables(taxa_tab_ikem,taxa_tab_norway)
```

Merging two countries based on the different matrices - Ileum, Colon.

**Terminal ileum**

```{r, warning=FALSE}
ileum_data <- merging_data(asv_tab_1=asv_tab_ikem,
                           asv_tab_2=asv_tab_norway,
                           taxa_tab_1=taxa_tab_ikem,
                           taxa_tab_2=taxa_tab_norway,
                           metadata_1=metadata_ikem,
                           metadata_2=metadata_norway,
                           segment="TI",Q="Q2")

ileum_asv_tab <- ileum_data[[1]]
ileum_taxa_tab <- ileum_data[[2]]
ileum_metadata <- ileum_data[[3]]
```

**Colon**

```{r, warning=FALSE}
colon_data <- merging_data(asv_tab_1=asv_tab_ikem,
                           asv_tab_2=asv_tab_norway,
                           taxa_tab_1=taxa_tab_ikem,
                           taxa_tab_2=taxa_tab_norway,
                           metadata_1=metadata_ikem,
                           metadata_2=metadata_norway,
                           segment="colon",Q="Q2")

colon_asv_tab <- colon_data[[1]]
colon_taxa_tab <- colon_data[[2]]
colon_metadata <- colon_data[[3]]
```

# Data Analysis - Terminal ileum

```{r, warning=FALSE}
segment="terminal_ileum"
```

## Filtering

Rules: 

  - sequencing depth \> 10000 
  
  - nearZeroVar() with default settings 

**Rarefaction Curve**

```{r, warning=FALSE}
path="../intermediate_files/rarecurves"
seq_depth_threshold <- 10000
```

```{r, warning=FALSE,  eval = FALSE}
ps <- construct_phyloseq(ileum_asv_tab,ileum_taxa_tab,ileum_metadata)
rareres <- get_rarecurve(obj=ps, chunks=500)
save(rareres,file = file.path(path,"rarefaction_ileum.Rdata"))
```

```{r, warning=FALSE}
load(file.path(path,"rarefaction_ileum.Rdata"))

prare <- ggrarecurve(obj=rareres,
                      factorNames="Country",
                      indexNames=c("Observe")) + 
  theme_bw() +
  theme(axis.text=element_text(size=8), panel.grid=element_blank(),
        strip.background = element_rect(colour=NA,fill="grey"),
        strip.text.x = element_text(face="bold")) + 
  geom_vline(xintercept = seq_depth_threshold, 
             linetype="dashed", 
             color = "red") + 
  xlim(0, 20000)

prare
```

**Library size**

```{r, warning=FALSE,  fig.width=5, fig.height=4, fig.fullwidth=TRUE, warning=FALSE}
read_counts(ileum_asv_tab, line = c(5000,10000))
```

### Sequencing depth

```{r, warning=FALSE}
data_filt <- seq_depth_filtering(ileum_asv_tab,
                                 ileum_taxa_tab,
                                 ileum_metadata,
                                 seq_depth_threshold = 10000)

filt_ileum_asv_tab <- data_filt[[1]]; alpha_ileum_asv_tab <- filt_ileum_asv_tab
filt_ileum_taxa_tab <- data_filt[[2]]; alpha_ileum_taxa_tab <- filt_ileum_taxa_tab
filt_ileum_metadata <- data_filt[[3]]; alpha_ileum_metadata <- filt_ileum_metadata

seq_step <- dim(filt_ileum_asv_tab)[1]
```

**Library size**

```{r, warning=FALSE,  fig.width=5, fig.height=4, fig.fullwidth=TRUE, warning=FALSE}
read_counts(filt_ileum_asv_tab,line = c(5000,10000))
```

### NearZeroVar

```{r, warning=FALSE}
data_filt <- nearzerovar_filtering(filt_ileum_asv_tab, 
                                   filt_ileum_taxa_tab,
                                   filt_ileum_metadata)

filt_ileum_asv_tab <- data_filt[[1]]
filt_ileum_taxa_tab <- data_filt[[2]]
nearzero_step <- dim(filt_ileum_asv_tab)[1]
```

**Library size**

```{r, warning=FALSE,  fig.width=5, fig.height=4, fig.fullwidth=TRUE, warning=FALSE}
read_counts(filt_ileum_asv_tab,line = c(5000,10000))
```

### Final Counts

```{r, warning=FALSE}
final_counts_filtering(ileum_asv_tab,
                       filt_ileum_asv_tab,
                       filt_ileum_metadata,
                       seq_step, 0, nearzero_step) %>% `colnames<-`("Count")
```

## Beta diversity

Calculating Aitchison distance (euclidean distance on clr-transformed
data), both at ASV and genus level.


### Supplementary analysis

```{r, warning=FALSE}
supplements_beta <- list()
```

#### Genus level

```{r, warning=FALSE}
level="genus"
```

##### Bray-Curtis

**PERMANOVA**

```{r, warning=FALSE}
pairwise_df <- filt_ileum_genus_tab %>% column_to_rownames("SeqID") %>% t()

# main effect
pp_main <- pairwise.adonis(pairwise_df,filt_ileum_metadata$Group,covariate = filt_ileum_metadata$Country, sim.method = "bray", p.adjust.m="BH")

# interaction
pp_int <- pairwise.adonis(pairwise_df,filt_ileum_metadata$Group,covariate = filt_ileum_metadata$Country, interaction = TRUE, sim.method = "bray", p.adjust.m="BH")

# tidy the results
pp_factor <- pp_main[[1]]
pp_cov <- pp_main[[2]]
pp_fac.cov <- pp_int[[3]]

cols <- c("pairs","Df","SumsOfSqs", "F.Model","R2","p.value", "p.adjusted", "sig")
colnames(pp_factor) <- cols; colnames(pp_cov) <- cols; colnames(pp_fac.cov) <- cols; 

# save raw results
supplements_beta[[paste("bray",level,segment)]] <- rbind(pp_factor,pp_cov,pp_fac.cov)
```

```{r, warning=FALSE}
# see the results
knitr::kable(pp_factor,digits = 3,caption = "PERMANOVA, GROUP separation")
knitr::kable(pp_cov,digits = 3,caption = "PERMANOVA, COUNTRY separation")
knitr::kable(pp_fac.cov,digits = 3,caption = "PERMANOVA, INTERACTION GROUP:Country")
```

Interaction check

```{r, warning=FALSE}
interaction_sig <- pp_fac.cov$pairs[pp_fac.cov$p.adjusted < 0.05]

for (i in 1:length(interaction_sig)){
  group1 <- unlist(strsplit(interaction_sig[i],split = " vs "))[1]
  group2 <- unlist(strsplit(interaction_sig[i],split = " vs "))[2]
  group2 <- unlist(strsplit(group2,split = " : "))[1]
  
  result_list <- adonis_postanalysis(x=pairwise_df,
                      factors = filt_ileum_metadata$Group,
                      covariate = filt_ileum_metadata$Country, 
                      group1 = group1,
                      group2 = group2,
                      sim.method = 'bray')
  print(result_list)
}

```

**Plots**

```{r, warning=FALSE}
p <- pca_plot_custom(filt_ileum_genus_tab,
                                 filt_ileum_genus_taxa,
                                 filt_ileum_metadata,
                                 measure = "bray",
                                 show_boxplots = TRUE,
                                 variable = "Group", size=3, show_legend=FALSE)

# save the results
supplements_beta[[paste("PCoA bray",level,segment)]] <- p

# see the results
p
```

##### Jaccard

**PERMANOVA**

```{r, warning=FALSE}
pairwise_df <- filt_ileum_genus_tab %>% column_to_rownames("SeqID") %>% t()

# main effect
pp_main <- pairwise.adonis(pairwise_df,filt_ileum_metadata$Group,covariate = filt_ileum_metadata$Country, sim.method = "jaccard", p.adjust.m="BH")

# interaction
pp_int <- pairwise.adonis(pairwise_df,filt_ileum_metadata$Group,covariate = filt_ileum_metadata$Country, interaction = TRUE, sim.method = "jaccard", p.adjust.m="BH")

# tidy the results
pp_factor <- pp_main[[1]]
pp_cov <- pp_main[[2]]
pp_fac.cov <- pp_int[[3]]

cols <- c("pairs","Df","SumsOfSqs", "F.Model","R2","p.value", "p.adjusted", "sig")
colnames(pp_factor) <- cols; colnames(pp_cov) <- cols; colnames(pp_fac.cov) <- cols; 

# save raw results
supplements_beta[[paste("jaccard",level,segment)]] <- rbind(pp_factor, 
                                                            pp_cov, 
                                                            pp_fac.cov)
```

```{r, warning=FALSE}
# see the results
knitr::kable(pp_factor,digits = 3,caption = "PERMANOVA, GROUP separation")
knitr::kable(pp_cov,digits = 3,caption = "PERMANOVA, COUNTRY separation")
knitr::kable(pp_fac.cov,digits = 3,caption = "PERMANOVA, INTERACTION GROUP:Country")
```

Interaction check

```{r, warning=FALSE}
interaction_sig <- pp_fac.cov$pairs[pp_fac.cov$p.adjusted < 0.05]

for (i in 1:length(interaction_sig)){
  group1 <- unlist(strsplit(interaction_sig[i],split = " vs "))[1]
  group2 <- unlist(strsplit(interaction_sig[i],split = " vs "))[2]
  group2 <- unlist(strsplit(group2,split = " : "))[1]
  
  result_list <- adonis_postanalysis(x=pairwise_df,
                      factors = filt_ileum_metadata$Group,
                      covariate = filt_ileum_metadata$Country, 
                      group1 = group1,
                      group2 = group2,
                      sim.method = 'jaccard')
  print(result_list)
}

```

**Plots**

```{r, warning=FALSE}
p <- pca_plot_custom(filt_ileum_genus_tab,
                                 filt_ileum_genus_taxa,
                                 filt_ileum_metadata,
                                 measure = "jaccard",
                                 show_boxplots = TRUE,
                                 variable = "Group", size=3, show_legend=FALSE)

# save the results
supplements_beta[[paste("PCoA jaccard",level,segment)]] <- p

# see the results
p
```

#### ASV level

```{r, warning=FALSE}
level="ASV"
```

##### Aitchison 

*PERMANOVA*

```{r, warning=FALSE}
# preparing data frame
pairwise_df <- filt_ileum_asv_tab %>% column_to_rownames("SeqID") %>% t()

# main effect
pp_main <- pairwise.adonis(pairwise_df,filt_ileum_metadata$Group,covariate = filt_ileum_metadata$Country, sim.method = "robust.aitchison", p.adjust.m="BH")

# interaction
pp_int <- pairwise.adonis(pairwise_df,filt_ileum_metadata$Group,covariate = filt_ileum_metadata$Country, interaction = TRUE, sim.method = "robust.aitchison", p.adjust.m="BH")

pp_factor <- pp_main[[1]]
pp_cov <- pp_main[[2]]
pp_fac.cov <- pp_int[[3]]

cols <- c("pairs","Df","SumsOfSqs", "F.Model","R2","p.value", "p.adjusted", "sig")
colnames(pp_factor) <- cols; colnames(pp_cov) <- cols; colnames(pp_fac.cov) <- cols; 

# save raw results
supplements_beta[[paste("aitchison",level,segment)]] <- rbind(pp_factor, 
                                                            pp_cov, 
                                                            pp_fac.cov)
```

```{r, warning=FALSE}
# see the results
knitr::kable(pp_factor,digits = 3,caption = "PERMANOVA, GROUP separation")
knitr::kable(pp_cov,digits = 3,caption = "PERMANOVA, COUNTRY separation")
knitr::kable(pp_fac.cov,digits = 3,caption = "PERMANOVA, INTERACTION GROUP:Country")
```

Interaction check

```{r, warning=FALSE}
interaction_sig <- pp_fac.cov$pairs[pp_fac.cov$p.adjusted < 0.05]

if (length(interaction_sig)>0){
for (i in 1:length(interaction_sig)){
  group1 <- unlist(strsplit(interaction_sig[i],split = " vs "))[1]
  group2 <- unlist(strsplit(interaction_sig[i],split = " vs "))[2]
  group2 <- unlist(strsplit(group2,split = " : "))[1]
  
  result_list <- adonis_postanalysis(x=pairwise_df,
                      factors = filt_ileum_metadata$Group,
                      covariate = filt_ileum_metadata$Country, 
                      group1 = group1,
                      group2 = group2)
  print(result_list)
}
}
```

**PCoA**

```{r, warning=FALSE}
p <- pca_plot_custom(filt_ileum_asv_tab,
                           filt_ileum_taxa_tab,
                           filt_ileum_metadata,
                           show_boxplots = TRUE,
                           variable = "Group", size=3, show_legend=FALSE)

# save the results
supplements_beta[[paste("PCoA aitchison",level,segment)]] <- p

# see the results
p
```

##### Bray-Curtis 

*PERMANOVA*

```{r, warning=FALSE}
# preparing data frame
pairwise_df <- filt_ileum_asv_tab %>% column_to_rownames("SeqID") %>% t()

# main effect
pp_main <- pairwise.adonis(pairwise_df,filt_ileum_metadata$Group,covariate = filt_ileum_metadata$Country, sim.method = "bray", p.adjust.m="BH")

# interaction
pp_int <- pairwise.adonis(pairwise_df,filt_ileum_metadata$Group,covariate = filt_ileum_metadata$Country, interaction = TRUE, sim.method = "bray", p.adjust.m="BH")

pp_factor <- pp_main[[1]]
pp_cov <- pp_main[[2]]
pp_fac.cov <- pp_int[[3]]

cols <- c("pairs","Df","SumsOfSqs", "F.Model","R2","p.value", "p.adjusted", "sig")
colnames(pp_factor) <- cols; colnames(pp_cov) <- cols; colnames(pp_fac.cov) <- cols; 

# save raw results
supplements_beta[[paste("bray",level,segment)]] <- rbind(pp_factor, 
                                                            pp_cov, 
                                                            pp_fac.cov)

```

```{r, warning=FALSE}
# see the results
knitr::kable(pp_factor,digits = 3,caption = "PERMANOVA, GROUP separation")
knitr::kable(pp_cov,digits = 3,caption = "PERMANOVA, COUNTRY separation")
knitr::kable(pp_fac.cov,digits = 3,caption = "PERMANOVA, INTERACTION GROUP:Country")
```

Interaction check

```{r, warning=FALSE}
interaction_sig <- pp_fac.cov$pairs[pp_fac.cov$p.adjusted < 0.05]

if (length(interaction_sig)>0){
for (i in 1:length(interaction_sig)){
  group1 <- unlist(strsplit(interaction_sig[i],split = " vs "))[1]
  group2 <- unlist(strsplit(interaction_sig[i],split = " vs "))[2]
  group2 <- unlist(strsplit(group2,split = " : "))[1]
  
  result_list <- adonis_postanalysis(x=pairwise_df,
                      factors = filt_ileum_metadata$Group,
                      covariate = filt_ileum_metadata$Country, 
                      group1 = group1,
                      group2 = group2,
                      sim.method = 'bray')
  print(result_list)
}
}
```

**PCoA**

```{r, warning=FALSE}
p <- pca_plot_custom(filt_ileum_asv_tab,
                     filt_ileum_taxa_tab,
                     filt_ileum_metadata,
                     measure = "bray",
                     show_boxplots = TRUE,
                     variable = "Group", size=3, show_legend=FALSE)

# save the results
supplements_beta[[paste("PCoA bray",level,segment)]] <- p

# see the results
p
```

##### Jaccard

*PERMANOVA*

```{r, warning=FALSE}
# preparing data frame
pairwise_df <- filt_ileum_asv_tab %>% column_to_rownames("SeqID") %>% t()

# main effect
pp_main <- pairwise.adonis(pairwise_df,filt_ileum_metadata$Group,covariate = filt_ileum_metadata$Country, sim.method = "jaccard", p.adjust.m="BH")

# interaction
pp_int <- pairwise.adonis(pairwise_df,filt_ileum_metadata$Group,covariate = filt_ileum_metadata$Country, interaction = TRUE, sim.method = "jaccard", p.adjust.m="BH")

pp_factor <- pp_main[[1]]
pp_cov <- pp_main[[2]]
pp_fac.cov <- pp_int[[3]]

cols <- c("pairs","Df","SumsOfSqs", "F.Model","R2","p.value", "p.adjusted", "sig")
colnames(pp_factor) <- cols; colnames(pp_cov) <- cols; colnames(pp_fac.cov) <- cols; 

# save raw results
supplements_beta[[paste("jaccard",level,segment)]] <- rbind(pp_factor, 
                                                            pp_cov, 
                                                            pp_fac.cov)
```

```{r, warning=FALSE}
# see the results
knitr::kable(pp_factor,digits = 3,caption = "PERMANOVA, GROUP separation")
knitr::kable(pp_cov,digits = 3,caption = "PERMANOVA, COUNTRY separation")
knitr::kable(pp_fac.cov,digits = 3,caption = "PERMANOVA, INTERACTION GROUP:Country")
```

Interaction check

```{r, warning=FALSE}
interaction_sig <- pp_fac.cov$pairs[pp_fac.cov$p.adjusted < 0.05]

if (length(interaction_sig)>0){
for (i in 1:length(interaction_sig)){
  group1 <- unlist(strsplit(interaction_sig[i],split = " vs "))[1]
  group2 <- unlist(strsplit(interaction_sig[i],split = " vs "))[2]
  group2 <- unlist(strsplit(group2,split = " : "))[1]
  
  result_list <- adonis_postanalysis(x=pairwise_df,
                      factors = filt_ileum_metadata$Group,
                      covariate = filt_ileum_metadata$Country, 
                      group1 = group1,
                      group2 = group2,
                      sim.method = 'jaccard')
  print(result_list)
}
}
```

**PCoA**

```{r, warning=FALSE}
p <- pca_plot_custom(filt_ileum_asv_tab,
                     filt_ileum_taxa_tab,
                     filt_ileum_metadata,
                     measure = "jaccard",
                     show_boxplots = TRUE,
                     variable = "Group", size=3, show_legend=FALSE)

# save the results
supplements_beta[[paste("PCoA jaccard",level,segment)]] <- p

# see the results
p
```

#### Saving results

```{r, warning=FALSE}
write.xlsx(supplements_beta[!grepl("PCoA",names(supplements_beta))],
           file = file.path(path,
           paste0("supplements_beta_diversity_", segment,".xlsx")))
```

## Univariate Analysis

## Machine learning

```{r, warning=FALSE}
path = "../results/Q2/models"
```

### ElasticNet

```{r, warning=FALSE}
model="enet"
```

#### Genus level

```{r, warning=FALSE}
level="genus"
```

Aggregate taxa

```{r, warning=FALSE}
genus_data <- aggregate_taxa(ileum_asv_tab,
                             ileum_taxa_tab,
                             taxonomic_level = level)

ileum_genus_tab <- genus_data[[1]]
ileum_genus_taxa_tab <- genus_data[[2]]
```

##### rPSC vs non-rPSC - all

```{r, warning=FALSE}
group <- c("rPSC","non-rPSC")
comparison_name <- paste0(group[1], " vs ",group[2])
```

```{r, warning=FALSE}
model_name <- paste(comparison_name,level,segment)

# prepare the data
filt_ileum_uni_data <- binomial_prep(ileum_genus_tab,
                                     ileum_genus_taxa_tab,
                                     ileum_metadata,
                                     group, 
                                     usage="ml_clr")

# fit the model
enet_model <- glmnet_binomial(filt_ileum_uni_data,
                              sample_method = "atypboot",
                              outcome="Group", 
                              N=500,
                              reuse=TRUE,
                              file=model_name,
                              Q="Q2")

# ROC
roc_c <- roc_curve(enet_model, group)

# save the results
models_summ <- list()
models_cm <- list()
roc_cs <- list()
betas <- list()

models_summ[[model_name]] <- enet_model$model_summary
models_cm[[model_name]] <- enet_model$conf_matrices$original
roc_cs[[model_name]] <- enet_model$kfold_rocobjs
betas[[model_name]] <- as.matrix(enet_model$betas)

# see the results
enet_model$model_summary %>% t()
enet_model$conf_matrices
enet_model$plot

roc_c
```

##### rPSC vs non-rPSC - subset

```{r, warning=FALSE}
group <- c("rPSC","non-rPSC")
comparison_name <- paste0(group[1], " vs ",group[2]," subset")
```

```{r, warning=FALSE}
taxa_to_test <- read.xlsx("../results/Q1/univariate_analysis/significant_taxa_terminal_ileum.xlsx", sheet=" genus healthy vs pre_ltx")

model_name <- paste(comparison_name,level,segment)

# prepare the data
filt_ileum_uni_data <- binomial_prep_psc_effect(ileum_genus_tab,
                                                ileum_genus_taxa_tab, 
                                                ileum_metadata,
                                                taxa_to_test,
                                                usage="ml_clr")

# fit the model
enet_model <- glmnet_binomial(filt_ileum_uni_data,
                              sample_method = "atypboot",
                              outcome="Group", 
                              N=500,
                              reuse=TRUE,
                              file=model_name,
                              Q="Q2")

# ROC
roc_c <- roc_curve(enet_model, group)

# save the results
models_summ[[model_name]] <- enet_model$model_summary
models_cm[[model_name]] <- enet_model$conf_matrices$original
roc_cs[[model_name]] <- enet_model$kfold_rocobjs
betas[[model_name]] <- as.matrix(enet_model$betas)

# see the results
enet_model$model_summary %>% t()
enet_model$conf_matrices
enet_model$plot

roc_c
```

#### Saving results

```{r, warning=FALSE}
models_summ_df_ileum <- do.call(rbind, 
  models_summ[grep(segment,names(models_summ),value = TRUE)])

write.csv(models_summ_df_ileum,file.path(path,paste0("elastic_net_",segment,".csv")))
```


### Supplementary models

```{r, warning=FALSE}
supplements_models <- list()
```

#### CLR-transformed data

##### kNN

```{r, warning=FALSE}
model="knn"
```

###### Genus level

```{r, warning=FALSE}
level="genus"
```

Aggregate taxa

```{r, warning=FALSE}
genus_data <- aggregate_taxa(ileum_asv_tab,
                             ileum_taxa_tab,
                             taxonomic_level = level)

ileum_genus_tab <- genus_data[[1]]
ileum_genus_taxa_tab <- genus_data[[2]]
```

**rPSC vs non-rPSC all**

```{r, warning=FALSE}
group <- c("rPSC","non-rPSC")
comparison_name <- paste0(group[1], " vs ",group[2])
```

```{r, warning=FALSE}
model_name <- paste(comparison_name,level,segment)

# prepare the data
filt_ileum_uni_data <- binomial_prep(ileum_genus_tab,
                                     ileum_genus_taxa_tab,
                                     ileum_metadata,
                                     group, 
                                     usage="ml_clr")

# fit the model
knn_model <- knn_binomial(filt_ileum_uni_data,
                              sample_method = "atypboot",
                              outcome="Group", 
                              N=500,
                              reuse=TRUE,
                              file=model_name,
                              Q="Q2")

# ROC curve
roc_c <- roc_curve(knn_model, group)

# save the results
supplements_models[["models_summ"]][["knn_model"]][[model_name]] <- knn_model$model_summary
supplements_models[["roc_cs"]][["knn_model"]][[model_name]] <- knn_model$kfold_rocobjs


# see the results
knn_model$model_summary %>% t()


roc_c
```

**rPSC vs non-rPSC - subset**

```{r, warning=FALSE}
group <- c("rPSC","non-rPSC")
comparison_name <- paste0(group[1], " vs ",group[2]," subset")
```

```{r, warning=FALSE}
taxa_to_test <- read.xlsx("../results/Q1/univariate_analysis/significant_taxa_terminal_ileum.xlsx", sheet=" genus healthy vs pre_ltx")

model_name <- paste(comparison_name,level,segment)

# prepare the data
filt_ileum_uni_data <- binomial_prep_psc_effect(ileum_genus_tab,
                                                ileum_genus_taxa_tab, 
                                                ileum_metadata,
                                                taxa_to_test,
                                                usage="ml_clr")

# fit the model
knn_model <- knn_binomial(filt_ileum_uni_data,
                              sample_method = "atypboot",
                              outcome="Group", 
                              N=500,
                              reuse=TRUE,
                              file=model_name,
                              Q="Q2")

# ROC
roc_c <- roc_curve(knn_model, group)

# save the results
supplements_models[["models_summ"]][["knn_model"]][[model_name]] <- knn_model$model_summary
supplements_models[["roc_cs"]][["knn_model"]][[model_name]] <- knn_model$kfold_rocobjs

# see the results
knn_model$model_summary %>% t()

roc_c
```

##### Random Forest

```{r, warning=FALSE}
model="rf"
```

###### Genus level

```{r, warning=FALSE}
level="genus"
```

Aggregate taxa

```{r, warning=FALSE}
genus_data <- aggregate_taxa(ileum_asv_tab,
                             ileum_taxa_tab,
                             taxonomic_level = level)

ileum_genus_tab <- genus_data[[1]]
ileum_genus_taxa_tab <- genus_data[[2]]
```

**rPSC vs non-rPSC** - all

```{r, warning=FALSE}
group <- c("rPSC","non-rPSC")
comparison_name <- paste0(group[1], " vs ",group[2])
```

```{r, warning=FALSE}
model_name <- paste(comparison_name,level,segment)

# prepare the data
filt_ileum_uni_data <- binomial_prep(ileum_genus_tab,
                                     ileum_genus_taxa_tab,
                                     ileum_metadata,
                                     group, 
                                     usage="ml_clr")

# fit the model
rf_model <- rf_binomial(filt_ileum_uni_data,
                              sample_method = "atypboot",
                              outcome="Group", 
                              N=500,
                              reuse=TRUE,
                              file=model_name,
                              Q="Q2")

# ROC curve
roc_c <- roc_curve(rf_model, group)

# save the results
supplements_models[["models_summ"]][["rf_model"]][[model_name]] <- rf_model$model_summary
supplements_models[["roc_cs"]][["rf_model"]][[model_name]] <- rf_model$kfold_rocobjs

# see the results
rf_model$model_summary %>% t()


roc_c
```

**rPSC vs non-rPSC** - subset

```{r, warning=FALSE}
group <- c("rPSC","non-rPSC")
comparison_name <- paste0(group[1], " vs ",group[2]," subset")
```

```{r, warning=FALSE}
taxa_to_test <- read.xlsx("../results/Q1/univariate_analysis/significant_taxa_terminal_ileum.xlsx", sheet=" genus healthy vs pre_ltx")

model_name <- paste(comparison_name,level,segment)

# prepare the data
filt_ileum_uni_data <- binomial_prep_psc_effect(ileum_genus_tab,
                                                ileum_genus_taxa_tab, 
                                                ileum_metadata,
                                                taxa_to_test,
                                                usage="ml_clr")

# fit the model
rf_model <- rf_binomial(filt_ileum_uni_data,
                              sample_method = "atypboot",
                              outcome="Group", 
                              N=500,
                              reuse=TRUE,
                              file=model_name,
                              Q="Q2")

# ROC curve
roc_c <- roc_curve(rf_model, group)

# save the results
supplements_models[["models_summ"]][["rf_model"]][[model_name]] <- rf_model$model_summary
supplements_models[["roc_cs"]][["rf_model"]][[model_name]] <- rf_model$kfold_rocobjs

# see the results
rf_model$model_summary %>% t()


roc_c
```

##### Gradient boosting

```{r, warning=FALSE}
model="gb"
```

###### Genus level

```{r, warning=FALSE}
level="genus"
```

Aggregate taxa

```{r, warning=FALSE}
genus_data <- aggregate_taxa(ileum_asv_tab,
                             ileum_taxa_tab,
                             taxonomic_level = level)

ileum_genus_tab <- genus_data[[1]]
ileum_genus_taxa_tab <- genus_data[[2]]
```

**rPSC vs non-rPSC** - all

```{r, warning=FALSE}
group <- c("rPSC","non-rPSC")
comparison_name <- paste0(group[1], " vs ",group[2])
```

```{r, warning=FALSE}
model_name <- paste(comparison_name,level,segment)

# prepare the data
filt_ileum_uni_data <- binomial_prep(ileum_genus_tab,
                                     ileum_genus_taxa_tab,
                                     ileum_metadata,
                                     group, 
                                     usage="ml_clr")

# fit the model
gbm_model <- gbm_binomial(filt_ileum_uni_data,
                              sample_method = "atypboot",
                              outcome="Group", 
                              N=500,
                              reuse=TRUE,
                              file=model_name,
                              Q="Q2")

# ROC curve
roc_c <- roc_curve(gbm_model, group)

# save the results
supplements_models[["models_summ"]][["gbm_model"]][[model_name]] <- gbm_model$model_summary
supplements_models[["roc_cs"]][["gbm_model"]][[model_name]] <- gbm_model$kfold_rocobjs

# see the results
gbm_model$model_summary %>% t()

roc_c
```

**rPSC vs non-rPSC** - subset

```{r, warning=FALSE}
group <- c("rPSC","non-rPSC")
comparison_name <- paste0(group[1], " vs ",group[2]," subset")
```

```{r, warning=FALSE}
taxa_to_test <- read.xlsx("../results/Q1/univariate_analysis/significant_taxa_terminal_ileum.xlsx", sheet=" genus healthy vs pre_ltx")

model_name <- paste(comparison_name,level,segment)

# prepare the data
filt_ileum_uni_data <- binomial_prep_psc_effect(ileum_genus_tab,
                                                ileum_genus_taxa_tab, 
                                                ileum_metadata,
                                                taxa_to_test,
                                                usage="ml_clr")

# fit the model
gbm_model <- gbm_binomial(filt_ileum_uni_data,
                              sample_method = "atypboot",
                              outcome="Group", 
                              N=500,
                              reuse=TRUE,
                              file=model_name,
                              Q="Q2")

# ROC curve
roc_c <- roc_curve(gbm_model, group)

# save the results
supplements_models[["models_summ"]][["gbm_model"]][[model_name]] <- gbm_model$model_summary
supplements_models[["roc_cs"]][["gbm_model"]][[model_name]] <- gbm_model$kfold_rocobjs

# see the results
gbm_model$model_summary %>% t()

roc_c
```


#### Saving results

```{r, warning=FALSE}
models_list <- list()

for (model_name in names(supplements_models$models_summ)){
  df <- do.call(rbind, supplements_models$models_summ[[model_name]])
  models_list[[model_name]] <- df
}

write.xlsx(models_list,
           file=file.path(path,paste0("supplements_models_",segment,".xlsx")),
           rowNames=TRUE)

```


## Results overview

#### Alpha diversity

**Models**

```{r,warning=FALSE}
knitr::kable(pc_observed[[segment]],
             digits = 3,
             caption = "Results of linear model testing ASV Richness")

knitr::kable(pc_shannon[[segment]],
             digits = 3,
             caption = "Results of linear model testing Shannon index")

knitr::kable(pc_simpson[[segment]],
             digits = 3,
             caption = "Results of linear model testing Simpson index")

knitr::kable(pc_pielou[[segment]],
             digits = 3,
             caption = "Results of linear model testing Pielou index")
```

**Plots**

```{r, warning=FALSE, fig.width=6, fig.height=4, fig.fullwidth=TRUE}
alpha_div_plots[[paste(segment,"Country")]]
```

#### Beta diversity

**Main results**

```{r,warning=FALSE}
knitr::kable(pairwise_aitchison_raw[[paste("genus", segment)]],
             digits = 3,
             caption = "Results of PERMANOVA - robust aitchison distance")
```

**PCA**

```{r, warning=FALSE,  fig.width=5, fig.height=3, fig.fullwidth=TRUE, eval=TRUE}
pca_plots_list[[paste(segment,"genus custom")]]
```

**Supplements**

```{r, warning=FALSE}
knitr::kable(supplements_beta[!grepl("PCoA",names(supplements_beta)) & (grepl("genus",names(supplements_beta)))],
             digits = 3,
             caption = "Supplementary PERMANOVA results: Bray-curtis, Jaccard distances")
```

**Supplements PCA**

```{r, warning=FALSE,  fig.width=8, fig.height=7, fig.fullwidth=TRUE, eval=TRUE}
ggarrange(plotlist = supplements_beta[grepl("PCoA",names(supplements_beta))],
          labels=names(supplements_beta[grepl("PCoA",names(supplements_beta))]),
          font.label = list(size=5,face="plain"),
          ncol=2,nrow=3)
```

#### Univariate analysis

**Number of significant taxa**

```{r, warning=FALSE}
knitr::kable(cbind(as.data.frame(lapply(list_intersections,nrow)),
      as.data.frame(lapply(rpsc_effect,nrow))) %>% t() %>% 
  `colnames<-`("Count") %>% 
  `rownames<-`(c(names(list_intersections),"PSC effect Genus")),caption="Number of significant taxa")
```

#### Machine learning

**Main models**

```{r, warning=FALSE}
knitr::kable(models_summ_df_ileum %>% dplyr::select(
"alpha","lambda",
"auc_optimism_corrected",
"auc_optimism_corrected_CIL",
"auc_optimism_corrected_CIU"),
             digits=2,caption="Elastic net results")
```

**ROC - Genus level**

```{r, warning=FALSE,  warning=FALSE}
roc_curve_all_custom(roc_cs[c(1:2)],Q="Q2",
                     model_name="enet_model",legend = FALSE)
```

**Supplementary models**


```{r, warning=FALSE}
# Build final dataframe
models_list[["enet_model"]] <- models_summ_df_ileum
final_df <- tibble(row_names = rownames(models_list[[1]]))

# Loop through models and extract required values
for (model_name in names(models_list)) {
  model_df <- models_list[[model_name]]
  
  # Combine AUC_optimism_corrected with its CI values
  final_df[[model_name]] <- paste0(
    round(model_df$auc_optimism_corrected, 3), 
    " (", round(model_df$auc_optimism_corrected_CIL, 3), "; ", 
    round(model_df$auc_optimism_corrected_CIU, 3), ")"
  )
}

knitr::kable(final_df, caption="All models")
```

**Supplementary ROC - genus**

```{r, warning=FALSE,  fig.width=10, fig.height=8, fig.fullwidth=TRUE, warning=FALSE}
rocs_list <- supplements_models$roc_cs
rocs_list[["enet_model"]] <- roc_cs
plot_list <- list()

for (model_name in names(rocs_list)) {
  plot_list[[model_name]] <- roc_curve_all_custom(rocs_list[[model_name]][c(1:2)],
                       Q="Q2",
                       model_name=model_name,legend = FALSE)
}

p <- ggarrange(plotlist = plot_list,labels = names(rocs_list),font.label = list(face="plain",size=7))
p
```

```{r, warning=FALSE, results='hide', echo=FALSE}
pdf("../figures/Q2/models_ileum.pdf",
    height =10,width = 10)
p
dev.off()
```

# Data Analysis - Right colon

```{r, warning=FALSE}
segment="right_colon"
```

## Filtering

Rules: 
  - sequencing depth \> 10000 
  
  - nearZeroVar() with default settings 
  
**Rarefaction Curve**

```{r, warning=FALSE}
path="../intermediate_files/rarecurves"
seq_depth_threshold <- 10000
```

```{r, warning=FALSE,  eval = FALSE}
ps <- construct_phyloseq(rcolon_asv_tab,rcolon_taxa_tab,rcolon_metadata)
rareres <- get_rarecurve(obj=ps, chunks=500)
save(rareres,file = file.path(path,"rarefaction_colon.Rdata"))
```

```{r, warning=FALSE}
load(file.path(path,"rarefaction_colon.Rdata"))
seq_depth_threshold <- 10000
prare <- ggrarecurve(obj=rareres,
                      factorNames="Country",
                      indexNames=c("Observe")) + 
        theme_bw()+
        theme(axis.text=element_text(size=8), 
              panel.grid=element_blank(),
              strip.background = element_rect(colour=NA,fill="grey"),
              strip.text.x = element_text(face="bold")) + 
        geom_vline(xintercept = seq_depth_threshold, 
                   linetype="dashed", color = "red") + 
        xlim(0, 20000)

prare
```

Library size

```{r, warning=FALSE,  fig.width=5, fig.height=4, fig.fullwidth=TRUE}
read_counts(rcolon_asv_tab, line = c(5000,10000))
```

### Sequencing depth

```{r, warning=FALSE}
data_filt <- seq_depth_filtering(rcolon_asv_tab,
                                 rcolon_taxa_tab,
                                 rcolon_metadata,
                                 seq_depth_threshold = 10000)

filt_rcolon_asv_tab <- data_filt[[1]]; alpha_rcolon_asv_tab <- filt_rcolon_asv_tab
filt_rcolon_taxa_tab <- data_filt[[2]]; alpha_rcolon_taxa_tab <- filt_rcolon_taxa_tab
filt_rcolon_metadata <- data_filt[[3]]; alpha_rcolon_metadata <- filt_rcolon_metadata

seq_step <- dim(filt_rcolon_asv_tab)[1]
```

Library size

```{r, warning=FALSE,  fig.width=5, fig.height=4, fig.fullwidth=TRUE}
read_counts(filt_rcolon_asv_tab,line = c(10000))
```

### NearZeroVar

```{r, warning=FALSE}
data_filt <- nearzerovar_filtering(filt_rcolon_asv_tab,
                                   filt_rcolon_taxa_tab,
                                   filt_rcolon_metadata)

filt_rcolon_asv_tab <- data_filt[[1]]
filt_rcolon_taxa_tab <- data_filt[[2]]
nearzero_step <- dim(filt_rcolon_asv_tab)[1]
```

Library size

```{r, warning=FALSE,  fig.width=5, fig.height=4, fig.fullwidth=TRUE}
read_counts(filt_rcolon_asv_tab,line = c(5000,10000))
```

Check zero depth

```{r, warning=FALSE}
data_filt <- check_zero_depth(filt_rcolon_asv_tab, 
                              filt_rcolon_taxa_tab, 
                              filt_rcolon_metadata)

filt_rcolon_asv_tab <- data_filt[[1]]; 
filt_rcolon_taxa_tab <- data_filt[[2]]; 
filt_rcolon_metadata <- data_filt[[3]]; 
```

Library size

```{r, warning=FALSE,  fig.width=5, fig.height=4, fig.fullwidth=TRUE}
read_counts(filt_rcolon_asv_tab,line = c(5000,10000))
```

### Final Counts

```{r, warning=FALSE}
final_counts_filtering(rcolon_asv_tab,
                       filt_rcolon_asv_tab,
                       filt_rcolon_metadata,
                       seq_step, 0, nearzero_step)
```


## Beta diversity

Calculating Aitchison distance (euclidean distance on clr-transformed
data), both at ASV and genus level.

### Supplementary analysis

#### Genus level

```{r, warning=FALSE}
level="genus"
```

##### Bray-Curtis

**PERMANOVA**

```{r, warning=FALSE}
pairwise_df <- filt_rcolon_genus_tab %>% column_to_rownames("SeqID") %>% t()

# main effect
pp_main <- pairwise.adonis(pairwise_df,
                           filt_rcolon_genus_metadata$Group,
                           covariate = filt_rcolon_genus_metadata$Country, 
                           patients = filt_rcolon_genus_metadata$Patient,
                           sim.method = "bray", p.adjust.m="BH")

# interaction
pp_int <- pairwise.adonis(pairwise_df,
                          filt_rcolon_genus_metadata$Group,
                          covariate = filt_rcolon_genus_metadata$Country, 
                          patients = filt_rcolon_genus_metadata$Patient,
                          interaction = TRUE, sim.method = "bray", p.adjust.m="BH")

# tidy the results
pp_factor <- pp_main[[1]]
pp_cov <- pp_main[[2]]
pp_fac.cov <- pp_int[[3]]

cols <- c("pairs","Df","SumsOfSqs", "F.Model","R2","p.value", "p.adjusted", "sig")
colnames(pp_factor) <- cols; colnames(pp_cov) <- cols; colnames(pp_fac.cov) <- cols; 

# save raw results
supplements_beta[[paste("bray",level,segment)]] <- rbind(pp_factor,pp_cov,pp_fac.cov)
```

```{r, warning=FALSE}
# see the results
knitr::kable(pp_factor,digits = 3,caption = "PERMANOVA, GROUP separation")
knitr::kable(pp_cov,digits = 3,caption = "PERMANOVA, COUNTRY separation")
knitr::kable(pp_fac.cov,digits = 3,caption = "PERMANOVA, INTERACTION GROUP:Country")
```

Interaction check

```{r, warning=FALSE}
interaction_sig <- pp_fac.cov$pairs[pp_fac.cov$p.adjusted < 0.05]

for (i in 1:length(interaction_sig)){
  group1 <- unlist(strsplit(interaction_sig[i],split = " vs "))[1]
  group2 <- unlist(strsplit(interaction_sig[i],split = " vs "))[2]
  group2 <- unlist(strsplit(group2,split = " : "))[1]
  
  result_list <- adonis_postanalysis(x=pairwise_df,
                      factors = filt_rcolon_genus_metadata$Group,
                      covariate = filt_rcolon_genus_metadata$Country, 
                      patients = filt_rcolon_genus_metadata$Patient,
                      group1 = group1,
                      group2 = group2,
                      sim.method = 'bray')
  print(result_list)
}

```

**Plots**

```{r, warning=FALSE}
p <- pca_plot_custom(filt_rcolon_genus_tab,
                                 filt_rcolon_genus_taxa,
                                 filt_rcolon_genus_metadata,
                                 measure = "bray",
                                 show_boxplots = TRUE,
                                 variable = "Group", size=3, show_legend=FALSE)

# save the results
supplements_beta[[paste("PCoA bray",level,segment)]] <- p

# see the results
p
```

##### Jaccard

**PERMANOVA**

```{r, warning=FALSE}
pairwise_df <- filt_rcolon_genus_tab %>% column_to_rownames("SeqID") %>% t()

# main effect
pp_main <- pairwise.adonis(pairwise_df,
                           filt_rcolon_genus_metadata$Group,
                           covariate = filt_rcolon_genus_metadata$Country,
                           patients = filt_rcolon_genus_metadata$Patient,
                           sim.method = "jaccard", p.adjust.m="BH")

# interaction
pp_int <- pairwise.adonis(pairwise_df,
                          filt_rcolon_genus_metadata$Group,
                          covariate = filt_rcolon_genus_metadata$Country,
                          patients = filt_rcolon_genus_metadata$Patient,
                          interaction = TRUE, sim.method = "jaccard", p.adjust.m="BH")

# tidy the results
pp_factor <- pp_main[[1]]
pp_cov <- pp_main[[2]]
pp_fac.cov <- pp_int[[3]]

cols <- c("pairs","Df","SumsOfSqs", "F.Model","R2","p.value", "p.adjusted", "sig")
colnames(pp_factor) <- cols; colnames(pp_cov) <- cols; colnames(pp_fac.cov) <- cols; 

# save raw results
supplements_beta[[paste("jaccard",level,segment)]] <- rbind(pp_factor, 
                                                            pp_cov, 
                                                            pp_fac.cov)
```

```{r, warning=FALSE}
# see the results
knitr::kable(pp_factor,digits = 3,caption = "PERMANOVA, GROUP separation")
knitr::kable(pp_cov,digits = 3,caption = "PERMANOVA, COUNTRY separation")
knitr::kable(pp_fac.cov,digits = 3,caption = "PERMANOVA, INTERACTION GROUP:Country")
```

Interaction check

```{r, warning=FALSE}
interaction_sig <- pp_fac.cov$pairs[pp_fac.cov$p.adjusted < 0.05]

for (i in 1:length(interaction_sig)){
  group1 <- unlist(strsplit(interaction_sig[i],split = " vs "))[1]
  group2 <- unlist(strsplit(interaction_sig[i],split = " vs "))[2]
  group2 <- unlist(strsplit(group2,split = " : "))[1]
  
  result_list <- adonis_postanalysis(x=pairwise_df,
                      factors = filt_rcolon_genus_metadata$Group,
                      covariate = filt_rcolon_genus_metadata$Country, 
                      patients = filt_rcolon_genus_metadata$Patient,
                      group1 = group1,
                      group2 = group2,
                      sim.method = 'jaccard')
  print(result_list)
}

```

**Plots**

```{r, warning=FALSE}
p <- pca_plot_custom(filt_rcolon_genus_tab,
                                 filt_rcolon_genus_taxa,
                                 filt_rcolon_genus_metadata,
                                 measure = "jaccard",
                                 show_boxplots = TRUE,
                                 variable = "Group", size=3, show_legend=FALSE)

# save the results
supplements_beta[[paste("PCoA jaccard",level,segment)]] <- p

# see the results
p
```

#### ASV level

```{r, warning=FALSE}
level="ASV"
```

##### Aitchison

```{r, warning=FALSE}
# preparing data frame
pairwise_df <- filt_rcolon_asv_tab %>% column_to_rownames("SeqID") %>% t()

# main effect
pp_main <- pairwise.adonis(x=pairwise_df,
                          filt_rcolon_metadata$Group,
                           covariate = filt_rcolon_metadata$Country, 
                           sim.method = "robust.aitchison", 
                           p.adjust.m="BH",
                           patients = filt_rcolon_metadata$Patient)

# interaction
pp_int <- pairwise.adonis(pairwise_df,filt_rcolon_metadata$Group,
                          covariate = filt_rcolon_metadata$Country, 
                          interaction = TRUE, 
                          sim.method = "robust.aitchison", 
                          p.adjust.m="BH",
                          patients = filt_rcolon_metadata$Patient)

pp_factor <- pp_main[[1]]
pp_cov <- pp_main[[2]]
pp_fac.cov <- pp_int[[3]]

cols <- c("pairs","Df","SumsOfSqs", "F.Model","R2","p.value", "p.adjusted", "sig")
colnames(pp_factor) <- cols; colnames(pp_cov) <- cols; colnames(pp_fac.cov) <- cols; 

# save raw results
supplements_beta[[paste("aitchison",level,segment)]] <- rbind(pp_factor, 
                                                            pp_cov, 
                                                            pp_fac.cov)

# see the results
pp_factor
pp_cov
pp_fac.cov

```

```{r, warning=FALSE}
# see the results
knitr::kable(pp_factor,digits = 3,caption = "PERMANOVA, GROUP separation")
knitr::kable(pp_cov,digits = 3,caption = "PERMANOVA, COUNTRY separation")
knitr::kable(pp_fac.cov,digits = 3,caption = "PERMANOVA, INTERACTION GROUP:Country")
```

Interaction check

```{r, warning=FALSE}
interaction_sig <- pp_fac.cov$pairs[pp_fac.cov$p.adjusted < 0.05]

if (length(interaction_sig)>0){
 for (i in 1:length(interaction_sig)){
  group1 <- unlist(strsplit(interaction_sig[i],split = " vs "))[1]
  group2 <- unlist(strsplit(interaction_sig[i],split = " vs "))[2]
  group2 <- unlist(strsplit(group2,split = " : "))[1]
  
  result_list <- adonis_postanalysis(x=pairwise_df,
                      factors = filt_rcolon_metadata$Group,
                      covariate = filt_rcolon_metadata$Country, 
                      group1 = group1,
                      group2 = group2,
                      patients = filt_rcolon_metadata$Patient)
  print(result_list)
} 
}
```

**PCoA**

```{r, warning=FALSE}
p <- pca_plot_custom(filt_rcolon_asv_tab,
                           filt_rcolon_taxa_tab,
                           filt_rcolon_metadata,
                           show_boxplots = TRUE,
                           variable = "Group", 
                           size=3, 
                           show_legend=FALSE)

# save the results
supplements_beta[[paste("PCoA aitchison",level,segment)]] <- p

# see the results
p
```

##### Bray-Curtis 

**PERMANOVA**

```{r, warning=FALSE}
# preparing data frame
pairwise_df <- filt_rcolon_asv_tab %>% column_to_rownames("SeqID") %>% t()

# main effect
pp_main <- pairwise.adonis(pairwise_df,
                           filt_rcolon_metadata$Group,
                           covariate = filt_rcolon_metadata$Country,
                           patients = filt_rcolon_metadata$Patient,
                           sim.method = "bray", p.adjust.m="BH")

# interaction
pp_int <- pairwise.adonis(pairwise_df,
                          filt_rcolon_metadata$Group,
                          covariate = filt_rcolon_metadata$Country, 
                          patients = filt_rcolon_metadata$Patient,
                          interaction = TRUE, sim.method = "bray", p.adjust.m="BH")

pp_factor <- pp_main[[1]]
pp_cov <- pp_main[[2]]
pp_fac.cov <- pp_int[[3]]

cols <- c("pairs","Df","SumsOfSqs", "F.Model","R2","p.value", "p.adjusted", "sig")
colnames(pp_factor) <- cols; colnames(pp_cov) <- cols; colnames(pp_fac.cov) <- cols; 

# save raw results
supplements_beta[[paste("bray",level,segment)]] <- rbind(pp_factor, 
                                                            pp_cov, 
                                                            pp_fac.cov)

```

```{r, warning=FALSE}
# see the results
knitr::kable(pp_factor,digits = 3,caption = "PERMANOVA, GROUP separation")
knitr::kable(pp_cov,digits = 3,caption = "PERMANOVA, COUNTRY separation")
knitr::kable(pp_fac.cov,digits = 3,caption = "PERMANOVA, INTERACTION GROUP:Country")
```

Interaction check

```{r, warning=FALSE}
interaction_sig <- pp_fac.cov$pairs[pp_fac.cov$p.adjusted < 0.05]

for (i in 1:length(interaction_sig)){
  group1 <- unlist(strsplit(interaction_sig[i],split = " vs "))[1]
  group2 <- unlist(strsplit(interaction_sig[i],split = " vs "))[2]
  group2 <- unlist(strsplit(group2,split = " : "))[1]
  
  result_list <- adonis_postanalysis(x=pairwise_df,
                      factors = filt_rcolon_metadata$Group,
                      covariate = filt_rcolon_metadata$Country, 
                      group1 = group1,
                      group2 = group2,
                      patients = filt_rcolon_metadata$Patient,
                      sim.method = 'bray')
  print(result_list)
}

```

**PCoA**

```{r, warning=FALSE}
p <- pca_plot_custom(filt_rcolon_asv_tab,
                     filt_rcolon_taxa_tab,
                     filt_rcolon_metadata,
                     measure = "bray",
                     show_boxplots = TRUE,
                     variable = "Group", size=3, show_legend=FALSE)

# save the results
supplements_beta[[paste("PCoA bray",level,segment)]] <- p

# see the results
p
```

##### Jaccard

**PERMANOVA**

```{r, warning=FALSE}
# preparing data frame
pairwise_df <- filt_rcolon_asv_tab %>% column_to_rownames("SeqID") %>% t()

# main effect
pp_main <- pairwise.adonis(pairwise_df,
                           filt_rcolon_metadata$Group,
                           covariate = filt_rcolon_metadata$Country,
                           patients = filt_rcolon_metadata$Patient,
                           sim.method = "jaccard", p.adjust.m="BH")

# interaction
pp_int <- pairwise.adonis(pairwise_df,
                          filt_rcolon_metadata$Group,
                          covariate = filt_rcolon_metadata$Country, 
                          patients = filt_rcolon_metadata$Patient,
                          interaction = TRUE, sim.method = "jaccard", p.adjust.m="BH")

pp_factor <- pp_main[[1]]
pp_cov <- pp_main[[2]]
pp_fac.cov <- pp_int[[3]]

cols <- c("pairs","Df","SumsOfSqs", "F.Model","R2","p.value", "p.adjusted", "sig")
colnames(pp_factor) <- cols; colnames(pp_cov) <- cols; colnames(pp_fac.cov) <- cols; 

# save raw results
supplements_beta[[paste("jaccard",level,segment)]] <- rbind(pp_factor, 
                                                            pp_cov, 
                                                            pp_fac.cov)
```

```{r, warning=FALSE}
# see the results
knitr::kable(pp_factor,digits = 3,caption = "PERMANOVA, GROUP separation")
knitr::kable(pp_cov,digits = 3,caption = "PERMANOVA, COUNTRY separation")
knitr::kable(pp_fac.cov,digits = 3,caption = "PERMANOVA, INTERACTION GROUP:Country")
```

Interaction check

```{r, warning=FALSE}
interaction_sig <- pp_fac.cov$pairs[pp_fac.cov$p.adjusted < 0.05]

for (i in 1:length(interaction_sig)){
  group1 <- unlist(strsplit(interaction_sig[i],split = " vs "))[1]
  group2 <- unlist(strsplit(interaction_sig[i],split = " vs "))[2]
  group2 <- unlist(strsplit(group2,split = " : "))[1]
  
  result_list <- adonis_postanalysis(x=pairwise_df,
                      factors = filt_rcolon_metadata$Group,
                      covariate = filt_rcolon_metadata$Country, 
                      patients = filt_rcolon_metadata$Patient,
                      group1 = group1,
                      group2 = group2,
                      sim.method = 'jaccard')
  print(result_list)
}

```

**PCoA**

```{r, warning=FALSE}
p <- pca_plot_custom(filt_rcolon_asv_tab,
                     filt_rcolon_taxa_tab,
                     filt_rcolon_metadata,
                     measure = "jaccard",
                     show_boxplots = TRUE,
                     variable = "Group", size=3, show_legend=FALSE)

# save the results
supplements_beta[[paste("PCoA jaccard",level,segment)]] <- p

# see the results
p
```

#### Saving results

```{r, warning=FALSE}
write.xlsx(supplements_beta[!grepl("PCoA",names(supplements_beta))],
           file = file.path(path,
           paste0("supplements_beta_diversity_", segment,".xlsx")))
```

## Univariate Analysis

## Machine learning

```{r, warning=FALSE}
path = "../results/Q2/models"
```

### ElasticNet

```{r, warning=FALSE}
model="enet"
```

#### Genus level

```{r, warning=FALSE}
level="genus"
```

Aggregate taxa

```{r, warning=FALSE}
genus_data <- aggregate_taxa(rcolon_asv_tab,
                             rcolon_taxa_tab,
                             taxonomic_level = level)

rcolon_genus_tab <- genus_data[[1]]
rcolon_genus_taxa_tab <- genus_data[[2]]
```

##### rPSC vs non-rPSC - all

```{r, warning=FALSE}
group <- c("rPSC","non-rPSC")
comparison_name <- paste0(group[1], " vs ",group[2])
```

```{r, warning=FALSE}
model_name <- paste(comparison_name,level,segment)

# prepare the data
filt_rcolon_uni_data <- binomial_prep(rcolon_genus_tab,
                                     rcolon_genus_taxa_tab,
                                     rcolon_metadata,
                                     group, 
                                     usage="ml_clr")

# fit the model
enet_model <- glmnet_binomial(filt_rcolon_uni_data,
                              sample_method = "atypboot",
                              outcome="Group", 
                              N=10,
                              reuse=FALSE,
                              file=model_name,
                              Q="Q2")

# ROC
roc_c <- roc_curve(enet_model, group)

# save the results
models_summ[[model_name]] <- enet_model$model_summary
models_cm[[model_name]] <- enet_model$conf_matrices$original
roc_cs[[model_name]] <- enet_model$kfold_rocobjs
betas[[model_name]] <- as.matrix(enet_model$betas)

# see the results
enet_model$model_summary %>% t()
enet_model$conf_matrices
enet_model$plot

roc_c
```

##### rPSC vs non-rPSC - subset

```{r, warning=FALSE}
group <- c("rPSC","non-rPSC")
comparison_name <- paste0(group[1], " vs ",group[2]," subset")
```

```{r, warning=FALSE}
taxa_to_test <- read.xlsx("../results/Q1/univariate_analysis/significant_taxa_right_colon.xlsx", sheet=" genus healthy vs pre_ltx")

model_name <- paste(comparison_name,level,segment)

# prepare the data
filt_rcolon_uni_data <- binomial_prep_psc_effect(rcolon_genus_tab,
                                                rcolon_genus_taxa_tab, 
                                                rcolon_metadata,
                                                taxa_to_test,
                                                usage="ml_clr")

# fit the model
enet_model <- glmnet_binomial(filt_rcolon_uni_data,
                              sample_method = "atypboot",
                              outcome="Group", 
                              N=10,
                              reuse=FALSE,
                              file=model_name,
                              Q="Q2")

# ROC
roc_c <- roc_curve(enet_model, group)

# save the results
models_summ[[model_name]] <- enet_model$model_summary
models_cm[[model_name]] <- enet_model$conf_matrices$original
roc_cs[[model_name]] <- enet_model$kfold_rocobjs
betas[[model_name]] <- as.matrix(enet_model$betas)

# see the results
enet_model$model_summary %>% t()
enet_model$conf_matrices
enet_model$plot

roc_c
```

#### Saving results

```{r, warning=FALSE}
models_summ_df_colon <- do.call(rbind, 
  models_summ[grep(segment,names(models_summ),value = TRUE)])

write.csv(models_summ_df_colon,file.path(path,paste0("elastic_net_",segment,".csv")))
```

### Supplementary models

#### CLR-transformed data

##### kNN

```{r, warning=FALSE}
model="knn"
```

###### Genus level

```{r, warning=FALSE}
level="genus"
```

Aggregate taxa

```{r, warning=FALSE}
genus_data <- aggregate_taxa(rcolon_asv_tab,
                             rcolon_taxa_tab,
                             taxonomic_level = level)

rcolon_genus_tab <- genus_data[[1]]
rcolon_genus_taxa_tab <- genus_data[[2]]
```

**rPSC vs non-rPSC - all**

```{r, warning=FALSE}
group <- c("rPSC","non-rPSC")
comparison_name <- paste0(group[1], " vs ",group[2])
```

```{r, warning=FALSE}
model_name <- paste(comparison_name,level,segment)

# prepare the data
filt_rcolon_uni_data <- binomial_prep(rcolon_genus_tab,
                                     rcolon_genus_taxa_tab,
                                     rcolon_metadata,
                                     group, 
                                     usage="ml_clr",
                                     patient = TRUE)

# fit the model
knn_model <- knn_binomial(filt_rcolon_uni_data,
                              sample_method = "atypboot",
                              outcome="Group", 
                              N=500,
                          clust_var="Patient",
                              reuse=TRUE,
                              file=model_name,
                              Q="Q2")

# ROC curve
roc_c <- roc_curve(knn_model, group)

# save the results
supplements_models[["models_summ"]][["knn_model"]][[model_name]] <- knn_model$model_summary
supplements_models[["roc_cs"]][["knn_model"]][[model_name]] <- knn_model$kfold_rocobjs


# see the results
knn_model$model_summary %>% t()


roc_c
```

**rPSC vs non-rPSC - subset**

```{r, warning=FALSE}
group <- c("rPSC","non-rPSC")
comparison_name <- paste0(group[1], " vs ",group[2]," subset")
```

```{r, warning=FALSE}
taxa_to_test <- read.xlsx("../results/Q1/univariate_analysis/significant_taxa_colon.xlsx", sheet=" genus healthy vs pre_ltx")

model_name <- paste(comparison_name,level,segment)

# prepare the data
filt_ileum_uni_data <- binomial_prep_psc_effect(colon_genus_tab,
                                                rcolon_genus_taxa_tab, 
                                                rcolon_metadata,
                                                taxa_to_test,
                                                usage="ml_clr",
                                                patient = TRUE)

# fit the model
knn_model <- knn_binomial(filt_ileum_uni_data,
                              sample_method = "atypboot",
                              outcome="Group", 
                              N=500,
                              reuse=TRUE,
                              file=model_name,
                              Q="Q2",clust_var="Patient")

# ROC
roc_c <- roc_curve(knn_model, group)

# save the results
supplements_models[["models_summ"]][["knn_model"]][[model_name]] <- knn_model$model_summary
supplements_models[["roc_cs"]][["knn_model"]][[model_name]] <- knn_model$kfold_rocobjs

# see the results
knn_model$model_summary %>% t()

roc_c
```


##### Random Forest

```{r, warning=FALSE}
model="rf"
```

###### Genus level

```{r, warning=FALSE}
level="genus"
```

Aggregate taxa

```{r, warning=FALSE}
genus_data <- aggregate_taxa(rcolon_asv_tab,
                             rcolon_taxa_tab,
                             taxonomic_level = level)

rcolon_genus_tab <- genus_data[[1]]
rcolon_genus_taxa_tab <- genus_data[[2]]
```

**rPSC vs non-rPSC - all**

```{r, warning=FALSE}
group <- c("rPSC","non-rPSC")
comparison_name <- paste0(group[1], " vs ",group[2])
```

```{r, warning=FALSE}
model_name <- paste(comparison_name,level,segment)

# prepare the data
filt_rcolon_uni_data <- binomial_prep(rcolon_genus_tab,
                                     rcolon_genus_taxa_tab,
                                     rcolon_metadata,
                                     group, 
                                     usage="ml_clr",
                                     patient = TRUE)

# fit the model
rf_model <- rf_binomial(filt_rcolon_uni_data,
                              sample_method = "atypboot",
                              outcome="Group", 
                              N=500,
                        clust_var="Patient",
                              reuse=TRUE,
                              file=model_name,
                              Q="Q2")

# ROC curve
roc_c <- roc_curve(rf_model, group)

# save the results
supplements_models[["models_summ"]][["rf_model"]][[model_name]] <- rf_model$model_summary
supplements_models[["roc_cs"]][["rf_model"]][[model_name]] <- rf_model$kfold_rocobjs

# see the results
rf_model$model_summary %>% t()

roc_c
```

**rPSC vs non-rPSC - subset**

```{r, warning=FALSE}
group <- c("rPSC","non-rPSC")
comparison_name <- paste0(group[1], " vs ",group[2]," subset")
```

```{r, warning=FALSE}
taxa_to_test <- read.xlsx("../results/Q1/univariate_analysis/significant_taxa_colon.xlsx", sheet=" genus healthy vs pre_ltx")

model_name <- paste(comparison_name,level,segment)

# prepare the data
filt_ileum_uni_data <- binomial_prep_psc_effect(rcolon_genus_tab,
                                                rcolon_genus_taxa_tab, 
                                                rcolon_metadata,
                                                taxa_to_test,
                                                usage="ml_clr",
                                                patient = TRUE)

# fit the model
rf_model <- rf_binomial(filt_ileum_uni_data,
                              sample_method = "atypboot",
                              outcome="Group", 
                              N=500,
                              reuse=TRUE,
                              file=model_name,
                              Q="Q2",
                        clust_var="Patient")

# ROC curve
roc_c <- roc_curve(rf_model, group)

# save the results
supplements_models[["models_summ"]][["rf_model"]][[model_name]] <- rf_model$model_summary
supplements_models[["roc_cs"]][["rf_model"]][[model_name]] <- rf_model$kfold_rocobjs

# see the results
rf_model$model_summary %>% t()

roc_c
```


##### Gradient boosting

```{r, warning=FALSE}
model="gb"
```

###### Genus level

```{r, warning=FALSE}
level="genus"
```

Aggregate taxa

```{r, warning=FALSE}
genus_data <- aggregate_taxa(rcolon_asv_tab,
                             rcolon_taxa_tab,
                             taxonomic_level = level)

rcolon_genus_tab <- genus_data[[1]]
rcolon_genus_taxa_tab <- genus_data[[2]]
```

**rPSC vs non-rPSC - all**

```{r, warning=FALSE}
group <- c("rPSC","non-rPSC")
comparison_name <- paste0(group[1], " vs ",group[2])
```

```{r, warning=FALSE}
model_name <- paste(comparison_name,level,segment)

# prepare the data
filt_rcolon_uni_data <- binomial_prep(rcolon_genus_tab,
                                     rcolon_genus_taxa_tab,
                                     rcolon_metadata,
                                     group, 
                                     usage="ml_clr",
                                     patient = TRUE)

# fit the model
gbm_model <- gbm_binomial(filt_rcolon_uni_data,
                              sample_method = "atypboot",
                              outcome="Group", 
                              N=500,
                          clust_var="Patient",
                              reuse=TRUE,
                              file=model_name,
                              Q="Q2")

# ROC curve
roc_c <- roc_curve(gbm_model, group)

# save the results
supplements_models[["models_summ"]][["gbm_model"]][[model_name]] <- gbm_model$model_summary
supplements_models[["roc_cs"]][["gbm_model"]][[model_name]] <- gbm_model$kfold_rocobjs

# see the results
gbm_model$model_summary %>% t()

roc_c
```

**rPSC vs non-rPSC - subset*

```{r, warning=FALSE}
group <- c("rPSC","non-rPSC")
comparison_name <- paste0(group[1], " vs ",group[2]," subset")
```

```{r, warning=FALSE}
taxa_to_test <- read.xlsx("../results/Q1/univariate_analysis/significant_taxa_colon.xlsx", sheet=" genus healthy vs pre_ltx")

model_name <- paste(comparison_name,level,segment)

# prepare the data
filt_rcolon_uni_data <- binomial_prep_psc_effect(rcolon_genus_tab,
                                                rcolon_genus_taxa_tab, 
                                                rcolon_metadata,
                                                taxa_to_test,
                                                usage="ml_clr",
                                                patient = TRUE)

# fit the model
gbm_model <- gbm_binomial(filt_rcolon_uni_data,
                              sample_method = "atypboot",
                              outcome="Group", 
                              N=500,
                              reuse=TRUE,
                              file=model_name,
                              Q="Q2",
                          clust_var = "Patient")

# ROC curve
roc_c <- roc_curve(gbm_model, group)

# save the results
supplements_models[["models_summ"]][["gbm_model"]][[model_name]] <- gbm_model$model_summary
supplements_models[["roc_cs"]][["gbm_model"]][[model_name]] <- gbm_model$kfold_rocobjs

# see the results
gbm_model$model_summary %>% t()

roc_c
```

#### Saving results

```{r, warning=FALSE}
models_list <- list()

for (model_name in names(supplements_models$models_summ)){
  df <- do.call(rbind, supplements_models$models_summ[[model_name]])
  models_list[[model_name]] <- df
}

write.xlsx(models_list,
           file=file.path(path,paste0("supplements_models_",segment,".xlsx")),
           rowNames=TRUE)

```

## Results overview

#### Alpha diversity

**Linear models**

```{r,warning=FALSE}
knitr::kable(pc_observed[[segment]],
             digits = 3,
             caption = "Results of linear model testing ASV Richness")

knitr::kable(pc_shannon[[segment]],
             digits = 3,
             caption = "Results of linear model testing Shannon index")

knitr::kable(pc_simpson[[segment]],
             digits = 3,
             caption = "Results of linear model testing Simpson index")

knitr::kable(pc_pielou[[segment]],
             digits = 3,
             caption = "Results of linear model testing Pielou index")
```

**Plots**

```{r, warning=FALSE, fig.width=6, fig.height=4, fig.fullwidth=TRUE}
alpha_div_plots[[paste(segment,"Country")]]
```

#### Beta diversity

**Main results**

```{r,warning=FALSE}
knitr::kable(pairwise_aitchison_raw[[paste("genus", segment)]],
             digits = 3,
             caption = "Results of PERMANOVA - robust aitchison distance")
```

**PCA**

```{r, warning=FALSE,  fig.width=5, fig.height=3, fig.fullwidth=TRUE, eval=TRUE}
pca_plots_list[[paste(segment,"genus custom")]]
```

**Supplements**

```{r, warning=FALSE}
knitr::kable(
  supplements_beta[!grepl("PCoA",names(supplements_beta)) &
                  (grepl("genus",names(supplements_beta))) & 
                    (grepl(segment,names(supplements_beta)))],
             digits = 3,
             caption = "Supplementary PERMANOVA results: Bray-curtis, Jaccard distances")
```

**Supplementary PCA**

```{r, warning=FALSE,  fig.width=8, fig.height=7, fig.fullwidth=TRUE, eval=TRUE}
plot_list <- supplements_beta[grepl("PCoA",names(supplements_beta)) &
                              grepl(segment,names(supplements_beta))]

ggarrange(plotlist = plot_list,
          labels=names(plot_list),
          font.label = list(size=5,face="plain"),
          ncol=2,nrow=3)
```

#### Univariate analysis

**Number of significant taxa**

```{r, warning=FALSE}
knitr::kable(as.data.frame(lapply(list_intersections,nrow)) %>% t() %>% 
  `colnames<-`("Count") )
```


#### Machine learning

**Main models**


```{r, warning=FALSE}
knitr::kable(models_summ_df_colon %>% dplyr::select(
"alpha","lambda",
"auc_optimism_corrected",
"auc_optimism_corrected_CIL",
"auc_optimism_corrected_CIU"),
             digits=3,caption="Elastic net results")
```

**ROC - Genus level**

```{r, warning=FALSE,  warning=FALSE}
roc_curve_all_custom(roc_cs[3:4],Q="Q2",
                     model_name="enet_model",legend=FALSE)
```

**Supplementary models**


```{r, warning=FALSE}
# Build final dataframe
models_list[["enet_model"]] <- rbind(models_summ_df_ileum,models_summ_df_colon)
final_df <- tibble(row_names = rownames(models_list[[1]]))

# Loop through models and extract required values
for (model_name in names(models_list)) {
  model_df <- models_list[[model_name]]
  
  # Combine AUC_optimism_corrected with its CI values
  final_df[[model_name]] <- paste0(
    round(model_df$auc_optimism_corrected, 3), 
    " (", round(model_df$auc_optimism_corrected_CIL, 3), "; ", 
    round(model_df$auc_optimism_corrected_CIU, 3), ")"
  )
}

knitr::kable(final_df, caption="All models")
write.csv(final_df,file=file.path(path,"AUC_all_models.csv"),row.names = FALSE)
```

**Supplementary ROC - genus**

```{r, warning=FALSE,  fig.width=10, fig.height=8, fig.fullwidth=TRUE, warning=FALSE}
rocs_list <- supplements_models$roc_cs
rocs_list[["enet_model"]] <- roc_cs
plot_list <- list()

for (model_name in names(rocs_list)) {
  plot_list[[model_name]] <- roc_curve_all_custom(rocs_list[[model_name]][c(2:3)],
                       Q="Q2",
                       model_name=model_name,legend=FALSE)
}

p <- ggarrange(plotlist = plot_list,labels = names(rocs_list),font.label = list(face="plain",size=7))

p
```

```{r, warning=FALSE, results='hide', echo=FALSE}
pdf("../figures/Q2/models_rcolon.pdf",
    height =10,width = 10)
p
dev.off()
```

# Data Analysis - Left colon

```{r, warning=FALSE}
segment="left_colon"
```

## Filtering

Rules: 
  - sequencing depth \> 10000 
  
  - nearZeroVar() with default settings 
  
**Rarefaction Curve**

```{r, warning=FALSE}
path="../intermediate_files/rarecurves"
seq_depth_threshold <- 10000
```

```{r, warning=FALSE,  eval = FALSE}
ps <- construct_phyloseq(lcolon_asv_tab,lcolon_taxa_tab,lcolon_metadata)
rareres <- get_rarecurve(obj=ps, chunks=500)
save(rareres,file = file.path(path,"rarefaction_colon.Rdata"))
```

```{r, warning=FALSE}
load(file.path(path,"rarefaction_colon.Rdata"))
seq_depth_threshold <- 10000
prare <- ggrarecurve(obj=rareres,
                      factorNames="Country",
                      indexNames=c("Observe")) + 
        theme_bw()+
        theme(axis.text=element_text(size=8), 
              panel.grid=element_blank(),
              strip.background = element_rect(colour=NA,fill="grey"),
              strip.text.x = element_text(face="bold")) + 
        geom_vline(xintercept = seq_depth_threshold, 
                   linetype="dashed", color = "red") + 
        xlim(0, 20000)

prare
```

Library size

```{r, warning=FALSE,  fig.width=5, fig.height=4, fig.fullwidth=TRUE}
read_counts(lcolon_asv_tab, line = c(5000,10000))
```

### Sequencing depth

```{r, warning=FALSE}
data_filt <- seq_depth_filtering(lcolon_asv_tab,
                                 lcolon_taxa_tab,
                                 lcolon_metadata,
                                 seq_depth_threshold = 10000)

filt_lcolon_asv_tab <- data_filt[[1]]; alpha_lcolon_asv_tab <- filt_lcolon_asv_tab
filt_lcolon_taxa_tab <- data_filt[[2]]; alpha_lcolon_taxa_tab <- filt_lcolon_taxa_tab
filt_lcolon_metadata <- data_filt[[3]]; alpha_lcolon_metadata <- filt_lcolon_metadata

seq_step <- dim(filt_lcolon_asv_tab)[1]
```

Library size

```{r, warning=FALSE,  fig.width=5, fig.height=4, fig.fullwidth=TRUE}
read_counts(filt_lcolon_asv_tab,line = c(10000))
```

### NearZeroVar

```{r, warning=FALSE}
data_filt <- nearzerovar_filtering(filt_lcolon_asv_tab,
                                   filt_lcolon_taxa_tab,
                                   filt_lcolon_metadata)

filt_lcolon_asv_tab <- data_filt[[1]]
filt_lcolon_taxa_tab <- data_filt[[2]]
nearzero_step <- dim(filt_lcolon_asv_tab)[1]
```

Library size

```{r, warning=FALSE,  fig.width=5, fig.height=4, fig.fullwidth=TRUE}
read_counts(filt_lcolon_asv_tab,line = c(5000,10000))
```

Check zero depth

```{r, warning=FALSE}
data_filt <- check_zero_depth(filt_lcolon_asv_tab, 
                              filt_lcolon_taxa_tab, 
                              filt_lcolon_metadata)

filt_lcolon_asv_tab <- data_filt[[1]]; 
filt_lcolon_taxa_tab <- data_filt[[2]]; 
filt_lcolon_metadata <- data_filt[[3]]; 
```

Library size

```{r, warning=FALSE,  fig.width=5, fig.height=4, fig.fullwidth=TRUE}
read_counts(filt_lcolon_asv_tab,line = c(5000,10000))
```

### Final Counts

```{r, warning=FALSE}
final_counts_filtering(lcolon_asv_tab,
                       filt_lcolon_asv_tab,
                       filt_lcolon_metadata,
                       seq_step, 0, nearzero_step)
```


## Beta diversity

Calculating Aitchison distance (euclidean distance on clr-transformed
data), both at ASV and genus level.

### Supplementary analysis

#### Genus level

```{r, warning=FALSE}
level="genus"
```

##### Bray-Curtis

**PERMANOVA**

```{r, warning=FALSE}
pairwise_df <- filt_lcolon_genus_tab %>% column_to_rownames("SeqID") %>% t()

# main effect
pp_main <- pairwise.adonis(pairwise_df,
                           filt_lcolon_genus_metadata$Group,
                           covariate = filt_lcolon_genus_metadata$Country, 
                           patients = filt_lcolon_genus_metadata$Patient,
                           sim.method = "bray", p.adjust.m="BH")

# interaction
pp_int <- pairwise.adonis(pairwise_df,
                          filt_lcolon_genus_metadata$Group,
                          covariate = filt_lcolon_genus_metadata$Country, 
                          patients = filt_lcolon_genus_metadata$Patient,
                          interaction = TRUE, sim.method = "bray", p.adjust.m="BH")

# tidy the results
pp_factor <- pp_main[[1]]
pp_cov <- pp_main[[2]]
pp_fac.cov <- pp_int[[3]]

cols <- c("pairs","Df","SumsOfSqs", "F.Model","R2","p.value", "p.adjusted", "sig")
colnames(pp_factor) <- cols; colnames(pp_cov) <- cols; colnames(pp_fac.cov) <- cols; 

# save raw results
supplements_beta[[paste("bray",level,segment)]] <- rbind(pp_factor,pp_cov,pp_fac.cov)
```

```{r, warning=FALSE}
# see the results
knitr::kable(pp_factor,digits = 3,caption = "PERMANOVA, GROUP separation")
knitr::kable(pp_cov,digits = 3,caption = "PERMANOVA, COUNTRY separation")
knitr::kable(pp_fac.cov,digits = 3,caption = "PERMANOVA, INTERACTION GROUP:Country")
```

Interaction check

```{r, warning=FALSE}
interaction_sig <- pp_fac.cov$pairs[pp_fac.cov$p.adjusted < 0.05]

for (i in 1:length(interaction_sig)){
  group1 <- unlist(strsplit(interaction_sig[i],split = " vs "))[1]
  group2 <- unlist(strsplit(interaction_sig[i],split = " vs "))[2]
  group2 <- unlist(strsplit(group2,split = " : "))[1]
  
  result_list <- adonis_postanalysis(x=pairwise_df,
                      factors = filt_lcolon_genus_metadata$Group,
                      covariate = filt_lcolon_genus_metadata$Country, 
                      patients = filt_lcolon_genus_metadata$Patient,
                      group1 = group1,
                      group2 = group2,
                      sim.method = 'bray')
  print(result_list)
}

```

**Plots**

```{r, warning=FALSE}
p <- pca_plot_custom(filt_lcolon_genus_tab,
                                 filt_lcolon_genus_taxa,
                                 filt_lcolon_genus_metadata,
                                 measure = "bray",
                                 show_boxplots = TRUE,
                                 variable = "Group", size=3, show_legend=FALSE)

# save the results
supplements_beta[[paste("PCoA bray",level,segment)]] <- p

# see the results
p
```

##### Jaccard

**PERMANOVA**

```{r, warning=FALSE}
pairwise_df <- filt_lcolon_genus_tab %>% column_to_rownames("SeqID") %>% t()

# main effect
pp_main <- pairwise.adonis(pairwise_df,
                           filt_lcolon_genus_metadata$Group,
                           covariate = filt_lcolon_genus_metadata$Country,
                           patients = filt_lcolon_genus_metadata$Patient,
                           sim.method = "jaccard", p.adjust.m="BH")

# interaction
pp_int <- pairwise.adonis(pairwise_df,
                          filt_lcolon_genus_metadata$Group,
                          covariate = filt_lcolon_genus_metadata$Country,
                          patients = filt_lcolon_genus_metadata$Patient,
                          interaction = TRUE, sim.method = "jaccard", p.adjust.m="BH")

# tidy the results
pp_factor <- pp_main[[1]]
pp_cov <- pp_main[[2]]
pp_fac.cov <- pp_int[[3]]

cols <- c("pairs","Df","SumsOfSqs", "F.Model","R2","p.value", "p.adjusted", "sig")
colnames(pp_factor) <- cols; colnames(pp_cov) <- cols; colnames(pp_fac.cov) <- cols; 

# save raw results
supplements_beta[[paste("jaccard",level,segment)]] <- rbind(pp_factor, 
                                                            pp_cov, 
                                                            pp_fac.cov)
```

```{r, warning=FALSE}
# see the results
knitr::kable(pp_factor,digits = 3,caption = "PERMANOVA, GROUP separation")
knitr::kable(pp_cov,digits = 3,caption = "PERMANOVA, COUNTRY separation")
knitr::kable(pp_fac.cov,digits = 3,caption = "PERMANOVA, INTERACTION GROUP:Country")
```

Interaction check

```{r, warning=FALSE}
interaction_sig <- pp_fac.cov$pairs[pp_fac.cov$p.adjusted < 0.05]

for (i in 1:length(interaction_sig)){
  group1 <- unlist(strsplit(interaction_sig[i],split = " vs "))[1]
  group2 <- unlist(strsplit(interaction_sig[i],split = " vs "))[2]
  group2 <- unlist(strsplit(group2,split = " : "))[1]
  
  result_list <- adonis_postanalysis(x=pairwise_df,
                      factors = filt_lcolon_genus_metadata$Group,
                      covariate = filt_lcolon_genus_metadata$Country, 
                      patients = filt_lcolon_genus_metadata$Patient,
                      group1 = group1,
                      group2 = group2,
                      sim.method = 'jaccard')
  print(result_list)
}

```

**Plots**

```{r, warning=FALSE}
p <- pca_plot_custom(filt_lcolon_genus_tab,
                                 filt_lcolon_genus_taxa,
                                 filt_lcolon_genus_metadata,
                                 measure = "jaccard",
                                 show_boxplots = TRUE,
                                 variable = "Group", size=3, show_legend=FALSE)

# save the results
supplements_beta[[paste("PCoA jaccard",level,segment)]] <- p

# see the results
p
```

#### ASV level

```{r, warning=FALSE}
level="ASV"
```

##### Aitchison

```{r, warning=FALSE}
# preparing data frame
pairwise_df <- filt_lcolon_asv_tab %>% column_to_rownames("SeqID") %>% t()

# main effect
pp_main <- pairwise.adonis(x=pairwise_df,
                          filt_lcolon_metadata$Group,
                           covariate = filt_lcolon_metadata$Country, 
                           sim.method = "robust.aitchison", 
                           p.adjust.m="BH",
                           patients = filt_lcolon_metadata$Patient)

# interaction
pp_int <- pairwise.adonis(pairwise_df,filt_lcolon_metadata$Group,
                          covariate = filt_lcolon_metadata$Country, 
                          interaction = TRUE, 
                          sim.method = "robust.aitchison", 
                          p.adjust.m="BH",
                          patients = filt_lcolon_metadata$Patient)

pp_factor <- pp_main[[1]]
pp_cov <- pp_main[[2]]
pp_fac.cov <- pp_int[[3]]

cols <- c("pairs","Df","SumsOfSqs", "F.Model","R2","p.value", "p.adjusted", "sig")
colnames(pp_factor) <- cols; colnames(pp_cov) <- cols; colnames(pp_fac.cov) <- cols; 

# save raw results
supplements_beta[[paste("aitchison",level,segment)]] <- rbind(pp_factor, 
                                                            pp_cov, 
                                                            pp_fac.cov)

# see the results
pp_factor
pp_cov
pp_fac.cov

```

```{r, warning=FALSE}
# see the results
knitr::kable(pp_factor,digits = 3,caption = "PERMANOVA, GROUP separation")
knitr::kable(pp_cov,digits = 3,caption = "PERMANOVA, COUNTRY separation")
knitr::kable(pp_fac.cov,digits = 3,caption = "PERMANOVA, INTERACTION GROUP:Country")
```

Interaction check

```{r, warning=FALSE}
interaction_sig <- pp_fac.cov$pairs[pp_fac.cov$p.adjusted < 0.05]

if (length(interaction_sig)>0){
 for (i in 1:length(interaction_sig)){
  group1 <- unlist(strsplit(interaction_sig[i],split = " vs "))[1]
  group2 <- unlist(strsplit(interaction_sig[i],split = " vs "))[2]
  group2 <- unlist(strsplit(group2,split = " : "))[1]
  
  result_list <- adonis_postanalysis(x=pairwise_df,
                      factors = filt_lcolon_metadata$Group,
                      covariate = filt_lcolon_metadata$Country, 
                      group1 = group1,
                      group2 = group2,
                      patients = filt_lcolon_metadata$Patient)
  print(result_list)
} 
}
```

**PCoA**

```{r, warning=FALSE}
p <- pca_plot_custom(filt_lcolon_asv_tab,
                           filt_lcolon_taxa_tab,
                           filt_lcolon_metadata,
                           show_boxplots = TRUE,
                           variable = "Group", 
                           size=3, 
                           show_legend=FALSE)

# save the results
supplements_beta[[paste("PCoA aitchison",level,segment)]] <- p

# see the results
p
```

##### Bray-Curtis 

**PERMANOVA**

```{r, warning=FALSE}
# preparing data frame
pairwise_df <- filt_lcolon_asv_tab %>% column_to_rownames("SeqID") %>% t()

# main effect
pp_main <- pairwise.adonis(pairwise_df,
                           filt_lcolon_metadata$Group,
                           covariate = filt_lcolon_metadata$Country,
                           patients = filt_lcolon_metadata$Patient,
                           sim.method = "bray", p.adjust.m="BH")

# interaction
pp_int <- pairwise.adonis(pairwise_df,
                          filt_lcolon_metadata$Group,
                          covariate = filt_lcolon_metadata$Country, 
                          patients = filt_lcolon_metadata$Patient,
                          interaction = TRUE, sim.method = "bray", p.adjust.m="BH")

pp_factor <- pp_main[[1]]
pp_cov <- pp_main[[2]]
pp_fac.cov <- pp_int[[3]]

cols <- c("pairs","Df","SumsOfSqs", "F.Model","R2","p.value", "p.adjusted", "sig")
colnames(pp_factor) <- cols; colnames(pp_cov) <- cols; colnames(pp_fac.cov) <- cols; 

# save raw results
supplements_beta[[paste("bray",level,segment)]] <- rbind(pp_factor, 
                                                            pp_cov, 
                                                            pp_fac.cov)

```

```{r, warning=FALSE}
# see the results
knitr::kable(pp_factor,digits = 3,caption = "PERMANOVA, GROUP separation")
knitr::kable(pp_cov,digits = 3,caption = "PERMANOVA, COUNTRY separation")
knitr::kable(pp_fac.cov,digits = 3,caption = "PERMANOVA, INTERACTION GROUP:Country")
```

Interaction check

```{r, warning=FALSE}
interaction_sig <- pp_fac.cov$pairs[pp_fac.cov$p.adjusted < 0.05]

for (i in 1:length(interaction_sig)){
  group1 <- unlist(strsplit(interaction_sig[i],split = " vs "))[1]
  group2 <- unlist(strsplit(interaction_sig[i],split = " vs "))[2]
  group2 <- unlist(strsplit(group2,split = " : "))[1]
  
  result_list <- adonis_postanalysis(x=pairwise_df,
                      factors = filt_lcolon_metadata$Group,
                      covariate = filt_lcolon_metadata$Country, 
                      group1 = group1,
                      group2 = group2,
                      patients = filt_lcolon_metadata$Patient,
                      sim.method = 'bray')
  print(result_list)
}

```

**PCoA**

```{r, warning=FALSE}
p <- pca_plot_custom(filt_lcolon_asv_tab,
                     filt_lcolon_taxa_tab,
                     filt_lcolon_metadata,
                     measure = "bray",
                     show_boxplots = TRUE,
                     variable = "Group", size=3, show_legend=FALSE)

# save the results
supplements_beta[[paste("PCoA bray",level,segment)]] <- p

# see the results
p
```

##### Jaccard

**PERMANOVA**

```{r, warning=FALSE}
# preparing data frame
pairwise_df <- filt_lcolon_asv_tab %>% column_to_rownames("SeqID") %>% t()

# main effect
pp_main <- pairwise.adonis(pairwise_df,
                           filt_lcolon_metadata$Group,
                           covariate = filt_lcolon_metadata$Country,
                           patients = filt_lcolon_metadata$Patient,
                           sim.method = "jaccard", p.adjust.m="BH")

# interaction
pp_int <- pairwise.adonis(pairwise_df,
                          filt_lcolon_metadata$Group,
                          covariate = filt_lcolon_metadata$Country, 
                          patients = filt_lcolon_metadata$Patient,
                          interaction = TRUE, sim.method = "jaccard", p.adjust.m="BH")

pp_factor <- pp_main[[1]]
pp_cov <- pp_main[[2]]
pp_fac.cov <- pp_int[[3]]

cols <- c("pairs","Df","SumsOfSqs", "F.Model","R2","p.value", "p.adjusted", "sig")
colnames(pp_factor) <- cols; colnames(pp_cov) <- cols; colnames(pp_fac.cov) <- cols; 

# save raw results
supplements_beta[[paste("jaccard",level,segment)]] <- rbind(pp_factor, 
                                                            pp_cov, 
                                                            pp_fac.cov)
```

```{r, warning=FALSE}
# see the results
knitr::kable(pp_factor,digits = 3,caption = "PERMANOVA, GROUP separation")
knitr::kable(pp_cov,digits = 3,caption = "PERMANOVA, COUNTRY separation")
knitr::kable(pp_fac.cov,digits = 3,caption = "PERMANOVA, INTERACTION GROUP:Country")
```

Interaction check

```{r, warning=FALSE}
interaction_sig <- pp_fac.cov$pairs[pp_fac.cov$p.adjusted < 0.05]

for (i in 1:length(interaction_sig)){
  group1 <- unlist(strsplit(interaction_sig[i],split = " vs "))[1]
  group2 <- unlist(strsplit(interaction_sig[i],split = " vs "))[2]
  group2 <- unlist(strsplit(group2,split = " : "))[1]
  
  result_list <- adonis_postanalysis(x=pairwise_df,
                      factors = filt_lcolon_metadata$Group,
                      covariate = filt_lcolon_metadata$Country, 
                      patients = filt_lcolon_metadata$Patient,
                      group1 = group1,
                      group2 = group2,
                      sim.method = 'jaccard')
  print(result_list)
}

```

**PCoA**

```{r, warning=FALSE}
p <- pca_plot_custom(filt_lcolon_asv_tab,
                     filt_lcolon_taxa_tab,
                     filt_lcolon_metadata,
                     measure = "jaccard",
                     show_boxplots = TRUE,
                     variable = "Group", size=3, show_legend=FALSE)

# save the results
supplements_beta[[paste("PCoA jaccard",level,segment)]] <- p

# see the results
p
```

#### Saving results

```{r, warning=FALSE}
write.xlsx(supplements_beta[!grepl("PCoA",names(supplements_beta))],
           file = file.path(path,
           paste0("supplements_beta_diversity_", segment,".xlsx")))
```

## Machine learning

```{r, warning=FALSE}
path = "../results/Q2/models"
```

### ElasticNet

```{r, warning=FALSE}
model="enet"
```

#### Genus level

```{r, warning=FALSE}
level="genus"
```

Aggregate taxa

```{r, warning=FALSE}
genus_data <- aggregate_taxa(lcolon_asv_tab,
                             lcolon_taxa_tab,
                             taxonomic_level = level)

lcolon_genus_tab <- genus_data[[1]]
lcolon_genus_taxa_tab <- genus_data[[2]]
```

##### rPSC vs non-rPSC - all

```{r, warning=FALSE}
group <- c("rPSC","non-rPSC")
comparison_name <- paste0(group[1], " vs ",group[2])
```

```{r, warning=FALSE}
model_name <- paste(comparison_name,level,segment)

# prepare the data
filt_lcolon_uni_data <- binomial_prep(lcolon_genus_tab,
                                     lcolon_genus_taxa_tab,
                                     lcolon_metadata,
                                     group, 
                                     usage="ml_clr",
                                     patient = TRUE)

# fit the model
enet_model <- glmnet_binomial(filt_lcolon_uni_data,
                              sample_method = "atypboot",
                              outcome="Group", 
                              N=10,
                              reuse=TRUE,
                              file=model_name,
                              Q="Q2",
                              clust_var = "Patient")

# ROC
roc_c <- roc_curve(enet_model, group)

# save the results
models_summ[[model_name]] <- enet_model$model_summary
models_cm[[model_name]] <- enet_model$conf_matrices$original
roc_cs[[model_name]] <- enet_model$kfold_rocobjs
betas[[model_name]] <- as.matrix(enet_model$betas)

# see the results
enet_model$model_summary %>% t()
enet_model$conf_matrices
enet_model$plot

roc_c
```

##### rPSC vs non-rPSC - subset

```{r, warning=FALSE}
group <- c("rPSC","non-rPSC")
comparison_name <- paste0(group[1], " vs ",group[2]," subset")
```

```{r, warning=FALSE}
taxa_to_test <- read.xlsx("../results/Q1/univariate_analysis/significant_taxa_right_colon.xlsx", sheet=" genus healthy vs pre_ltx")

model_name <- paste(comparison_name,level,segment)

# prepare the data
filt_lcolon_uni_data <- binomial_prep_psc_effect(lcolon_genus_tab,
                                                lcolon_genus_taxa_tab, 
                                                lcolon_metadata,
                                                taxa_to_test,
                                                usage="ml_clr",
                                                patient = TRUE)

# fit the model
enet_model <- glmnet_binomial(filt_lcolon_uni_data,
                              sample_method = "atypboot",
                              outcome="Group", 
                              N=10,
                              reuse=TRUE,
                              file=model_name,
                              Q="Q2",clust_var = "Patient")

# ROC
roc_c <- roc_curve(enet_model, group)

# save the results
models_summ[[model_name]] <- enet_model$model_summary
models_cm[[model_name]] <- enet_model$conf_matrices$original
roc_cs[[model_name]] <- enet_model$kfold_rocobjs
betas[[model_name]] <- as.matrix(enet_model$betas)

# see the results
enet_model$model_summary %>% t()
enet_model$conf_matrices
enet_model$plot

roc_c

```

#### Saving results

```{r, warning=FALSE}
models_summ_df_colon <- do.call(rbind, 
  models_summ[grep(segment,names(models_summ),value = TRUE)])

write.csv(models_summ_df_colon,file.path(path,paste0("elastic_net_",segment,".csv")))
```

### Supplementary models

#### CLR-transformed data

##### kNN

```{r, warning=FALSE}
model="knn"
```

###### Genus level

```{r, warning=FALSE}
level="genus"
```

Aggregate taxa

```{r, warning=FALSE}
genus_data <- aggregate_taxa(lcolon_asv_tab,
                             lcolon_taxa_tab,
                             taxonomic_level = level)

lcolon_genus_tab <- genus_data[[1]]
lcolon_genus_taxa_tab <- genus_data[[2]]
```

**rPSC vs non-rPSC - all**

```{r, warning=FALSE}
group <- c("rPSC","non-rPSC")
comparison_name <- paste0(group[1], " vs ",group[2])
```

```{r, warning=FALSE}
model_name <- paste(comparison_name,level,segment)

# prepare the data
filt_lcolon_uni_data <- binomial_prep(lcolon_genus_tab,
                                     lcolon_genus_taxa_tab,
                                     lcolon_metadata,
                                     group, 
                                     usage="ml_clr",
                                     patient = TRUE)

# fit the model
knn_model <- knn_binomial(filt_lcolon_uni_data,
                              sample_method = "atypboot",
                              outcome="Group", 
                              N=500,
                          clust_var="Patient",
                              reuse=TRUE,
                              file=model_name,
                              Q="Q2")

# ROC curve
roc_c <- roc_curve(knn_model, group)

# save the results
supplements_models[["models_summ"]][["knn_model"]][[model_name]] <- knn_model$model_summary
supplements_models[["roc_cs"]][["knn_model"]][[model_name]] <- knn_model$kfold_rocobjs


# see the results
knn_model$model_summary %>% t()


roc_c
```

**rPSC vs non-rPSC - subset**

```{r, warning=FALSE}
group <- c("rPSC","non-rPSC")
comparison_name <- paste0(group[1], " vs ",group[2]," subset")
```

```{r, warning=FALSE}
taxa_to_test <- read.xlsx("../results/Q1/univariate_analysis/significant_taxa_colon.xlsx", sheet=" genus healthy vs pre_ltx")

model_name <- paste(comparison_name,level,segment)

# prepare the data
filt_ileum_uni_data <- binomial_prep_psc_effect(colon_genus_tab,
                                                lcolon_genus_taxa_tab, 
                                                lcolon_metadata,
                                                taxa_to_test,
                                                usage="ml_clr",
                                                patient = TRUE)

# fit the model
knn_model <- knn_binomial(filt_ileum_uni_data,
                              sample_method = "atypboot",
                              outcome="Group", 
                              N=500,
                              reuse=TRUE,
                              file=model_name,
                              Q="Q2",clust_var="Patient")

# ROC
roc_c <- roc_curve(knn_model, group)

# save the results
supplements_models[["models_summ"]][["knn_model"]][[model_name]] <- knn_model$model_summary
supplements_models[["roc_cs"]][["knn_model"]][[model_name]] <- knn_model$kfold_rocobjs

# see the results
knn_model$model_summary %>% t()

roc_c
```


##### Random Forest

```{r, warning=FALSE}
model="rf"
```

###### Genus level

```{r, warning=FALSE}
level="genus"
```

Aggregate taxa

```{r, warning=FALSE}
genus_data <- aggregate_taxa(lcolon_asv_tab,
                             lcolon_taxa_tab,
                             taxonomic_level = level)

lcolon_genus_tab <- genus_data[[1]]
lcolon_genus_taxa_tab <- genus_data[[2]]
```

**rPSC vs non-rPSC - all**

```{r, warning=FALSE}
group <- c("rPSC","non-rPSC")
comparison_name <- paste0(group[1], " vs ",group[2])
```

```{r, warning=FALSE}
model_name <- paste(comparison_name,level,segment)

# prepare the data
filt_lcolon_uni_data <- binomial_prep(lcolon_genus_tab,
                                     lcolon_genus_taxa_tab,
                                     lcolon_metadata,
                                     group, 
                                     usage="ml_clr",
                                     patient = TRUE)

# fit the model
rf_model <- rf_binomial(filt_lcolon_uni_data,
                              sample_method = "atypboot",
                              outcome="Group", 
                              N=500,
                        clust_var="Patient",
                              reuse=TRUE,
                              file=model_name,
                              Q="Q2")

# ROC curve
roc_c <- roc_curve(rf_model, group)

# save the results
supplements_models[["models_summ"]][["rf_model"]][[model_name]] <- rf_model$model_summary
supplements_models[["roc_cs"]][["rf_model"]][[model_name]] <- rf_model$kfold_rocobjs

# see the results
rf_model$model_summary %>% t()

roc_c
```

**rPSC vs non-rPSC - subset**

```{r, warning=FALSE}
group <- c("rPSC","non-rPSC")
comparison_name <- paste0(group[1], " vs ",group[2]," subset")
```

```{r, warning=FALSE}
taxa_to_test <- read.xlsx("../results/Q1/univariate_analysis/significant_taxa_colon.xlsx", sheet=" genus healthy vs pre_ltx")

model_name <- paste(comparison_name,level,segment)

# prepare the data
filt_ileum_uni_data <- binomial_prep_psc_effect(lcolon_genus_tab,
                                                lcolon_genus_taxa_tab, 
                                                lcolon_metadata,
                                                taxa_to_test,
                                                usage="ml_clr",
                                                patient = TRUE)

# fit the model
rf_model <- rf_binomial(filt_ileum_uni_data,
                              sample_method = "atypboot",
                              outcome="Group", 
                              N=500,
                              reuse=TRUE,
                              file=model_name,
                              Q="Q2",
                        clust_var="Patient")

# ROC curve
roc_c <- roc_curve(rf_model, group)

# save the results
supplements_models[["models_summ"]][["rf_model"]][[model_name]] <- rf_model$model_summary
supplements_models[["roc_cs"]][["rf_model"]][[model_name]] <- rf_model$kfold_rocobjs

# see the results
rf_model$model_summary %>% t()

roc_c
```


##### Gradient boosting

```{r, warning=FALSE}
model="gb"
```

###### Genus level

```{r, warning=FALSE}
level="genus"
```

Aggregate taxa

```{r, warning=FALSE}
genus_data <- aggregate_taxa(lcolon_asv_tab,
                             lcolon_taxa_tab,
                             taxonomic_level = level)

lcolon_genus_tab <- genus_data[[1]]
lcolon_genus_taxa_tab <- genus_data[[2]]
```

**rPSC vs non-rPSC - all**

```{r, warning=FALSE}
group <- c("rPSC","non-rPSC")
comparison_name <- paste0(group[1], " vs ",group[2])
```

```{r, warning=FALSE}
model_name <- paste(comparison_name,level,segment)

# prepare the data
filt_lcolon_uni_data <- binomial_prep(lcolon_genus_tab,
                                     lcolon_genus_taxa_tab,
                                     lcolon_metadata,
                                     group, 
                                     usage="ml_clr",
                                     patient = TRUE)

# fit the model
gbm_model <- gbm_binomial(filt_lcolon_uni_data,
                              sample_method = "atypboot",
                              outcome="Group", 
                              N=500,
                          clust_var="Patient",
                              reuse=TRUE,
                              file=model_name,
                              Q="Q2")

# ROC curve
roc_c <- roc_curve(gbm_model, group)

# save the results
supplements_models[["models_summ"]][["gbm_model"]][[model_name]] <- gbm_model$model_summary
supplements_models[["roc_cs"]][["gbm_model"]][[model_name]] <- gbm_model$kfold_rocobjs

# see the results
gbm_model$model_summary %>% t()

roc_c
```

**rPSC vs non-rPSC - subset*

```{r, warning=FALSE}
group <- c("rPSC","non-rPSC")
comparison_name <- paste0(group[1], " vs ",group[2]," subset")
```

```{r, warning=FALSE}
taxa_to_test <- read.xlsx("../results/Q1/univariate_analysis/significant_taxa_colon.xlsx", sheet=" genus healthy vs pre_ltx")

model_name <- paste(comparison_name,level,segment)

# prepare the data
filt_lcolon_uni_data <- binomial_prep_psc_effect(lcolon_genus_tab,
                                                lcolon_genus_taxa_tab, 
                                                lcolon_metadata,
                                                taxa_to_test,
                                                usage="ml_clr",
                                                patient = TRUE)

# fit the model
gbm_model <- gbm_binomial(filt_lcolon_uni_data,
                              sample_method = "atypboot",
                              outcome="Group", 
                              N=500,
                              reuse=TRUE,
                              file=model_name,
                              Q="Q2",
                          clust_var = "Patient")

# ROC curve
roc_c <- roc_curve(gbm_model, group)

# save the results
supplements_models[["models_summ"]][["gbm_model"]][[model_name]] <- gbm_model$model_summary
supplements_models[["roc_cs"]][["gbm_model"]][[model_name]] <- gbm_model$kfold_rocobjs

# see the results
gbm_model$model_summary %>% t()

roc_c
```

#### Saving results

```{r, warning=FALSE}
models_list <- list()

for (model_name in names(supplements_models$models_summ)){
  df <- do.call(rbind, supplements_models$models_summ[[model_name]])
  models_list[[model_name]] <- df
}

write.xlsx(models_list,
           file=file.path(path,paste0("supplements_models_",segment,".xlsx")),
           rowNames=TRUE)

```

## Results overview

#### Alpha diversity

**Linear models**

```{r,warning=FALSE}
knitr::kable(pc_observed[[segment]],
             digits = 3,
             caption = "Results of linear model testing ASV Richness")

knitr::kable(pc_shannon[[segment]],
             digits = 3,
             caption = "Results of linear model testing Shannon index")

knitr::kable(pc_simpson[[segment]],
             digits = 3,
             caption = "Results of linear model testing Simpson index")

knitr::kable(pc_pielou[[segment]],
             digits = 3,
             caption = "Results of linear model testing Pielou index")
```

**Plots**

```{r, warning=FALSE, fig.width=6, fig.height=4, fig.fullwidth=TRUE}
alpha_div_plots[[paste(segment,"Country")]]
```

#### Beta diversity

**Main results**

```{r,warning=FALSE}
knitr::kable(pairwise_aitchison_raw[[paste("genus", segment)]],
             digits = 3,
             caption = "Results of PERMANOVA - robust aitchison distance")
```

**PCA**

```{r, warning=FALSE,  fig.width=5, fig.height=3, fig.fullwidth=TRUE, eval=TRUE}
pca_plots_list[[paste(segment,"genus custom")]]
```

**Supplements**

```{r, warning=FALSE}
knitr::kable(
  supplements_beta[!grepl("PCoA",names(supplements_beta)) &
                  (grepl("genus",names(supplements_beta))) & 
                    (grepl(segment,names(supplements_beta)))],
             digits = 3,
             caption = "Supplementary PERMANOVA results: Bray-curtis, Jaccard distances")
```

**Supplementary PCA**

```{r, warning=FALSE,  fig.width=8, fig.height=7, fig.fullwidth=TRUE, eval=TRUE}
plot_list <- supplements_beta[grepl("PCoA",names(supplements_beta)) &
                              grepl(segment,names(supplements_beta))]

ggarrange(plotlist = plot_list,
          labels=names(plot_list),
          font.label = list(size=5,face="plain"),
          ncol=2,nrow=3)
```

#### Univariate analysis

**Number of significant taxa**

```{r, warning=FALSE}
knitr::kable(as.data.frame(lapply(list_intersections,nrow)) %>% t() %>% 
  `colnames<-`("Count") )
```


#### Machine learning

**Main models**


```{r, warning=FALSE}
knitr::kable(models_summ_df_colon %>% dplyr::select(
"alpha","lambda",
"auc_optimism_corrected",
"auc_optimism_corrected_CIL",
"auc_optimism_corrected_CIU"),
             digits=3,caption="Elastic net results")
```

**ROC - Genus level**

```{r, warning=FALSE,  warning=FALSE}
roc_curve_all_custom(roc_cs[3:4],Q="Q2",
                     model_name="enet_model",legend=FALSE)
```

**Supplementary models**


```{r, warning=FALSE}
# Build final dataframe
models_list[["enet_model"]] <- rbind(models_summ_df_ileum,models_summ_df_colon)
final_df <- tibble(row_names = rownames(models_list[[1]]))

# Loop through models and extract required values
for (model_name in names(models_list)) {
  model_df <- models_list[[model_name]]
  
  # Combine AUC_optimism_corrected with its CI values
  final_df[[model_name]] <- paste0(
    round(model_df$auc_optimism_corrected, 3), 
    " (", round(model_df$auc_optimism_corrected_CIL, 3), "; ", 
    round(model_df$auc_optimism_corrected_CIU, 3), ")"
  )
}

knitr::kable(final_df, caption="All models")
write.csv(final_df,file=file.path(path,"AUC_all_models.csv"),row.names = FALSE)
```

**Supplementary ROC - genus**

```{r, warning=FALSE,  fig.width=10, fig.height=8, fig.fullwidth=TRUE, warning=FALSE}
rocs_list <- supplements_models$roc_cs
rocs_list[["enet_model"]] <- roc_cs
plot_list <- list()

for (model_name in names(rocs_list)) {
  plot_list[[model_name]] <- roc_curve_all_custom(rocs_list[[model_name]][c(2:3)],
                       Q="Q2",
                       model_name=model_name,legend=FALSE)
}

p <- ggarrange(plotlist = plot_list,labels = names(rocs_list),font.label = list(face="plain",size=7))

p
```

```{r, warning=FALSE, results='hide', echo=FALSE}
pdf("../figures/Q2/models_colon.pdf",
    height =10,width = 10)
p
dev.off()
```


# Session info

```{r, warning=FALSE}
sessionInfo()
```

