---
title: "Geography-independent mucosal microbiota alterations in primary sclerosing cholangitis persist after liver transplantation"
subtitle: "5. Hypothesis: PSC disease is associated with fecal calprotectin"
output:
  html_document:
    df_print: paged
    toc: true
    toc_float: true
    toc_depth: 6
editor_options: 
  markdown: 
    wrap: 72
---

# Introduction

## About 

An IBD vs. no-IBD comparison was performed within PSC patients (pre-LTx and rPSC individuals combined) to discover IBD-specific pattern. 

**IBD vs no_IBD analysis on merged data**

-   Alpha diversity –\> group effect, country effect, interaction effect

-   Beta diversity – PERMANOVA, PCA -\> group effect, country effect,
    interaction effect

-   DAA -\>

    -   o Group effect – linDA + MaAsLin2 intersection

    -   o Country effect – – taxa with significant interaction effect were excluded based on individual post-hoc analysis. Taxa had to have the same direction in both countries 

**IBD vs no_IBD** – finding IBD-specific pattern in PSC patients

*Importing libraries and custom functions built for this analysis*

```{r, warning=FALSE}
source("custom_functions.R")
```

## Data Import

Importing ASV, taxa and metadata tables for both Czech and Norway
samples.

**Czech**

*MICROBIOME*

```{r, warning=FALSE}
path = "../../../../data/analysis_ready_data/ikem/"
asv_tab_ikem <- as.data.frame(fread(file.path(path,"asv_table_ikem.csv"),
                                    check.names = FALSE))
taxa_tab_ikem <- as.data.frame(fread(file.path(path,"taxa_table_ikem.csv"),
                                     check.names = FALSE))
metadata_ikem <- as.data.frame(fread(file.path(path,"metadata_ikem.csv"),
                                     check.names = FALSE))
```

*Fecal calprotectin info*

```{r, warning=FALSE}
# clinical metadata
metadata_clinical <- read.csv("../../../../data/clinical_data/clinical_metadata_cz.csv")
metadata_clinical$PatientID <- as.character(metadata_clinical$PatientID)
```

**Norway**

*MICROBIOME*

```{r, warning=FALSE}
path = "../../../../data/analysis_ready_data/norway/"
asv_tab_norway <- as.data.frame(fread(file.path(path,"asv_table_norway.csv"),
                                    check.names = FALSE))
taxa_tab_norway <- as.data.frame(fread(file.path(path,"taxa_table_norway.csv"),
                                    check.names = FALSE))
metadata_norway <- as.data.frame(fread(file.path(path,"metadata_norway.csv"),
                                    check.names = FALSE))
```

*Fecal calprotectin info*

```{r, warning=FALSE}
# clinical metadata
metadata_clinical_no <- read.csv("../../../../data/clinical_data/clinical_metadata_no.csv")
metadata_clinical_no$PatientID <- paste0("NO_", as.character(metadata_clinical_no$subjectid))
```

### Merging

**Clinical data**

```{r}
metadata_ikem <- merge(metadata_ikem,metadata_clinical[,c("SampleID","Fecal.calprotectin")],by="SampleID",all.x=TRUE) %>%
  dplyr::mutate(Calprotectin=Fecal.calprotectin) %>% 
  dplyr::mutate(Calprotectin = case_when(
        Calprotectin == ">6000" ~ "6000",
        TRUE ~ Calprotectin)) %>%
  dplyr::mutate(Calprotectin= as.numeric(Calprotectin))

metadata_norway <- merge(metadata_norway,metadata_clinical_no[,c("SampleID","Calprotectin")],
                         by="SampleID",all.x=TRUE) %>%
  dplyr::mutate(Calprotectin= as.numeric(Calprotectin))
```

**Terminal ileum**

```{r, warning=FALSE}
ileum_data <- merging_data(asv_tab_1=asv_tab_ikem,
                           asv_tab_2= asv_tab_norway,
                           taxa_tab_1=taxa_tab_ikem,
                           taxa_tab_2=taxa_tab_norway,
                           metadata_1=metadata_ikem,
                           metadata_2=metadata_norway,
                           segment="TI",Q="Q5")

ileum_asv_tab <- ileum_data[[1]]
ileum_taxa_tab <- ileum_data[[2]]
ileum_metadata <- ileum_data[[3]]

```


```{r}
ileum_metadata %<>%  dplyr::mutate(
  log_Calprotectin=log(Calprotectin,base = 10))

median(ileum_metadata$Calprotectin,na.rm = TRUE)
median(ileum_metadata$log_Calprotectin,na.rm = TRUE)

p1 <- ggplot(ileum_metadata,aes(log_Calprotectin)) + 
  geom_histogram() + 
  geom_vline(xintercept = log(250,base=10), linetype="dashed", 
               color = "red")
```

*Splitting to groups*


```{r}
ileum_metadata %<>% dplyr::mutate(Calprotectin_groups = case_when(
    Calprotectin <= 250 ~ "low",
    Calprotectin > 250 ~ "high"))

ileum_metadata %>% group_by(Calprotectin_groups) %>%
  count()
```

```{r}
p2 <- ggplot(ileum_metadata %>% dplyr::filter(!is.na(Calprotectin_groups)),aes(Calprotectin_groups)) + 
  geom_bar()
```

```{r, fig.width=10,fig.height=4}
p <- ggarrange(p1,p2,ncol = 2)
p
```


```{r, warning=FALSE}
ileum_metadata %<>% drop_na(Calprotectin) %>% dplyr::filter(Group!="healthy")
ileum_asv_tab %<>% dplyr::select("SeqID",ileum_metadata$SampleID)
ileum_taxa_tab %<>% dplyr::filter(SeqID %in% ileum_asv_tab$SeqID)
```

**Colon**

```{r, warning=FALSE}
colon_data <- merging_data(asv_tab_1=asv_tab_ikem,
                           asv_tab_2=asv_tab_norway,
                           taxa_tab_1=taxa_tab_ikem,
                           taxa_tab_2=taxa_tab_norway,
                           metadata_1=metadata_ikem,
                           metadata_2=metadata_norway,
                           segment="colon",Q="Q5")

colon_asv_tab <- colon_data[[1]]
colon_taxa_tab <- colon_data[[2]]
colon_metadata <- colon_data[[3]]
```

```{r}
colon_metadata %<>%  dplyr::mutate(
  log_Calprotectin=log(Calprotectin,base = 10)) 

colon_metadata_forplot <- colon_metadata %>%
        group_by(Patient) %>%
        distinct(Patient, .keep_all = TRUE) %>%
        as.data.frame() 

median(colon_metadata_forplot$Calprotectin,na.rm = TRUE)
median(colon_metadata_forplot$log_Calprotectin,na.rm = TRUE)

p1 <- ggplot(colon_metadata_forplot,aes(log_Calprotectin)) + 
  geom_histogram() + 
  geom_vline(xintercept = log(250,base=10), linetype="dashed", 
               color = "red")
```

*Splitting to groups*


```{r}
colon_metadata %<>% dplyr::mutate(Calprotectin_groups = case_when(
    Calprotectin <= 250 ~ "low",
    Calprotectin > 250 ~ "high"))

colon_metadata_forplot %<>% dplyr::mutate(Calprotectin_groups = case_when(
    Calprotectin <= 250 ~ "low",
    Calprotectin > 250 ~ "high"))

colon_metadata_forplot %>% group_by(Calprotectin_groups) %>%
  count()
```

```{r}
p2 <- ggplot(colon_metadata_forplot %>% dplyr::filter(!is.na(Calprotectin_groups)),aes(Calprotectin_groups)) + 
  geom_bar()
```

```{r, fig.width=10,fig.height=4}
p <- ggarrange(p1,p2,ncol = 2)
p
```


```{r, warning=FALSE}
colon_metadata %<>% drop_na(Calprotectin) %>% dplyr::filter(Group!="healthy")

colon_asv_tab %<>% dplyr::select("SeqID",colon_metadata$SampleID)
colon_taxa_tab %<>% dplyr::filter(SeqID %in% colon_asv_tab$SeqID)
```

## CHI SQUARE TEST

### pre_LTx, rPSC, non-rPSC

```{r}
metadata <- rbind(ileum_metadata,colon_metadata) %>% 
        group_by(Patient) %>%
        distinct(Patient, .keep_all = TRUE) %>%
        as.data.frame() 
```

```{r, warning=FALSE}
df <- metadata %>% dplyr::group_by(Calprotectin_groups,Group) %>%
  distinct(Patient, .keep_all = TRUE) %>%
  count() %>% drop_na() 

# Summarize by Nancy and Group
data <- df %>%
  group_by(Calprotectin_groups, Group) %>%
  summarise(n = sum(n), .groups = "drop") %>%
  pivot_wider(names_from = Group, values_from = n) %>%
  column_to_rownames("Calprotectin_groups") %>%
  as.matrix()

chi_res <- chisq.test(data)
data
chi_res$expected
chi_res
```


# Data Analysis - Terminal ileum

```{r, warning=FALSE}
segment="terminal_ileum"
```

## Filtering

Rules: 

  - sequencing depth \> 10000 
  
  - nearZeroVar() with default settings 

**Rarefaction Curve**

```{r, warning=FALSE}
path="../../../intermediate_files/rarecurves"
seq_depth_threshold <- 10000
```

```{r, eval = FALSE}
ps <- construct_phyloseq(ileum_asv_tab,ileum_taxa_tab,ileum_metadata)
rareres <- get_rarecurve(obj=ps, chunks=500)
save(rareres,file = file.path(path,"rarefaction_ileum.Rdata"))
```

```{r, warning=FALSE}
load(file.path(path,"rarefaction_ileum.Rdata"))

prare <- ggrarecurve(obj=rareres,
                      factorNames="Country",
                      indexNames=c("Observe")) + 
  theme_bw() +
  theme(axis.text=element_text(size=8), panel.grid=element_blank(),
        strip.background = element_rect(colour=NA,fill="grey"),
        strip.text.x = element_text(face="bold")) + 
  geom_vline(xintercept = seq_depth_threshold, 
             linetype="dashed", 
             color = "red") + 
  xlim(0, 20000)

prare
```

**Library size**

```{r, fig.width=5, fig.height=4, fig.fullwidth=TRUE, warning=FALSE}
read_counts(ileum_asv_tab, line = c(5000,10000))
```

### Sequencing depth

```{r, warning=FALSE}
data_filt <- seq_depth_filtering(ileum_asv_tab,
                                 ileum_taxa_tab,
                                 ileum_metadata,
                                 seq_depth_threshold = 10000)

filt_ileum_asv_tab <- data_filt[[1]]; alpha_ileum_asv_tab <- filt_ileum_asv_tab
filt_ileum_taxa_tab <- data_filt[[2]]; alpha_ileum_taxa_tab <- filt_ileum_taxa_tab
filt_ileum_metadata <- data_filt[[3]]; alpha_ileum_metadata <- filt_ileum_metadata

seq_step <- dim(filt_ileum_asv_tab)[1]
```

**Library size**

```{r, fig.width=5, fig.height=4, fig.fullwidth=TRUE, warning=FALSE}
read_counts(filt_ileum_asv_tab,line = c(5000,10000))
```

### NearZeroVar

```{r, warning=FALSE}
data_filt <- nearzerovar_filtering(filt_ileum_asv_tab, 
                                   filt_ileum_taxa_tab,
                                   filt_ileum_metadata)

filt_ileum_asv_tab <- data_filt[[1]]
filt_ileum_taxa_tab <- data_filt[[2]]
nearzero_step <- dim(filt_ileum_asv_tab)[1]
```

Library size

```{r, fig.width=5, fig.height=4, fig.fullwidth=TRUE, warning=FALSE}
read_counts(filt_ileum_asv_tab,line = c(5000,10000))
```

### Final Counts

```{r, warning=FALSE}
final_counts_filtering(ileum_asv_tab,
                       filt_ileum_asv_tab,
                       filt_ileum_metadata,
                       seq_step, 0, nearzero_step) %>% `colnames<-`("Count")
```

## Alpha diversity

```{r, warning=FALSE}
path = "../../../results/merged_sites/main_analysis/Q5/alpha_diversity"
```

**Calculation**

```{r, warning=FALSE}
# Construct MPSE object
alpha_ileum_metadata$Sample <- alpha_ileum_metadata$SampleID
ileum_mpse <- as.MPSE(construct_phyloseq(alpha_ileum_asv_tab,
                                         alpha_ileum_taxa_tab,
                                         alpha_ileum_metadata))

ileum_mpse %<>% mp_rrarefy(raresize = 10000,seed = 123)

# Calculate alpha diversity - rarefied counts
ileum_mpse %<>% mp_cal_alpha(.abundance=RareAbundance, force=TRUE)
```

```{r, warning=FALSE}

alpha_div_plots <- list()

# preparing data frame
alpha_data <- data.frame(SampleID=ileum_mpse$Sample.x,
                         Observe=ileum_mpse$Observe,
                         Shannon=ileum_mpse$Shannon,
                         Simpson=ileum_mpse$Simpson,
                         Pielou=ileum_mpse$Pielou,
                         Group=ileum_mpse$Calprotectin_groups,
                         Country=ileum_mpse$Country,
                         Patient=ileum_mpse$Patient)

write.csv(alpha_data,file.path(path,paste0("alpha_indices_",segment,".csv")),
          row.names = FALSE)

alpha_data %<>% drop_na()
```

### Plots

```{r, fig.width=4, fig.height=5, fig.fullwidth=TRUE}
p_boxplot_alpha <- alpha_diversity_countries(alpha_data,
                                             show_legend = FALSE,
                                             size=1,gap=FALSE)

# save the results
alpha_div_plots[[paste(segment,"Country")]] <- p_boxplot_alpha

# see the results
p_boxplot_alpha
```

```{r,results='hide', echo=FALSE}
pdf("../../../figures/merged_sites/Q5/alpha_diversity_terminal_ileum.pdf",
    height =4,width = 6)
p_boxplot_alpha
dev.off()
```


### Linear Model

```{r, warning=FALSE}
path = "../../../results/merged_sites/main_analysis/Q5/alpha_diversity"
alpha_data <- read.csv(file.path(path,paste0("alpha_indices_",segment,".csv")))
```

**Richness**

```{r, warning=FALSE}
results_model <- pairwise.lm(formula = "Observe ~ Group * Country",
                             factors=alpha_data$Group,
                             data=alpha_data)

# check interaction
if (!is.data.frame(results_model)){
  results_model_observe <- results_model[[1]]
  results_model_observe_emeans <- results_model[[2]]
} else {
  results_model_observe <- results_model
  results_model_observe_emeans <- NA
}

# save the results
pc_observed <- list(); 
pc_observed[[segment]] <- results_model_observe
```

```{r, warning=FALSE}
# see the results
knitr::kable(results_model_observe,digits = 3,
caption = "Raw results of linear model of richness estimation.")

knitr::kable(results_model_observe_emeans,digits = 3,
caption = "Raw results of independent country analysis")
```

**Shannon**

```{r, warning=FALSE}
results_model <- pairwise.lm(formula = "Shannon ~ Group * Country",
                             factors=alpha_data$Group,
                             data=alpha_data)

# check interaction
if (!is.data.frame(results_model)){
  results_model_shannon <- results_model[[1]]
  results_model_shannon_emeans <- results_model[[2]]
} else {
  results_model_shannon <- results_model
  results_model_shannon_emeans <- NA
}

# save the results
pc_shannon <- list(); 
pc_shannon[[segment]] <- as.data.frame(results_model_shannon)

```

```{r, warning=FALSE}
# see the results
knitr::kable(results_model_shannon,digits = 3,
caption = "Raw results of linear model of Shannon estimation.")

knitr::kable(results_model_shannon_emeans,digits = 3,
caption = "Raw results of independent country analysis")
```

**Simpson**

```{r, warning=FALSE}
results_model <- pairwise.lm(formula = "Simpson ~ Group * Country",
                                     factors=alpha_data$Group,
                                     data=alpha_data)

# check interaction
if (!is.data.frame(results_model)){
  results_model_simpson <- results_model[[1]]
  results_model_simpson_emeans <- results_model[[2]]
} else {
  results_model_simpson <- results_model
  results_model_simpson_emeans <- NA
}


# save the results
pc_simpson <- list(); 
pc_simpson[[segment]] <- as.data.frame(results_model_simpson)
```

```{r, warning=FALSE}
# see the results
knitr::kable(results_model_simpson,digits = 3,
caption = "Raw results of linear model of Simpson estimation.")

knitr::kable(results_model_simpson_emeans,digits = 3,
caption = "Raw results of independent country analysis")
```

### Saving results

```{r, warning=FALSE}
alpha_list <- list(
  Richness=pc_observed[[segment]],
  Shannon=pc_shannon[[segment]])
                   
write.xlsx(alpha_list, 
           file = file.path(path,paste0("alpha_diversity_results_",segment,".xlsx")))
```

## Beta diversity

Calculating Aitchison distance (euclidean distance on clr-transformed
data), both at ASV and genus level.

### Main analysis

**Genus level, Aitchison distance**

```{r, warning=FALSE}
level="genus"
```

```{r, warning=FALSE}
path = "../../../results/merged_sites/main_analysis/Q5/beta_diversity"
```

```{r, warning=FALSE}
pairwise_aitchison_raw <- list()
pca_plots_list <- list()
```

Aggregation, filtering

```{r, warning=FALSE}
# Aggregation
genus_data <- aggregate_taxa(ileum_asv_tab,
                             ileum_taxa_tab,
                             taxonomic_level=level,
                             names=TRUE)
# Filtration
filt_data <- filtering_steps(genus_data[[1]],
                             genus_data[[2]],
                             ileum_metadata,
                             seq_depth_threshold=10000)

filt_ileum_genus_tab <- filt_data[[1]]
filt_ileum_genus_taxa <- filt_data[[2]]
filt_ileum_metadata <- filt_data[[3]] %>% dplyr::mutate(Group=Calprotectin_groups)
```

##### PERMANOVA

```{r, warning=FALSE}
pairwise_df <- filt_ileum_genus_tab %>% column_to_rownames("SeqID") %>% t()

# main effect
pp_main <- pairwise.adonis(pairwise_df,
                           filt_ileum_metadata$Group,
                           covariate = filt_ileum_metadata$Country, sim.method = "robust.aitchison", p.adjust.m="BH")

# interaction
pp_int <- pairwise.adonis(pairwise_df,filt_ileum_metadata$Group,covariate = filt_ileum_metadata$Country, interaction = TRUE, sim.method = "robust.aitchison", p.adjust.m="BH")

# tidy the results
pp_factor <- pp_main[[1]]
pp_cov <- pp_main[[2]]
pp_fac.cov <- pp_int[[3]]

cols <- c("pairs","Df","SumsOfSqs", "F.Model","R2","p.value", "p.adjusted", "sig")
colnames(pp_factor) <- cols; colnames(pp_cov) <- cols; colnames(pp_fac.cov) <- cols; 

# save raw results
pairwise_aitchison_raw[[paste(level, segment)]] <- rbind(pp_factor,pp_cov,pp_fac.cov)
```

```{r, warning=FALSE}
# see the results
knitr::kable(pp_factor,digits = 3,caption = "PERMANOVA, GROUP separation")
knitr::kable(pp_cov,digits = 3,caption = "PERMANOVA, COUNTRY separation")
knitr::kable(pp_fac.cov,digits = 3,caption = "PERMANOVA, INTERACTION GROUP:Country")
```

Interaction check

```{r, warning=FALSE}
interaction_sig <- pp_fac.cov$pairs[pp_fac.cov$p.adjusted < 0.1]

if (length(interaction_sig)>0){
  for (i in 1:length(interaction_sig)){
  group1 <- unlist(strsplit(interaction_sig[i],split = " vs "))[1]
  group2 <- unlist(strsplit(interaction_sig[i],split = " vs "))[2]
  group2 <- unlist(strsplit(group2,split = " : "))[1]
  
  result_list <- adonis_postanalysis(x=pairwise_df,
                      factors = filt_ileum_metadata$v,
                      covariate = filt_ileum_metadata$Country, 
                      group1 = group1,
                      group2 = group2)
  print(result_list)
}
}

```

##### Plots

**PCA**

```{r, warning=FALSE}
p <- pca_plot_custom(filt_ileum_genus_tab,
                                 filt_ileum_genus_taxa,
                                 filt_ileum_metadata,
                                 show_boxplots = TRUE,
                                 variable = "Group", size=3, show_legend=FALSE)

# save the results
pca_plots_list[[paste(segment,level,"custom")]] <- p

# see the results
p
```

```{r,results='hide', echo=FALSE}
pdf("../../../figures/merged_sites/Q5/beta_diversity_terminal_ileum.pdf",
    height =5,width = 5)
p
dev.off()
```

#### Saving results

```{r, warning=FALSE}
write.xlsx(pairwise_aitchison_raw[[paste(level, segment)]], 
           file = file.path(path,
           paste0("beta_diversity_results_", segment,".xlsx")))
```

## Univariate Analysis

### Main analysis

```{r, warning=FALSE}
level="genus"
```

```{r, warning=FALSE}
# needed paths
path = "../../../results/merged_sites/main_analysis/Q5/univariate_analysis"
path_maaslin=file.path("../../../intermediate_files/maaslin/Q5",level)
```

```{r, warning=FALSE}
# variables
raw_linda_results_genus <- list();
raw_linda_results_genus[[segment]] <- list()
linda_results_genus <- list(); 
linda_results_genus[[segment]] <- list()

# country and interaction problems
list_intersections <- list()
list_venns <- list()
uni_statistics <- list()

# workbook for final df
wb <- createWorkbook()

```

Aggregate taxa

```{r, warning=FALSE}
genus_data <- aggregate_taxa(ileum_asv_tab,
                             ileum_taxa_tab,
                             taxonomic_level = level)

ileum_genus_tab <- genus_data[[1]]
ileum_genus_taxa_tab <- genus_data[[2]]

ileum_genus_asv_taxa_tab <- create_asv_taxa_table(ileum_genus_tab,
                                                  ileum_genus_taxa_tab)
```

#### low vs high

```{r, warning=FALSE}
group <- c("low","high")
comparison_name <- paste0(group[1], " vs ",group[2])
```

```{r, echo=FALSE}
comparison_title <- sub("(\\w+) vs (\\w+)", "\\2 and \\1", comparison_name)
```

##### linDA

```{r, fig.width=15, fig.height=5, fig.fullwidth=TRUE}

ileum_metadata %<>% dplyr::mutate(Group=Calprotectin_groups)

# prepare the data
linda_data <- binomial_prep(ileum_genus_tab,
                            ileum_genus_taxa_tab,
                            ileum_metadata,
                            group, usage="linDA")

filt_ileum_uni_data <- linda_data[[1]]
filt_ileum_uni_taxa <- linda_data[[2]]
filt_ileum_uni_metadata <- linda_data[[3]]

# fit the model
linda.obj <- linda(filt_ileum_uni_data, 
                   filt_ileum_uni_metadata, 
                   formula = '~ Group * Country')

linda.output <- linda.obj$output
linda.output <- linda_renaming(linda.output, group)

# save the results
group1 <- paste0(group[1], " vs ","Group",group[2])
group2 <- paste0(group[1], " , ",group[2], " - CZ vs NO") 
group3 <- paste0(group[1], " vs ","Group",group[2], ":CountryNO")

for (grp in c(group1,group2,group3)){
  raw_linda_results_genus[[segment]][[grp]] <- 
    rawlinda.df(linda.output,
                grp,
                filt_ileum_uni_data,
                filt_ileum_uni_taxa)
  
  linda_results_genus[[segment]][[grp]] <- 
    linda.df(linda.output,
             grp,
             filt_ileum_uni_data,
             filt_ileum_uni_taxa)
}

```

```{r, fig.width=15, fig.height=5, fig.fullwidth=TRUE}
# volcano plot
volcano_1 <- volcano_plot_linda(linda.output, 
                                group1, 
                                taxa_table = filt_ileum_uni_taxa) + 
            ggtitle(comparison_title)

volcano_2  <- volcano_plot_linda(linda.output, group2, 
                                taxa_table = filt_ileum_uni_taxa) + 
            ggtitle("NO vs CZ")

volcano_3 <- volcano_plot_linda(linda.output, group3, 
                                taxa_table = filt_ileum_uni_taxa) +
            ggtitle("Interaction effect")

volcano <- ggarrange(volcano_1,volcano_2,volcano_3, ncol=3)
volcano
```

##### MaAsLin2

```{r, echo=FALSE,results='hide',message=FALSE,warning=FALSE}
log_file <- file(tempfile(), open = "wt")
sink(log_file)  # Redirect standard output
sink(log_file, type = "message")  # Redirect messages and warnings

fit_data = Maaslin2(
    input_data = filt_ileum_uni_data, 
    input_metadata = filt_ileum_uni_metadata, min_abundance = 0,
    min_prevalence = 0,min_variance = 0,
    output = file.path(path_maaslin,group1), max_significance = 0.05,
    fixed_effects = c('Group', 'Country'),correction = "BH")

sink()  # Restore standard output
sink(type = "message")  # Restore messages
```

```{r, warning=FALSE}
volcano1 <- volcano_plot_maaslin(fit_data,filt_ileum_uni_taxa) + 
            ggtitle(comparison_title)

volcano2 <- volcano_plot_maaslin(fit_data,filt_ileum_uni_taxa,variable="Country") + 
            ggtitle("Country effect")

volcano <- ggarrange(volcano1,volcano2, ncol=2)
volcano
```

##### Group - Intersection

```{r, warning=FALSE}
intersection_results <- group_intersection(group, list_intersections, list_venns,
                                           linda.output, fit_data,
                                           raw_linda_results_genus, 
                                           segment = segment,
                                           level=level)

list_intersections <- intersection_results[[1]]
list_venns <- intersection_results[[2]]
venn <- intersection_results[[3]]

# show the results
venn
```

##### Interaction effect

```{r, warning=FALSE}
list_interaction_significant <- country_interaction(group,
                                                    linda.output, 
                                                    list_intersections,
                                          filt_ileum_uni_data,
                                          filt_ileum_uni_metadata,
                                          segment=segment,
                                          level=level)

# see the result
## significant interaction effect
list_interaction_significant[[1]]

## results for czech cohort
list_interaction_significant[[2]]

## results for norwegian cohort
list_interaction_significant[[3]]
```

Removing problematic taxa

```{r, warning=FALSE}
list_intersections <- removing_interaction_problems(group,
                                                    list_interaction_significant,
                                                    list_intersections,
                                                    segment=segment,
                                                    level=level)
```

##### Basic statistics

```{r, warning=FALSE}
uni_df <-  merge(basic_univariate_statistics(linda_data,group),
                 raw_linda_results_genus[[segment]][[group1]],
                 by="SeqID",all=TRUE)

uni_df[["final_sig"]] <- uni_df$SeqID %in% list_intersections[[paste(segment,level,comparison_name)]][["SeqID"]]
uni_statistics[[segment]][[paste(level,comparison_name)]] <- uni_df

# for comparison
new_name <- comparison_name
addWorksheet(wb, sheetName = new_name)
writeData(wb, sheet = new_name, uni_df, rowNames=FALSE)
```

#### Visualization

Heatmap visualizing the linDA's logFoldChange for taxa with p \< 0.1.

```{r, fig.width=10, fig.height=17, fig.fullwidth=TRUE}
list_heatmap <- list_intersections[grep(paste(segment,level),
                                  names(list_intersections),value=TRUE)]

if (length((list_heatmap[[1]][[1]]))>1){
  p_heatmap_linda <- heatmap_linda(list_heatmap,ileum_taxa_tab)
  p_heatmap_linda
}

```

Dotheatmap

```{r, warning=FALSE, fig.height=5, fig.width=4}
if (length((list_heatmap[[1]][[1]]))>0){
dotheatmap_linda <- dot_heatmap_linda(list_heatmap,
                                      uni_statistics$terminal_ileum[grepl(level,names(uni_statistics$terminal_ileum))],
                                      ileum_taxa_tab)
dotheatmap_linda

}
```

**Horizontal bar plot**

```{r, warning=FALSE, fig.height=5, fig.width=4}
if (length((list_heatmap[[1]][[1]]))>0){
p_prevalence <- horizontal_barplot(wb,taxa=levels(dotheatmap_linda$data$SeqID))

p_prevalence_final <- ggarrange(p_prevalence,
                                ggplot() + theme_minimal(),
                                nrow = 2,heights = c(1,0.085))
p <- ggarrange(dotheatmap_linda + theme(legend.position = "none"),p_prevalence_final,ncol=2,widths = c(1,0.2))
p
}
```

```{r,results='hide',eval=FALSE, echo=FALSE}
pdf("../../../figures/merged_sites/Q5/dotplot_terminal_ileum.pdf",
    height =10,width = 4)
p
dev.off()
```

#### Saving results

```{r, warning=FALSE}
# ALL DATA
saveWorkbook(wb,file.path(path,paste0("uni_analysis_wb_",segment,".xlsx")),
             overwrite = TRUE)

# SIGNIFICANT taxa

write.xlsx(list_intersections[grepl(segment,names(list_intersections))] %>%
            `names<-`(gsub(segment, "", names(
              list_intersections[grepl(segment,names(list_intersections))]))),
           file.path(path,paste0("significant_taxa_",segment,".xlsx")))
```


## Machine learning

```{r, warning=FALSE}
path = "../../../results/merged_sites/main_analysis/Q5/models"
```

### ElasticNet

```{r, warning=FALSE}
model="enet"
```

#### ASV level

```{r, warning=FALSE}
level="ASV"
```

##### low vs high

```{r, warning=FALSE}
group <- c("low","high")
comparison_name <- paste0(group[1], " vs ",group[2])
```

```{r, warning=FALSE}
model_name <- paste(comparison_name,level,segment)

# prepare the data
filt_ileum_uni_data <- binomial_prep(ileum_asv_tab,
                                     ileum_taxa_tab,
                                     ileum_metadata,
                                     group, usage="ml_clr")

# fit the model
enet_model <- glmnet_binomial(filt_ileum_uni_data,
                              sample_method = "atypboot",
                              outcome="Group", 
                              N=10,
                              reuse=TRUE,
                              file=model_name,
                              Q="Q5")

# ROC curve
roc_c <- roc_curve(enet_model, group)

# save the results
models_summ <- list()
models_cm <- list()
betas <- list()
roc_cs <- list()

models_summ[[model_name]] <- enet_model$model_summary
models_cm[[model_name]] <- enet_model$conf_matrices$original
roc_cs[[model_name]] <- enet_model$kfold_rocobjs
betas[[model_name]] <- as.matrix(enet_model$betas)

# see the results
enet_model$model_summary %>% t()
enet_model$conf_matrices
enet_model$plot
roc_c
```

#### Genus level

```{r, warning=FALSE}
level="genus"
```

Aggregate taxa

```{r, warning=FALSE}
genus_data <- aggregate_taxa(ileum_asv_tab,
                             ileum_taxa_tab,
                             taxonomic_level = level)

ileum_genus_tab <- genus_data[[1]]
ileum_genus_taxa_tab <- genus_data[[2]]
```

##### low vs high

```{r, warning=FALSE}
group <- c("low","high")
comparison_name <- paste0(group[1], " vs ",group[2])
```

```{r, warning=FALSE}
model_name <- paste(comparison_name,level,segment)

# prepare the data
filt_ileum_uni_data <- binomial_prep(ileum_genus_tab,
                                     ileum_genus_taxa_tab,
                                     ileum_metadata,
                                     group, 
                                     usage="ml_clr")

# fit the model
enet_model <- glmnet_binomial(filt_ileum_uni_data,
                              sample_method = "atypboot",
                              outcome="Group", 
                              N=10,
                              reuse=TRUE,
                              file=model_name,
                              Q="Q5")

# ROC
roc_c <- roc_curve(enet_model, group)

# save the results
models_summ[[model_name]] <- enet_model$model_summary
models_cm[[model_name]] <- enet_model$conf_matrices$original
roc_cs[[model_name]] <- enet_model$kfold_rocobjs
betas[[model_name]] <- as.matrix(enet_model$betas)

# see the results
enet_model$model_summary %>% t()
enet_model$conf_matrices
enet_model$plot

roc_c
```

#### Saving results

```{r, warning=FALSE}
models_summ_df_ileum <- do.call(rbind, 
  models_summ[grep(segment,names(models_summ),value = TRUE)])

write.csv(models_summ_df_ileum,file.path(path,paste0("elastic_net_",segment,".csv")))
```

## Results overview

#### Alpha diversity

```{r, warning=FALSE}
pc_observed[[segment]]
pc_shannon[[segment]]
pc_simpson[[segment]]
```

Plots

```{r, fig.width=4, fig.height=5, fig.fullwidth=TRUE}
alpha_div_plots[[paste(segment,"Country")]]
```


#### Beta diversity

**Main results**

```{r, warning=FALSE}
pairwise_aitchison_raw[[paste("genus", segment)]]
```

*PCA*

```{r, fig.width=5, fig.height=3, fig.fullwidth=TRUE, eval=TRUE}
pca_plots_list[[paste(segment,"genus custom")]]
```


#### Univariate analysis

**Number of significant taxa**

```{r, warning=FALSE}
knitr::kable(as.data.frame(lapply(list_intersections,nrow)) %>% t() %>% 
  `colnames<-`("Count") %>% 
  `rownames<-`(c(names(list_intersections))),caption="Number of significant taxa")
```

#### Machine learning

**Main models**

```{r, warning=FALSE}
knitr::kable(models_summ_df_ileum %>% dplyr::select(
"alpha","lambda",
"auc_optimism_corrected",
"auc_optimism_corrected_CIL",
"auc_optimism_corrected_CIU"),
             digits=2,caption="Elastic net results")
```

**ROC - Genus level**

```{r, warning=FALSE}
p <- roc_curve_all_custom(roc_cs[2],Q="Q5",
                     model_name="enet_model")

p
```


```{r,results='hide', echo=FALSE, warning=FALSE}
pdf("../../../figures/merged_sites/Q5/enet_terminal_ileum.pdf",
    height =5,width = 5)
p
dev.off()
```

# Data Analysis - Colon

```{r, warning=FALSE}
segment="colon"
```

## Filtering

Rules: 

  - sequencing depth \> 10000 
  
  - nearZeroVar() with default settings 

**Rarefaction Curve**

```{r, warning=FALSE}
path="../../../intermediate_files/rarecurves"
seq_depth_threshold <- 10000
```

```{r, eval = FALSE}
ps <- construct_phyloseq(colon_asv_tab,colon_taxa_tab,colon_metadata)
rareres <- get_rarecurve(obj=ps, chunks=500)
save(rareres,file = file.path(path,"rarefaction_colon.Rdata"))
```

```{r, warning=FALSE}
load(file.path(path,"rarefaction_colon.Rdata"))

prare <- ggrarecurve(obj=rareres,
                      factorNames="Country",
                      indexNames=c("Observe")) + 
  theme_bw() +
  theme(axis.text=element_text(size=8), panel.grid=element_blank(),
        strip.background = element_rect(colour=NA,fill="grey"),
        strip.text.x = element_text(face="bold")) + 
  geom_vline(xintercept = seq_depth_threshold, 
             linetype="dashed", 
             color = "red") + 
  xlim(0, 20000)

prare
```

**Library size**

```{r, fig.width=5, fig.height=4, fig.fullwidth=TRUE, warning=FALSE}
read_counts(colon_asv_tab, line = c(5000,10000))
```

### Sequencing depth

```{r, warning=FALSE}
data_filt <- seq_depth_filtering(colon_asv_tab,
                                 colon_taxa_tab,
                                 colon_metadata,
                                 seq_depth_threshold = 10000)

filt_colon_asv_tab <- data_filt[[1]]; alpha_colon_asv_tab <- filt_colon_asv_tab
filt_colon_taxa_tab <- data_filt[[2]]; alpha_colon_taxa_tab <- filt_colon_taxa_tab
filt_colon_metadata <- data_filt[[3]]; alpha_colon_metadata <- filt_colon_metadata

seq_step <- dim(filt_colon_asv_tab)[1]
```

**Library size**

```{r, fig.width=5, fig.height=4, fig.fullwidth=TRUE, warning=FALSE}
read_counts(filt_colon_asv_tab,line = c(5000,10000))
```

### NearZeroVar

```{r, warning=FALSE}
data_filt <- nearzerovar_filtering(filt_colon_asv_tab, 
                                   filt_colon_taxa_tab,
                                   filt_colon_metadata)

filt_colon_asv_tab <- data_filt[[1]]
filt_colon_taxa_tab <- data_filt[[2]]
nearzero_step <- dim(filt_colon_asv_tab)[1]
```

Library size

```{r, fig.width=5, fig.height=4, fig.fullwidth=TRUE, warning=FALSE}
read_counts(filt_colon_asv_tab,line = c(5000,10000))
```

### Final Counts

```{r, warning=FALSE}
final_counts_filtering(colon_asv_tab,
                       filt_colon_asv_tab,
                       filt_colon_metadata,
                       seq_step, 0, nearzero_step) %>% `colnames<-`("Count")
```

## Alpha diversity

```{r, warning=FALSE}
path = "../../../results/merged_sites/main_analysis/Q5/alpha_diversity"
```

**Calculation**

```{r, warning=FALSE}
# Construct MPSE object
alpha_colon_metadata$Sample <- alpha_colon_metadata$SampleID
colon_mpse <- as.MPSE(construct_phyloseq(alpha_colon_asv_tab,
                                         alpha_colon_taxa_tab,
                                         alpha_colon_metadata))

colon_mpse %<>% mp_rrarefy(raresize = 10000,seed = 123)

# Calculate alpha diversity - rarefied counts
colon_mpse %<>% mp_cal_alpha(.abundance=RareAbundance, force=TRUE)
```

```{r, warning=FALSE}

# preparing data frame
alpha_data <- data.frame(SampleID=colon_mpse$Sample.x,
                         Observe=colon_mpse$Observe,
                         Shannon=colon_mpse$Shannon,
                         Simpson=colon_mpse$Simpson,
                         Pielou=colon_mpse$Pielou,
                         Group=colon_mpse$Calprotectin_groups,
                         Country=colon_mpse$Country,
                         Patient=colon_mpse$Patient)

write.csv(alpha_data,file.path(path,paste0("alpha_indices_",segment,".csv")),
          row.names = FALSE)

alpha_data %<>% drop_na()
```

### Plots

```{r, fig.width=6, fig.height=4, fig.fullwidth=TRUE}
p_boxplot_alpha <- alpha_diversity_countries(alpha_data,
                                             show_legend = FALSE,
                                             size=1,gap=FALSE)

# save the results
alpha_div_plots[[paste(segment,"Country")]] <- p_boxplot_alpha

# see the results
p_boxplot_alpha
```

```{r,results='hide', echo=FALSE}
pdf("../../../figures/merged_sites/Q5/alpha_diversity_colon.pdf",
    height =4,width = 6)
p_boxplot_alpha
dev.off()
```


### Linear Model

```{r, warning=FALSE}
path = "../../../results/merged_sites/main_analysis/Q5/alpha_diversity"
alpha_data <- read.csv(file.path(path,paste0("alpha_indices_",segment,".csv")))
```

**Richness**

```{r, warning=FALSE}
results_model <- pairwise.lmer(formula = "Observe ~ Group * Country +  (1|Patient)",
                             factors=alpha_data$Group,
                             data=alpha_data)

# check interaction
if (!is.data.frame(results_model)){
  results_model_observe <- results_model[[1]]
  results_model_observe_emeans <- results_model[[2]]
} else {
  results_model_observe <- results_model
  results_model_observe_emeans <- NA
}

# save the results
pc_observed[[segment]] <- results_model_observe
```

```{r, warning=FALSE}
# see the results
knitr::kable(results_model_observe,digits = 3,
caption = "Raw results of linear model of richness estimation.")

knitr::kable(results_model_observe_emeans,digits = 3,
caption = "Raw results of independent country analysis")
```

**Shannon**

```{r, warning=FALSE}
results_model <- pairwise.lmer(formula = "Shannon ~ Group * Country + (1|Patient)",
                             factors=alpha_data$Group,
                             data=alpha_data)

# check interaction
if (!is.data.frame(results_model)){
  results_model_shannon <- results_model[[1]]
  results_model_shannon_emeans <- results_model[[2]]
} else {
  results_model_shannon <- results_model
  results_model_shannon_emeans <- NA
}

# save the results
pc_shannon[[segment]] <- as.data.frame(results_model_shannon)

```

```{r, warning=FALSE}
# see the results
knitr::kable(results_model_shannon,digits = 3,
caption = "Raw results of linear model of Shannon estimation.")

knitr::kable(results_model_shannon_emeans,digits = 3,
caption = "Raw results of independent country analysis")
```

**Simpson**

```{r, warning=FALSE}
results_model <- pairwise.lmer(formula = "Simpson ~ Group * Country + (1|Patient)",
                                     factors=alpha_data$Group,
                                     data=alpha_data)

# check interaction
if (!is.data.frame(results_model)){
  results_model_simpson <- results_model[[1]]
  results_model_simpson_emeans <- results_model[[2]]
} else {
  results_model_simpson <- results_model
  results_model_simpson_emeans <- NA
}


# save the results
pc_simpson[[segment]] <- as.data.frame(results_model_simpson)
```

```{r, warning=FALSE}
# see the results
knitr::kable(results_model_simpson,digits = 3,
caption = "Raw results of linear model of Simpson estimation.")

knitr::kable(results_model_simpson_emeans,digits = 3,
caption = "Raw results of independent country analysis")
```

### Saving results

```{r, warning=FALSE}
alpha_list <- list(
  Richness=pc_observed[[segment]],
  Shannon=pc_shannon[[segment]])
                   
write.xlsx(alpha_list, 
           file = file.path(path,paste0("alpha_diversity_results_",segment,".xlsx")))
```


## Beta diversity

Calculating Aitchison distance (euclidean distance on clr-transformed
data), both at ASV and genus level.

### Main analysis

**Genus level, Aitchison distance**

```{r, warning=FALSE}
level="genus"
```

```{r, warning=FALSE}
path = "../../../results/merged_sites/main_analysis/Q5/beta_diversity"
```

Aggregation, filtering

```{r, warning=FALSE}
# Aggregation
genus_data <- aggregate_taxa(colon_asv_tab,
                             colon_taxa_tab,
                             taxonomic_level=level,
                             names=TRUE)
# Filtration
filt_data <- filtering_steps(genus_data[[1]],
                             genus_data[[2]],
                             colon_metadata,
                             seq_depth_threshold=10000)

filt_colon_genus_tab <- filt_data[[1]]
filt_colon_genus_taxa <- filt_data[[2]]
filt_colon_metadata <- filt_data[[3]]

filt_colon_metadata %<>% dplyr::mutate(Group=Calprotectin_groups)
```

##### PERMANOVA

```{r, warning=FALSE}
pairwise_df <- filt_colon_genus_tab %>% column_to_rownames("SeqID") %>% t()

# main effect
pp_main <- pairwise.adonis(pairwise_df,filt_colon_metadata$Group,
                           covariate = filt_colon_metadata$Country, 
                           patients = filt_colon_metadata$Patient,
                           sim.method = "robust.aitchison", p.adjust.m="BH")

# interaction
pp_int <- pairwise.adonis(pairwise_df,filt_colon_metadata$Group,
                          covariate = filt_colon_metadata$Country, 
                          interaction = TRUE, 
                          patients = filt_colon_metadata$Patient,
                          sim.method = "robust.aitchison", p.adjust.m="BH")

# tidy the results
pp_factor <- pp_main[[1]]
pp_cov <- pp_main[[2]]
pp_fac.cov <- pp_int[[3]]

cols <- c("pairs","Df","SumsOfSqs", "F.Model","R2","p.value", "p.adjusted", "sig")
colnames(pp_factor) <- cols; colnames(pp_cov) <- cols; colnames(pp_fac.cov) <- cols; 

# save raw results
pairwise_aitchison_raw[[paste(level, segment)]] <-rbind(pp_factor,pp_cov,pp_fac.cov)

```

```{r, warning=FALSE}
# see the results
knitr::kable(pp_factor,digits = 3,caption = "PERMANOVA, GROUP separation")
knitr::kable(pp_cov,digits = 3,caption = "PERMANOVA, COUNTRY separation")
knitr::kable(pp_fac.cov,digits = 3,caption = "PERMANOVA, INTERACTION GROUP:Country")
```

Interaction check

```{r, warning=FALSE}
interaction_sig <- pp_fac.cov$pairs[pp_fac.cov$p.adjusted < 0.1]

if (length(interaction_sig)>0){
 for (i in 1:length(interaction_sig)){
  group1 <- unlist(strsplit(interaction_sig[i],split = " vs "))[1]
  group2 <- unlist(strsplit(interaction_sig[i],split = " vs "))[2]
  group2 <- unlist(strsplit(group2,split = " : "))[1]
  
  result_list <- adonis_postanalysis(x=pairwise_df,
                      factors = filt_colon_genus_metadata$Group,
                      covariate = filt_colon_genus_metadata$Country, 
                      group1 = group1,
                      group2 = group2,
                      patients = filt_colon_genus_metadata$Patient)
  print(result_list)
} 
}

```

##### Plots

**PCoA custom**

```{r, warning=FALSE}
p <- pca_plot_custom(filt_colon_genus_tab,
                                 filt_colon_genus_taxa,
                                 filt_colon_metadata,
                                 show_boxplots = TRUE,
                                 variable = "Group", size=2, 
                                 show_legend=FALSE)

# save the results
pca_plots_list[[paste(segment,level,"custom")]] <- p

# see the results
p

```

```{r,results='hide', echo=FALSE}
pdf("../../../figures/merged_sites/Q5/beta_diversity_colon.pdf",
    height =5,width = 5)
p
dev.off()
```

#### Saving results

```{r, warning=FALSE}
write.xlsx(pairwise_aitchison_raw[[paste(level, segment)]], 
           file = file.path(path,
           paste0("beta_diversity_results_", segment,".xlsx")))
```

## Univariate Analysis

### Main - Genus level

```{r, warning=FALSE}
level="genus"
```

```{r, warning=FALSE}
# needed paths
path = "../../../results/merged_sites/main_analysis/Q5/univariate_analysis"
path_maaslin=file.path("../../../intermediate_files/maaslin/Q5",level)
```

```{r, warning=FALSE}
# variables
raw_linda_results_genus[[segment]] <- list()
linda_results_genus[[segment]] <- list()

# country and interaction problems
list_intersections <- list()
list_venns <- list()
uni_statistics <- list()

# workbook for final df
wb <- createWorkbook()

```

#### Genus level

```{r, warning=FALSE}
level="genus"
```

Aggregate taxa

```{r, warning=FALSE}
genus_data <- aggregate_taxa(colon_asv_tab,
                             colon_taxa_tab,
                             taxonomic_level = level)

colon_genus_tab <- genus_data[[1]]
colon_genus_taxa_tab <- genus_data[[2]]

colon_genus_asv_taxa_tab <- create_asv_taxa_table(colon_genus_tab,
                                                  colon_genus_taxa_tab)
```

##### low vs high

```{r, warning=FALSE}
group <- c("low","high")
comparison_name <- paste0(group[1], " vs ",group[2])
```

```{r, echo=FALSE}
comparison_title <- sub("(\\w+) vs (\\w+)", "\\2 and \\1", comparison_name)
```

###### linda

```{r, fig.width=15, fig.height=5, fig.fullwidth=TRUE}

colon_metadata %<>% dplyr::mutate(Group=Calprotectin_groups)

# prepare the data
linda_data <- binomial_prep(colon_genus_tab,
                            colon_genus_taxa_tab,
                            colon_metadata,
                            group, usage="linDA")

filt_colon_uni_data <- linda_data[[1]]
filt_colon_uni_taxa <- linda_data[[2]]
filt_colon_uni_metadata <- linda_data[[3]]

# fit the model
linda.obj <- linda(filt_colon_uni_data,
                   filt_colon_uni_metadata,
                   formula = '~ Group * Country + (1|Patient)')

linda.output <- linda.obj$output
linda.output <- linda_renaming(linda.output, group)

# save the results
group1 <- paste0(group[1], " vs ","Group",group[2])
group2 <- paste0(group[1], " , ",group[2], " - CZ vs NO") 
group3 <- paste0(group[1], " vs ","Group",group[2], ":CountryNO")


for (grp in c(group1,group2,group3)){
  raw_linda_results_genus[[segment]][[grp]] <- 
    rawlinda.df(linda.output,
                grp,
                filt_colon_uni_data,
                filt_colon_uni_taxa)
  
  linda_results_genus[[segment]][[grp]] <- 
    linda.df(linda.output,
             grp,
             filt_colon_uni_data,
             filt_colon_uni_taxa)
}
```

```{r, fig.width=15, fig.height=5, fig.fullwidth=TRUE}
# volcano plot
volcano_1 <- volcano_plot_linda(linda.output, group1, 
                                taxa_table = filt_colon_uni_taxa) + 
            ggtitle(paste(group,collapse=" vs "))

volcano_2  <- volcano_plot_linda(linda.output, group2, 
                                taxa_table = filt_colon_uni_taxa) + 
            ggtitle("NO vs CZ")

volcano_3 <- volcano_plot_linda(linda.output, group3, 
                                taxa_table = filt_colon_uni_taxa) +
            ggtitle("Interaction effect")

volcano <- ggarrange(volcano_1,volcano_2,volcano_3, ncol=3)

# see the plot
volcano
```

###### MaAsLin2

```{r, echo=FALSE,results='hide',message=FALSE,warning=FALSE}
log_file <- file(tempfile(), open = "wt")
sink(log_file)  # Redirect standard output
sink(log_file, type = "message")  # Redirect messages and warnings

fit_data = Maaslin2(
    input_data = filt_colon_uni_data, 
    input_metadata = filt_colon_uni_metadata, min_abundance = 0,
    min_prevalence = 0,min_variance = 0,
    output = file.path(path_maaslin,group1), max_significance = 0.05,
    fixed_effects = c('Group', 'Country'),random_effects = "Patient",
    correction = "BH")

sink()  # Restore standard output
sink(type = "message")  # Restore messages
```

```{r, warning=FALSE}
volcano1 <- volcano_plot_maaslin(fit_data,filt_colon_uni_taxa) + 
            ggtitle(paste(group[1], "vs", group[2]))

volcano2 <- volcano_plot_maaslin(fit_data,filt_colon_uni_taxa,variable="Country") + 
            ggtitle("Country effect")

volcano <- ggarrange(volcano1,volcano2, ncol=2)
volcano
```

###### Group - Intersection

```{r, warning=FALSE}
intersection_results <- group_intersection(group, list_intersections, list_venns,
                                           linda.output, fit_data,
                                           raw_linda_results_genus, 
                                           segment = segment,
                                           level=level)

list_intersections <- intersection_results[[1]]
list_venns <- intersection_results[[2]]
venn <- intersection_results[[3]]

# show the results
venn
```

###### Interaction effect

```{r, warning=FALSE}
list_interaction_significant <- country_interaction(group,
                                                    linda.output, 
                                                    list_intersections,
                                          filt_colon_uni_data,
                                          filt_colon_uni_metadata,
                                          segment=segment,
                                          level=level)

# see the result
## significant interaction effect
list_interaction_significant[[1]]

## results for czech cohort
list_interaction_significant[[2]]

## results for norwegian cohort
list_interaction_significant[[3]]
```

Removing problematic taxa

```{r, warning=FALSE}
list_intersections <- removing_interaction_problems(group,
                                                    list_interaction_significant,
                                                    list_intersections,
                                                    segment=segment,
                                                    level=level)
```

##### Basic statistics

```{r, warning=FALSE}
uni_df <-  merge(basic_univariate_statistics(linda_data,group),
                 raw_linda_results_genus[[segment]][[group1]],
                 by="SeqID",all=TRUE)

uni_df[["final_sig"]] <- uni_df$SeqID %in% list_intersections[[paste(segment,level,comparison_name)]][["SeqID"]]
uni_statistics[[segment]][[paste(level,comparison_name)]] <- uni_df

# for comparison
new_name <- comparison_name
addWorksheet(wb, sheetName = new_name)
writeData(wb, sheet = new_name, uni_df, rowNames=FALSE)
```

##### Visualization

Heatmap visualizing the linDA's logFoldChange for taxa with p \< 0.1.

```{r, fig.width=10, fig.height=17, fig.fullwidth=TRUE}
list_heatmap <- list_intersections[grep(paste(segment,level),
                                  names(list_intersections),value=TRUE)]
if (length((list_heatmap[[1]][[1]]))>1){
p_heatmap_linda <- heatmap_linda(list_heatmap,colon_taxa_tab)
p_heatmap_linda
}
```

Dot heatmap

```{r, fig.width=5, fig.height=17, fig.fullwidth=TRUE}
if (length((list_heatmap[[1]][[1]]))>0){
dotheatmap_linda <- dot_heatmap_linda(list_heatmap,
                                      uni_statistics$terminal_ileum[grepl(level,names(uni_statistics$terminal_ileum))],
                                      colon_taxa_tab)
dotheatmap_linda
}
```

**Horizontal bar plot**

```{r, warning=FALSE}
if (length((list_heatmap[[1]][[1]]))>0){
p_prevalence <- horizontal_barplot(wb,taxa=levels(dotheatmap_linda$data$SeqID))

p_prevalence_final <- ggarrange(p_prevalence,
                                ggplot() + theme_minimal(),
                                nrow = 2,heights = c(1,0.085))
p <- ggarrange(dotheatmap_linda + theme(legend.position = "none"),p_prevalence_final,ncol=2,widths = c(1,0.2))
p
}
```

```{r,results='hide',eval=FALSE, echo=FALSE}
pdf("../../../figures/merged_sites/Q5/dotplot_colon.pdf",
    height =10,width = 4)
p
dev.off()
```


#### Saving results

```{r, warning=FALSE}
# ALL DATA
saveWorkbook(wb,file.path(path,paste0("uni_analysis_wb_",segment,".xlsx")),
             overwrite = TRUE)

# SIGNIFICANT taxa

write.xlsx(list_intersections[grepl(segment,names(list_intersections))] %>%
            `names<-`(gsub(segment, "", names(
              list_intersections[grepl(segment,names(list_intersections))]))),
           file.path(path,paste0("significant_taxa_",segment,".xlsx")))
```

## Machine learning

```{r, warning=FALSE}
path = "../../../results/merged_sites/main_analysis/Q5/models"
```

### ElasticNet

```{r, warning=FALSE}
model="enet"
```

#### ASV level

```{r, warning=FALSE}
level="ASV"
```

#####low vs high

```{r, warning=FALSE}
group <- c("low","high")
comparison_name <- paste0(group[1], " vs ",group[2])
```

```{r, warning=FALSE}
model_name <- paste(comparison_name,level,segment)

# prepare the data
filt_colon_uni_data <- binomial_prep(colon_asv_tab,
                                     colon_taxa_tab,
                                     colon_metadata,
                                     group, usage="ml_clr",patient = TRUE)

# fit the model
enet_model <- glmnet_binomial(filt_colon_uni_data,
                              sample_method = "atypboot",
                              outcome="Group", 
                              N=10,
                              reuse=TRUE,
                              file=model_name,clust_var = "Patient",
                              Q="Q5")

# ROC curve
roc_c <- roc_curve(enet_model, group)

# save the results
models_summ <- list()
models_cm <- list()
betas <- list()
roc_cs <- list()

models_summ[[model_name]] <- enet_model$model_summary
models_cm[[model_name]] <- enet_model$conf_matrices$original
roc_cs[[model_name]] <- enet_model$kfold_rocobjs
betas[[model_name]] <- as.matrix(enet_model$betas)

# see the results
enet_model$model_summary %>% t()
enet_model$conf_matrices
enet_model$plot
roc_c
```

#### Genus level

```{r, warning=FALSE}
level="genus"
```

Aggregate taxa

```{r, warning=FALSE}
genus_data <- aggregate_taxa(colon_asv_tab,
                             colon_taxa_tab,
                             taxonomic_level = level)

colon_genus_tab <- genus_data[[1]]
colon_genus_taxa_tab <- genus_data[[2]]
```

##### low vs high

```{r, warning=FALSE}
group <- c("low","high")
comparison_name <- paste0(group[1], " vs ",group[2])
```

```{r, warning=FALSE}
model_name <- paste(comparison_name,level,segment)

# prepare the data
filt_colon_uni_data <- binomial_prep(colon_genus_tab,
                                     colon_genus_taxa_tab,
                                     colon_metadata,
                                     group, 
                                     usage="ml_clr",patient = TRUE)

# fit the model
enet_model <- glmnet_binomial(filt_colon_uni_data,
                              sample_method = "atypboot",
                              outcome="Group", 
                              N=10,
                              reuse=TRUE,clust_var = "Patient",
                              file=model_name,
                              Q="Q5")

# ROC
roc_c <- roc_curve(enet_model, group)

# save the results
models_summ[[model_name]] <- enet_model$model_summary
models_cm[[model_name]] <- enet_model$conf_matrices$original
roc_cs[[model_name]] <- enet_model$kfold_rocobjs
betas[[model_name]] <- as.matrix(enet_model$betas)

# see the results
enet_model$model_summary %>% t()
enet_model$conf_matrices
enet_model$plot

roc_c
```

#### Saving results

```{r, warning=FALSE}
models_summ_df_colon <- do.call(rbind, 
  models_summ[grep(segment,names(models_summ),value = TRUE)])

write.csv(models_summ_df_colon,file.path(path,paste0("elastic_net_",segment,".csv")))
```

## Results overview

#### Alpha diversity

```{r, warning=FALSE}
pc_observed[[segment]]
pc_shannon[[segment]]
pc_simpson[[segment]]
```

Plots

```{r, fig.width=4, fig.height=5, fig.fullwidth=TRUE}
alpha_div_plots[[paste(segment,"Country")]]
```

#### Beta diversity

**Main results**

```{r, warning=FALSE}
pairwise_aitchison_raw[[paste("genus", segment)]]
```

PCA

```{r, fig.width=5, fig.height=3, fig.fullwidth=TRUE, eval=TRUE}
pca_plots_list[[paste(segment,"genus custom")]]
```


#### Univariate analysis

**Number of significant taxa**

```{r, warning=FALSE}
knitr::kable(as.data.frame(lapply(list_intersections,nrow)) %>% t() %>% 
  `colnames<-`("Count") %>% 
  `rownames<-`(c(names(list_intersections))),caption="Number of significant taxa")
```


#### Machine learning

**Main models**

Summary

```{r, warning=FALSE}
knitr::kable(models_summ_df_colon %>% dplyr::select(
"alpha","lambda",
"auc_optimism_corrected",
"auc_optimism_corrected_CIL",
"auc_optimism_corrected_CIU"),
             digits=3,caption="Elastic net results")
```

**ROC - Genus level**

```{r, warning=FALSE}
p <- roc_curve_all_custom(roc_cs[2],Q="Q5",
                     model_name="enet_model")

p
```


```{r,results='hide', echo=FALSE, warning=FALSE}
pdf("../../../figures/merged_sites/Q5/enet_colon.pdf",
    height =5,width = 5)
p
dev.off()
```

# Paper-ready visualizations

## Alpha diversity

```{r, warning=FALSE, fig.width=10, fig.height=4, fig.fullwidth=TRUE}
p_A <- alpha_div_plots$`terminal_ileum Country` +
  ggtitle("Terminal ileum")+
  theme(plot.title = element_text(hjust=0.5,face = "bold",size = 15)) 

p_B <-  alpha_div_plots$`colon Country` +
  ggtitle("Colon") +
  theme(plot.title = element_text(hjust=0.5,face = "bold",size = 15)) 

Q5_alpha <- ggarrange(p_A,ggplot() + theme_minimal(),p_B,nrow=1, ncol=3,
                      widths = c(1,0.1,1))
Q5_alpha
```

## Beta diversity

```{r, warning=FALSE, fig.width=10, fig.height=4, fig.fullwidth=TRUE}
pca_ti <- pca_plots_list$`terminal_ileum genus custom` 
pca_colon <- pca_plots_list$`colon genus custom` 

genus_Q5_beta <- ggarrange(pca_ti,
                           ggplot() + theme_minimal(),
                           pca_colon,ncol=3,
                           widths = c(1,0.1,1))
genus_Q5_beta
```

```{r}
save(Q5_alpha,genus_Q5_beta,file="../../../results/merged_sites/main_analysis/Q5/figures.RData")
```

**Alpha + Beta diversity**

```{r, fig.width=10, fig.height=8, fig.fullwidth=TRUE}
alpha_beta <- ggarrange(Q5_alpha,genus_Q5_beta,
                        ncol = 1,nrow=2,labels = c("A","B"))
alpha_beta
```

## Supplement Figure

```{r, fig.width=10, fig.height=8, fig.fullwidth=TRUE}
alpha_beta <- ggarrange(alpha_beta,ggplot() + theme_minimal(),ncol=2,
                                widths = c(1,0.15))

alpha_beta
```

```{r,results='hide', echo=FALSE}
pdf("../../../figures/merged_sites/Q5/Supplement_figure_2B.pdf",paper = "a4",height = 7,width = 12)
alpha_beta
dev.off()
```

**Q3+Q5**

```{r}
load("../../../results/merged_sites/main_analysis/Q3/figures.RData")
```

```{r, warning=FALSE,  fig.width=10, fig.height=12, fig.fullwidth=TRUE}

p <- ggarrange(Q3_alpha,Q5_alpha,
               genus_Q3_beta,genus_Q5_beta,ncol=1,labels = LETTERS)
p <- ggarrange(p,ncol=2,widths = c(1,0.15))

p
```

```{r, warning=FALSE, results='hide', echo=FALSE}
pdf("../../../figures/merged_sites/Q5/Supplementary_Figure2.pdf",paper = "a4",height = 10,width = 12)
p
dev.off()
```

## DAA

```{r,  warning=FALSE, fig.width=10, fig.height=13, fig.fullwidth=TRUE}
if (exists("dot_heatmap_ileum")){
 p_ileum <- dot_heatmap_ileum + 
  ggtitle("Terminal ileum") +
  theme(plot.title = element_text(hjust=0.5,face = "bold",size = 15),
        legend.position = "none") 
}

if (exists("dot_heatmap_colon")){
  p_colon <- dot_heatmap_colon  +
  ggtitle("Colon") +
  theme(plot.title = element_text(hjust=0.5,face = "bold",size = 15),
        legend.position = "none")
}

if (exists("p_ileum") | exists("p_colon")){
  heatmap_plot <- ggarrange(p_ileum,
                          p_colon,
                          ncol = 2,labels=c("A","B"))
  heatmap_plot
}


```

```{r,results='hide', eval=FALSE, echo=FALSE}
pdf("../../../figures/merged_sites/Q5/supplement_2.pdf",
    height =10,width = 8,paper="a4")
heatmap_plot
dev.off()
```

# Session info

```{r, warning=FALSE}
sessionInfo()
```
