---
title: "Geography-independent mucosal microbiota alterations in primary sclerosing cholangitis persist after liver transplantation"
subtitle: "Additional disease control"
format: 
  html:  
    embed-resources: true
    keep-md: false
    toc: true
    toc-expand: 3
    toc-depth: 6
    toc-location: left
    number-sections: true
    number-depth: 6
    code-fold: false
    code-tools: true
    code-summary: "Open code"
    grid: 
      body-width: 1000px
      sidebar-width: 600px
      margin-width: 100px
editor: source
project:
  type: default
  output-dir: output
fontsize: 12 px
---

# Introduction

**PSC vs ALD**

In previous analyses, we showed specific mucosal composition in PSC patients which persist after liver transplantation. Here, we added another positive control, to explore if this condition is disease specific, or it can be credited to (liver) disease in general.  

Fo this purpose, we conduct the univariate analysis on Czech ALD patients after LTx. By comparing post_LTx PSC and ALD patients, we can check if the proposed PSC-associated bacteria are at least partially overlapping.

If so, this would confirm our previous conclusions that these bacteria are specific for PSC.


**Importing libraries and custom functions built for this analysis**

```{r,warning=FALSE}
source("custom_functions.R")
```

# Data Import

Importing ASV, taxa and metadata tables for both Czech ALD samples
samples.

```{r,warning=FALSE}
path = "../../../../data/analysis_ready_data/ikem/ALD/"
asv_tab_ALD <- as.data.frame(fread(file.path(path,"asv_table_ALD.tsv"),
                                    check.names = FALSE))
taxa_tab_ALD <- as.data.frame(fread(file.path(path,"taxa_table_ALD.tsv"),
                                     check.names = FALSE))
metadata_ALD <- as.data.frame(fread(file.path(path,"metadata_ALD.tsv"),
                                     check.names = FALSE))
```

Importing previous Czech dataset with PSC patients.

```{r,warning=FALSE}
path = "../../../../data/analysis_ready_data/ikem/"
asv_tab_ikem <- as.data.frame(fread(file.path(path,"asv_table_ikem.csv"),
                                    check.names = FALSE))
taxa_tab_ikem <- as.data.frame(fread(file.path(path,"taxa_table_ikem.csv"),
                                     check.names = FALSE))
metadata_ikem <- as.data.frame(fread(file.path(path,"metadata_ikem.csv"),
                                     check.names = FALSE))
```

Importing proposed PSC-associated taxa.

```{r}
psc_genus_ti <- read.xlsx("../../../results/merged_sites/main_analysis/Q1/univariate_analysis/psc_effect_terminal_ileum.xlsx")

psc_genus_colon <- read.xlsx("../../../results/merged_sites/main_analysis/Q1/univariate_analysis/psc_effect_colon.xlsx")
```

# Terminal ileum

```{r}
segment="terminal_ileum"

ileum_data <- merging_data(asv_tab_1=asv_tab_ALD,
                           asv_tab_2=asv_tab_ikem,
                           taxa_tab_1=taxa_tab_ALD,
                           taxa_tab_2=taxa_tab_ikem,
                           metadata_1=metadata_ALD,
                           metadata_2=metadata_ikem,
                           segment="TI",Q="QALD")

ileum_asv_tab <- ileum_data[[1]]
ileum_taxa_tab <- ileum_data[[2]]
ileum_metadata <- ileum_data[[3]]
```


## Filtering

Rules: 

  - sequencing depth \> 10000 
  
  - nearZeroVar() with default settings 



**Library size**

```{r, fig.width=5, fig.height=4, fig.fullwidth=TRUE, warning=FALSE}
read_counts(ileum_asv_tab, line = c(5000,10000))
```

### Sequencing depth

```{r, warning=FALSE}
data_filt <- seq_depth_filtering(ileum_asv_tab,
                                 ileum_taxa_tab,
                                 ileum_metadata,
                                 seq_depth_threshold = 10000)

filt_ileum_asv_tab <- data_filt[[1]]; alpha_ileum_asv_tab <- filt_ileum_asv_tab
filt_ileum_taxa_tab <- data_filt[[2]]; alpha_ileum_taxa_tab <- filt_ileum_taxa_tab
filt_ileum_metadata <- data_filt[[3]]; alpha_ileum_metadata <- filt_ileum_metadata

seq_step <- dim(filt_ileum_asv_tab)[1]
```

**Library size**

```{r, fig.width=5, fig.height=4, fig.fullwidth=TRUE, warning=FALSE}
read_counts(filt_ileum_asv_tab,line = c(5000,10000))
```

### NearZeroVar

```{r, warning=FALSE}
data_filt <- nearzerovar_filtering(filt_ileum_asv_tab, 
                                   filt_ileum_taxa_tab,
                                   filt_ileum_metadata)

filt_ileum_asv_tab <- data_filt[[1]]
filt_ileum_taxa_tab <- data_filt[[2]]
nearzero_step <- dim(filt_ileum_asv_tab)[1]
```

Library size

```{r, fig.width=5, fig.height=4, fig.fullwidth=TRUE, warning=FALSE}
read_counts(filt_ileum_asv_tab,line = c(5000,10000))
```

### Final Counts

```{r, warning=FALSE}
final_counts_filtering(ileum_asv_tab,
                       filt_ileum_asv_tab,
                       filt_ileum_metadata,
                       seq_step, 0, nearzero_step) %>% `colnames<-`("Count")
```

## Alpha diversity

```{r, warning=FALSE}
path = "../../../results/merged_sites/main_analysis/QALD/alpha_diversity"
```

**Calculation**

```{r, warning=FALSE}
# Construct MPSE object
alpha_ileum_metadata$Sample <- alpha_ileum_metadata$SampleID
ileum_mpse <- as.MPSE(construct_phyloseq(alpha_ileum_asv_tab,
                                         alpha_ileum_taxa_tab,
                                         alpha_ileum_metadata))

ileum_mpse %<>% mp_rrarefy(raresize = 10000,seed = 123)

# Calculate alpha diversity - rarefied counts
ileum_mpse %<>% mp_cal_alpha(.abundance=RareAbundance, force=TRUE)
```

```{r, warning=FALSE}

alpha_div_plots <- list()

# preparing data frame
alpha_data <- data.frame(SampleID=ileum_mpse$Sample.x,
                         Observe=ileum_mpse$Observe,
                         Shannon=ileum_mpse$Shannon,
                         Simpson=ileum_mpse$Simpson,
                         Pielou=ileum_mpse$Pielou,
                         Group=ileum_mpse$Group,
                         Country=ileum_mpse$Country,
                         Patient=ileum_mpse$Patient)

write.csv(alpha_data,file.path(path,paste0("alpha_indices_",segment,".csv")),
          row.names = FALSE)
```

### Plots

```{r, fig.width=4, fig.height=5, fig.fullwidth=TRUE}
p_boxplot_alpha <- alpha_diversity_countries(alpha_data,
                                             show_legend = FALSE,
                                             size=1,gap=FALSE,jitter=1.5)

# save the results
alpha_div_plots[[paste(segment,"Country")]] <- p_boxplot_alpha

# see the results
p_boxplot_alpha
```

```{r,results='hide', echo=FALSE}
pdf("../../../figures/merged_sites/QALD/alpha_diversity_terminal_ileum.pdf",
    height =4,width = 6)
p_boxplot_alpha
dev.off()
```

### Linear Model

```{r, warning=FALSE}
path = "../../../results/merged_sites/main_analysis/QALD/alpha_diversity"
alpha_data <- read.csv(file.path(path,paste0("alpha_indices_",segment,".csv")))
```

**Richness**

```{r, warning=FALSE}
results_model <- pairwise.lm(formula = "Observe ~ Group",
                             factors=alpha_data$Group,
                             data=alpha_data)

# check interaction
if (!is.data.frame(results_model)){
  results_model_observe <- results_model[[1]]
  results_model_observe_emeans <- results_model[[2]]
} else {
  results_model_observe <- results_model
  results_model_observe_emeans <- NA
}

# save the results
pc_observed <- list(); 
pc_observed[[segment]] <- results_model_observe
```

```{r, warning=FALSE}
# see the results
knitr::kable(results_model_observe,digits = 3,
caption = "Raw results of linear model of richness estimation.")

knitr::kable(results_model_observe_emeans,digits = 3,
caption = "Raw results of independent country analysis")
```

**Shannon**

```{r, warning=FALSE}
results_model <- pairwise.lm(formula = "Shannon ~ Group",
                             factors=alpha_data$Group,
                             data=alpha_data)

# check interaction
if (!is.data.frame(results_model)){
  results_model_shannon <- results_model[[1]]
  results_model_shannon_emeans <- results_model[[2]]
} else {
  results_model_shannon <- results_model
  results_model_shannon_emeans <- NA
}

# save the results
pc_shannon <- list(); 
pc_shannon[[segment]] <- as.data.frame(results_model_shannon)

```

```{r, warning=FALSE}
# see the results
knitr::kable(results_model_shannon,digits = 3,
caption = "Raw results of linear model of Shannon estimation.")

knitr::kable(results_model_shannon_emeans,digits = 3,
caption = "Raw results of independent country analysis")
```

**Simpson**

```{r, warning=FALSE}
results_model <- pairwise.lm(formula = "Simpson ~ Group",
                                     factors=alpha_data$Group,
                                     data=alpha_data)

# check interaction
if (!is.data.frame(results_model)){
  results_model_simpson <- results_model[[1]]
  results_model_simpson_emeans <- results_model[[2]]
} else {
  results_model_simpson <- results_model
  results_model_simpson_emeans <- NA
}


# save the results
pc_simpson <- list(); 
pc_simpson[[segment]] <- as.data.frame(results_model_simpson)
```

```{r, warning=FALSE}
# see the results
knitr::kable(results_model_simpson,digits = 3,
caption = "Raw results of linear model of Simpson estimation.")

knitr::kable(results_model_simpson_emeans,digits = 3,
caption = "Raw results of independent country analysis")
```

**Pielou**

```{r, warning=FALSE}
results_model <- pairwise.lm(formula = "Pielou ~ Group",
                                     factors=alpha_data$Group,
                                     data=alpha_data)

# check interaction

if (!is.data.frame(results_model)){
  results_model_pielou <- results_model[[1]]
  results_model_pielou_emeans <- results_model[[2]]
} else {
  results_model_pielou <- results_model
  results_model_pielou_emeans <- NA
}

# save the results
pc_pielou <- list(); 
pc_pielou[[segment]] <- as.data.frame(results_model_pielou)
```

```{r, warning=FALSE}
# see the results
knitr::kable(results_model_pielou,digits = 3,
caption = "Raw results of linear model of Pielou estimation.")

knitr::kable(results_model_pielou_emeans,digits = 3,
caption = "Raw results of independent country analysis")
```

### Saving results

```{r, warning=FALSE}
alpha_list <- list(
  Richness=pc_observed[[segment]],
  Shannon=pc_shannon[[segment]])
                   
write.xlsx(alpha_list, 
           file = file.path(path,paste0("alpha_diversity_results_",segment,".xlsx")))
```


## Beta diversity

Calculating Aitchison distance (euclidean distance on clr-transformed
data), both at ASV and genus level.

### Main analysis

**Genus level, Aitchison distance**

```{r, warning=FALSE}
level="genus"
```

```{r, warning=FALSE}
path = "../../../results/merged_sites/main_analysis/QALD/beta_diversity"
```

```{r, warning=FALSE}
pairwise_aitchison_raw <- list()
pca_plots_list <- list()
```

Aggregation, filtering

```{r, warning=FALSE}
# Aggregation
genus_data <- aggregate_taxa(ileum_asv_tab,
                             ileum_taxa_tab,
                             taxonomic_level=level,
                             names=TRUE)
# Filtration
filt_data <- filtering_steps(genus_data[[1]],
                             genus_data[[2]],
                             ileum_metadata,
                             seq_depth_threshold=10000)

filt_ileum_genus_tab <- filt_data[[1]]
filt_ileum_genus_taxa <- filt_data[[2]]
filt_ileum_metadata <- filt_data[[3]]
```

##### PERMANOVA

```{r, warning=FALSE}
pairwise_df <- filt_ileum_genus_tab %>% column_to_rownames("SeqID") %>% t()

# main effect
pp_main <- pairwise.adonis(pairwise_df,
                           filt_ileum_metadata$Group,
                           sim.method = "robust.aitchison", p.adjust.m="BH")

# tidy the results
pp_factor <- pp_main

cols <- c("pairs","Df","SumsOfSqs", "F.Model","R2","p.value", "p.adjusted", "sig")
colnames(pp_factor) <- cols

# save raw results
pairwise_aitchison_raw[[paste(level, segment)]] <- pp_factor
```

```{r, warning=FALSE}
# see the results
knitr::kable(pp_factor,digits = 3,caption = "PERMANOVA, GROUP separation")
```

##### Plots

**PCA**

```{r, warning=FALSE}
p <- pca_plot_custom(filt_ileum_genus_tab,
                                 filt_ileum_genus_taxa,
                                 filt_ileum_metadata,
                                 show_boxplots = TRUE,
                                 variable = "Group", size=3, show_legend=FALSE)

# save the results
pca_plots_list[[paste(segment,level,"custom")]] <- p

# see the results
p
```

```{r,results='hide', echo=FALSE}
pdf("../../../figures/merged_sites/QALD/beta_diversity_terminal_ileum.pdf",
    height =5,width = 5)
p
dev.off()
```

#### Saving results

```{r,warning=FALSE}
write.xlsx(pairwise_aitchison_raw[[paste(level, segment)]], 
           file = file.path(path,
           paste0("beta_diversity_results_", segment,".xlsx")))
```

## Univariate Analysis

Similar to previous approaches, we will use two DAA tools - linDA and Maaslin. Only the intersection of sets
selected by each tool are shown to minimize false positives. As this analysis can only be conducted on Czech cohort, no interaction effect can be shown, therefore only Group effect is investigated. 

This analysis is performed at genus level only.

```{r,warning=FALSE}
level="genus"

# needed paths
path = "../../../results/merged_sites/main_analysis/QALD/univariate_analysis"
path_maaslin=file.path("../../../intermediate_files/maaslin/QALD",level)
```

```{r,warning=FALSE}
# variables
raw_linda_results_genus <- list();
raw_linda_results_genus[[segment]] <- list()
linda_results_genus <- list(); 
linda_results_genus[[segment]] <- list()

# country and interaction problems
list_country_union <- list()
list_intersections <- list()
list_venns <- list()
uni_statistics <- list()

# workbook for final df
wb <- createWorkbook()
```

Aggregate taxa

```{r,warning=FALSE}
genus_data <- aggregate_taxa(ileum_asv_tab,
                             ileum_taxa_tab,
                             taxonomic_level = level)

ileum_genus_tab <- genus_data[[1]]
ileum_genus_taxa_tab <- genus_data[[2]]

ileum_genus_asv_taxa_tab <- create_asv_taxa_table(ileum_genus_tab,
                                                  ileum_genus_taxa_tab)
```

### PSC vs ALD

We can assume, that the differentially abundant taxa between these two
groups are caused by PSC disease and other factors, like liver cirrhosis
and overall bad clinical condition.

```{r,warning=FALSE}
group <- c("ALD post_ltx","PSC post_ltx")
comparison_name <- paste0(group[1], " vs ",group[2])
```

```{r, echo=FALSE}
comparison_title <- sub('"(.+?)" vs "(.+?)"', '"\\2" and "\\1"', comparison_name)
```

#### linDA

```{r, warning=FALSE, fig.width=15, fig.height=5, fig.fullwidth=TRUE}

# prepare the data
linda_data <- binomial_prep(ileum_genus_tab,
                            ileum_genus_taxa_tab,
                            ileum_metadata,
                            group, usage="linDA")

filt_ileum_uni_data <- linda_data[[1]]
filt_ileum_uni_taxa <- linda_data[[2]]
filt_ileum_uni_metadata <- linda_data[[3]]

# fit the model
linda.obj <- linda(filt_ileum_uni_data, 
                   filt_ileum_uni_metadata, 
                   formula = '~ Group')

linda.output <- linda.obj$output
linda.output <- linda_renaming(linda.output, group)

# save the results
group1 <- paste0(group[1], " vs ","Group",group[2])

for (grp in c(group1)){
  raw_linda_results_genus[[segment]][[grp]] <- 
    rawlinda.df(linda.output,
                grp,
                filt_ileum_uni_data,
                filt_ileum_uni_taxa)
  
  linda_results_genus[[segment]][[grp]] <- 
    linda.df(linda.output,
             grp,
             filt_ileum_uni_data,
             filt_ileum_uni_taxa)
}

```

```{r, warning=FALSE, fig.width=5, fig.height=4, fig.fullwidth=TRUE}
# volcano plots
volcano <- volcano_plot_linda(linda.output, 
                                group1, 
                                taxa_table = filt_ileum_uni_taxa) + 
            ggtitle(comparison_title)

volcano
```

#### MaAsLin2

```{r, warning=FALSE, echo=FALSE,results='hide',message=FALSE,warning=FALSE}
log_file <- file(tempfile(), open = "wt")
sink(log_file)  # Redirect standard output
sink(log_file, type = "message")  # Redirect messages and warnings

fit_data <-  Maaslin2(
    input_data = filt_ileum_uni_data, 
    input_metadata = filt_ileum_uni_metadata, min_abundance = 0,
    min_prevalence = 0,min_variance = 0,
    output = file.path(path_maaslin,group1), max_significance = 0.05,
    fixed_effects = c('Group'),correction = "BH")

sink()  # Restore standard output
sink(type = "message")  # Restore messages
```

```{r,warning=FALSE}
volcano <- volcano_plot_maaslin(fit_data,filt_ileum_uni_taxa) + 
            ggtitle(comparison_title)


volcano
```

#### Group - Intersection

```{r,warning=FALSE}
intersection_results <- group_intersection(group, list_intersections, list_venns,
                                           linda.output, fit_data,
                                           raw_linda_results_genus, 
                                           segment = segment,
                                           level=level)

list_intersections <- intersection_results[[1]]
list_venns <- intersection_results[[2]]
venn <- intersection_results[[3]]

# show the results
venn
```

#### Basic statistics

```{r,warning=FALSE}
uni_df <-  merge(basic_univariate_statistics(linda_data,group),
                 raw_linda_results_genus[[segment]][[group1]],
                 by="SeqID",all=TRUE)

uni_df[["final_sig"]] <- uni_df$SeqID %in% list_intersections[[paste(segment,level,comparison_name)]][["SeqID"]]
uni_statistics[[segment]][[paste(level,comparison_name)]] <- uni_df

# for comparison
new_name <- comparison_name
addWorksheet(wb, sheetName = new_name)
writeData(wb, sheet = new_name, uni_df, rowNames=FALSE)
```

#### Comparison with PSC-associated taxa

```{r}
psc_ald_sig_ti <- uni_df %>%
  dplyr::filter(final_sig==TRUE) %>%
  dplyr::select(SeqID,Taxonomy,log2FoldChange)

psc_genus_ti <- psc_genus_ti
```

```{r}
list_core <- list()
list_core[["PSC vs ALD"]] <- psc_ald_sig_ti$Taxonomy
list_core[["PSC-associated taxa"]] <- psc_genus_ti$Taxonomy
  
# venn diagram
venn <- ggvenn(list_core, fill_color = c("blue", "red")) + 
    ggtitle("Intersection between PSC-associated taxa and (PSC vs ALD)") + 
    guides(fill = "none")

venn
```
```{r}
intersection_df <- intersect(list_core$`PSC vs ALD`,list_core$`PSC-associated taxa`)

intersection_df
```


### ALD vs HC

We can assume, that the differentially abundant taxa between these two
groups are caused by PSC disease and other factors, like liver cirrhosis
and overall bad clinical condition.

```{r,warning=FALSE}
group <- c("HC","ALD post_ltx")
comparison_name <- paste0(group[1], " vs ",group[2])
```

```{r, echo=FALSE}
comparison_title <- sub('"(.+?)" vs "(.+?)"', '"\\2" and "\\1"', comparison_name)
```

#### linDA

```{r, warning=FALSE, fig.width=15, fig.height=5, fig.fullwidth=TRUE}

# prepare the data
linda_data <- binomial_prep(ileum_genus_tab,
                            ileum_genus_taxa_tab,
                            ileum_metadata,
                            group, usage="linDA")

filt_ileum_uni_data <- linda_data[[1]]
filt_ileum_uni_taxa <- linda_data[[2]]
filt_ileum_uni_metadata <- linda_data[[3]]

# fit the model
linda.obj <- linda(filt_ileum_uni_data, 
                   filt_ileum_uni_metadata, 
                   formula = '~ Group')

linda.output <- linda.obj$output
linda.output <- linda_renaming(linda.output, group)

# save the results
group1 <- paste0(group[1], " vs ","Group",group[2])

for (grp in c(group1)){
  raw_linda_results_genus[[segment]][[grp]] <- 
    rawlinda.df(linda.output,
                grp,
                filt_ileum_uni_data,
                filt_ileum_uni_taxa)
  
  linda_results_genus[[segment]][[grp]] <- 
    linda.df(linda.output,
             grp,
             filt_ileum_uni_data,
             filt_ileum_uni_taxa)
}

```

```{r, warning=FALSE, fig.width=5, fig.height=4, fig.fullwidth=TRUE}
# volcano plots
volcano <- volcano_plot_linda(linda.output, 
                                group1, 
                                taxa_table = filt_ileum_uni_taxa) + 
            ggtitle(comparison_title)

volcano
```

#### MaAsLin2

```{r, warning=FALSE, echo=FALSE,results='hide',message=FALSE,warning=FALSE}
log_file <- file(tempfile(), open = "wt")
sink(log_file)  # Redirect standard output
sink(log_file, type = "message")  # Redirect messages and warnings

fit_data <-  Maaslin2(
    input_data = filt_ileum_uni_data, 
    input_metadata = filt_ileum_uni_metadata, min_abundance = 0,
    min_prevalence = 0,min_variance = 0,
    output = file.path(path_maaslin,group1), max_significance = 0.05,
    fixed_effects = c('Group'),correction = "BH")

sink()  # Restore standard output
sink(type = "message")  # Restore messages
```

```{r,warning=FALSE}
volcano <- volcano_plot_maaslin(fit_data,filt_ileum_uni_taxa) + 
            ggtitle(comparison_title)


volcano
```

#### Group - Intersection

```{r,warning=FALSE}
intersection_results <- group_intersection(group, list_intersections, list_venns,
                                           linda.output, fit_data,
                                           raw_linda_results_genus, 
                                           segment = segment,
                                           level=level)

list_intersections <- intersection_results[[1]]
list_venns <- intersection_results[[2]]
venn <- intersection_results[[3]]

# show the results
venn
```

#### Basic statistics

```{r,warning=FALSE}
uni_df <-  merge(basic_univariate_statistics(linda_data,group),
                 raw_linda_results_genus[[segment]][[group1]],
                 by="SeqID",all=TRUE)

uni_df[["final_sig"]] <- uni_df$SeqID %in% list_intersections[[paste(segment,level,comparison_name)]][["SeqID"]]
uni_statistics[[segment]][[paste(level,comparison_name)]] <- uni_df

# for comparison
new_name <- comparison_name
addWorksheet(wb, sheetName = new_name)
writeData(wb, sheet = new_name, uni_df, rowNames=FALSE)
```

#### Comparison with PSC-associated taxa

```{r}
hc_ald_sig_ti <- uni_df %>%
  dplyr::filter(final_sig==TRUE) %>%
  dplyr::select(SeqID,Taxonomy,log2FoldChange)

psc_genus_ti <- psc_genus_ti
```

```{r}
list_core <- list()
list_core[["ALD vs HC"]] <- hc_ald_sig_ti$Taxonomy
list_core[["PSC-associated taxa"]] <- psc_genus_ti$Taxonomy
  
# venn diagram
venn <- ggvenn(list_core, fill_color = c("blue", "red")) + 
    ggtitle("Intersection between PSC-associated taxa and (ALD vs HC)") + 
    guides(fill = "none")

venn
```

From this Venn diagram, we are interested in EXLUSIVE RIGH JOIN

```{r}
A <- list_core$`ALD vs HC`
B <- list_core$`PSC-associated taxa`
remaining_psc <- B[!(B %in% A)]

psc_genus_ti %>%
  dplyr::filter(Taxonomy %in% remaining_psc)

uni_df %>%
  dplyr::filter(Taxonomy %in% remaining_psc)
```

#### Visualization

Heatmap visualizing the linDA's logFoldChange for taxa with p \< 0.1.

```{r, warning=FALSE, fig.width=10, fig.height=17, fig.fullwidth=TRUE}
list_heatmap <- list_intersections[grep(paste(segment,level),
                                  names(list_intersections),value=TRUE)]

p_heatmap_linda <- heatmap_linda(list_heatmap,ileum_taxa_tab)
p_heatmap_linda
```

**Dot heatmap**

```{r, warning=FALSE, fig.width=5, fig.height=10, fig.fullwidth=TRUE}
dotheatmap_linda <- dot_heatmap_linda(list_heatmap,
                                      uni_statistics$terminal_ileum[grepl(level,names(uni_statistics$terminal_ileum))],
                                      ileum_taxa_tab) + xlab("") + ylab("")
dotheatmap_linda
```

**Horizontal bar plot**

```{r,fig.height=10, fig.width=4}
p_prevalence <- horizontal_barplot(wb,taxa=levels(dotheatmap_linda$data$SeqID))

p_prevalence_final <- ggarrange(p_prevalence,
                                ggplot() + theme_minimal(),
                                nrow = 2,heights = c(1,0.15))

p <- ggarrange(dotheatmap_linda + theme(legend.position = "none"),
               p_prevalence_final,
               ncol=2,widths = c(0.8,0.3))
p

dot_heatmap_ileum <- p
```

```{r, warning=FALSE,results='hide',echo=FALSE}
pdf("../../../figures/merged_sites/QALD/dotplot_terminal_ileum.pdf",
    height =7,width = 4)
p
dev.off()
```

# Data Analysis - Colon

```{r}
segment="colon"

colon_data <- merging_data(asv_tab_1=asv_tab_ALD,
                           asv_tab_2=asv_tab_ikem,
                           taxa_tab_1=taxa_tab_ALD,
                           taxa_tab_2=taxa_tab_ikem,
                           metadata_1=metadata_ALD,
                           metadata_2=metadata_ikem,
                           segment=segment,Q="QALD")

colon_asv_tab <- colon_data[[1]]
colon_taxa_tab <- colon_data[[2]]
colon_metadata <- colon_data[[3]]
```

## Filtering

Rules: - sequencing depth \> 10000 - nearZeroVar() with default settings

Library size

```{r, warning=FALSE, fig.width=5, fig.height=4, fig.fullwidth=TRUE}
read_counts(colon_asv_tab, line = c(5000,10000))
```

### Sequencing depth

```{r,warning=FALSE}
data_filt <- seq_depth_filtering(colon_asv_tab,
                                 colon_taxa_tab,
                                 colon_metadata,
                                 seq_depth_threshold = 10000)

filt_colon_asv_tab <- data_filt[[1]]; alpha_colon_asv_tab <- filt_colon_asv_tab
filt_colon_taxa_tab <- data_filt[[2]]; alpha_colon_taxa_tab <- filt_colon_taxa_tab
filt_colon_metadata <- data_filt[[3]]; alpha_colon_metadata <- filt_colon_metadata

seq_step <- dim(filt_colon_asv_tab)[1]
```

Library size

```{r, warning=FALSE, fig.width=5, fig.height=4, fig.fullwidth=TRUE}
read_counts(filt_colon_asv_tab,line = c(10000))
```

### NearZeroVar

```{r,warning=FALSE}
data_filt <- nearzerovar_filtering(filt_colon_asv_tab,
                                   filt_colon_taxa_tab,
                                   filt_colon_metadata)

filt_colon_asv_tab <- data_filt[[1]]
filt_colon_taxa_tab <- data_filt[[2]]
nearzero_step <- dim(filt_colon_asv_tab)[1]
```

Library size

```{r, warning=FALSE, fig.width=5, fig.height=4, fig.fullwidth=TRUE}
read_counts(filt_colon_asv_tab,line = c(5000,10000))
```

Check zero depth

```{r,warning=FALSE}
data_filt <- check_zero_depth(filt_colon_asv_tab, 
                              filt_colon_taxa_tab, 
                              filt_colon_metadata)

filt_colon_asv_tab <- data_filt[[1]]; 
filt_colon_taxa_tab <- data_filt[[2]]; 
filt_colon_metadata <- data_filt[[3]]; 
```

Library size

```{r, warning=FALSE, fig.width=5, fig.height=4, fig.fullwidth=TRUE}
read_counts(filt_colon_asv_tab,line = c(5000,10000))
```

### Final Counts

```{r,warning=FALSE}
final_counts_filtering(colon_asv_tab,
                       filt_colon_asv_tab,
                       filt_colon_metadata,
                       seq_step, 0, nearzero_step)
```

## Alpha diversity

Calculating Richness, Shannon, Simpson, Pielou indexes on raw
(unfiltered) rarefied data. Samples with sequencing depth \< 10000 were
excluded. Testing by linear mixed-effect model.

```{r,warning=FALSE}
path = "../../../results/merged_sites/main_analysis/QALD/alpha_diversity"
```

**Calculation**

```{r,warning=FALSE}
# Construct MPSE object
alpha_colon_metadata$Sample <- alpha_colon_metadata$SampleID
colon_mpse <- as.MPSE(construct_phyloseq(alpha_colon_asv_tab,
                                         alpha_colon_taxa_tab,
                                         alpha_colon_metadata))

colon_mpse %<>% mp_rrarefy(raresize = 10000,seed = 123)

# Calculate alpha diversity - rarefied counts
colon_mpse %<>% mp_cal_alpha(.abundance=RareAbundance, force=TRUE)
```

```{r,warning=FALSE}
alpha_data <- data.frame(SampleID=colon_mpse$Sample.x,
                         Observe=colon_mpse$Observe,
                         Shannon=colon_mpse$Shannon,
                         Simpson=colon_mpse$Simpson,
                         Pielou=colon_mpse$Pielou,
                         Group=colon_mpse$Group,
                         Country=colon_mpse$Country,
                         Patient=colon_mpse$Patient)

write.csv(alpha_data,file.path(path,paste0("alpha_indices_",segment,".csv")),
          row.names = FALSE)
```

### Plots

```{r, fig.width=4, fig.height=5, fig.fullwidth=TRUE}
p_boxplot_alpha <- alpha_diversity_countries(alpha_data,
                                             show_legend = FALSE,
                                             size=1,gap=TRUE,jitter = 1.5)

# save the results
alpha_div_plots[[paste(segment,"Country")]] <- p_boxplot_alpha

# see the results
p_boxplot_alpha
```

```{r, warning=FALSE,results='hide',echo=FALSE}
pdf("../../../figures/merged_sites/QALD/alpha_diversity_colon.pdf",
    height =4,width = 7)
p_boxplot_alpha
dev.off()
```

### Linear Model

```{r,warning=FALSE}
path = "../../../results/merged_sites/main_analysis/QALD/alpha_diversity"
alpha_data <- read.csv(file.path(path,paste0("alpha_indices_",segment,".csv")))
```

**Richness**

```{r,warning=FALSE}
results_model <- pairwise.lmer(
  formula = "Observe ~ Group + (1|Patient)",
  factors=alpha_data$Group,
  data=alpha_data)

# check interaction
if (!is.data.frame(results_model)){
  results_model_observe <- results_model[[1]]
  results_model_observe_detailed <- results_model[[2]]
} else {
  results_model_observe <- results_model
  results_model_observe_detailed <- NA
}

# save the results
pc_observed[[segment]] <- results_model_observe
```

```{r,warning=FALSE}
# see the results
knitr::kable(results_model_observe,digits = 3,
caption = "Raw results of linear model of richness estimation.")

knitr::kable(results_model_observe_detailed,digits = 3,
caption = "Raw results of independent country analysis")
```

**Shannon**

```{r,warning=FALSE}
results_model <- pairwise.lmer(
  formula = "Shannon ~ Group  + (1|Patient)",
  factors=alpha_data$Group,
  data=alpha_data)

# check interaction
if (!is.data.frame(results_model)){
  results_model_shannon <- results_model[[1]]
  results_model_shannon_detailed <- results_model[[2]]
} else {
  results_model_shannon <- results_model
  results_model_shannon_detailed <- NA
}

# save the results
pc_shannon[[segment]] <- as.data.frame(results_model_shannon)

```

```{r,warning=FALSE}
# see the results
knitr::kable(results_model_shannon,digits = 3,
caption = "Raw results of linear model of Shannon estimation.")

knitr::kable(results_model_shannon_detailed,digits = 3,
caption = "Raw results of independent country analysis")
```

**Simpson**

```{r,warning=FALSE}
results_model <- pairwise.lmer(
  formula = "Simpson ~ Group + (1|Patient)",
  factors=alpha_data$Group,
  data=alpha_data)

# check interaction
if (!is.data.frame(results_model)){
  results_model_simpson <- results_model[[1]]
  results_model_simpson_detailed <- results_model[[2]]
} else {
  results_model_simpson <- results_model
  results_model_simpson_detailed <- NA
}

# save the results
pc_simpson[[segment]] <- as.data.frame(results_model_simpson)

```

```{r,warning=FALSE}
# see the results
knitr::kable(results_model_simpson,digits = 3,
caption = "Raw results of linear model of Simpson estimation.")

knitr::kable(results_model_simpson_detailed,digits = 3,
caption = "Raw results of independent country analysis")
```

**Pielou**

```{r,warning=FALSE}
results_model <- pairwise.lmer(
  formula = "Pielou ~ Group + (1|Patient)",
  factors=alpha_data$Group,
  data=alpha_data)

# check interaction
if (!is.data.frame(results_model)){
  results_model_pielou <- results_model[[1]]
  results_model_pielou_detailed <- results_model[[2]]
} else {
  results_model_pielou <- results_model
  results_model_pielou_detailed <- NA
}

# save the results
pc_pielou[[segment]] <- as.data.frame(results_model_pielou)

```

```{r,warning=FALSE}
# see the results
knitr::kable(results_model_pielou,digits = 3,
caption = "Raw results of linear model of Pielou estimation.")

knitr::kable(results_model_pielou_detailed,digits = 3,
caption = "Raw results of independent country analysis")
```

### Saving results

```{r,warning=FALSE}
alpha_list <- list(
  Richness=pc_observed[[segment]] %>% rownames_to_column("Comparison"),
  Shannon=pc_shannon[[segment]] %>% rownames_to_column("Comparison"),
  Simpson=pc_simpson[[segment]] %>% rownames_to_column("Comparison"),
  Pielou=pc_pielou[[segment]] %>% rownames_to_column("Comparison"))
                   
write.xlsx(alpha_list, 
           file = file.path(path,paste0("alpha_diversity_results_",segment,".xlsx")))
```

## Beta diversity

Calculating Aitchison distance (euclidean distance on clr-transformed
data) at genus level (main analysis). In supplements, there are also
bray-curtis and Jaccard distance at ASV and genus levels. Testing by
PERMANOVA.

### Main analysis

**Genus level, Aitchison distance**

```{r,warning=FALSE}
level="genus"
```

```{r,warning=FALSE}
path = "../../../results/merged_sites/main_analysis/QALD/beta_diversity"
```

Aggregation, filtering

```{r,warning=FALSE}
genus_data <- aggregate_taxa(colon_asv_tab,
                             colon_taxa_tab,
                             taxonomic_level=level,
                             names=TRUE)

filt_data <- filtering_steps(genus_data[[1]],
                             genus_data[[2]],
                             colon_metadata,
                            seq_depth_threshold=10000)

filt_colon_genus_tab <- filt_data[[1]]
filt_colon_genus_taxa <- filt_data[[2]]
filt_colon_genus_metadata <- filt_data[[3]]
```

### PERMANOVA

```{r,warning=FALSE}
pairwise_df <- filt_colon_genus_tab %>% column_to_rownames("SeqID") %>% t()

# main effect
pp_main <- pairwise.adonis(pairwise_df,filt_colon_genus_metadata$Group,
                           patients = filt_colon_genus_metadata$Patient,
                           sim.method = "robust.aitchison", p.adjust.m="BH")


# tidy the results
pp_factor <- pp_main

cols <- c("pairs","Df","SumsOfSqs", "F.Model","R2","p.value", "p.adjusted", "sig")
colnames(pp_factor) <- cols

# save raw results
pairwise_aitchison_raw[[paste(level, segment)]] <-rbind(pp_factor
                                                        )

```

```{r,warning=FALSE}
# see the results
knitr::kable(pp_factor,digits = 3,caption = "PERMANOVA, GROUP separation")

```


##### Plots

```{r,warning=FALSE}
p <- pca_plot_custom(filt_colon_genus_tab,
                                 filt_colon_genus_taxa,
                                 filt_colon_genus_metadata,
                                 show_boxplots = TRUE,
                                 variable = "Group", size=2, 
                                 show_legend= FALSE)

# save the results
pca_plots_list[[paste(segment,level,"custom")]] <- p

# see the results
p

```

```{r, warning=FALSE,results='hide',echo=FALSE}
pdf("../../../figures/merged_sites/QALD/beta_diversity_colon.pdf",
    height =5,width = 5)
p
dev.off()
```

#### Saving results

```{r,warning=FALSE}
write.xlsx(pairwise_aitchison_raw[[paste(level, segment)]], 
           file = file.path(path,
           paste0("beta_diversity_results_", segment,".xlsx")))
```

## Univariate Analysis


```{r,warning=FALSE}
level="genus"
```

```{r,warning=FALSE}
# needed paths
path = "../../../results/merged_sites/main_analysis/QALD/univariate_analysis"
path_maaslin=file.path("../../../intermediate_files/maaslin/QALD",level)

# variables
raw_linda_results_genus[[segment]] <- list()
linda_results_genus[[segment]] <- list()

# country and interaction problems
list_country_union <- list()
list_venns <- list()

# workbook for final df
wb <- createWorkbook()

```

```{r,warning=FALSE}
level="genus"
```

Aggregate taxa

```{r,warning=FALSE}
genus_data <- aggregate_taxa(colon_asv_tab,
                             colon_taxa_tab,
                             taxonomic_level = level)

colon_genus_tab <- genus_data[[1]]
colon_genus_taxa_tab <- genus_data[[2]]

colon_genus_asv_taxa_tab <- create_asv_taxa_table(colon_genus_tab,
                                                  colon_genus_taxa_tab)
```

##### PSC vs ALD

###### linDA

```{r,warning=FALSE}
group <- c("ALD post_ltx","PSC post_ltx")
comparison_name <- paste0(group[1], " vs ",group[2])
```

```{r, echo=FALSE}
comparison_title <- sub('"(.+?)" vs "(.+?)"', '"\\2" and "\\1"', comparison_name)
```

```{r, warning=FALSE, fig.width=15, fig.height=5, fig.fullwidth=TRUE}

# prepare the data
linda_data <- binomial_prep(colon_genus_tab,
                            colon_genus_taxa_tab,
                            colon_metadata,
                            group, usage="linDA")

filt_colon_uni_data <- linda_data[[1]]
filt_colon_uni_taxa <- linda_data[[2]]
filt_colon_uni_metadata <- linda_data[[3]]

# fit the model
linda.obj <- linda(filt_colon_uni_data,
                   filt_colon_uni_metadata,
                   formula = '~ Group + (1|Patient)')

linda.output <- linda.obj$output
linda.output <- linda_renaming(linda.output, group)

# save the results
group1 <- paste0(group[1], " vs ","Group",group[2])

for (grp in c(group1)){
  raw_linda_results_genus[[segment]][[grp]] <- 
    rawlinda.df(linda.output,
                grp,
                filt_colon_uni_data,
                filt_colon_uni_taxa)
  
  linda_results_genus[[segment]][[grp]] <- 
    linda.df(linda.output,
             grp,
             filt_colon_uni_data,
             filt_colon_uni_taxa)
}
```

```{r, warning=FALSE, fig.width=5, fig.height=4, fig.fullwidth=TRUE}
# volcano plot
volcano <- volcano_plot_linda(linda.output, group1, 
                                taxa_table = filt_colon_uni_taxa) + 
            ggtitle(paste(group,collapse=" vs "))


# see the plot
volcano
```

###### MaAsLin2

```{r, warning=FALSE, echo=FALSE,results='hide',message=FALSE,warning=FALSE}

log_file <- file(tempfile(), open = "wt")
sink(log_file)  # Redirect standard output
sink(log_file, type = "message")  # Redirect messages and warnings

fit_data = Maaslin2(
    input_data = filt_colon_uni_data, 
    input_metadata = filt_colon_uni_metadata, min_abundance = 0,
    min_prevalence = 0,min_variance = 0,
    output = file.path(path_maaslin,group1), max_significance = 0.05,
    fixed_effects = c('Group'),random_effects = "Patient",
    correction = "BH")

sink()  # Restore standard output
sink(type = "message")  # Restore messages
```

```{r,warning=FALSE}
volcano <- volcano_plot_maaslin(fit_data,filt_colon_uni_taxa) + 
            ggtitle(paste(group[1], "vs", group[2]))

volcano
```

###### Group - Intersection

```{r,warning=FALSE}
intersection_results <- group_intersection(group, list_intersections, list_venns,
                                           linda.output, fit_data,
                                           raw_linda_results_genus, 
                                           segment = segment,
                                           level=level)

list_intersections <- intersection_results[[1]]
list_venns <- intersection_results[[2]]
venn <- intersection_results[[3]]

# show the results
venn
```


###### Basic statistics

```{r,warning=FALSE}
uni_df <-  merge(basic_univariate_statistics(linda_data,group),
                 raw_linda_results_genus[[segment]][[group1]],
                 by="SeqID",all=TRUE)

uni_df[["final_sig"]] <- uni_df$SeqID %in% list_intersections[[paste(segment,level,comparison_name)]][["SeqID"]]
uni_statistics[[segment]][[paste(level,comparison_name)]] <- uni_df

# for comparison
new_name <- comparison_name
addWorksheet(wb, sheetName = new_name)
writeData(wb, sheet = new_name, uni_df, rowNames=FALSE)
```

###### Comparison with PSC-associated taxa

```{r}
psc_ald_sig_colon <- uni_df %>%
  dplyr::filter(final_sig==TRUE) %>%
  dplyr::select(SeqID,Taxonomy,log2FoldChange)

psc_genus_colon <- psc_genus_colon
```

```{r}
list_core <- list()
list_core[["PSC vs ALD"]] <- psc_ald_sig_colon$Taxonomy
list_core[["PSC-associated taxa"]] <- psc_genus_colon$Taxonomy

# venn diagram
venn <- ggvenn(list_core, fill_color = c("blue", "red")) + 
    ggtitle("Intersection between PSC-associated taxa and (PSC vs ALD)") + 
    guides(fill = "none")

venn
```

```{r}
intersection_df <- intersect(list_core$`PSC vs ALD`,list_core$`PSC-associated taxa`)

intersection_df
```

##### ALD vs HC

###### linDA

```{r,warning=FALSE}
group <- c("HC","ALD post_ltx")
comparison_name <- paste0(group[1], " vs ",group[2])
```

```{r, echo=FALSE}
comparison_title <- sub('"(.+?)" vs "(.+?)"', '"\\2" and "\\1"', comparison_name)
```

```{r, warning=FALSE, fig.width=15, fig.height=5, fig.fullwidth=TRUE}

# prepare the data
linda_data <- binomial_prep(colon_genus_tab,
                            colon_genus_taxa_tab,
                            colon_metadata,
                            group, usage="linDA")

filt_colon_uni_data <- linda_data[[1]]
filt_colon_uni_taxa <- linda_data[[2]]
filt_colon_uni_metadata <- linda_data[[3]]

# fit the model
linda.obj <- linda(filt_colon_uni_data,
                   filt_colon_uni_metadata,
                   formula = '~ Group + (1|Patient)')

linda.output <- linda.obj$output
linda.output <- linda_renaming(linda.output, group)

# save the results
group1 <- paste0(group[1], " vs ","Group",group[2])

for (grp in c(group1)){
  raw_linda_results_genus[[segment]][[grp]] <- 
    rawlinda.df(linda.output,
                grp,
                filt_colon_uni_data,
                filt_colon_uni_taxa)
  
  linda_results_genus[[segment]][[grp]] <- 
    linda.df(linda.output,
             grp,
             filt_colon_uni_data,
             filt_colon_uni_taxa)
}
```

```{r, warning=FALSE, fig.width=5, fig.height=4, fig.fullwidth=TRUE}
# volcano plot
volcano <- volcano_plot_linda(linda.output, group1, 
                                taxa_table = filt_colon_uni_taxa) + 
            ggtitle(paste(group,collapse=" vs "))


# see the plot
volcano
```

###### MaAsLin2

```{r, warning=FALSE, echo=FALSE,results='hide',message=FALSE,warning=FALSE}

log_file <- file(tempfile(), open = "wt")
sink(log_file)  # Redirect standard output
sink(log_file, type = "message")  # Redirect messages and warnings

fit_data = Maaslin2(
    input_data = filt_colon_uni_data, 
    input_metadata = filt_colon_uni_metadata, min_abundance = 0,
    min_prevalence = 0,min_variance = 0,
    output = file.path(path_maaslin,group1), max_significance = 0.05,
    fixed_effects = c('Group'),random_effects = "Patient",
    correction = "BH")

sink()  # Restore standard output
sink(type = "message")  # Restore messages
```

```{r,warning=FALSE}
volcano <- volcano_plot_maaslin(fit_data,filt_colon_uni_taxa) + 
            ggtitle(paste(group[1], "vs", group[2]))

volcano
```

###### Group - Intersection

```{r,warning=FALSE}
intersection_results <- group_intersection(group, list_intersections, list_venns,
                                           linda.output, fit_data,
                                           raw_linda_results_genus, 
                                           segment = segment,
                                           level=level)

list_intersections <- intersection_results[[1]]
list_venns <- intersection_results[[2]]
venn <- intersection_results[[3]]

# show the results
venn
```


###### Basic statistics

```{r,warning=FALSE}
uni_df <-  merge(basic_univariate_statistics(linda_data,group),
                 raw_linda_results_genus[[segment]][[group1]],
                 by="SeqID",all=TRUE)

uni_df[["final_sig"]] <- uni_df$SeqID %in% list_intersections[[paste(segment,level,comparison_name)]][["SeqID"]]
uni_statistics[[segment]][[paste(level,comparison_name)]] <- uni_df

# for comparison
new_name <- comparison_name
addWorksheet(wb, sheetName = new_name)
writeData(wb, sheet = new_name, uni_df, rowNames=FALSE)
```

###### Comparison with PSC-associated taxa

```{r}
hc_ald_sig_colon <- uni_df %>%
  dplyr::filter(final_sig==TRUE) %>%
  dplyr::select(SeqID,Taxonomy,log2FoldChange)

psc_genus_colon <- psc_genus_colon
```

```{r}
list_core <- list()
list_core[["ALD vs HC"]] <- hc_ald_sig_colon$Taxonomy
list_core[["PSC-associated taxa"]] <- psc_genus_colon$Taxonomy

# venn diagram
venn <- ggvenn(list_core, fill_color = c("blue", "red")) + 
    ggtitle("Intersection between PSC-associated taxa and (ALD vs HC)") + 
    guides(fill = "none")

venn
```

From this Venn diagram, we are interested in EXLUSIVE RIGH JOIN

```{r}
A <- list_core$`ALD vs HC`
B <- list_core$`PSC-associated taxa`
remaining_psc <- B[!(B %in% A)]

psc_genus_colon %>%
  dplyr::filter(Taxonomy %in% remaining_psc)

uni_df %>%
  dplyr::filter(Taxonomy %in% remaining_psc)
```

##### Visualization

Heatmap visualizing the linDA's logFoldChange for taxa with p \< 0.1.

```{r, warning=FALSE, fig.width=10, fig.height=17, fig.fullwidth=TRUE}
list_heatmap <- list_intersections[grep(paste(segment,level),
                                  names(list_intersections),value=TRUE)]

p_heatmap_linda <- heatmap_linda(list_heatmap,colon_taxa_tab)
p_heatmap_linda
```

Dot heatmap

```{r, warning=FALSE, fig.width=5, fig.height=17, fig.fullwidth=TRUE}
dotheatmap_linda <- dot_heatmap_linda(list_heatmap,
                                      uni_statistics$colon[grepl(level,names(uni_statistics$colon))],
                                      colon_taxa_tab) + xlab("") + ylab("")
dotheatmap_linda
```

**Horizontal bar plot**

```{r,fig.height=10, fig.width=4}
p_prevalence <- horizontal_barplot(wb,taxa=levels(dotheatmap_linda$data$SeqID))
```

```{r,fig.height=10,fig.width=4,warning=FALSE}
p_prevalence_final <- ggarrange(p_prevalence,
                                ggplot() + theme_minimal(),
                                nrow = 2,heights = c(1,0.115))
p <- ggarrange(dotheatmap_linda + theme(legend.position = "none"),p_prevalence_final,ncol=2,widths = c(1,0.3))

dot_heatmap_colon <- p
dot_heatmap_colon
```

```{r, warning=FALSE,results='hide',echo=FALSE}
pdf("../../../figures/merged_sites/QALD/dotplot_colon.pdf",
    height =10,width = 4)
p
dev.off()
```

# Paper-ready visualizations

**Alpha diversity**

```{r, warning=FALSE, fig.width=10, fig.height=4, fig.fullwidth=TRUE,eval=FALSE}
p_A <- alpha_div_plots$`terminal_ileum Country` +
  ggtitle("Terminal ileum")+
  theme(plot.title = element_text(hjust=0.5,face = "bold",size = 15),
        legend.position = "none")  

p_B <-  alpha_div_plots$`colon Country` +
  ggtitle("Colon") +
  theme(plot.title = element_text(hjust=0.5,face = "bold",size = 15),
        legend.position = "none") 

QALD_alpha <- ggarrange(p_A, ggplot() + theme_minimal(),
                      p_B,
                      nrow=1, ncol=3,widths = c(1,0.1,1))

QALD_alpha
```

**Beta diversity**

```{r, warning=FALSE, fig.width=10, fig.height=4, fig.fullwidth=TRUE}
pca_ti <- pca_plots_list$`terminal_ileum genus custom`  +
  ggtitle("Terminal ileum")+
  theme(plot.title = element_text(hjust=0.5,face = "bold",size = 15),
        legend.position = "none")  


pca_colon <- pca_plots_list$`colon genus custom` + 
  ggtitle("Colon")+
  theme(plot.title = element_text(hjust=0.5,face = "bold",size = 15),
        legend.position = "none")  

genus_QALD_beta <- ggarrange(pca_ti,
                           ggplot() + theme_minimal(),
                           pca_colon,
                           ncol=3,
                           widths = c(1,0.1,1))
genus_QALD_beta
```

**Alpha + Beta diversity**

```{r, warning=FALSE, fig.width=10, fig.height=8, fig.fullwidth=TRUE,eval=FALSE}
alpha_beta <- ggarrange(QALD_alpha,genus_QALD_beta,
                        ncol = 1,nrow=2,labels = c("A","B"))
alpha_beta
```

## Supplement

```{r, warning=FALSE, fig.width=10, fig.height=4, fig.fullwidth=TRUE}
genus_QALD_beta <- ggarrange(genus_QALD_beta,ggplot() + theme_minimal(),ncol=2,
                                widths = c(1,0.15))

genus_QALD_beta
```

```{r, warning=FALSE,results='hide',echo=FALSE}
pdf("../../../figures/merged_sites/QALD/supplement3.pdf",paper="a4",height = 4,width = 12)
genus_QALD_beta
dev.off()
```

# Session info

```{r, warning=FALSE}
sessionInfo()
```