---
title: "PSC article analysis Q2 CZ"
output:
  html_document:
    df_print: paged
    toc: true
    toc_float: true
    toc_depth: 6
editor_options: 
  markdown: 
    wrap: 72
---
      # Custom Functions


```{r}
source("custom_functions.R")
```


# 2. Hypothesis: 2. PSC recurrence is associated with a specific composition of the gut microbiome

**rPSC vs non-rPSC vs Healthy analysis on merged data**

-   Alpha diversity –\> group effect, country effect, interaction effect

-   Beta diversity – PERMANOVA, PCA -\> group effect, country effect,
    interaction effect

-   DAA -\>

    -   Group effect – linDA + MaAsLin2 intersection

    -   Country effect – linDA + MaAsLin2 union

**rPSC vs non-rPSC** –by this comparison, we will find the effect of
recurrence (whether the microbial composition is associated with it)

**non-rPSC vs Healthy** – by this comparison, we will find if the
non-rPSC samples are „closer“ to the healthy samples then rPSC samples
in terms of the microbial composition. In other words, does non-rPSC
samples have „healthy microbiome“?

**rPSC vs Healthy** – by this comparison, we will find how much are the
rPSC samples different from healthy samples.

However, the rPSC and non-rPSC samples are not different at all (see
results). Therefore, we can do EXCLUSIVE LEFT JOIN of differentially
abundant taxa from the analyses (rPSC vs Healthy, non-rPSC vs Healthy)


## Data Import

Importing ASV, taxa and metadata tables for both czech and norway
samples.

**Norway**

```{r}
path = "../../data/analysis_ready_data/norway/"
asv_tab_norway <- as.data.frame(fread(file.path(path,"asv_table_norway.csv"),
                                      check.names = FALSE))
taxa_tab_norway <- as.data.frame(fread(file.path(path,"taxa_table_norway.csv"),
                                       check.names = FALSE))
metadata_norway <- as.data.frame(fread(file.path(path,"metadata_norway.csv"),
                                       check.names = FALSE))
```

# Spliting to segments

**TO DO: STATISTICS OF READS**

Merging two countries based on the different matrices - Ileum, Colon.

**Terminal ileum**

```{r}
ileum_data <- merging_data(asv_tab_1=asv_tab_norway,
                           asv_tab_2=NULL,
                           taxa_tab_1=taxa_tab_norway,
                           taxa_tab_2=NULL,
                           metadata_1=metadata_norway,
                           metadata_2=NULL,
                           segment="TI",Q="Q2")

ileum_asv_tab <- ileum_data[[1]]
ileum_taxa_tab <- ileum_data[[2]]
ileum_metadata <- ileum_data[[3]]
```

**Colon**

```{r}
colon_data <- merging_data(asv_tab_1=asv_tab_norway,
                           asv_tab_2=NULL,
                           taxa_tab_1=taxa_tab_norway,
                           taxa_tab_2=NULL,
                           metadata_1=metadata_norway,
                           metadata_2=NULL,
                           segment="colon",Q="Q2")

colon_asv_tab <- colon_data[[1]]
colon_taxa_tab <- colon_data[[2]]
colon_metadata <- colon_data[[3]]
```

# Data Analysis - Terminal ileum

```{r}
segment="terminal_ileum"
```

## Filtering

Rules: - prevalence \> 5% (per group) - nearZeroVar with default
settings - sequencing depth \> 5000 - taxonomic assignment at least
order

**Library size**

```{r, fig.width=5, fig.height=4, fig.fullwidth=TRUE}
read_counts(ileum_asv_tab, line = c(5000,10000))
```

### Sequencing depth

```{r}
data_filt <- seq_depth_filtering(ileum_asv_tab,
                                 ileum_taxa_tab,
                                 ileum_metadata,
                                 seq_depth_threshold = 10000)

filt_ileum_asv_tab <- data_filt[[1]]; alpha_ileum_asv_tab <- filt_ileum_asv_tab
filt_ileum_taxa_tab <- data_filt[[2]]; alpha_ileum_taxa_tab <- filt_ileum_taxa_tab
filt_ileum_metadata <- data_filt[[3]]; alpha_ileum_metadata <- filt_ileum_metadata

seq_step <- dim(filt_ileum_asv_tab)[1]
```

**Library size**

```{r, fig.width=5, fig.height=4, fig.fullwidth=TRUE}
read_counts(filt_ileum_asv_tab,line = c(5000,10000))
```

### NearZeroVar

```{r}
data_filt <- nearzerovar_filtering(filt_ileum_asv_tab, 
                                   filt_ileum_taxa_tab,
                                   filt_ileum_metadata)

filt_ileum_asv_tab <- data_filt[[1]]
filt_ileum_taxa_tab <- data_filt[[2]]
nearzero_step <- dim(filt_ileum_asv_tab)[1]
```

Library size

```{r, fig.width=5, fig.height=4, fig.fullwidth=TRUE}
read_counts(filt_ileum_asv_tab,line = c(5000,10000))
```

### Final Counts

```{r}
final_counts_filtering(ileum_asv_tab,
                       filt_ileum_asv_tab,
                       filt_ileum_metadata,
                       seq_step, 0, nearzero_step) %>% `colnames<-`("Count")
```


## Machine learning

```{r}
path = "../results/Q2_norway/models"
```

### ElasticNet

```{r}
model="enet"
```


#### Genus level

```{r}
level="genus"
```

Aggregate taxa

```{r}
genus_data <- aggregate_taxa(ileum_asv_tab,
                             ileum_taxa_tab,
                             taxonomic_level = level)

ileum_genus_tab <- genus_data[[1]]
ileum_genus_taxa_tab <- genus_data[[2]]
```

##### rPSC vs non-rPSC - all

```{r}
group <- c("rPSC","non-rPSC")
comparison_name <- paste0(group[1], " vs ",group[2])
```

```{r}
model_name <- paste(comparison_name,level,segment)

# prepare the data
filt_ileum_uni_data <- binomial_prep(ileum_genus_tab,
                                     ileum_genus_taxa_tab,
                                     ileum_metadata,
                                     group, 
                                     usage="ml_clr")

# fit the model
enet_model <- glmnet_binomial(filt_ileum_uni_data,
                              sample_method = "atypboot",
                              outcome="Group", 
                              N=10,
                              reuse=TRUE,
                              file=model_name,
                              Q="Q2_norway")

# ROC
roc_c <- roc_curve(enet_model, group)

# save the results
models_summ <- list()
models_cm <- list()
betas <- list()
roc_cs <- list()

# save the results
models_summ[[model_name]] <- enet_model$model_summary
models_cm[[model_name]] <- enet_model$conf_matrices$original
roc_cs[[model_name]] <- enet_model$kfold_rocobjs
betas[[model_name]] <- as.matrix(enet_model$betas)

# see the results
enet_model$model_summary %>% t()
enet_model$conf_matrices
enet_model$plot

roc_c
```

##### rPSC vs non-rPSC - subset

```{r}
group <- c("rPSC","non-rPSC")
comparison_name <- paste0(group[1], " vs ",group[2]," subset")
```

```{r, warning=FALSE}
taxa_to_test <- read.xlsx("../results/Q1_norway/univariate_analysis/significant_taxa_terminal_ileum.xlsx", sheet=" genus healthy vs pre_ltx")

model_name <- paste(comparison_name,level,segment)

# prepare the data
filt_ileum_uni_data <- binomial_prep_psc_effect(ileum_genus_tab,
                                                ileum_genus_taxa_tab, 
                                                ileum_metadata,
                                                taxa_to_test,
                                                usage="ml_clr")

# fit the model
enet_model <- glmnet_binomial(filt_ileum_uni_data,
                              sample_method = "atypboot",
                              outcome="Group", 
                              N=500,
                              reuse=FALSE,
                              file=model_name,
                              Q="Q2_norway")

# ROC
roc_c <- roc_curve(enet_model, group)

# save the results
models_summ <- list()
models_cm <- list()
betas <- list()
roc_cs <- list()

# save the results
models_summ[[model_name]] <- enet_model$model_summary
models_cm[[model_name]] <- enet_model$conf_matrices$original
roc_cs[[model_name]] <- enet_model$kfold_rocobjs
betas[[model_name]] <- as.matrix(enet_model$betas)

# see the results
enet_model$model_summary %>% t()
enet_model$conf_matrices
enet_model$plot

roc_c
```

#### Saving results

```{r}
models_summ_df_ileum <- do.call(rbind, 
  models_summ[grep(segment,names(models_summ),value = TRUE)])

write.csv(models_summ_df_ileum,file.path(path,paste0("elastic_net_",segment,".csv")))
```


### Supplementary models

```{r}
supplements_models <- list()
```

#### CLR-transformed data

##### kNN

```{r}
model="knn"
```

```{r}
level="genus"
```

Aggregate taxa

```{r}
genus_data <- aggregate_taxa(ileum_asv_tab,
                             ileum_taxa_tab,
                             taxonomic_level = level)

ileum_genus_tab <- genus_data[[1]]
ileum_genus_taxa_tab <- genus_data[[2]]
```

**rPSC vs non-rPSC - all**

```{r}
group <- c("rPSC","non-rPSC")
comparison_name <- paste0(group[1], " vs ",group[2])
```

```{r}
model_name <- paste(comparison_name,level,segment)

# prepare the data
filt_ileum_uni_data <- binomial_prep(ileum_genus_tab,
                                     ileum_genus_taxa_tab,
                                     ileum_metadata,
                                     group, 
                                     usage="ml_clr")

# fit the model
knn_model <- knn_binomial(filt_ileum_uni_data,
                              sample_method = "atypboot",
                              outcome="Group", 
                              N=500,
                              reuse=TRUE,
                              file=model_name,
                              Q="Q2_norway")

# ROC curve
roc_c <- roc_curve(knn_model, group)

# save the results
supplements_models[["models_summ"]][["knn_model"]][[model_name]] <- knn_model$model_summary
supplements_models[["roc_cs"]][["knn_model"]][[model_name]] <- knn_model$kfold_rocobjs


# see the results
knn_model$model_summary %>% t()
roc_c
```

**rPSC vs non-rPSC - subset**

```{r}
group <- c("rPSC","non-rPSC")
comparison_name <- paste0(group[1], " vs ",group[2]," subset")
```

```{r, warning=FALSE}
taxa_to_test <- read.xlsx("../results/Q1_norway/univariate_analysis/significant_taxa_terminal_ileum.xlsx", sheet=" genus healthy vs pre_ltx")

model_name <- paste(comparison_name,level,segment)

# prepare the data
filt_ileum_uni_data <- binomial_prep_psc_effect(ileum_genus_tab,
                                                ileum_genus_taxa_tab, 
                                                ileum_metadata,
                                                taxa_to_test,
                                                usage="ml_clr")

# fit the model
knn_model <- knn_binomial(filt_ileum_uni_data,
                              sample_method = "atypboot",
                              outcome="Group", 
                              N=500,
                              reuse=FALSE,
                              file=model_name,
                              Q="Q2_norway")

# ROC curve
roc_c <- roc_curve(knn_model, group)

# save the results
supplements_models[["models_summ"]][["knn_model"]][[model_name]] <- knn_model$model_summary
supplements_models[["roc_cs"]][["knn_model"]][[model_name]] <- knn_model$kfold_rocobjs


# see the results
knn_model$model_summary %>% t()
roc_c
```

##### Random Forest

```{r}
model="rf"
```


```{r}
level="genus"
```

Aggregate taxa

```{r}
genus_data <- aggregate_taxa(ileum_asv_tab,
                             ileum_taxa_tab,
                             taxonomic_level = level)

ileum_genus_tab <- genus_data[[1]]
ileum_genus_taxa_tab <- genus_data[[2]]
```

**rPSC vs non-rPSC - all**

```{r}
group <- c("rPSC","non-rPSC")
comparison_name <- paste0(group[1], " vs ",group[2])
```

```{r}
model_name <- paste(comparison_name,level,segment)

# prepare the data
filt_ileum_uni_data <- binomial_prep(ileum_genus_tab,
                                     ileum_genus_taxa_tab,
                                     ileum_metadata,
                                     group, 
                                     usage="ml_clr")

# fit the model
rf_model <- rf_binomial(filt_ileum_uni_data,
                              sample_method = "atypboot",
                              outcome="Group", 
                              N=500,
                              reuse=TRUE,
                              file=model_name,
                              Q="Q2_norway")

# ROC curve
roc_c <- roc_curve(rf_model, group)

# save the results
supplements_models[["models_summ"]][["rf_model"]][[model_name]] <- rf_model$model_summary
supplements_models[["roc_cs"]][["rf_model"]][[model_name]] <- rf_model$kfold_rocobjs

# see the results
rf_model$model_summary %>% t()
roc_c
```

**rPSC vs non-rPSC - subset**

```{r}
group <- c("rPSC","non-rPSC")
comparison_name <- paste0(group[1], " vs ",group[2]," subset")
```

```{r, warning=FALSE}
taxa_to_test <- read.xlsx("../results/Q1_norway/univariate_analysis/significant_taxa_terminal_ileum.xlsx", sheet=" genus healthy vs pre_ltx")

model_name <- paste(comparison_name,level,segment)

# prepare the data
filt_ileum_uni_data <- binomial_prep_psc_effect(ileum_genus_tab,
                                                ileum_genus_taxa_tab, 
                                                ileum_metadata,
                                                taxa_to_test,
                                                usage="ml_clr")

# fit the model
rf_model <- rf_binomial(filt_ileum_uni_data,
                              sample_method = "atypboot",
                              outcome="Group", 
                              N=500,
                              reuse=FALSE,
                              file=model_name,
                              Q="Q2_norway")

# ROC curve
roc_c <- roc_curve(rf_model, group)

# save the results
supplements_models[["models_summ"]][["rf_model"]][[model_name]] <- rf_model$model_summary
supplements_models[["roc_cs"]][["rf_model"]][[model_name]] <- rf_model$kfold_rocobjs

# see the results
rf_model$model_summary %>% t()
roc_c
```

##### Gradient boosting

```{r}
model="gb"
```

```{r}
level="genus"
```

Aggregate taxa

```{r}
genus_data <- aggregate_taxa(ileum_asv_tab,
                             ileum_taxa_tab,
                             taxonomic_level = level)

ileum_genus_tab <- genus_data[[1]]
ileum_genus_taxa_tab <- genus_data[[2]]
```

**rPSC vs non-rPSC - all**

```{r}
group <- c("rPSC","non-rPSC")
comparison_name <- paste0(group[1], " vs ",group[2])
```

```{r}
model_name <- paste(comparison_name,level,segment)

# prepare the data
filt_ileum_uni_data <- binomial_prep(ileum_genus_tab,
                                     ileum_genus_taxa_tab,
                                     ileum_metadata,
                                     group, 
                                     usage="ml_clr")

# fit the model
gbm_model <- gbm_binomial(filt_ileum_uni_data,
                              sample_method = "atypboot",
                              outcome="Group", 
                              N=10,
                              reuse=TRUE,
                              file=model_name,
                              Q="Q2_norway")

# ROC curve
roc_c <- roc_curve(gbm_model, group)

# save the results
supplements_models[["models_summ"]][["gbm_model"]][[model_name]] <- gbm_model$model_summary
supplements_models[["roc_cs"]][["gbm_model"]][[model_name]] <- gbm_model$kfold_rocobjs

# see the results
gbm_model$model_summary %>% t()
roc_c
```

**rPSC vs non-rPSC - subset**

```{r}
group <- c("rPSC","non-rPSC")
comparison_name <- paste0(group[1], " vs ",group[2]," subset")
```

```{r, warning=FALSE}
taxa_to_test <- read.xlsx("../results/Q1_norway/univariate_analysis/significant_taxa_terminal_ileum.xlsx", sheet=" genus healthy vs pre_ltx")

model_name <- paste(comparison_name,level,segment)

# prepare the data
filt_ileum_uni_data <- binomial_prep_psc_effect(ileum_genus_tab,
                                                ileum_genus_taxa_tab, 
                                                ileum_metadata,
                                                taxa_to_test,
                                                usage="ml_clr")


# fit the model
gbm_model <- gbm_binomial(filt_ileum_uni_data,
                              sample_method = "atypboot",
                              outcome="Group", 
                              N=10,
                              reuse=TRUE,
                              file=model_name,
                              Q="Q2_norway")

# ROC curve
roc_c <- roc_curve(gbm_model, group)

# save the results
supplements_models[["models_summ"]][["gbm_model"]][[model_name]] <- gbm_model$model_summary
supplements_models[["roc_cs"]][["gbm_model"]][[model_name]] <- gbm_model$kfold_rocobjs

# see the results
gbm_model$model_summary %>% t()
roc_c
```



#### Saving results

```{r}
models_list <- list()

for (model_name in names(supplements_models$models_summ)){
  df <- do.call(rbind, supplements_models$models_summ[[model_name]])
  models_list[[model_name]] <- df
}

write.xlsx(models_list,
           file=file.path(path,paste0("supplements_models_",segment,".xlsx")),
           rowNames=TRUE)

```


## Results overview

#### Machine learning

**Main models**

Summary

```{r}
knitr::kable(models_summ_df_ileum %>% dplyr::select(
"alpha","lambda",
"auc_optimism_corrected",
"auc_optimism_corrected_CIL",
"auc_optimism_corrected_CIU"),
             digits=2,caption="Elastic net results")
```

**ROC - Genus level**

```{r}
roc_curve_all_custom(roc_cs[c(1:4)], 
                     Q="Q2_norway",
                     model_name="enet_model")
```

**ROC - Genus level**

```{r, eval=FALSE}
roc_curve_all_custom(roc_cs[c(5:8)],Q="Q2_norway",
                     model_name="enet_model")
```

**Supplementary models**

Summary

```{r}
# Build final dataframe
models_list[["enet_model"]] <- models_summ_df_ileum
final_df <- tibble(row_names = rownames(models_list[[1]]))

# Loop through models and extract required values
for (model_name in names(models_list)) {
  model_df <- models_list[[model_name]]
  
  # Combine AUC_optimism_corrected with its CI values
  final_df[[model_name]] <- paste0(
    round(model_df$auc_optimism_corrected, 3), 
    " (", round(model_df$auc_optimism_corrected_CIL, 3), "; ", 
    round(model_df$auc_optimism_corrected_CIU, 3), ")"
  )
}

knitr::kable(final_df, caption="All models")
```

ROC - ASV

```{r, fig.width=10, fig.height=8, fig.fullwidth=TRUE}
rocs_list <- supplements_models$roc_cs
rocs_list[["enet_model"]] <- roc_cs

plot_list <- list()

for (model_name in names(rocs_list)) {
  plot_list[[model_name]] <- roc_curve_all_custom(rocs_list[[model_name]][c(1:4)],
                       Q="Q2_norway",
                       model_name=model_name)
}

ggarrange(plotlist = plot_list,labels = names(rocs_list),font.label = list(face="plain",size=7))
```

ROC - genus

```{r, fig.width=10, fig.height=8, fig.fullwidth=TRUE, eval=FALSE}
plot_list <- list()

for (model_name in names(rocs_list)) {
  plot_list[[model_name]] <- roc_curve_all_custom(rocs_list[[model_name]][c(5:8)],
                       Q="Q2_norway",
                       model_name=model_name)
}

p <- ggarrange(plotlist = plot_list,labels = names(rocs_list),font.label = list(face="plain",size=7))
p
```

```{r,results='hide'}
pdf("../figures/Q2_norway/models_ileum.pdf",
    height =10,width = 10)
p
dev.off()
```

# Analysis - Colon

```{r}
segment="colon"
```

## Filtering

Rules: - prevalence \> 5% (per group) - nearZeroVar with default
settings - sequencing depth \> 5000 - taxonomic assignment at least
order



Library size

```{r, fig.width=5, fig.height=4, fig.fullwidth=TRUE}
read_counts(colon_asv_tab, line = c(5000,10000))
```

### Sequencing depth

```{r}
data_filt <- seq_depth_filtering(colon_asv_tab,
                                 colon_taxa_tab,
                                 colon_metadata,
                                 seq_depth_threshold = 10000)

filt_colon_asv_tab <- data_filt[[1]]; alpha_colon_asv_tab <- filt_colon_asv_tab
filt_colon_taxa_tab <- data_filt[[2]]; alpha_colon_taxa_tab <- filt_colon_taxa_tab
filt_colon_metadata <- data_filt[[3]]; alpha_colon_metadata <- filt_colon_metadata

seq_step <- dim(filt_colon_asv_tab)[1]
```

Library size

```{r, fig.width=5, fig.height=4, fig.fullwidth=TRUE}
read_counts(filt_colon_asv_tab,line = c(10000))
```

### NearZeroVar

```{r}
data_filt <- nearzerovar_filtering(filt_colon_asv_tab,
                                   filt_colon_taxa_tab,
                                   filt_colon_metadata)

filt_colon_asv_tab <- data_filt[[1]]
filt_colon_taxa_tab <- data_filt[[2]]
nearzero_step <- dim(filt_colon_asv_tab)[1]
```

Library size

```{r, fig.width=5, fig.height=4, fig.fullwidth=TRUE}
read_counts(filt_colon_asv_tab,line = c(5000,10000))
```

Check zero depth

```{r}
data_filt <- check_zero_depth(filt_colon_asv_tab, 
                              filt_colon_taxa_tab, 
                              filt_colon_metadata)

filt_colon_asv_tab <- data_filt[[1]]; 
filt_colon_taxa_tab <- data_filt[[2]]; 
filt_colon_metadata <- data_filt[[3]]; 
```

Library size

```{r, fig.width=5, fig.height=4, fig.fullwidth=TRUE}
read_counts(filt_colon_asv_tab,line = c(5000,10000))
```

### Final Counts

```{r}
final_counts_filtering(colon_asv_tab,
                       filt_colon_asv_tab,
                       filt_colon_metadata,
                       seq_step, 0, nearzero_step)
```


## Machine learning

```{r}
path = "../results/Q2_norway/models"
```

### ElasticNet

```{r}
model="enet"
```


#### Genus level

```{r}
level="genus"
```

Aggregate taxa

```{r}
genus_data <- aggregate_taxa(colon_asv_tab,
                             colon_taxa_tab,
                             taxonomic_level = level)

colon_genus_tab <- genus_data[[1]]
colon_genus_taxa_tab <- genus_data[[2]]
```

##### rPSC vs non-rPSC - all

```{r}
group <- c("rPSC","non-rPSC")
comparison_name <- paste0(group[1], " vs ",group[2])
```

```{r}
model_name <- paste(comparison_name,level,segment)

# prepare the data
filt_colon_uni_data <- binomial_prep(colon_genus_tab,
                                     colon_genus_taxa_tab,
                                     colon_metadata,
                                     group, 
                                     usage="ml_clr",
                                     patient = TRUE)

# fit the model
enet_model <- glmnet_binomial(filt_colon_uni_data,
                              sample_method = "atypboot",
                              outcome="Group", 
                              N=500,
                              clust_var="Patient",
                              reuse=TRUE,
                              file=model_name,
                              Q="Q2_norway")

# ROC
roc_c <- roc_curve(enet_model, group)

# save the results
models_summ[[model_name]] <- enet_model$model_summary
models_cm[[model_name]] <- enet_model$conf_matrices$original
roc_cs[[model_name]] <- enet_model$kfold_rocobjs
betas[[model_name]] <- as.matrix(enet_model$betas)

# see the results
enet_model$model_summary %>% t()
enet_model$conf_matrices
enet_model$plot

roc_c
```

##### rPSC vs non-rPSC - subset

```{r}
group <- c("rPSC","non-rPSC")
comparison_name <- paste0(group[1], " vs ",group[2]," subset")
```

```{r, warning=FALSE}
taxa_to_test <- read.xlsx("../results/Q1_norway/univariate_analysis/significant_taxa_colon.xlsx", sheet=" genus healthy vs pre_ltx")

model_name <- paste(comparison_name,level,segment)

# prepare the data
filt_colon_uni_data <- binomial_prep_psc_effect(colon_genus_tab,
                                                colon_genus_taxa_tab, 
                                                colon_metadata,
                                                taxa_to_test,
                                                usage="ml_clr",
                                                patient=TRUE)

# fit the model
enet_model <- glmnet_binomial(filt_colon_uni_data,
                              sample_method = "atypboot",
                              outcome="Group", 
                              N=500,
                              clust_var="Patient",
                              reuse=FALSE,
                              file=model_name,
                              Q="Q2_norway")

# ROC
roc_c <- roc_curve(enet_model, group)

# save the results
models_summ[[model_name]] <- enet_model$model_summary
models_cm[[model_name]] <- enet_model$conf_matrices$original
roc_cs[[model_name]] <- enet_model$kfold_rocobjs
betas[[model_name]] <- as.matrix(enet_model$betas)

# see the results
enet_model$model_summary %>% t()
enet_model$conf_matrices
enet_model$plot

roc_c
```
#### Saving results

```{r}
models_summ_df_colon <- do.call(rbind, 
  models_summ[grep(segment,names(models_summ),value = TRUE)])

write.csv(models_summ_df_colon,file.path(path,paste0("elastic_net_",segment,".csv")))
```

### Supplementary models

#### CLR-transformed data

##### kNN

```{r}
model="knn"
```


```{r}
level="genus"
```

Aggregate taxa

```{r}
genus_data <- aggregate_taxa(colon_asv_tab,
                             colon_taxa_tab,
                             taxonomic_level = level)

colon_genus_tab <- genus_data[[1]]
colon_genus_taxa_tab <- genus_data[[2]]
```

**rPSC vs non-rPSC - all**

```{r}
group <- c("rPSC","non-rPSC")
comparison_name <- paste0(group[1], " vs ",group[2])
```

```{r}
model_name <- paste(comparison_name,level,segment)

# prepare the data
filt_colon_uni_data <- binomial_prep(colon_genus_tab,
                                     colon_genus_taxa_tab,
                                     colon_metadata,
                                     group, 
                                     usage="ml_clr",
                                     patient = TRUE)

# fit the model
knn_model <- knn_binomial(filt_colon_uni_data,
                              sample_method = "atypboot",
                              outcome="Group", 
                              N=500,
                          clust_var="Patient",
                              reuse=TRUE,
                              file=model_name,
                              Q="Q2_norway")

# ROC curve
roc_c <- roc_curve(knn_model, group)

# save the results
supplements_models[["models_summ"]][["knn_model"]][[model_name]] <- knn_model$model_summary
supplements_models[["roc_cs"]][["knn_model"]][[model_name]] <- knn_model$kfold_rocobjs


# see the results
knn_model$model_summary %>% t()
roc_c
```

**rPSC vs non-rPSC - subset**

```{r}
group <- c("rPSC","non-rPSC")
comparison_name <- paste0(group[1], " vs ",group[2]," subset")
```

```{r, warning=FALSE}
taxa_to_test <- read.xlsx("../results/Q1_norway/univariate_analysis/significant_taxa_colon.xlsx", sheet=" genus healthy vs pre_ltx")

model_name <- paste(comparison_name,level,segment)

# prepare the data
filt_colon_uni_data <- binomial_prep_psc_effect(colon_genus_tab,
                                                colon_genus_taxa_tab, 
                                                colon_metadata,
                                                taxa_to_test,
                                                usage="ml_clr",
                                                patient=TRUE)

# fit the model
knn_model <- knn_binomial(filt_colon_uni_data,
                              sample_method = "atypboot",
                              outcome="Group", 
                              N=500,
                          clust_var="Patient",
                              reuse=FALSE,
                              file=model_name,
                              Q="Q2_norway")

# ROC curve
roc_c <- roc_curve(knn_model, group)

# save the results
supplements_models[["models_summ"]][["knn_model"]][[model_name]] <- knn_model$model_summary
supplements_models[["roc_cs"]][["knn_model"]][[model_name]] <- knn_model$kfold_rocobjs


# see the results
knn_model$model_summary %>% t()
roc_c
```

##### Random Forest

```{r}
model="rf"
```


```{r}
level="genus"
```

Aggregate taxa

```{r}
genus_data <- aggregate_taxa(colon_asv_tab,
                             colon_taxa_tab,
                             taxonomic_level = level)

colon_genus_tab <- genus_data[[1]]
colon_genus_taxa_tab <- genus_data[[2]]
```

**rPSC vs non-rPSC - all**

```{r}
group <- c("rPSC","non-rPSC")
comparison_name <- paste0(group[1], " vs ",group[2])
```

```{r}
model_name <- paste(comparison_name,level,segment)

# prepare the data
filt_colon_uni_data <- binomial_prep(colon_genus_tab,
                                     colon_genus_taxa_tab,
                                     colon_metadata,
                                     group, 
                                     usage="ml_clr",
                                     patient = TRUE)

# fit the model
rf_model <- rf_binomial(filt_colon_uni_data,
                              sample_method = "atypboot",
                              outcome="Group", 
                              N=500,
                        clust_var="Patient",
                              reuse=TRUE,
                              file=model_name,
                              Q="Q2_norway")

# ROC curve
roc_c <- roc_curve(rf_model, group)

# save the results
supplements_models[["models_summ"]][["rf_model"]][[model_name]] <- rf_model$model_summary
supplements_models[["roc_cs"]][["rf_model"]][[model_name]] <- rf_model$kfold_rocobjs

# see the results
rf_model$model_summary %>% t()
roc_c
```

**rPSC vs non-rPSC - subset**

```{r}
group <- c("rPSC","non-rPSC")
comparison_name <- paste0(group[1], " vs ",group[2]," subset")
```

```{r, warning=FALSE}
taxa_to_test <- read.xlsx("../results/Q1_norway/univariate_analysis/significant_taxa_colon.xlsx", sheet=" genus healthy vs pre_ltx")

model_name <- paste(comparison_name,level,segment)

# prepare the data
filt_colon_uni_data <- binomial_prep_psc_effect(colon_genus_tab,
                                                colon_genus_taxa_tab, 
                                                colon_metadata,
                                                taxa_to_test,
                                                usage="ml_clr",
                                                patient=TRUE)
# fit the model
rf_model <- rf_binomial(filt_colon_uni_data,
                              sample_method = "atypboot",
                              outcome="Group", 
                              N=500,
                        clust_var="Patient",
                              reuse=FALSE,
                              file=model_name,
                              Q="Q2_norway")

# ROC curve
roc_c <- roc_curve(rf_model, group)

# save the results
supplements_models[["models_summ"]][["rf_model"]][[model_name]] <- rf_model$model_summary
supplements_models[["roc_cs"]][["rf_model"]][[model_name]] <- rf_model$kfold_rocobjs

# see the results
rf_model$model_summary %>% t()
roc_c
```


##### Gradient boosting

```{r}
model="gb"
```


```{r}
level="genus"
```

Aggregate taxa

```{r}
genus_data <- aggregate_taxa(colon_asv_tab,
                             colon_taxa_tab,
                             taxonomic_level = level)

colon_genus_tab <- genus_data[[1]]
colon_genus_taxa_tab <- genus_data[[2]]
```

**rPSC vs non-rPSC - all**

```{r}
group <- c("rPSC","non-rPSC")
comparison_name <- paste0(group[1], " vs ",group[2])
```

```{r}
model_name <- paste(comparison_name,level,segment)

# prepare the data
filt_colon_uni_data <- binomial_prep(colon_genus_tab,
                                     colon_genus_taxa_tab,
                                     colon_metadata,
                                     group, 
                                     usage="ml_clr",
                                     patient = TRUE)

# fit the model
gbm_model <- gbm_binomial(filt_colon_uni_data,
                              sample_method = "atypboot",
                              outcome="Group", 
                              N=500,
                          clust_var="Patient",
                              reuse=TRUE,
                              file=model_name,
                              Q="Q2_norway")

# ROC curve
roc_c <- roc_curve(gbm_model, group)

# save the results
supplements_models[["models_summ"]][["gbm_model"]][[model_name]] <- gbm_model$model_summary
supplements_models[["roc_cs"]][["gbm_model"]][[model_name]] <- gbm_model$kfold_rocobjs

# see the results
gbm_model$model_summary %>% t()
roc_c
```

**rPSC vs non-rPSC - subset**

```{r}
group <- c("rPSC","non-rPSC")
comparison_name <- paste0(group[1], " vs ",group[2]," subset")
```

```{r, warning=FALSE}
taxa_to_test <- read.xlsx("../results/Q1_norway/univariate_analysis/significant_taxa_colon.xlsx", sheet=" genus healthy vs pre_ltx")

model_name <- paste(comparison_name,level,segment)

# prepare the data
filt_colon_uni_data <- binomial_prep_psc_effect(colon_genus_tab,
                                                colon_genus_taxa_tab, 
                                                colon_metadata,
                                                taxa_to_test,
                                                usage="ml_clr",
                                                patient=TRUE)

# fit the model
gbm_model <- gbm_binomial(filt_colon_uni_data,
                              sample_method = "atypboot",
                              outcome="Group", 
                              N=500,
                          clust_var="Patient",
                              reuse=FALSE,
                              file=model_name,
                              Q="Q2_norway")

# ROC curve
roc_c <- roc_curve(gbm_model, group)

# save the results
supplements_models[["models_summ"]][["gbm_model"]][[model_name]] <- gbm_model$model_summary
supplements_models[["roc_cs"]][["gbm_model"]][[model_name]] <- gbm_model$kfold_rocobjs

# see the results
gbm_model$model_summary %>% t()
roc_c
```

#### Saving results

```{r}
models_list <- list()

for (model_name in names(supplements_models$models_summ)){
  df <- do.call(rbind, supplements_models$models_summ[[model_name]])
  models_list[[model_name]] <- df
}

write.xlsx(models_list,
           file=file.path(path,paste0("supplements_models_",segment,".xlsx")),
           rowNames=TRUE)

```

## Results overview

#### Machine learning

**Main models**

Summary

```{r}
knitr::kable(models_summ_df_colon %>% dplyr::select(
"alpha","lambda",
"auc_optimism_corrected",
"auc_optimism_corrected_CIL",
"auc_optimism_corrected_CIU"),
             digits=3,caption="Elastic net results")
```

**ROC - ASV level**

```{r}
roc_curve_all_custom(roc_cs[c(5:8)], 
                     Q="Q2_norway",
                     model_name="enet_model")
```

**ROC - Genus level**

```{r, eval=FALSE}
roc_curve_all_custom(roc_cs[c(13:16)],Q="Q2_norway",
                     model_name="enet_model")
```

**Supplementary models**

Summary

```{r}
# Build final dataframe
models_list[["enet_model"]] <- rbind(models_summ_df_ileum,models_summ_df_colon)
final_df <- tibble(row_names = rownames(models_list[[1]]))

# Loop through models and extract required values
for (model_name in names(models_list)) {
  model_df <- models_list[[model_name]]
  
  # Combine AUC_optimism_corrected with its CI values
  final_df[[model_name]] <- paste0(
    round(model_df$auc_optimism_corrected, 3), 
    " (", round(model_df$auc_optimism_corrected_CIL, 3), "; ", 
    round(model_df$auc_optimism_corrected_CIU, 3), ")"
  )
}

knitr::kable(final_df, caption="All models")
write.csv(final_df,file=file.path(path,"AUC_all_models.csv"),row.names = FALSE)
```

ROC - ASV

```{r, fig.width=10, fig.height=8, fig.fullwidth=TRUE}
rocs_list <- supplements_models$roc_cs
rocs_list[["enet_model"]] <- roc_cs

plot_list <- list()

for (model_name in names(rocs_list)) {
  plot_list[[model_name]] <- roc_curve_all_custom(rocs_list[[model_name]][c(1:4)],
                       Q="Q2_norway",
                       model_name=model_name)
}

ggarrange(plotlist = plot_list,labels = names(rocs_list),font.label = list(face="plain",size=7))
```

ROC - genus

```{r, fig.width=10, fig.height=8, fig.fullwidth=TRUE}
plot_list <- list()

for (model_name in names(rocs_list)) {
  plot_list[[model_name]] <- roc_curve_all_custom(rocs_list[[model_name]][c(5:8)],
                       Q="Q2_norway",
                       model_name=model_name)
}

p <- ggarrange(plotlist = plot_list,labels = names(rocs_list),font.label = list(face="plain",size=7))
```

```{r,results='hide'}
pdf("../figures/Q2_norway/models_colon.pdf",
    height =10,width = 10)
p
dev.off()
```

