---
title: "PSC, clinical impact"
output:
  html_document:
    df_print: paged
    toc: true
    toc_float: true
    toc_depth: 6
editor_options: 
  markdown: 
    wrap: 72
---

```{r}
library(openxlsx)
library(tidyverse)
library(reshape2)
library(picante)
library(tidyr)
library(ggpubr)
library(mice)
source("custom_functions.R")
```

# Functions

```{r}
clinical_boxplot <- function(metadata,variable){
  metadata_var <- metadata %>% dplyr::select(all_of(c("PatientID",
                                                       variable, "Group","Country"))) %>% as.data.frame()
  metadata_var[,variable] <- as.numeric(metadata_var[,variable])
  metadata_var_without_na <- metadata_var[!is.na(metadata_var[,variable]),]
  colors <- c("#309f87","#f9c675","#F08080","#A00000")
  if (length(unique(metadata_var_without_na$Group))==1) colors <- c("#f9c675")
  if (!"healthy" %in% unique(metadata_var_without_na$Group)) colors <- c("#f9c675","#F08080","#A00000")
  
p <- ggplot(metadata_var_without_na) + 
  geom_boxplot(aes(x=Group, y=!!sym(variable)),outliers = FALSE) + 
    geom_jitter(width = 0.2,height = 0,aes(x=Group, y=!!sym(variable), color=Group,shape=Country),size=2) +
  scale_fill_manual(values=colors) + 
scale_color_manual(values=colors) + 
   theme_minimal() + 
    theme(panel.border = element_rect(color = "black", fill = NA, size = 0),
          axis.ticks.x = element_line(size=0.3,color = "black"),
          axis.ticks.y = element_line(size=0.3,color="black"),
          axis.ticks.length = unit(4,"pt"),
          panel.grid = element_blank())
return(p)
}
```

```{r}
clinical_correlation <- function(metadata,variable,level,
                                 segment="ileum"){
  set.seed(123)
  level <- tolower(level)
  
  metadata_var <- metadata %>% 
    dplyr::select(all_of(c("SampleID", "PatientID",
                paste0("dys_filtered_",level),
                variable, "Group")))
  
  metadata_var[,variable] <- as.numeric(metadata_var[,variable])
  metadata_var <- drop_na(metadata_var)
  
  if (segment=="colon"){
    n_iterations <- 100
    corrs <- list()
    for (i in 1:n_iterations){
      random_sampled_df <- metadata_var %>%
      group_by(PatientID) %>%  # Group by PatientID
      slice_sample(n = 1) %>% # Randomly pick one row per PatientID
      ungroup()
      
      corr <- cor.table(
        random_sampled_df[,c(variable,paste0("dys_filtered_",level))], 
        cor.method="spearman"
        )
      
      corr$P <- corr$P[2]
      corr$r <- corr$r[2]
      corrs[[i]] <- corr
    }
    
    r_values_list <- unlist(map(corrs, ~ .x$r))  # List of r matrices
    p_values_list <- unlist(map(corrs, ~ .x$P))  # List of P matrices
    mean_r <- round(median(r_values_list),2)
    
    p_value <- quantile(p_values_list,0.9)
    corr <- list()
    corr$r <- mean_r
    corr$P <- p_value
  } else {
    corr <- cor.table(
      metadata_var[,c(variable,paste0("dys_filtered_",level))],
      cor.method="spearman")
    
    corr$P <- corr$P[2]
    corr$r <- round(corr$r[2],2)
  }
  
  return(corr)
}

clinical_correlation_abundances <- function(metadata,variable,taxon,level,
                                            segment="ileum"){
  set.seed(123)
  level <- tolower(level)
  
  metadata_var <- metadata %>% dplyr::select(all_of(c("SampleID", "PatientID",
                                                       variable,taxon,
                                                      "Group")))
  metadata_var[,variable] <- as.numeric(metadata_var[,variable])
  metadata_var <- drop_na(metadata_var)
  
  if (segment=="colon"){
    n_iterations <- 100
    corrs <- list()
    for (i in 1:n_iterations){
      random_sampled_df <- metadata_var %>%
      group_by(PatientID) %>%  # Group by PatientID
      slice_sample(n = 1) %>% # Randomly pick one row per PatientID
      ungroup()
      
      corr <- cor.table(random_sampled_df[,c(variable,taxon)], cor.method="spearman")
      corr$P <- corr$P[2]
      corr$r <- corr$r[2]
      corrs[[i]] <- corr
    }
     r_values_list <- unlist(map(corrs, ~ .x$r))  # List of r matrices
    p_values_list <- unlist(map(corrs, ~ .x$P))  # List of P matrices
    mean_r <- round(median(r_values_list),2)
    
    p_value <- quantile(p_values_list,0.9)
    #min_p <- min(p_values_list)
    #min_p <- round(min_p,2)
    #min_p <- ifelse(min_p<=0.05,"< 0.05",ifelse(min_p<=0.01,"< 0.01",ifelse(min_p<=0.001,"< 0.001",paste0("= ",min_p))))
    
    #max_p <- max(p_values_list)
    #max_p <- round(max_p,2)
    #max_p <- ifelse(max_p<=0.05,"< 0.05",ifelse(max_p<=0.01,"< 0.01",ifelse(max_p<=0.001,"< 0.001",max_p)))
    
    #if (min_p==max_p) combined_df <- min_p
    #else {
    #  max_p <- round(quantile(corr$P,0.9),2)
    #  combined_df <- ifelse(max_p<=0.05,"< 0.05",ifelse(max_p<=0.01,"< 0.01",ifelse(max_p<=0.001,"< 0.001",paste("=",max_p) )))
    #}
    corr <- list()
    corr$r <- mean_r
    #corr$P <- combined_df
    corr$P <- p_value
  } else {
    corr <- cor.table(metadata_var[,c(variable,taxon)], cor.method="spearman")
    #max_p <- round(corr$P[2],2)
    #max_p <- 
    #  max_p <- ifelse(max_p<=0.05,"< 0.05",ifelse(max_p<=0.01,"< 0.01",ifelse(max_p<=0.001,"< 0.001",paste0("= ",max_p))))
    #corr$P <- max_p
    corr$r <- round(corr$r[2],2)
    corr$P <- corr$P[2]
  }
  
  return(corr)
}

clinical_scatter <- function(corr,metadata,variable,level){
  level <- tolower(level)
  df_for_scatter_plot <- metadata %>% 
    dplyr::select(all_of(c("SampleID", "PatientID",
                           paste0("dys_filtered_",level),
                           variable, "Group","Country")))
  
  df_for_scatter_plot[,variable] <- as.numeric(df_for_scatter_plot[,variable])
  
  df_for_scatter_plot <- drop_na(df_for_scatter_plot)
  
  colors <- c("#309f87","#f9c675","#F08080","#A00000")
  if (!"healthy" %in% unique(df_for_scatter_plot$Group)) colors <- c("#f9c675","#F08080","#A00000")
  
  r_corr <- corr$r
  p_corr <- corr$P
  p_corr <- ifelse(p_corr < 0.001, "<0.001", ifelse(p_corr < 0.01, "<0.01", ifelse(p_corr < 0.05, "<0.05", paste0("=",round(p_corr,3)))))
  
  x_var <- paste0("dys_filtered_",level)
  x_position <- (max(df_for_scatter_plot[,x_var])) - 0.2*(max(df_for_scatter_plot[,x_var]) - min(df_for_scatter_plot[,x_var]))
  
  y_position <- max(df_for_scatter_plot[,variable]) - 0.05*(max(df_for_scatter_plot[,variable]) - min(df_for_scatter_plot[,variable]))
  
  labels <- paste0("r = ",r_corr, "\\np ",p_corr)
  p <- ggplot(df_for_scatter_plot) + 
  geom_point(aes(x=!!sym(x_var), y=!!sym(variable), color=Group,shape=Country)) + 
  geom_smooth(aes(x=!!sym(x_var), y=!!sym(variable)),method=lm, se=TRUE) +
    geom_text(aes(x=x_position, y=y_position),
                  label=gsub("\\\\n", "\n", labels),hjust = 0) + 
  scale_color_manual(values=colors) + 
  theme_minimal() + 
    theme(panel.border = element_rect(color = "black", fill = NA, size = 0),
          axis.ticks.x = element_line(size=0.3,color = "black"),
          axis.ticks.y = element_line(size=0.3,color="black"),
          axis.ticks.length = unit(4,"pt"),
          panel.grid = element_blank()) + 
  xlab("MRI")
  return(p)
}

clinical_scatter_abundances <- function(corr,metadata,variable,taxon,level,
                                        size=4){
  level <- tolower(level)
  df_for_scatter_plot <- metadata %>% dplyr::select(all_of(c("SampleID", "PatientID","Country",
                                                       taxon,
                                                       variable, "Group")))
  df_for_scatter_plot[,variable] <- as.numeric(df_for_scatter_plot[,variable])
  df_for_scatter_plot <- drop_na(df_for_scatter_plot)
  colors <- c("#309f87","#f9c675","#F08080","#A00000")
  if (!"healthy" %in% unique(df_for_scatter_plot$Group)) colors <- c("#f9c675","#F08080","#A00000")
  
  x_var <- taxon
  r_corr <- corr$r
  p_corr <- corr$P
  
  x_position <- (max(df_for_scatter_plot[,x_var])) - 0.2*(max(df_for_scatter_plot[,x_var]) - min(df_for_scatter_plot[,x_var]))
  y_position <- max(df_for_scatter_plot[,variable]) - 0.05*(max(df_for_scatter_plot[,variable]) - min(df_for_scatter_plot[,variable]))
  labels <- paste0("r = ",r_corr, "\\np ",p_corr)
  p <- ggplot(df_for_scatter_plot) + 
  geom_point(aes(x=!!sym(x_var), y=!!sym(variable), color=Group,shape=Country)) + 
  geom_smooth(aes(x=!!sym(x_var), y=!!sym(variable)),method=lm, se=TRUE) +
    geom_text(aes(x=x_position, y=y_position,label=gsub("\\\\n", "\n", labels)),size=size) + 
  scale_color_manual(values=colors) + 
  theme_minimal() + 
    theme(panel.border = element_rect(color = "black", fill = NA, size = 0),
          axis.ticks.x = element_line(size=0.3,color = "black"),
          axis.ticks.y = element_line(size=0.3,color="black"),
          axis.ticks.length = unit(4,"pt"),
          panel.grid = element_blank()) 
  return(p)
}

```

```{r}
clinical_contingency <- function(metadata,variable){
  metadata_var <- metadata %>% dplyr::select(all_of(c("PatientID",
                                                       variable, "Group"))) %>% as.data.frame()
  metadata_var_without_na <- metadata_var[!is.na(metadata_var[,variable]),]
  cont_table <- as.data.frame(table(metadata_var_without_na$Group, metadata_var_without_na[,variable]))
  
  p <- ggplot(cont_table, aes(x = Var1, y = Var2, fill = Freq)) +
  geom_tile(color = "white") +  # Tile background color
  geom_text(aes(label = Freq), color = "black", size = 6) +  # Add numbers
  scale_fill_gradient(low = "white", high = "blue") +  # Color gradient
  labs(title = "Contingency Table with Counts", 
       x = "Group", y = variable) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
  return(p)
}

clinical_boxplot_categorical <- function(metadata, variable,level){
  level=tolower(level)
  metadata_var <- metadata %>% dplyr::select(all_of(c("SampleID", "PatientID",
                                                       paste0("dys_filtered_",level),
                                                       variable, "Group")))
  metadata_var <- drop_na(metadata_var)
  metadata_var[,variable] <- as.factor(metadata_var[,variable])

  y_var <- paste0("dys_filtered_",level)
  p <- ggplot(metadata_var) + 
    geom_boxplot(aes(x=!!sym(variable), y=!!sym(y_var)),outliers = FALSE) + 
      geom_jitter(width = 0.2,height = 0,aes(x=!!sym(variable), y=!!sym(y_var),
                                             color=Group),size=2) +
    scale_fill_manual(values=c("#f9c675","#F08080","#A00000")) + 
  scale_color_manual(values=c("#f9c675","#F08080","#A00000")) + 
    theme_minimal() + 
    theme(panel.border = element_rect(color = "black", fill = NA, size = 0),
          axis.ticks.x = element_line(size=0.3,color = "black"),
          axis.ticks.y = element_line(size=0.3,color="black"),
          axis.ticks.length = unit(4,"pt"),
          panel.grid = element_blank())
  return(p)
p
}
```

```{r}
subprepare_for_heatmap <- function(corrs_segment,MDI=TRUE){
    corrs_segment <- corrs_segment[!grepl("_log_",names(corrs_segment))]
    names(corrs_segment) <- gsub("score_","",names(corrs_segment))
    # Example list structure for demonstration
    # Initialize empty lists for r and P values
    r_values <- list()
    p_values <- list()
    
    # Extract segment and clinical variables
    for (name in names(corrs_segment)) {
      if (MDI) {
        clinical_var <- sub("^(.*?_){2}", "",  name)
        if (all(grepl("ileum",names(corrs_segment)))) {
        segment <- "MRI terminal ileum"
      } else if (all(grepl("colon",names(corrs_segment)))) {
        segment <- "MRI colon"
      } else (message("PROBLEM with segment"))
        
      }
      else {
        segment <- sub("^(.*?_){3}", "",  name)
        clinical_var <- sub(segment,"", sub("^(.*?_){2}", "",  name))
        clinical_var <- gsub("_","",clinical_var)
      }
  

      # Fill the r and P lists
      if (!is.null(corrs_segment[[name]]$r)) {
        r_values[[segment]][clinical_var] <- corrs_segment[[name]]$r
      }
      
      if (!is.null(corrs_segment[[name]]$P)) {
        p_values[[segment]][clinical_var] <- corrs_segment[[name]]$P
      }
    }

    # Convert the lists to data frames
    r_df <- do.call(rbind, lapply(r_values, function(x) setNames(as.data.frame(t(x)), names(x))))
    p_df <- do.call(rbind, lapply(p_values, function(x) setNames(as.data.frame(t(x)), names(x))))
    
    # Add row names as segment names
    rownames(r_df) <- names(r_values)
    rownames(p_df) <- names(p_values)
    
    if (MDI) pd_df_corrected <- as.data.frame(p.adjust(p_df,method="BH"))
    else pd_df_corrected <- as.data.frame(apply(p_df,2,function(x) p.adjust(x, method="BH")))
    p_df_sig <- pd_df_corrected
    p_df_sig[,] <- ""
    
    p_df_sig[pd_df_corrected < 0.05] <- "*"
    p_df_sig[pd_df_corrected < 0.01] <- "**"
    p_df_sig[pd_df_corrected < 0.001] <- "***"
    
    if (MDI) p_df_sig <- t(p_df_sig) %>% 
      as.data.frame()  %>% 
      `rownames<-`(segment)
    
    return(list(p_df_sig,r_df))
}

prepare_for_heatmap <- function(corrs_ileum,corrs_colon){

  ileum_list <- subprepare_for_heatmap(corrs_ileum)
  colon_list <- subprepare_for_heatmap(corrs_colon)

  p_df_sig <- rbind(ileum_list[[1]],colon_list[[1]])
  r_df <- rbind(ileum_list[[2]],colon_list[[2]])

  return(list(p_df_sig,r_df))
  
}

```


# Import clinical data

## CZ

```{r}
# clinical metadata
metadata_clinical <- read.csv("../../data/clinical_data/clinical_metadata_cz.csv")
metadata_clinical$PatientID <- as.character(metadata_clinical$PatientID)

# DYSBIOSIS
metadata_dysbiosis <- read.csv("../../data/clinical_data/recurrence_metadata.csv") %>%
  dplyr::filter(Country=="CZ")

# ALPHA DIVERSITY
metadata_alpha_ileum <- read.csv(
  "../results/Q1/alpha_diversity/alpha_indices_terminal_ileum.csv") %>%
  dplyr::filter(Country=="CZ")

metadata_alpha_colon <- read.csv(
  "../results/Q1/alpha_diversity/alpha_indices_colon.csv") %>%
  dplyr::filter(Country=="CZ")

metadata_alpha <- rbind(metadata_alpha_ileum,metadata_alpha_colon) %>%
  mutate(PatientID=Patient) %>%
  dplyr::select(-c(Patient, Group))

# MERGING
metadata_cz <- full_join(metadata_clinical, metadata_dysbiosis, by=c("SampleID","Matrix","PatientID","Group","Country"))

metadata_cz <- full_join(metadata_cz, metadata_alpha, by=c("SampleID","PatientID","Country"))

metadata_cz$Group <- factor(metadata_cz$Group,levels = c("healthy","pre_ltx","non-rPSC","rPSC"))
```

## NO

```{r}
# clinical metadata
metadata_clinical <- read.csv("../../data/clinical_data/clinical_metadata_no.csv")
metadata_clinical %<>% mutate(PatientID=subjectid,
                              Matrix=segment) %>%
  dplyr::select(-subjectid,-segment)
metadata_clinical$PatientID <- as.character(paste0("NO_",metadata_clinical$PatientID))

# DYSBIOSIS
metadata_dysbiosis <- read.csv("../../data/clinical_data/dysbiosis_metadata.csv") %>%
  dplyr::filter(Country=="NO")  %>%
  dplyr::select(-c(Patient))

# ALPHA DIVERSITY
metadata_alpha_ileum <- read.csv(
  "../results/Q1/alpha_diversity/alpha_indices_terminal_ileum.csv") %>%
  dplyr::filter(Country=="NO")

metadata_alpha_colon <- read.csv(
  "../results/Q1/alpha_diversity/alpha_indices_colon.csv") %>%
  dplyr::filter(Country=="NO")

metadata_alpha <- rbind(metadata_alpha_ileum,metadata_alpha_colon) %>%
  mutate(PatientID=Patient) %>%
  dplyr::select(-c(Patient, Group))

# MERGING
metadata_no <- full_join(metadata_clinical, metadata_dysbiosis, by=c("SampleID","Matrix","PatientID","Group","Country"))

metadata_no <- full_join(metadata_no, metadata_alpha, by=c("SampleID","PatientID","Country"))

metadata_no$Group <- factor(metadata_no$Group,levels = c("healthy","pre_ltx","non-rPSC","rPSC"))
```

## Merging metadata

```{r}
metadata_cz %<>% dplyr::mutate(Calprotectin=Fecal.calprotectin,
                             AOM=AOM_score,
                             APRI=APRI_score,
                             FIB4=FIB4_score) %>%
  dplyr::select(SampleID,Matrix,PatientID,Group,Country,Bilirubin,ALP,Calprotectin,
                MAYO_PSC,AOM,APRI,FIB4,Platelets,AST,Creatinine,
                Albumin,ALT,PSC_IBD,GGT,INR,Albumin,Nancy_max,eMAYO,MAYO_dai,
                dys_unfiltered_asv,dys_unfiltered_genus,
                dys_filtered_asv,dys_filtered_genus,
                Observe,Shannon,Simpson,Pielou)

metadata_no %<>% dplyr::mutate(Platelets=TRC,
                             Creatinine=Kreatinin,
                             MAYO_PSC=Mayo_score,
                             AST=ASAT/60,
                             ALT=ALAT/60,
                             PSC_IBD=IBD,
                             Bilirubin=Bilirubin*17.1,
                             ALP=ALP/60,
                             Albumin=Albumin*10) %>%
  dplyr::mutate(PSC_IBD = case_when(
      PSC_IBD == "no_ibd" ~ "0",
      PSC_IBD == "ibd" ~ "1",
      TRUE ~ Group  # Keep other values as is
    )) %>%
  dplyr::select(SampleID,Matrix,PatientID,Group,Country,Bilirubin,ALP,Calprotectin,
                MAYO_PSC,AOM,APRI,FIB4,Platelets,AST,Creatinine,
                Albumin,ALT,PSC_IBD,
                dys_unfiltered_asv,dys_unfiltered_genus,
                dys_filtered_asv,dys_filtered_genus,
                Observe,Shannon,Simpson,Pielou) 

metadata_final <- merge(metadata_cz,metadata_no,all = TRUE)
```


```{r}
variables <- c("PatientID", "Group", "Country",
               "Bilirubin", "Albumin",
               "ALP", "Platelets",
               "Calprotectin",
               "MAYO_PSC","AOM", 
               "APRI","INR",
               "FIB4",
                     "Creatinine",
                     "ALT",
                     "AST","GGT","Nancy_max","eMAYO","MAYO_dai")
```

```{r}
metadata_final$Calprotectin[metadata_final$Group=="healthy"] <- NA
metadata_final$Calprotectin[metadata_final$Calprotectin==">6000"] <- 6000
metadata_final$Calprotectin <- as.numeric(metadata_final$Calprotectin)
  
metadata_ileum <- metadata_final %>% subset(Matrix=="TI") %>% as.data.frame()
metadata_colon <- metadata_final %>% subset(Matrix!="TI") %>% as.data.frame()

# TO DO!!!!!!!!!!!!!! nefunguje ako ma
metadata_for_boxplots <- metadata_final[,variables] %>%
  group_by(PatientID) %>%
  distinct(PatientID, .keep_all = TRUE) %>%
  as.data.frame()
```

# Import microbiome data

**Data Import**

Importing ASV, taxa and metadata tables for both Czech and Norway
samples.

*Czech*

```{r}
path = "../../data/analysis_ready_data/ikem/"
asv_tab_ikem <- as.data.frame(fread(file.path(path,"asv_table_ikem.csv"),
                                    check.names = FALSE))
taxa_tab_ikem <- as.data.frame(fread(file.path(path,"taxa_table_ikem.csv"),
                                     check.names = FALSE))
metadata_ikem <- as.data.frame(fread(file.path(path,"metadata_ikem.csv"),
                                     check.names = FALSE))
```

*Norway*

```{r}
path = "../../data/analysis_ready_data/norway/"
asv_tab_norway <- as.data.frame(fread(file.path(path,"asv_table_norway.csv"),
                                    check.names = FALSE))
taxa_tab_norway <- as.data.frame(fread(file.path(path,"taxa_table_norway.csv"),
                                     check.names = FALSE))
metadata_norway <- as.data.frame(fread(file.path(path,"metadata_norway.csv"),
                                     check.names = FALSE))
```

**Spliting to segments**

*Terminal ileum*

```{r}
ileum_data <- merging_data(asv_tab_1=asv_tab_ikem,
                           asv_tab_2=asv_tab_norway,
                           taxa_tab_1=taxa_tab_ikem,
                           taxa_tab_2=taxa_tab_norway,
                           metadata_1=metadata_ikem,
                           metadata_2=metadata_norway,
                           segment="TI",Q="clinical")

ileum_asv_tab <- ileum_data[[1]]
ileum_taxa_tab <- ileum_data[[2]]
ileum_metadata <- ileum_data[[3]]
```

*Colon*

```{r}
colon_data <- merging_data(asv_tab_1=asv_tab_ikem,
                           asv_tab_2=asv_tab_norway,
                           taxa_tab_1=taxa_tab_ikem,
                           taxa_tab_2=taxa_tab_norway,
                           metadata_1=metadata_ikem,
                           metadata_2=metadata_norway,
                           segment="colon",Q="clinical")

colon_asv_tab <- colon_data[[1]]
colon_taxa_tab <- colon_data[[2]]
colon_metadata <- colon_data[[3]]
```

**Filtering**

Rules: - prevalence \> 5% (per group) - nearZeroVar with default
settings - sequencing depth \> 5000 - taxonomic assignment at least
order

*Sequencing depth*

```{r}
data_filt <- seq_depth_filtering(ileum_asv_tab,
                                 ileum_taxa_tab,
                                 ileum_metadata,
                                 seq_depth_threshold = 10000)

filt_ileum_asv_tab <- data_filt[[1]]; alpha_ileum_asv_tab <- filt_ileum_asv_tab
filt_ileum_taxa_tab <- data_filt[[2]]; alpha_ileum_taxa_tab <- filt_ileum_taxa_tab
filt_ileum_metadata <- data_filt[[3]]; alpha_ileum_metadata <- filt_ileum_metadata

seq_step <- dim(filt_ileum_asv_tab)[1]
```

*NearZeroVar*

```{r}
data_filt <- nearzerovar_filtering(filt_ileum_asv_tab, 
                                   filt_ileum_taxa_tab,
                                   filt_ileum_metadata)

filt_ileum_asv_tab <- data_filt[[1]]
filt_ileum_taxa_tab <- data_filt[[2]]
nearzero_step <- dim(filt_ileum_asv_tab)[1]
```

**Filtering**

Rules: - prevalence \> 5% (per group) - nearZeroVar with default
settings - sequencing depth \> 5000 - taxonomic assignment at least
order

*Sequencing depth*

```{r}
data_filt <- seq_depth_filtering(colon_asv_tab,
                                 colon_taxa_tab,
                                 colon_metadata,
                                 seq_depth_threshold = 10000)

filt_colon_asv_tab <- data_filt[[1]]; alpha_colon_asv_tab <- filt_colon_asv_tab
filt_colon_taxa_tab <- data_filt[[2]]; alpha_colon_taxa_tab <- filt_colon_taxa_tab
filt_colon_metadata <- data_filt[[3]]; alpha_colon_metadata <- filt_colon_metadata

seq_step <- dim(filt_colon_asv_tab)[1]
```

*NearZeroVar*

```{r}
data_filt <- nearzerovar_filtering(filt_colon_asv_tab,
                                   filt_colon_taxa_tab,
                                   filt_colon_metadata)

filt_colon_asv_tab <- data_filt[[1]]
filt_colon_taxa_tab <- data_filt[[2]]
nearzero_step <- dim(filt_colon_asv_tab)[1]
```

*Check zero depth*

```{r}
data_filt <- check_zero_depth(filt_colon_asv_tab, 
                              filt_colon_taxa_tab, 
                              filt_colon_metadata)

filt_colon_asv_tab <- data_filt[[1]]; 
filt_colon_taxa_tab <- data_filt[[2]]; 
filt_colon_metadata <- data_filt[[3]]; 
```


# RECURRENCE INDEX

## Ileum

**All indices**

```{r,fig.width=12, fig.height=4}
metadata_ileum_melted <- melt(metadata_ileum %>% 
                                dplyr::select("Group", 
                                              "dys_unfiltered_asv",
                                              "dys_unfiltered_genus",
                                              "dys_filtered_asv",
                                              "dys_filtered_genus"))

p_il <- ggplot(metadata_ileum_melted) + 
  geom_boxplot(aes(x=Group, y=value),outliers = FALSE) + 
  geom_jitter(width = 0.2,height = 0,aes(x=Group, y=value, color=Group), 
              size=2) +
  facet_wrap(~variable, ncol = 4,scales = "free") + 
  theme_minimal() + 
    theme(panel.border = element_rect(color = "black", fill = NA, size = 0),
          axis.ticks.x = element_line(size=0.3,color = "black"),
          axis.ticks.y = element_line(size=0.3,color="black"),
          axis.ticks.length = unit(4,"pt"),
          panel.grid = element_blank()) + 
  scale_fill_manual(values=c("#309f87","#f9c675","#F08080","#A00000")) + 
  scale_color_manual(values=c("#309f87","#f9c675","#F08080","#A00000"))

p_il    
```

**FILTERED GENUS**

```{r}
dys_limit <- max(metadata_ileum$dys_filtered_genus,na.rm = TRUE) + 0.6*max(metadata_ileum$dys_filtered_genus,na.rm = TRUE)
dys_min_limit <-  min(metadata_ileum$dys_filtered_genus)

p_il_dys_genus <- ggplot(metadata_ileum %>%
                           mutate(`MDI`=dys_filtered_genus)) + 
  geom_boxplot(aes(x=Group, y=`MDI`),
               outliers = FALSE) + 
  geom_jitter(width = 0.3,height = 0,
              aes(x=Group, y=`MDI`, color=Group,shape=Country),
              size=1.5) +
  theme_minimal() + 
    theme(panel.border = element_rect(color = "black", fill = NA, size = 0),
          axis.ticks.x = element_line(size=0.3,color = "black"),
          axis.ticks.y = element_line(size=0.3,color="black"),
          axis.ticks.length = unit(4,"pt"),
          panel.grid = element_blank()) + 
  ylim(dys_min_limit,dys_limit) + 
  scale_fill_manual(values=c("#309f87","#f9c675","#F08080","#A00000")) + 
  scale_color_manual(values=c("#309f87","#f9c675","#F08080","#A00000")) + 
  theme(axis.text.x = element_text(angle = 45,face = "bold",vjust = 0.5))

p_il_dys_genus
```

```{r,results='hide'}
pdf("../figures/clinical/MRI_terminal_ileum.pdf",
    height =5,width = 5)
p_il_dys_genus
dev.off()
```

### AUC


```{r}
comparisons <- combn(unique(metadata_ileum$Group), 2, simplify = TRUE)

comparisons <- comparisons[,c(3,2,1,6,5,4)]
  
auc_df <- data.frame(Comparison=rep(NA,ncol(comparisons)),
                     AUC=rep(NA,ncol(comparisons)))

roc_curve_list <- list()
roc_curve_data_list <- list()

for (i in 1:ncol(comparisons)){
  name <- paste0(comparisons[1,i]," vs ", comparisons[2,i])
  metadata_sub <- subset(metadata_ileum,metadata_ileum$Group %in% comparisons[,i])
  metadata_sub$Group <- factor(metadata_sub$Group,
                               levels=c(comparisons[1,i],comparisons[2,i]))
  roc_curve_data <- roc(metadata_sub$Group, metadata_sub$dys_unfiltered_genus) 
  auc_value <- auc(roc_curve_data)  # Compute AUC
  auc_df[i,] <- c(name,auc_value)
  roc_curve_list[[name]] <- ggroc(roc_curve_data) + 
    theme_minimal() + 
    theme(legend.position = "none") + 
    ggtitle(paste0(name,' (AUC = ', round(auc_value,2))) + 
    theme(plot.title = element_text(size = 7))
  roc_curve_data_list[[name]] <- roc_curve_data
}

auc_df
roc_ileum <- roc_curve_all(roc_curve_data_list)
roc_ileum
```

```{r,results='hide'}
pdf("../figures/clinical/MRI_ROC_terminal_ileum.pdf",
    height =5,width = 7)
roc_ileum
dev.off()
```

## Colon

**All indices**

```{r,fig.width=12, fig.height=4}
colon_metadata_melted <- melt(metadata_colon %>% 
                                dplyr::select("Group", 
                                              "dys_unfiltered_asv",
                                              "dys_unfiltered_genus",
                                              "dys_filtered_asv",
                                              "dys_filtered_genus"))
p_col <- ggplot(colon_metadata_melted) + 
  geom_boxplot(aes(x=Group, y=value),outliers = FALSE) + 
  geom_jitter(width = 0.2,height = 0,
              aes(x=Group, y=value, color=Group),
              size=2) +
  facet_wrap(~variable, ncol = 4,scales = "free") + 
  theme_minimal() + 
 theme(panel.border = element_rect(color = "black", fill = NA, size = 0)) +
  scale_fill_manual(values=c("#309f87","#f9c675","#F08080","#A00000")) + 
  scale_color_manual(values=c("#309f87","#f9c675","#F08080","#A00000"))

p_col  
```

**FILTERED GENUS**

```{r}
dys_limit <- max(metadata_colon$dys_filtered_genus,na.rm = TRUE) + 0.6*max(metadata_colon$dys_filtered_genus,na.rm = TRUE)
dys_min_limit <-  min(metadata_colon$dys_filtered_genus)

p_col_dys_genus <- ggplot(metadata_colon %>%
                            mutate(`MDI`=dys_filtered_genus)) + 
  geom_boxplot(aes(x=Group, y=`MDI`),outliers = FALSE) + 
  geom_jitter(width = 0.3,height = 0,
              aes(x=Group, y=`MDI`, color=Group,shape=Country),
              size=1) +
  theme_minimal() + 
    theme(panel.border = element_rect(color = "black", fill = NA, size = 0),
          axis.ticks.x = element_line(size=0.3,color = "black"),
          axis.ticks.y = element_line(size=0.3,color="black"),
          axis.ticks.length = unit(4,"pt"),
          panel.grid = element_blank()) + 
  ylim(dys_min_limit,dys_limit) + 
  scale_fill_manual(values=c("#309f87","#f9c675","#F08080","#A00000")) + 
  scale_color_manual(values=c("#309f87","#f9c675","#F08080","#A00000")) + 
  theme(axis.text.x = element_text(angle = 45,face = "bold",vjust = 0.5))

p_col_dys_genus
```

```{r,results='hide'}
pdf("../figures/clinical/MRI_colon.pdf",
    height =5,width = 5)
p_col_dys_genus
dev.off()
```

### AUC

```{r}
metadata_colon_auc <- metadata_colon %>%
  group_by(PatientID) %>%
  distinct(PatientID, .keep_all = TRUE) %>%
  as.data.frame()

comparisons <- combn(unique(metadata_colon_auc$Group), 2, simplify = TRUE)
comparisons <- comparisons[,c(3,2,1,6,5,4)]

auc_df <- data.frame(Comparison=rep(NA,ncol(comparisons)),
                     AUC=rep(NA,ncol(comparisons)))

roc_curve_list <- list()
roc_curve_data_list <- list()

for (i in 1:ncol(comparisons)){
  name <- paste0(comparisons[1,i]," vs ", comparisons[2,i])
  metadata_sub <- subset(metadata_colon_auc,metadata_colon_auc$Group %in% comparisons[,i])
  metadata_sub$Group <- factor(metadata_sub$Group,
                               levels=c(comparisons[1,i],comparisons[2,i]))
  roc_curve_data <- roc(metadata_sub$Group, metadata_sub$dys_unfiltered_genus) 
  auc_value <- auc(roc_curve_data)  # Compute AUC
  auc_df[i,] <- c(name,auc_value)
  roc_curve_list[[name]] <- ggroc(roc_curve_data) + 
    theme_minimal() + 
    theme(legend.position = "none") + 
    ggtitle(paste0(name,' (AUC = ', round(auc_value,2),")")) + 
    theme(plot.title = element_text(size = 7))
  roc_curve_data_list[[name]] <- roc_curve_data
}

auc_df
roc_colon <- roc_curve_all(roc_curve_data_list)
roc_colon
```

```{r,results='hide'}
pdf("../figures/clinical/MRI_ROC_colon.pdf",
    height =5,width = 7)
roc_colon
dev.off()
```


## Linear models

**Ileum**

```{r}
results_model <- pairwise.lm(
  formula = "dys_filtered_genus ~ Group * Country",
  factors=metadata_ileum$Group,
  data=metadata_ileum)

knitr::kable(results_model[[1]][,c("Estimate", "p.adj","sig")],
             digits=3,caption="Results of linear modeling")
```

```{r}
results_model <- pairwise.lmer(
  formula = "dys_filtered_genus ~ Group * Country + (1|PatientID)",
  factors=metadata_colon$Group,
  data=metadata_colon)

knitr::kable(results_model[[1]][,c("Estimate", "p.adj","sig")],
             digits=3,caption="Results of linear mixed-effect modeling")

```


```{r, eval=FALSE}
dys <- ggarrange(p_il + ggtitle("Ileum"),
                 p_col + ggtitle("Colon"),nrow = 2)

pdf("results/clinical/dysbiosis_index_boxplots.pdf",
    height = 8,width =15)

dys
dev.off()
```

# ALPHA DIVERSITY

```{r}
alpha_div_plots <- list()
alpha_boxplots <- list()
corrs <- list()
```

## Ileum

```{r}
ileum_metadata_alpha_final <- metadata_ileum %>% 
  dplyr::select("SampleID", "PatientID","Group","Country",
                "Observe","Shannon","Simpson","Pielou",
                "dys_filtered_asv", "dys_filtered_genus")
```

```{r}
variables <- c("Observe","Shannon","Simpson","Pielou")

for (clinical_variable in variables){
  
  # ILEUM
  # boxplot
  p <- clinical_boxplot(ileum_metadata_alpha_final,
                      variable=clinical_variable)

  alpha_boxplots[[paste0("ileum_",clinical_variable)]] <- p
  
  # correlation
  ## ASV
  level="ASV"
  corr <- clinical_correlation(
    ileum_metadata_alpha_final,
    clinical_variable,
    level)
  
  corrs[[paste0("ileum_asv_",clinical_variable)]] <- corr
  
  p <- clinical_scatter(corr,
                        ileum_metadata_alpha_final,
                        clinical_variable,
                        level)
  
  alpha_div_plots[[paste0("ileum_asv_",clinical_variable)]] <- p

  ## Genus
  level="genus"
  corr <- clinical_correlation(ileum_metadata_alpha_final,
                               clinical_variable,
                               level)
  corrs[[paste0("ileum_genus_",clinical_variable)]] <- corr
  
  p <- clinical_scatter(corr,
                        ileum_metadata_alpha_final,
                        clinical_variable,
                        level)
  
  alpha_div_plots[[paste0("ileum_genus_",clinical_variable)]] <- p
}
```

### Boxplot

```{r,fig.width=12, fig.height=4}
alpha_boxplots_ileum <- ggarrange(
  plotlist = alpha_boxplots[grepl("ileum",names(alpha_boxplots))],
  ncol = 4,
  common.legend = TRUE, 
  legend = "right")

alpha_boxplots_ileum
```


## Colon

```{r}
colon_metadata_alpha_final <- metadata_colon %>% 
  dplyr::select("SampleID", "PatientID", "Group","Country",
                "Observe","Shannon","Simpson","Pielou",
                "dys_filtered_asv","dys_filtered_genus")
```

```{r}
variables <- c("Observe","Shannon","Simpson","Pielou")

for (clinical_variable in variables){

  # COLUMN
  p <- clinical_boxplot(colon_metadata_alpha_final,
                        variable=clinical_variable)

  alpha_boxplots[[paste0("colon_",clinical_variable)]] <- p
  
  # correlation
  ## ASV
  level="ASV"
  corr <- clinical_correlation(colon_metadata_alpha_final,
                               clinical_variable,
                               level,
                               segment="colon")
  
  corrs[[paste0("colon_asv_",clinical_variable)]] <- corr
  
  p <- clinical_scatter(corr,
                        colon_metadata_alpha_final,
                        clinical_variable,
                        level)
  
  alpha_div_plots[[paste0("colon_asv_",clinical_variable)]] <- p

  ## Genus
  level="genus"
  corr <- clinical_correlation(colon_metadata_alpha_final,
                               clinical_variable,
                               level,
                               segment="colon")
  
  corrs[[paste0("colon_genus_",clinical_variable)]] <- corr
  
  p <- clinical_scatter(corr,
                        colon_metadata_alpha_final,
                        clinical_variable,
                        level)
  
  alpha_div_plots[[paste0("colon_genus_",clinical_variable)]] <- p
}

```

### Boxplot

```{r,fig.width=12, fig.height=4}
alpha_boxplots_colon <- ggarrange(
  plotlist = alpha_boxplots[grepl("colon",names(alpha_boxplots))],
  ncol = 4,
  common.legend = TRUE, 
  legend = "right")

alpha_boxplots_colon
```

```{r, eval=FALSE}
alpha_boxplot <- ggarrange(alpha_boxplots_ileum,
                           alpha_boxplots_colon,
                           nrow = 2)
pdf("results/clinical/alpha_diversity_boxplots.pdf",height = 7,width =17)
alpha_boxplot
dev.off()
```

## Scatterplots

**ASV**

```{r,fig.width=20, fig.height=8}
alpha_dys_asv <- ggarrange(
  plotlist = alpha_div_plots[grepl("asv",names(alpha_div_plots))],
  ncol=4,
  nrow=2,
  common.legend = TRUE,
  legend="right")

alpha_dys_asv
```

**Genus**

```{r,fig.width=20, fig.height=8}
alpha_dys_genus <- ggarrange(
  plotlist = alpha_div_plots[grepl("genus",names(alpha_div_plots))],
  ncol=4,
  nrow=2,
  common.legend = TRUE,
  legend="right")

alpha_dys_genus
```

**Saving plots**

```{r, eval=FALSE}
pdf("results/clinical/alpha_diversity.pdf",height = 8,width =18)
alpha_dys_asv
dev.off()
```

```{r, eval=FALSE}
pdf("results/clinical/alpha_diversity_genus.pdf",height = 8,width =18)
alpha_dys_genus
dev.off()
```

## Linear models

### Richness

**Ileum**

```{r}
results_model <- pairwise.lm(formula = "Observe ~ Group * Country",
                                     factors=metadata_ileum$Group,
                                     data=metadata_ileum)

knitr::kable(results_model[[1]][,c("Estimate", "p.adj","sig")],
             digits=3,caption="Results of linear modeling")
```

**Colon**

```{r}
results_model <- pairwise.lmer(
  formula = "Observe ~ Group * Country + (1|Patient)",
  factors=metadata_colon$Group,
  data=metadata_colon %>% mutate(Patient=PatientID))

knitr::kable(results_model[[1]][,c("Estimate", "p.adj","sig")],
             digits=3,caption="Results of linear modeling")
```

### Shannon

**Ileum**

```{r}
results_model <- pairwise.lm(formula = "Shannon ~ Group * Country",
                                     factors=metadata_ileum$Group,
                                     data=metadata_ileum)

knitr::kable(results_model[[1]][,c("Estimate", "p.adj","sig")],
             digits=3,caption="Results of linear modeling")
```

**Colon**

```{r}
results_model <- pairwise.lmer(
  formula = "Shannon ~ Group * Country + (1|Patient)",
  factors=metadata_colon$Group,
  data=metadata_colon %>% mutate(Patient=PatientID))

knitr::kable(results_model[[1]][,c("Estimate", "p.adj","sig")],
             digits=3,caption="Results of linear modeling")
```

# MRI and alpha diversity plot 


```{r}
terminal_ileum_list <- list(
  ileum_genus_mdi_boxplot=p_il_dys_genus + 
    theme(legend.position = "none") + 
    ggtitle("Terminal ileum") +
    theme(plot.title = element_text(hjust=0.5,face = "bold",size = 15)),
  ileum_genus_observe=                          alpha_div_plots[grepl("ileum_genus",names(alpha_div_plots))][[1]] +
    theme(legend.position = "none"),
  ileum_genus_shannon=alpha_div_plots[grepl("ileum_genus",names(alpha_div_plots))][[2]] + 
    theme(legend.position = "none"),
  ileum_roc=roc_ileum + theme(legend.position = "none")
  )
  
p_terminal_ileum <- ggarrange(
  plotlist = terminal_ileum_list,
  ncol=1,
  nrow=4,heights = c(1.2,1,1,1),labels = c("A","B","C","D"))

colon_list <- list(
  colon_genus_mdi_boxplot=p_col_dys_genus + 
    theme(legend.position = "none") +
  ggtitle("Colon") + 
    theme(plot.title = element_text(hjust=0.5,face = "bold",size = 15)),
  colon_genus_observe=                          alpha_div_plots[grepl("colon_genus",names(alpha_div_plots))][[1]] + 
    theme(legend.position = "none"),
  colon_genus_shannon=alpha_div_plots[grepl("colon_genus",names(alpha_div_plots))][[2]] + 
    theme(legend.position = "none"),
  colon_roc=roc_colon + 
    theme(legend.position = "none")
  )
  
p_colon <- ggarrange(
  plotlist = colon_list,
  ncol=1,
  nrow=4,heights = c(1.2,1,1,1))

p <- ggarrange(p_terminal_ileum,p_colon,ncol=2,common.legend = TRUE,
  legend="right")

p <- ggarrange(p,ggplot() + theme_minimal(),ncol=2,
                                widths = c(1,0.15))
```

```{r,fig.width=10,fig.height=10}
p
```

```{r,results='hide'}
pdf("../figures/clinical/FIGURE_MRI.pdf",
    height =12,width = 8,paper="a4")
p
dev.off()
```