---
title: "Halo halo TITLE"
subtitle: "Halo halo subtitle"
output:
  html_document:
    df_print: paged
    toc: true
    toc_float: true
    toc-expand: 6
    toc_depth: 6
    embed-resources: true
    keep-md: false
    toc-location: left
    number-sections: true
    number-depth: 6
    code-fold: true
    code-tools: true
    code-summary: "Open code"
theme: darkly
highlight-style: github-dark
fontsize: 12 px
code-block-bg: "#2D2D31"
fontcolor: "#D1D1D1"
monobackgroundcolor: "#3B3B3B"
---


------------------------------------------------------------------------


# 1. Hypothesis: PSC is associated with the gut microbiome changes.

## Introduction

**Pre_LTx vs Post_LTx vs Healthy analysis on merged data**

-   Alpha diversity – Richness, Shannon, Simpson, and Pielou indices
    tested by linear fixed effect or mixed effect model (group effect,
    country effect, interaction effect)

-   Beta diversity – PERMANOVA (group effect, country effect,
    interaction effect), PCA visualization

-   Differential abundance testing:

    -   o Group effect – linDA + MaAsLin2 intersection

    -   o Country effect – taxa with significant interaction effect were
        excluded based on individual post-hoc analysis. Taxa had to have
        the same direction in both countries

-   ML models, trained and validated using out-of-sample bootstrap (500
    resamplings):

   o ENET

   o kNN

   o GBoost

   o RF

------------------------------------------------------

Pairwise comparisons:

**Pre_ltx vs Healthy** – an effect of disease, liver cirrhosis and
overall bad clinical condition

**Pre_ltx vs post_ltx** – an effect of transplantation + general
improvement of the clinical state

**Post_ltx vs Healthy** – does LTx lead to „healthy microbiome”?

**Report structure** The analysis is divided into two segments: terminal
ileum and colon. Each part begins with data filtering, where sequencing
depth is assessed and nearZeroVar() is applied to remove taxa with no
variance in the dataset. This is followed by a series of analytical
steps:

1.  Alpha diversity

2.  Beta diversity

3.  Univariate testing

4.  Machine learning models training and validation

At the end of the analysis, a brief summary of the results is provided.

**Importing libraries and custom functions built for this analysis**

```{r,warning=FALSE}
source("custom_functions.R")
```

## Clinical characteristics

```{r,warning=FALSE}
clinical_char <- read.xlsx("../results/clinical_characteristics.xlsx",
                           colNames = FALSE)
colnames(clinical_char) <- clinical_char[3,]
clinical_char <- clinical_char[-c(1,2,3),]
header_groups <- c(" " = 2, "HC" = 2, "pre_LTx" = 2, "post_LTx_non-rPSC"=2, "post_LTx_rPSC"=2)
```

```{r,warning=FALSE}
knitr::kable(clinical_char, align="l", escape=FALSE,digits=3) %>% 
  kable_styling(bootstrap_options = c("hover", "striped", "responsive")) %>% 
  kable_styling(full_width = TRUE) %>%
  add_header_above(header_groups)
```

## Data Import

Importing ASV, taxa and metadata tables for both Czech and Norway
samples.

**Czech**

```{r,warning=FALSE}
path = "../../data/analysis_ready_data/ikem/"
asv_tab_ikem <- as.data.frame(fread(file.path(path,"asv_table_ikem.csv"),
                                    check.names = FALSE))
taxa_tab_ikem <- as.data.frame(fread(file.path(path,"taxa_table_ikem.csv"),
                                     check.names = FALSE))
metadata_ikem <- as.data.frame(fread(file.path(path,"metadata_ikem.csv"),
                                     check.names = FALSE))
```

**Norway**

```{r,warning=FALSE}
path = "../../data/analysis_ready_data/norway/"
asv_tab_norway <- as.data.frame(fread(file.path(path,"asv_table_norway.csv"),
                                    check.names = FALSE))
taxa_tab_norway <- as.data.frame(fread(file.path(path,"taxa_table_norway.csv"),
                                    check.names = FALSE))
metadata_norway <- as.data.frame(fread(file.path(path,"metadata_norway.csv"),
                                    check.names = FALSE))
```

## Merging

Merging two countries based on the different matrices - Ileum, Colon.

**Terminal ileum**

```{r,warning=FALSE}
ileum_data <- merging_data(asv_tab_1=asv_tab_ikem,
                           asv_tab_2=asv_tab_norway,
                           taxa_tab_1=taxa_tab_ikem,
                           taxa_tab_2=taxa_tab_norway,
                           metadata_1=metadata_ikem,
                           metadata_2=metadata_norway,
                           segment="TI",Q="Q1")

ileum_asv_tab <- ileum_data[[1]]
ileum_taxa_tab <- ileum_data[[2]]
ileum_metadata <- ileum_data[[3]]
```

**Reads statistics**

```{r,warning=FALSE}
summary(colSums(ileum_asv_tab[,-1]))
```

**Colon**

```{r,warning=FALSE}
colon_data <- merging_data(asv_tab_1=asv_tab_ikem,
                           asv_tab_2=asv_tab_norway,
                           taxa_tab_1=taxa_tab_ikem,
                           taxa_tab_2=taxa_tab_norway,
                           metadata_1=metadata_ikem,
                           metadata_2=metadata_norway,
                           segment="colon",Q="Q1")

colon_asv_tab <- colon_data[[1]]
colon_taxa_tab <- colon_data[[2]]
colon_metadata <- colon_data[[3]]
```

**Reads statistics**

```{r,warning=FALSE}
summary(colSums(colon_asv_tab[,-1]))
```

# Data Analysis - Terminal ileum

```{r,warning=FALSE}
segment="terminal_ileum"
```

## Filtering

Rules:

-   sequencing depth \> 10000

-   nearZeroVar() with default settings

**Rarefaction Curve**

```{r,warning=FALSE}
path="../intermediate_files/rarecurves"
seq_depth_threshold <- 10000
```

```{r, warning=FALSE, eval = FALSE}
ps <- construct_phyloseq(ileum_asv_tab,ileum_taxa_tab,ileum_metadata)
rareres <- get_rarecurve(obj=ps, chunks=500)
save(rareres,file = file.path(path,"rarefaction_ileum.Rdata"))
```

```{r, warning=FALSE,warning=FALSE}
load(file.path(path,"rarefaction_ileum.Rdata"))

prare <- ggrarecurve(obj=rareres,
                      factorNames="Country",
                      indexNames=c("Observe")) + 
  theme_bw() +
  theme(axis.text=element_text(size=8), panel.grid=element_blank(),
        strip.background = element_rect(colour=NA,fill="grey"),
        strip.text.x = element_text(face="bold")) + 
  geom_vline(xintercept = seq_depth_threshold, 
             linetype="dashed", 
             color = "red") + 
  xlim(0, 20000)

prare
```